---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicRanges)

```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_dft2Som"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
```

## 0.3 Custom functions

# 1 Data
## 1.1 VCF files
```{r}
dft2Som <- read_table(paste0(dataDir, "Data_S1B_DFT2_somatic_variants.txt"),
                      col_types = cols(.default = "c"))
dft2SomLong <- dft2Som %>%
  pivot_longer(
    cols = -c(CHROM, POS, REF, ALT),
    names_to = "tumor",
    values_to = "ratio"
  ) %>%
  mutate(chrom = as.character(CHROM),
         location = paste0(CHROM, ":", POS),
         position = as.numeric(POS)) %>%
  filter(nchar(REF) == 1, nchar(ALT) == 1)

rm(dft2Som)
gc()
```

### 1.1.3 Germline variants
```{r}
# germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"),
#                       col_types = cols(.default = "c"))
# 
# germlineLong <- head(germline) %>%
#   as.data.frame() %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "sample",
#     values_to = "ratio"
#   ) %>%
#   mutate(chrom = as.character(CHROM),
#          location = paste0(CHROM, ":", POS),
#          position = as.numeric(POS)) %>%
#   filter(nchar(REF) == 1, nchar(ALT) == 1)
# 
# rm(germline)
# gc()
```


## 1.2 CNVs
```{r}
dft2Cnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx", 
                     sheet = "Table S6C", 
                     startRow = 3)

dft2CnvLong <- dft2Cnv %>%
  pivot_longer(-c(1:10),
              names_to = "tumor", 
              values_to = "presence") %>%
  mutate(presence = as.numeric(presence)) %>%
  mutate(
    START = as.numeric(START),
    END = as.numeric(END),
    CHROM = as.character(CHROM),
    tumor = as.character(tumor)
  )

rm(dft2Cnv)
gc()
```


## 1.3 Purities
```{r}
dft2Purities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(dft2Purities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(dft2Purities),
  paste0(colnames(dft2Purities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(dft2Purities) <- newNames
dft2Purities <- dft2Purities[-1, ]
```

## 1.4 Allele counter data
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- Adjust names using metadata! 

### 1.4.1 Metadata
- Mapping AlCo results to VCF file names
- FOR now, filter where mapping = 1. There are sample names that I am certain match the name in Max's supplementary table. 
- Go back and check once workflow is done
```{r}
metadata <- read.xlsx(paste0(resDir,
                          "08_deconvTable_metadata_purityIrodsMergedManualEdit.xlsx")) %>%
  filter(mapping==1)  %>%
              dplyr::rename("tumorId"="TUMOUR.INFORMATION_TUMOUR.ID", 
                     "tumorIdRnaSeq"="rnaSeqManual")
```

### 1.4.2 DFT2
#### 1.4.2.1 Read in AlCo results
```{r}
dft2SomAlCo <- list.files(paste0(resDir, "/063_Results/06_deconvTable_alCo_dft2SomRefAltExport"), 
                          recursive = T)

# Read all files and combine
dft2SomAlCoDf <- rbindlist(lapply(dft2SomAlCo, function(f) {
  # Extract tumor ID between "dft2SomRefAltExport_" and "_realigned"
  tumorId <- sub(".*dft2SomRefAltExport_(.+)_realigned.*", "\\1", basename(f))
  
  # Read file
  dt <- fread(paste0(resDir, "/063_Results/06_deconvTable_alCo_dft2SomRefAltExport/", f),
              select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
  
  # Add tumor ID column
  dt[, tumorId := tumorId]
  
  return(dt)
})) 

length(unique(dft2SomAlCoDf$tumorId))
```
#### 1.4.2.2 Adjust tumorId of RNAseq data to match Max's metadata
- Make tumorId the metadata id.
- Save RNAseq data name.
- Adjust tumorLocationAlt
```{r}
dft2SomAlCoCombined <- dft2SomAlCoDf %>%
  filter(tumorId %in% metadata$tumorIdRnaSeq) %>%
  left_join(metadata %>%
              dplyr::select(tumorId,
                            tumorIdRnaSeq)) %>%
  mutate(tumorLocation = paste0(tumorId, "_", CHROM, ":", POS), 
         tumorLocationAlt = paste0(tumorId, "_", CHROM, ":", POS, "_", ALT))

length(unique(dft2SomAlCoCombined$tumorId))
```

# 2 Filter data
- Only variants for which the alternate allele is found exclusively in one of the samples (host or tumor), in at least one host-tumor pair.
- The same variant in each single host-tumor pair is considered independent (independent CNV + tumor purity value associated)
## 2.3 VCF + Purities + Ploidy
- Merge info on purity and ploidy from metadata
- Create seperate columns for Alt/Total
- Filter out Alt > 0, this variant was not present in the tumor (reduces the size of the table).
```{r}
dft2PuritiesPloidy <- dft2Purities %>%
  mutate(ploidy = ifelse(X15_PLOIDY == "diploid",
                 2, 
                 ifelse(X15_PLOIDY == "tetraploid", 
                        4, 
                        "toCheck"))) %>%
  filter(ploidy != "toCheck")


dft2SomLongVaf <- dft2SomLong %>%
  mutate(
    Alt = as.integer(sub("/.*", "", ratio)),
    Total = as.integer(sub(".*/", "", ratio))
  ) %>%
  
  # If nAlt = 0, this was a variant in a different tumor sample, so filter out any 0 vafs!
  filter(Alt > 0) %>%
  left_join(dft2PuritiesPloidy %>%
              dplyr::rename("tumor"="TUMOUR.INFORMATION_TUMOUR ID", 
                     "purity"="X14_PURITY (VAF-based)") %>%
              mutate(purity = as.numeric(purity)) %>%
              dplyr::select(tumor, 
                            ploidy, 
                            purity)) %>%
  filter(tumor %in% dft2PuritiesPloidy$`TUMOUR.INFORMATION_TUMOUR ID`) %>%
  mutate(
    position = as.numeric(position),
    CHROM = as.character(CHROM),
    tumor = as.character(tumor)
  ) 
```

#### 2.3.1.2 Adjust ploidy at variant using CNV status. 
- Per variant per tumor, find all CNV events spanning that position
- This means that if you filter for presence > 0, there is a change in tumor CN. 
- For some variants, there may be more than one change in CN, to check which these are, filter for the rows where this is the case. 
##### 2.3.1.2.1 Find multiplets
```{r}
dft2CnvJoin <- dft2SomLongVaf %>%
  left_join(
    dft2CnvLong,
    by = join_by(
      CHROM,
      tumor,
      position >= START,
      position <= END
    )
  ) %>%
  filter(presence > 0) %>%
  distinct() %>%
  mutate(tumorLocation = paste0(tumor, 
                                "_", 
                                location),
         tumorLocationAlt = paste0(tumor, 
                                "_", 
                                location, 
                                "_", 
                                ALT)) 

# Tumor_position combinations where there is more than one CNV event
check <- which(table(dft2CnvJoin$tumorLocation)>1)
checkAlt <- which(table(dft2CnvJoin$tumorLocationAlt)>1)

### Reconstruct table but keep all variants, even when presence == 0, because these are instances where no CNV exists at that variant. 
dft2CnvJoinMultiplets <- dft2SomLongVaf %>%
  left_join(
    dft2CnvLong,
    by = join_by(
      CHROM,
      tumor,
      position >= START,
      position <= END
    )
  ) %>%
  # filter(presence > 0) %>%
  distinct() %>%
  mutate(tumorLocation = paste0(tumor, 
                                "_", 
                                location), 
         tumorLocationAlt = paste0(tumor, 
                                "_", 
                                location, 
                                "_", 
                                ALT))  %>%
  mutate(multiplets = ifelse(tumorLocation %in% names(check), 
                1, 
                0), 
         multipletsAlt = ifelse(tumorLocationAlt %in% names(checkAlt), 
                1, 
                0)) %>%
  mutate(variantType = "dft2Som") 

## File is too large to save.
# write.xlsx(dft2CnvJoinMultiplets %>% filter(presence > 0), 
#            paste0(resDir, 
#                   fprefix, 
#                   "_dft2CnvJoinMultiplets.xlsx"))

rm(dft2CnvJoin)
gc()

# Special cases where looked like a multiplet, but this is actually because there are two different variants at the same position in the same tumor!
checkMulti <- dft2CnvJoinMultiplets %>% 
  filter(multiplets == "1" & multipletsAlt == "0" & presence > 0)
```

##### 2.3.1.2.2 Variants with no change
- ploidyAdjusted = ploidy
- number of unique variants (taking into account alt) = number of rows, because no changes to cnv. YES? --> multiallelic variants possible (i.e., subclones, because these variants don't occur in any of the host samples.)
```{r}
dft2CnvJoinNoChanges <- dft2CnvJoinMultiplets %>%
  # Group by tumorLocation. If all rows presence = 0 (and, by default, not a multiplet), no change to CNV. 
  group_by(tumorLocationAlt) %>%
  filter(multipletsAlt == 0 & all(presence==0)) %>%
  mutate(
    nGain = sum(EVENT.TYPE == "GAIN"),
    nLoss = sum(EVENT.TYPE == "LOSS"),
    nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
    nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
    nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
    nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)")
  ) %>%
  ungroup() %>%
  mutate(copyNumberAdjusted = ploidy) %>%
  # remove CNV table columns
  dplyr::select(-c(setdiff(colnames(dft2CnvLong),
                           "tumor"))) %>%
  mutate(cnvAdjAnno = "noChanges")


# length(unique(dft2CnvJoinNoChanges$tumorLocation))==nrow(dft2CnvJoinNoChanges)
checkNoChange <- names(which(table(dft2CnvJoinNoChanges$tumorLocation)>1)) # 15 samples. This is because there are no changes, but you get duplicate rows if there is a diff Alt allele. 
checkNoChangeAlt <- names(which(table(dft2CnvJoinNoChanges$tumorLocationAlt)>1)) # 0 samples. 
```

##### 2.3.1.2.3 Variants with single change
- change ploidy according to single step
- number of unique variants = number of rows, because only 1 change to cnv per variant. 5 tumorLocation where more than one variant with a single CNV change.
- All presences should be 1 because single change
```{r}
dft2CnvJoinSingleChange <- dft2CnvJoinMultiplets %>%
  # Group by tumorLocation, if any presence not equals 0, keep all rows not equal to 0. If all rows presence = 0, no change to CNV. 
  mutate(ploidy=as.numeric(ploidy)) %>%
  filter(presence != 0 ) %>%
  group_by(tumorLocationAlt) %>%
  filter(multipletsAlt == 0) %>%
  mutate(
    nGain = sum(EVENT.TYPE == "GAIN"),
    nLoss = sum(EVENT.TYPE == "LOSS"),
    nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
    nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
    nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
    nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)")
  ) %>%
   # remove CNV table columns, because this has been aggregated by summing rows.
  dplyr::select(-c(setdiff(colnames(dft2CnvLong),
                           "tumor"))) %>%
  # take unique rows so that you lose the multiple rows/var b/c of multiple CNV steps
  distinct() %>%
  mutate(
    copyNumberAdjusted = case_when(
      # Case 1: diploid
      ploidy == 2 ~ ploidy + nGain - nLoss,
      
      # Case 2: tetraploid with pre-tetra events present
      ploidy == 4 & (nGainPre > 0 | nLossPre > 0) ~ 
        (2 + nGainPre - nLossPre) * 2 + 
        (nGain + nGainPost) - (nLoss + nLossPost),
      
      # Case 3: tetraploid without pre-tetra events present
      ploidy == 4 & nGainPre == 0 & nLossPre == 0 ~
        ploidy + (nGain + nGainPost) - (nLoss + nLossPost),
      TRUE ~ NA_real_
    )
  ) %>%
  mutate(cnvAdjAnno = "singleChange")
  
checkSingle <- names(which(table(dft2CnvJoinSingleChange$tumorLocation)>1)) # 0 
checkAltSingle <- names(which(table(dft2CnvJoinSingleChange$tumorLocationAlt)>1)) # 0 
```


##### 2.3.1.2.4 Variants with multiple changes
- Aggregate steps using same logic as before, only this time ploidy will change using sum of all steps rather than single change of CNV status.
```{r}
dft2CnvJoinMultipleChanges <- dft2CnvJoinMultiplets %>%
  mutate(ploidy=as.numeric(ploidy)) %>%
  # Group by tumorLocationAlt
  group_by(tumorLocationAlt) %>%
  # Remove variants where presence > 0, these rows are irrelevant for counting. 
  filter(multipletsAlt == 1 & presence > 0) %>%
    mutate(
    nGain = sum(EVENT.TYPE == "GAIN"),
    nLoss = sum(EVENT.TYPE == "LOSS"),
    nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
    nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
    nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
    nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)")
  ) %>%
  # remove CNV table columns, because this has been aggregated by summing rows.
  dplyr::select(-c(setdiff(colnames(dft2CnvLong),
                           "tumor"))) %>%
  # take unique rows so that you lose the multiple rows/var b/c of multiple CNV steps
  distinct() %>%
  mutate(
    copyNumberAdjusted = case_when(
      # Case 1: diploid
      ploidy == 2 ~ ploidy + nGain - nLoss,
      
      # Case 2: tetraploid with pre-tetra events present
      ploidy == 4 & (nGainPre > 0 | nLossPre > 0) ~ 
        (2 + nGainPre - nLossPre) * 2 + 
        (nGain + nGainPost) - (nLoss + nLossPost),
      
      # Case 3: tetraploid without pre-tetra events present
      ploidy == 4 & nGainPre == 0 & nLossPre == 0 ~
        ploidy + (nGain + nGainPost) - (nLoss + nLossPost),
      TRUE ~ NA_real_
    )
  ) %>%
  mutate(cnvAdjAnno = "multipleChanges")

checkMultipleChangesAlt <- names(which(table(dft2CnvJoinMultipleChanges$tumorLocationAlt)>1)) # 0 samples duplicated.
```

#### 2.3.1.3 Rejoin tables
- TRIPLE check no variants are missing because of filtering!
- Add host columns for merging with germline variants later on. 
```{r}
dft2SomCnvCorrected <- rbind(dft2CnvJoinNoChanges,
                             dft2CnvJoinSingleChange,
                             dft2CnvJoinMultipleChanges) %>%
  mutate(tumorVaf = Alt/Total, 
         tumorVafPurity = tumorVaf * purity, 
         hostVaf = 0, 
         hostVafPurity = 0) %>%
  distinct()
# write.xlsx(dft2SomCnvCorrected, 
#            paste0(resDir,
#                   fprefix,
#                   "_dft2SomCnvCorrected.xlsx"))
rm(dft2CnvJoinNoChanges,
   dft2CnvJoinSingleChange,
   dft2CnvJoinMultipleChanges,
   dft2CnvJoinMultiplets)


# dft2SomCnvCorrected <- read.xlsx(paste0(resDir,
#                   fprefix,
#                   "_dft2SomCnvCorrected.xlsx"))
# 

saveRDS(dft2SomCnvCorrected,
        paste0(resDir,
                  fprefix,
                  "_dft2SomCnvCorrected.rds"))

dft2SomCnvCorrected <- readRDS(paste0(resDir,
                  fprefix,
                  "_dft2SomCnvCorrected.rds")) 
```



# 3 Construct final table
- Join adjusted CNV status with AlleleCounter results
- Of note, NOT all samples have an associated RNA-seq file! See metadata script.
- Merge with gtf file on informative positions (exons)
- Adjust VAF
## 3.1 Merge AlleleCounter results
- 32 tumors (DFT2) with matched RNA seq data using mapping name between supp id and manual id
- Number of vars drops because not all vars in a tumor for which we have matched RNAseq.
```{r}
length(intersect(dft2SomCnvCorrected$tumor, metadata$tumorId))

dft2SomVarAlCoMerged <- dft2SomCnvCorrected %>%
  filter(tumorLocationAlt %in% dft2SomAlCoCombined$tumorLocationAlt) %>%
  left_join(dft2SomAlCoCombined %>%
              dplyr::select(TOTAL,
                            NREF, 
                            NALT, 
                            tumorLocationAlt,
                            tumorIdRnaSeq)) %>%
  distinct()

nrow(dft2SomVarAlCoMerged %>% 
       filter(TOTAL > 0)) # 1478 variants
```

### 3.2.2 GTF annotations
- noisyr has a function to extract all exons and positions from a gtf file.
- alternatively annotate using 
```{r}
gtfFile <- "~/gitClones/01_Deconvolution/02_Deconvolution_DFT1/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf"

# gtf <- makeTxDbFromGFF("~/gitClones/01_Deconvolution/02_Deconvolution_DFT2/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf")
# annoId <- 	cast_gtf_to_genes("~/gitClones/01_Deconvolution/02_Deconvolution_DFT2/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf", 
#                            feature = "exon")
# 
# annoNames <- 	cast_gtf_to_genes("~/gitClones/01_Deconvolution/02_Deconvolution_DFT2/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf", 
#                            feature = "exon", 
#                            "gene_name")
# 
# 
# annoGeneName <- 	cast_gtf_to_genes("~/gitClones/01_Deconvolution/02_Deconvolution_DFT2/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf", 
#                            feature = "gene", 
#                            "gene_name")

gtf <- import(gtfFile)

genes <- gtf[gtf$type == "gene"]

exons <- gtf[gtf$type == "exon"]

genes <- genes[genes$gene_biotype == "protein_coding"]


variants <- GRanges(
  seqnames = dft2SomVarAlCoMerged$chrom,
  ranges = IRanges(start = dft2SomVarAlCoMerged$position, 
                   end = dft2SomVarAlCoMerged$position)
)

# Which variants fall in an exon
exonHits <- findOverlaps(variants, exons)

# Map exon hits to gene IDs
exonGeneIds <- exons$gene_id[subjectHits(exonHits)]
exonGeneNames <- exons$gene_name[subjectHits(exonHits)]
exonBiotypes <- exons$gene_biotype[subjectHits(exonHits)]

hitsDf <- data.frame(
  query = queryHits(exonHits),
  gene_id = exonGeneIds,
  gene_name = exonGeneNames,
  biotype = exonBiotypes
)

collapseFun <- function(x) paste(unique(x), collapse = ";")

collapsed <- aggregate(. ~ query, 
                       data = hitsDf, 
                       FUN = collapseFun)

# Compare with assigned gene_id
dft2SomVarAlCoMergedAnno <- dft2SomVarAlCoMerged

dft2SomVarAlCoMergedAnno$exonGeneId <- NA
dft2SomVarAlCoMergedAnno$exonGeneName <- NA
dft2SomVarAlCoMergedAnno$exonBiotype <- NA

dft2SomVarAlCoMergedAnno$exonGeneId[collapsed$query] <- collapsed$gene_id
dft2SomVarAlCoMergedAnno$exonGeneName[collapsed$query] <- collapsed$gene_name
dft2SomVarAlCoMergedAnno$exonBiotype[collapsed$query] <- collapsed$biotype

length(unique(dft2SomVarAlCoMergedAnno %>%
         filter(is.na(exonGeneId)==F) %>%
         distinct() %>%
         pull(tumorLocationAlt)))

# [1] 913 unique vars intersecting a gene

length(unique(unique(dft2SomVarAlCoMergedAnno %>%
         filter(is.na(exonGeneId)==F) %>%
         distinct() %>%
         pull(exonGeneId))))

# [1] 112 genes

saveRDS(dft2SomVarAlCoMergedAnno, 
        paste0(resDir, 
               fprefix, 
               "_dft2SomVarAlCoMergedAnno.rds"))

write.xlsx(dft2SomVarAlCoMergedAnno, 
        paste0(resDir, 
               fprefix, 
               "_dft2SomVarAlCoMergedAnno.xlsx"))
```


