---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
```{r}
# install.packages("slider")
```

### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(arrow)
library(furrr)
library(patchwork)
library(slider)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_germline_dft2"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
source("~/gitClones/08_deconvTable/082_Scripts/genotype_germline_variants.R")
histoDir <- paste0(plotDir, 
                   "histo/")
dir.create(histoDir)
barDir <- paste0(plotDir, 
                 "bar/")
dir.create(barDir)


```

## 0.3 Custom functions
### 0.3.1 Compute tAlt
```{r}
computeTAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         hTotal, 
                         hAlt) {
  
    print("Solving for tAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    tAlt = ((obsVaf*denominator)-(1-purityTumor)*hAlt)/purityTumor
  return(tAlt)
}

assign_nAlt_from_vaf <- function(vaf, total_cn) {
  
  if (is.na(total_cn) || is.na(vaf)) return(NA_integer_)
  
  if (total_cn == 0) return(0)
  
  if (total_cn == 1) {
    return(ifelse(vaf < 0.5, 0, 1))
  }
  
  if (total_cn == 2) {
    if (vaf < 0.25) return(0)
    else if (vaf < 0.75) return(1)
    else return(2)
  }
  
  if (total_cn == 3) {
    if (vaf < 0.15) return(0)
    else if (vaf < 0.5) return(1)
    else if (vaf < 0.82) return(2)
    else return(3)
  }
  
  if (total_cn == 4) {
    if (vaf < 0.12) return(0)
    else if (vaf < 0.39) return(1)
    else if (vaf < 0.64) return(2)
    else if (vaf < 0.86) return(3)
    else return(4)
  }
  
  return(NA_integer_)  # fallback
}
```
### 0.3.2 Compute hAlt
```{r}
computeHAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         tAlt,
                         hTotal) {
  
  # Cases where hAlt = 0, solve for tAlt
 
    print("Solving for hAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    hAlt = (obsVaf*denominator)/(1-purityTumor)
  return(hAlt)
}

computeHAlt(obsVaf = 0.5, 
            purityTumor = 0.8, 
            tTotal = 4, 
            hTotal = 2, 
            tAlt=0)



```
### 0.3.3 Geometric mean
- Kevin's scripts
  - To calculate size factors: 
  - Compute the geom.mean of totals for each variant (optional: add pseudocount)
  - Divide each variant's read total by the geometric mean for each variant
  - Then take the median of these ratios for each sample
  - (Optional) restrict to variants with presence in all tumors
```{r}
geomean <- function(x) {
  if (all(x == 0)) {
    NA_real_
  } else {
    exp(mean(log(x[x > 0])))
  }
}
```

### 0.3.4 Summarize germline variants
```{r}
library(dplyr)
library(tidyr)

library(data.table)

summarise_variants_dt <- function(df) {
  dt <- as.data.table(df)
  
  # ensure uniqueness per tumor per location
  dt <- unique(dt[, .(locationAlt, sampleTumor, statusUpdate)])
  
  # count per location + status
  counts <- dt[, .N, by = .(locationAlt, statusUpdate)]
  
  # total tumors per location
  totals <- dt[, .(total_tumors = uniqueN(sampleTumor)), by = locationAlt]
  
  # merge counts with totals
  counts <- counts[totals, on = "locationAlt"]
  
  # compute percentage
  counts[, pct := N / total_tumors]
  
  # reshape wide (one column per statusUpdate)
  wide <- dcast(
    counts,
    locationAlt + total_tumors ~ statusUpdate,
    value.var = "pct",
    fill = 0
  )
  
  # compute majority
  status_cols <- setdiff(names(wide), c("locationAlt", "total_tumors"))
  
  wide[, majority := {
    vals <- unlist(.SD)
    max_val <- max(vals)
    winners <- names(vals)[vals == max_val]
    
    if (length(winners) > 1) "flag" else winners
  }, by = locationAlt, .SDcols = status_cols]
  
  return(wide)
}
```

### 0.3.4 Copy number state sequence
```{r}
possible_cn_absorbing <- function(start_cn, nGain, nLoss) {
  # Each state is a pair: CN, remaining events (gains and losses)
  states <- list(list(cn = start_cn, g = nGain, l = nLoss))
  
  final_cns <- c()
  
  while (length(states) > 0) {
    new_states <- list()
    
    for (s in states) {
      cn <- s$cn
      g  <- s$g
      l  <- s$l
      
      if (cn == 0 || (g == 0 && l == 0)) {
        # CN is 0 (absorbing) or no events left → record final CN
        final_cns <- c(final_cns, cn)
      } else {
        # Apply a gain if available
        if (g > 0 && cn > 0) {
          new_states <- append(new_states, list(list(cn = cn + 1, g = g - 1, l = l)))
        }
        # Apply a loss if available
        if (l > 0) {
          new_cn <- max(0, cn - 1)
          new_states <- append(new_states, list(list(cn = new_cn, g = g, l = l - 1)))
        }
      }
    }
    
    states <- new_states
  }
  sort(unique((final_cns)))
}

possible_cn_absorbing(3, 2, 3)
```
# 1 Data
## 1.1 Purities and ploidy
```{r}
germlinePurities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(germlinePurities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(germlinePurities),
  paste0(colnames(germlinePurities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(germlinePurities) <- newNames
germlinePurities <- germlinePurities[-1, ]

germlinePuritiesPloidy <- germlinePurities %>%
  mutate(ploidy = ifelse(X15_PLOIDY == "diploid",
                 2, 
                 ifelse(X15_PLOIDY == "tetraploid", 
                        4, 
                        "toCheck") 
                 )) %>%
  filter(ploidy != "toCheck") %>%
  dplyr::rename("sampleTumor"="TUMOUR.INFORMATION_TUMOUR ID", 
                     "purity"="X14_PURITY (VAF-based)", 
                "dftType"="X9_LINEAGE") %>%
              mutate(purity = as.numeric(purity), 
                     ploidy = as.numeric(ploidy), 
                     sex=ifelse(HOST.SEX == "♂", 
                                "male", 
                                ifelse(HOST.SEX == "♀", 
                                       "female", 
                                       "toCheck"))) %>%
              dplyr::select(sampleTumor, 
                            ploidy, 
                            purity, 
                            sex, 
                            dftType
                            ) 
  
table(germlinePuritiesPloidy$sex, germlinePuritiesPloidy$dftType)

males <- germlinePuritiesPloidy %>%
  filter(sex == "male") 

dft2 <- germlinePuritiesPloidy %>%
  filter(dftType == "DFT2") 
```
## 1.2 VCF files
- Read in germline vcf file, all variants. 
- Pivot longer so that each variant and each sample is a single row with the ALT/TOTAL ratio.
```{r}
# germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"),
#                       col_types = cols(.default = "c")) 
# 
# # Starting with autosomal germline and somatic substitution variants, we identified genomic
# # intervals containing germline absolute copy number polymorphisms (see section 7.1) and removed
# # germline substitutions occurring either within these intervals, or within 10 kb on either side of
# # these intervals.
# # This alignment included 1,175,235 substitution variants
# # (1,070,436 germline, 104,799 somatic).
# # > 1175235-1070436
# # [1] 104799 --> fall into a CNV region??
# # # 
# 
# sampleInfo <- tibble(sample = colnames(germline)[-(1:4)]) %>%
#   mutate(
#     sampleType = str_extract(sample, "T|H"),
#     tcgId = str_extract(sample, "^[^TH]+")
#   )
# 
# 
# germlineLong <- germline %>%
#   filter(nchar(REF) == 1,
#          nchar(ALT) == 1) %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "sample",
#     values_to = "ratio"
#   ) %>%
#   left_join(sampleInfo, by = "sample") %>%
#   transmute(
#     CHROM,
#     POS,
#     REF,
#     ALT,
#     tcgId,
#     sampleType,
#     ratio,
#     sample
#   )
# rm(germline)
# saveRDS(germlineLong,
#         paste0(resDir,
#                "08_deconvTable_germline_germlineLong.rds"))
# 
# 
# germlineLong <- readRDS(paste0(resDir,
#                "08_deconvTable_germline_germlineLong.rds"))
```

## 1.3 Pair tumor/host samples
- Using shared ID's pivot wider so that the ratio for matched tumor/host samples is side by side.
- Filter on variants where alt allele is present in only the host or the tumor sample within a pair! --> so that we have info. on tAlt or hAlt at that position.
- Tumor samples = 112
- Host samples = 81
- Paired = 103 --> 100 if I filter Max's table? Maybe the non-transmissible cases. 
```{r}
# tumorSamples <- germlineLong %>%
#   filter(sampleType == "T") %>%
#   dplyr::select(c("tcgId", "sample")) %>%
#   distinct()
# 
# normalSamples <- germlineLong %>%
#   filter(sampleType == "H") %>%
#   dplyr::select(c("tcgId", "sample")) %>%
#   distinct()
# 
# pairedSamples <- tumorSamples %>%
#   inner_join(normalSamples,
#             by= c("tcgId"))
# 
# # > setdiff(tumorSamples$tcgId, pairedSamples$tcgId). Checked all manually in Max's S2 table. None have a matched host. 
# #  [1] "49" --> no host.   "86" --> cell line + no host  "88" --> cell line + no host "102" --> no host "140" --> no host  "208"--> no host  "367"  "368"  "378"  "384"  "837"  "876"  "1071" "1515" "1524" "1553" "2692" "2695".
# 
# write.xlsx(pairedSamples,
#            paste0(paste0(resDir,
#                "08_deconvTable_germline_germlinePaired.xlsx")))
# 
# normal <- germlineLong %>%
#   filter(sampleType == "H") %>%
#   dplyr::rename(
#     vafHost = ratio,
#     sampleHost = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# gc()
# 
# tumor <- germlineLong %>%
#   filter(sampleType == "T") %>%
#   dplyr::rename(
#     vafTumor = ratio,
#     sampleTumor = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# gc()
# 
# paired <- tumor %>%
#   inner_join(normal,
#              by = c("CHROM",
#                     "POS", 
#                     "REF", 
#                     "ALT", 
#                     "tcgId"))
# 
# gc()
# 
# rm(tumor, normal, germlineLong)
# gc()
# 
# pairedNas <- paired %>%
#   filter(is.na(vafTumor) | is.na(vafHost)) # --> 0 rows!
# 
# write_dataset(paired,
#               format = "parquet",
#         paste0(resDir,
#                "08_deconvTable_germline_germlinePaired/"),
#         partitioning = "tcgId")
# 
# 
# germlinePaired <- open_dataset(paste0(resDir,
#                "08_deconvTable_germline_germlinePaired/")) %>%
#   mutate(
#     tumorAlt = as.integer(sub("/.*", "", vafTumor)),
#     tumorTotal = as.integer(sub(".*/", "", vafTumor)),
#     hostAlt = as.integer(sub("/.*", "", vafHost)),
#     hostTotal = as.integer(sub(".*/", "", vafHost))
#   ) %>%
#   mutate(obsVafTumor = tumorAlt/tumorTotal,
#          obsVafHost = hostAlt/hostTotal)%>%
#  mutate(locationRef = paste0(CHROM, ":", POS, "_", REF, "_", ALT),
#         tumorLocationRef = paste0(sampleTumor, 
#                                   "_", CHROM, ":", POS, "_", REF),
#         tumorLocationRefAlt = paste0(sampleTumor, 
#                                   "_", CHROM, ":", POS, "_", REF, "_", ALT), 
#         position = as.numeric(POS), 
#         tumor = as.character(sampleTumor))
# 
# # Overwrite data. 
# write_dataset(germlinePaired,
#               format = "parquet",
#               paste0(resDir,
#                      "08_deconvTable_germline_germlinePaired_computedVafs/"),
#               partitioning = "tcgId")

germlinePaired <- open_dataset(paste0(resDir,
                                      "08_deconvTable_germline_germlinePaired_computedVafs/")) %>%
  filter(sampleTumor %in% dft2$sampleTumor) %>%
  mutate(locationRef = paste0(CHROM, ":", POS, "_", REF),
         tumorLocation = paste0(sampleTumor, "_", CHROM, ":", POS), 
         tumorLocationAlt =paste0(sampleTumor, "_", CHROM, ":", POS, "_", ALT)) 

gc()
```

## 1.4 Subset germline vcfs by host, tumor, or shared vaf's.
- Germline (host+tumour) variants are defined as those with VAF >0.25 in any host.
- That could mean it has 0 VAF in a given tumor sample, but this was not observed in all samples and is therefore not somatic. 
- How to account for this when calling "true" informative variants?
```{r}
# germlinePairedFiltered <- germlinePaired 
```

## 1.5 CNVs - DFT2
### 1.5.1 Read in CNV data, join with VAF's
- Any germline variants falling within a CNV event were filtered out, so no need to adjust ploidy! 
```{r}
dft2Cnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx",
                     sheet = "Table S6C",
                     startRow = 3)

dft2CnvLong <- dft2Cnv %>%
  pivot_longer(-c(1:10),
              names_to = "tumor",
              values_to = "presence") %>%
  mutate(presence = as.numeric(presence)) %>%
  mutate(
    START = as.numeric(START),
    END = as.numeric(END),
    CHROM = as.character(CHROM),
    tumor = as.character(tumor)
  )

rm(dft2Cnv)
gc()

# dft2CnvJoin <- germlinePaired %>%
#   left_join(
#     dft2CnvLong,
#     by = join_by(
#       CHROM,
#       tumor,
#       position >= START,
#       position <= END
#     )
#    ) %>%
#   # filter(presence > 0) %>%
#   distinct() 
# 
# write_dataset(dft2CnvJoin,
#               format = "parquet",
#         paste0(resDir,
#                fprefix,
#                "_dft2CnvJoin/"),
#         partitioning = "tcgId")

dft2CnvJoin <- open_dataset(paste0(resDir,
               fprefix,
               "_dft2CnvJoin/"))

# # --- 1. Find duplicate keys via aggregation (cheap, no full collect) ----------
# dup_keys <- dft2CnvJoin %>%
#   count(tumorLocation) %>%          # pushed down to Arrow
#   collect() %>%                      # only the counts land in R
#   filter(n > 1) %>%
#   pull(tumorLocation)
# 
# dup_keys_refalt <- dft2CnvJoin %>%
#   count(tumorLocationRefAlt) %>%
#   collect() %>%
#   filter(n > 1) %>%
#   pull(tumorLocationRefAlt)
# 
# dft2CnvJoinMultiplets <- dft2CnvJoin %>%
#   mutate(multiplets = ifelse(tumorLocation %in% dup_keys, 
#                 1, 
#                 0), 
#          multipletsRefAlt = ifelse(tumorLocationRefAlt %in% dup_keys_refalt, 
#                 1, 
#                 0)) %>%
#   mutate(variantType = "dft2Germline") %>%
#   left_join(germlinePuritiesPloidy)
# 
# rm(dft2CnvJoin)
# gc()
# 
# write_dataset(dft2CnvJoinMultiplets,
#               format = "parquet",
#               paste0(resDir,
#                fprefix,
#                "_dft2CnvJoinMultiplets/"),
#               partitioning = "tcgId")

# checkMulti <- open_dataset(paste0(resDir,
#                fprefix,
#                "_dft2CnvJoinMultiplets/")) %>%
#   filter(multipletsRefAlt > 0 & presence > 0) %>%
#   collect()
# 
# checkSinglets <- open_dataset(paste0(resDir,
#                fprefix,
#                "_dft2CnvJoinMultiplets/")) %>%
#   filter(multipletsRefAlt == 0 & presence > 0) %>%
#   collect()
```

### 1.5.2 Adjust CN
```{r}
# dft2CnvJoinMultiplets <- open_dataset(paste0(resDir,
#                fprefix,
#                "_dft2CnvJoinMultiplets/"))
# 
# schema(dft2CnvJoinMultiplets)
# 
# multiplets <- open_dataset(paste0(resDir,
#                fprefix,
#                "_dft2CnvJoinMultiplets/")) %>%
#   filter(multipletsRefAlt == 1 & presence > 0) %>%
#   collect() %>%
#   pull(tumorLocationRefAlt)
# 
# singlets <- dft2CnvJoinMultiplets %>%
#   filter(multiplets =multipletsRefAlt == 0 & presence > 0) %>%
#   collect() %>%
#   pull(tumorLocationRefAlt)
# remaining <- dft2CnvJoinMultiplets %>%
#   dplyr::select(tumorLocationRefAlt) %>%
#   distinct() %>%
#   collect() %>%
#   pull(tumorLocationRefAlt)

# length(remaining) == length(singlets) + length(multiplets)

# # Filter on variants where I know ploidy = 2 and nGain = 1 and nLoss = 2. This is because this should lead to two possible final copy numbers (1 or 0).
# dft2Test <- open_dataset(paste0(resDir,
#                fprefix,
#                "_dft2CnvJoin/")) %>%
# 
#   mutate(ploidy = as.numeric(ploidy)) %>%
# 
#   group_by(tumorLocationRefAlt) %>%
# 
#   filter(multipletsRefAlt == 1 & presence > 0) %>%
#   
#   collect() %>%
# 
#   mutate(
#     nGain = sum(EVENT.TYPE == "GAIN"),
#     nLoss = sum(EVENT.TYPE == "LOSS"),
#     nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
#     nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
#     nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
#     nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)")
#   ) %>%
# 
#   dplyr::select(-c(setdiff(colnames(dft2CnvLong), "tumor"))) %>%
#   distinct() %>%
#   rowwise()  %>%
#   filter(ploidy == 2 & nGain == 1 & nLoss == 2) 
# 
# # Test functions
# dft2TestFunction <- dft2Test %>%
#   
#   mutate(
#     # ---- CASE 1: diploid ----
#     possibleCN = list(
#       if (ploidy == 2) {
#         possible_cn_absorbing(ploidy, nGain, nLoss)
#         
#       # ---- CASE 2: tetraploid with pre-events ----
#       } else if (ploidy == 4 & (nGainPre > 0 | nLossPre > 0)) {
#         
#         pre_cn <- possible_cn_absorbing(2, nGainPre, nLossPre)
#         pre_cn <- pre_cn * 2
#         
#         # apply post events to each possible pre state
#         unique(unlist(
#           lapply(pre_cn, function(cn) {
#             possible_cn_absorbing(
#               cn,
#               nGain + nGainPost,
#               nLoss + nLossPost
#             )
#           })
#         ))
#         
#       # ---- CASE 3: tetraploid without pre-events ----
#       } else if (ploidy == 4) {
#         
#         possible_cn_absorbing(
#           ploidy,
#           nGain + nGainPost,
#           nLoss + nLossPost
#         )
#         
#       } else {
#         NA_real_
#       }
#     ),
#     
#     minCN = min(possibleCN),
#     maxCN = max(possibleCN)
#     
#   ) %>%
#   ungroup() %>%
#   mutate(cnvAdjAnno = "multipleChangesTest")

```



#### 1.2.2.1 Variants with no change
- copyNumberAdjusted = ploidy
- number of unique variants (taking into account alt) = number of rows, because no changes to cnv. YES? --> multiallelic variants possible (i.e., subclones, because these variants don't occur in any of the host samples.)
```{r}
# rm(germlinePaired)
# 
# dft2CnvJoinNoChanges <- open_dataset(paste0(resDir,
#                                       "08_deconvTable_germline_germlinePaired_computedVafs/")) %>%
#   filter(sampleTumor %in% dft2$sampleTumor) %>%
#   mutate(locationRef = paste0(CHROM, ":", POS, "_", REF),
#          tumorLocation = paste0(sampleTumor, "_", CHROM, ":", POS), 
#          tumorLocationAlt =paste0(sampleTumor, "_", CHROM, ":", POS, "_", ALT)) %>%
#   filter(!tumorLocationRefAlt %in% c(singlets, multiplets)) %>%
#   # Group by tumorLocation. If all rows presence = 0 (and, by default, not a multiplet), no change to CNV. 
#   left_join(germlinePuritiesPloidy) %>%
#   mutate(presence = 0) %>%
#   mutate(
#     nGain = 0,
#     nLoss = 0,
#     nGainPost = 0,
#     nGainPre = 0,
#     nLossPost = 0,
#     nLossPre = 0,
#     sumPresence = 0
#   ) %>%
#   mutate(copyNumberAdjusted = ploidy) %>%
#   # # remove CNV table columns
#   # dplyr::select(-c(setdiff(colnames(dft2CnvLong),
#   #                          "tumor"))) %>%
#   mutate(cnvAdjAnno = "noChanges") %>%
#   collect()
# 
# 
# dft2CnvJoinNoChanges$rowId <- 1:nrow(dft2CnvJoinNoChanges)
# 
# chunk_size <- 5e4  # adjust to your RAM
# 
# dft2CnvJoinNoChanges <- dft2CnvJoinNoChanges %>%
#   mutate(chunk_id = (rowId - 1) %/% chunk_size)
# 
# write_dataset(dft2CnvJoinNoChanges,
#               paste0(resDir, "dft2CnvJoinNoChanges/"), 
#               partitioning = "chunk_id")
# 
# rm(dft2CnvJoinNoChanges)
# # Check distribution of adjusted CN per chromosome. Sanity check before running genotyper.
# check <- open_dataset(paste0(resDir, "dft2CnvJoinNoChanges/")) %>%
#   dplyr::select(copyNumberAdjusted, CHROM) %>%
#   collect()
# 
# table(check$copyNumberAdjusted, check$CHROM)
```

#### 1.2.2.2 Variants with single change
- change ploidy according to single step
- number of unique variants = number of rows, because only 1 change to cnv per variant. 5 tumorLocation where more than one variant with a single CNV change.
- All presences should be 1 because single change
- If CN goes to 0, it should stay at 0. (Applies for multistep, not here.) Running same code as multiplets as a sanity check, none of the maxCn should differ from minCn. 
```{r}
# dft2CnvJoinSingleChange <- dft2CnvJoinMultiplets %>%
#   mutate(ploidy=as.numeric(ploidy)) %>%
#   filter(presence > 0 & multipletsRefAlt == 0) %>%
#   group_by(tumorLocationAlt) %>%
#   mutate(
#     nGain = sum(EVENT.TYPE == "GAIN"),
#     nLoss = sum(EVENT.TYPE == "LOSS"),
#     nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
#     nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
#     nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
#     nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)"),
#     sumPresence = sum(presence)
#   ) %>%
#    # remove CNV table columns, because this has been aggregated by summing rows.
#   dplyr::select(-c(setdiff(colnames(dft2CnvLong),
#                            "tumor"))) %>%
#   mutate(
#     copyNumberAdjusted = pmax(0,
#     case_when(
#       # Case 1: At sex chromosomes, starting CN of tumor is 1 because male sample. 
#       ploidy == 1 ~ ploidy + nGain - nLoss,
#       
#       # Case 1: diploid
#       ploidy == 2 ~ ploidy + nGain - nLoss,
#       
#       # Case 2: tetraploid with pre-tetra events present
#       ploidy == 4 & (nGainPre > 0 | nLossPre > 0) ~ 
#         (2 + nGainPre - nLossPre) * 2 + 
#         (nGain + nGainPost) - (nLoss + nLossPost),
#       
#       # Case 3: tetraploid without pre-tetra events present
#       ploidy == 4 & nGainPre == 0 & nLossPre == 0 ~
#         ploidy + (nGain + nGainPost) - (nLoss + nLossPost),
#       
#       TRUE ~ NA_real_
#     )
#    )
#   ) %>%
#   mutate(cnvAdjAnno = "singleChange") 
# 
# 
# dft2CnvJoinSingleChange <- dft2CnvJoinSingleChange %>%
#   collect()
# 
# dft2CnvJoinSingleChange$rowId <- 1:nrow(dft2CnvJoinSingleChange)
# 
# chunk_size <- 5e4  # adjust to your RAM
# 
# dft2CnvJoinSingleChange <- dft2CnvJoinSingleChange %>%
#   mutate(chunk_id = (rowId - 1) %/% chunk_size)
# 
# write_dataset(dft2CnvJoinSingleChange,
#               paste0(resDir, "dft2CnvJoinSingleChange/"), 
#               partitioning = "chunk_id")
# 
# 
# # Check distribution of adjusted CN per chromosome. Sanity check before running genotyper.
# check <- open_dataset(paste0(resDir, "dft2CnvJoinSingleChange/")) %>%
#   left_join(germlinePaired %>%
#               filter(tumorLocationRefAlt %in% singlets) %>%
#               dplyr::select(CHROM, tumorLocationRefAlt),
#             by="tumorLocationRefAlt") %>%
#   dplyr::select(copyNumberAdjusted, CHROM) %>%
#   collect()
# 
# table(check$copyNumberAdjusted, check$CHROM)
#    
#          1      2      3      4      5      6      X      Y
#   1  23755 587361 315056  14980 788075 167703      0     17
#   3  67477 804412  15343 188067  26482  69869   7947      1
#   5    470  43331    823    996     90      0      0      0
```

#### 1.2.2.3 Variants with multiple changes
- Aggregate steps using same logic as before, only this time ploidy will change using sum of all steps rather than single change of CNV status.
- Issue: we don't know the order of events. Depending on order of losses and gains, the final CN could differ. Adding a permutation function which returns how many times a different final CN state could occur. 
- None of the tetraploid tumors have nLossPre > 1, therefore cannot go to 0.
- The only variable cases are diploid tumors whose variants have two losses and one gain. 
- Filter out the five variants that could have been CN 0 or 1!
  - Check these variants against the counts + gtf file.
```{r}
# dft2CnvJoinMultipleChanges <- dft2CnvJoinMultiplets %>%
#   
#   mutate(ploidy = as.numeric(ploidy)) %>%
#   
#   group_by(tumorLocationRefAlt) %>%
#   
#   filter(multipletsRefAlt == 1 & presence > 0) %>%
#   
#   mutate(
#     nGain = sum(EVENT.TYPE == "GAIN"),
#     nLoss = sum(EVENT.TYPE == "LOSS"),
#     nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
#     nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
#     nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
#     nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)"),
#     sumPresence = sum(presence)
#   ) %>%
#   
#   dplyr::select(-c(setdiff(colnames(dft2CnvLong), "tumor"))) %>%
#   distinct() %>%
#   collect() %>%
#   rowwise() %>%
#   
#   mutate(
#     
#     # ---- CASE 1: diploid ----
#     possibleCn = list(
#       if (ploidy == 1) {
#         possible_cn_absorbing(ploidy, nGain, nLoss)
#         
#       # ---- CASE 2: tetraploid with pre-events ----
#       }
#       else if (ploidy == 2) {
#         possible_cn_absorbing(ploidy, nGain, nLoss)
#         
#       # ---- CASE 2: tetraploid with pre-events ----
#       } else if (ploidy == 4 & (nGainPre > 0 | nLossPre > 0)) {
#         
#         pre_cn <- possible_cn_absorbing(2, nGainPre, nLossPre)
#         pre_cn <- pre_cn * 2
#         
#         # apply post events to each possible pre state
#         unique(unlist(
#           lapply(pre_cn, function(cn) {
#             possible_cn_absorbing(
#               cn,
#               nGain + nGainPost,
#               nLoss + nLossPost
#             )
#           })
#         ))
#         
#       # ---- CASE 3: tetraploid without pre-events ----
#       } else if (ploidy == 4) {
#         
#         possible_cn_absorbing(
#           ploidy,
#           nGain + nGainPost,
#           nLoss + nLossPost
#         )
#         
#       } else {
#         NA_real_
#       }
#     ),
#     
#     minCn = min(possibleCn),
#     maxCn = max(possibleCn)
#     
#   ) %>%
#   ungroup() %>%
#   mutate(cnvAdjAnno = "multipleChanges")
# 
# # 0 conflicting rows
# multipleCns <- dft2CnvJoinMultipleChanges %>% 
#   filter(minCn != maxCn)
# 
# dft2CnvJoinMultipleChangesFiltered <- dft2CnvJoinMultipleChanges %>%
#   filter(minCn == maxCn) %>%
#   mutate(copyNumberAdjusted = minCn) %>%
#   dplyr::select(-c(possibleCn, 
#                    maxCn, 
#                    minCn))
# 
# dft2CnvJoinMultipleChangesFiltered$rowId <- 1:nrow(dft2CnvJoinMultipleChangesFiltered)
# 
# chunk_size <- 5e4  # adjust to your RAM
# 
# dft2CnvJoinMultipleChangesFiltered <- dft2CnvJoinMultipleChangesFiltered %>%
#   mutate(chunk_id = (rowId - 1) %/% chunk_size)
# 
# write_dataset(dft2CnvJoinMultipleChangesFiltered,
#               paste0(resDir, "dft2CnvJoinMultipleChangesFiltered/"), 
#               partitioning = "chunk_id")

## All tumorLocationRefAlt overlap with germlinePaired, as they should. 
# checkMulti <- germlinePaired %>% 
#               filter(tumorLocationRefAlt %in% multiplets) %>% 
#               dplyr::select(CHROM, tumorLocationRefAlt) %>%
#   collect()

## Check distribution of adjusted CN per chromosome. Sanity check before running genotyper. 
# check <- open_dataset(paste0(resDir, "dft2CnvJoinMultipleChangesFiltered/")) %>%
#   left_join(germlinePaired %>% 
#               filter(tumorLocationRefAlt %in% multiplets) %>% 
#               dplyr::select(CHROM, tumorLocationRefAlt), 
#             by="tumorLocationRefAlt") %>%
#   dplyr::select(copyNumberAdjusted, CHROM) %>%
#   collect()

# table(check$copyNumberAdjusted, check$CHROM)
#         1     2     3     4     5     6
#   0     3     0   264     0 22680     0
#   2     0  6574     8   683   810     0
#   4     0     0     0  5984     0  3360
#   5     0     0     0   374     0   210
#   6     0     0     0   366     0   210
#   7     0     0     0   366     0   210
```





## 1.6 Allele counter data
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- Adjust names using metadata! 

### 1.6.1 Metadata
- Mapping AlCo results to VCF file names
- FOR now, don't filter where mapping = 1. There are sample names that I am certain match the name in Max's supplementary table. --> do this at the stage where we merge with AlCo results so that everything is genotyped
  - Filter on na in DFT status column, as these we likely highly host contaminated or non-DFT tumors.
```{r}
metadata <- read.xlsx(paste0(resDir,
                          "08_deconvTable_metadata_purityIrodsMergedManualEdit.xlsx")) %>%
              dplyr::rename("tumorId"="TUMOUR.INFORMATION_TUMOUR.ID",
                     "tumorIdRnaSeq"="rnaSeqManual") %>%
  filter(X9_LINEAGE %in% c("DFT1", 
                           "DFT2"))

# table(metadata$X9_LINEAGE)
# 
# DFT1 DFT2 
#   78   41 

metadataPairs <- metadata %>%
  filter(HOST.INFORMATION_HOST.ID!="N/A")

# DFT1 DFT2 
#   63   37 
```

### 1.6.2 Read in AlCo results
- Here, tumorId = the name of the RNAseq file. Need to join this with the metadata table and switch out for the VCF tumor name (Max's supp data.)
- We have AlCo data from 115 tumors, calling this the tumorRnaSeqId. 
  - This can be left joined with the paired tumorId from the metadata.

```{r}
# germlineAlCo <- list.files(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport"),
#                           recursive = T)
# 
# # Read all files and combine
# germlineAlCoDf <- rbindlist(lapply(germlineAlCo, function(f) {
#   # Extract tumor ID between "germlineRefAltExport_" and "_realigned"
#   tumorIdRnaSeq <- sub(".*germlineRefAltExport_(.+)_realigned.*", "\\1", basename(f))
# 
#   # Read file
#   dt <- fread(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport/", f),
#               select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
# 
#   # Add tumor ID column
#   dt[, tumorIdRnaSeq := tumorIdRnaSeq]
# 
#   return(dt)
# }))
# 
# length(unique(germlineAlCoDf$tumorIdRnaSeq)) # 115 RNAseq samples were genotyped.
# 
# length(intersect(germlineAlCoDf$tumorIdRnaSeq, metadata$tumorIdRnaSeq)) # 53 samples for which we have a paired RNAseq-DNAseq file.
# 
# chunk_size <- 5e4  # adjust to your RAM
# 
# ## Left join metadata
# germlineAlCoDf <- germlineAlCoDf %>%
#   mutate(rowId = row_number(),
#          chunk_id = (rowId - 1) %/% chunk_size) 
# 
# write_dataset(
#   germlineAlCoDf,
#   path = paste0(resDir, "germlineAlCoDf/"),
#   format = "parquet",
#   partitioning = "chunk_id"
# )

# Perform filtering and mutate on arrow, less memory intensive.
# Rows left after filter informative samples = 58966002
# germlineAlCoDfFiltered <- open_dataset(paste0(resDir, "germlineAlCoDf/")) %>%
#   filter(tumorIdRnaSeq %in% metadata$tumorIdRnaSeq) %>%
#   left_join(metadata) %>%
#   # Tumor ID = sample ID from germline vcf.
#   mutate(tumorLocationRef = paste0(tumorId, "_", CHROM, ":", POS, "_", REF),
#          tumorLocationRefAlt = paste0(tumorId, "_", CHROM, ":", POS, "_", REF, "_", ALT),
#          sampleTumor = tumorId) %>%
#   dplyr::select(-c(tumorId))
# 
# write_dataset(
#   germlineAlCoDfFiltered,
#   path = paste0(resDir, "germlineAlCoDfFiltered/"),
#   format = "parquet", 
#   partitioning = "sampleTumor"
#   
# )
```

### 1.6.3 Adjust tumorId of RNAseq data to match Max's metadata
- Filter to keep only samples for which we have RNAseq data.
  - Filter tumorId in AlCo for those in metadata$tumorIdRnaSeq
- Left join with metadata table, to get the matched tumorId
- Create new column, tumorId_CHROM_POS_ALT
- Filter for informative AlCo positions in germlinePaired, otherwise there are too many values in the table (slow to work with).
  - 53 / 115 tumor samples have matched RNA (need to check this.)
  - 46/75 samples have an informative variant in the germline vcf file AND a matched RNAseq position. 
  - /46 have at least one variant with TOTAL AlCo > 0. 
```{r}
# setDT(germlineAlCoDf)
# 
# # fast filtering
# germlineAlCoCombined <- germlineAlCoDf[tumorIdRnaSeq %in% metadata$tumorIdRnaSeq] %>%
#   left_join(metadata %>%
#               dplyr::select(tumorId, tumorIdRnaSeq))
# 
# germlineAlCoCombined[, c("tumorLocation", "tumorLocationRef") := .(
#   paste0(tumorId, "_", CHROM, ":", POS),
#   paste0(tumorId, "_", CHROM, ":", POS, "_", REF)
# )]
# 
# saveRDS(germlineAlCoCombined,
#         paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombined.rds"))
# 
# germlineAlCoCombined <- readRDS(paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombined.rds"))
# 
# 
# germlineAlCoCombinedFiltered <- germlineAlCoCombined[tumorLocationAlt %in% germlinePaired$tumorLocationAlt]
# 
# 
# saveRDS(germlineAlCoCombinedFiltered,
#         paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombinedFiltered.rds"))
# 
# rm(germlineAlCoCombined,
#    germlineAlCoDf,
#    germlinePairedFiltered)
# 
# gc()
# 
# germlineAlCoCombinedFiltered <- readRDS(paste0(resDir, 
#                fprefix, 
#                "_germlineAlCoCombinedFiltered.rds")) 
```

# 3 Construct final table
## 3.1 Compute genotypes
- Adjust total CN for X and Y chromosomes in male samples
- In a pure sample, vafH = 0 + (1)*Halt/(0 + 1*Htotal). Therefore, Halt = vafH*Htotal. 
- How do deal with ploidy for X and Y in male samples in 
### 3.1.1 No changes
- If host sample is a male, total CN at sex chromosomes == 1
- If tumor sample is DFT2, total CN at sex chromosomes == 1 or 2 depending on original ploidy.
- Don't filter out any of the tumors! Only do this at the stage where we merge with RNAseq data.
```{r}
# genotypedNoChanges <- open_dataset(paste0(resDir,
#                                       "dft2CnvJoinNoChanges/")) %>%
#   left_join(germlinePuritiesPloidy) %>%
#   filter(sampleTumor %in% dft2$sampleTumor) %>%
#   mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"),
#                              1,
#                              2),
#        nTumorTotal = case_when(
#                     !sampleTumor %in% dft2$sampleTumor & CHROM == "Y" ~ 0,          # DFT1 female, no Y
#                     sampleTumor %in% dft2$sampleTumor & CHROM %in% c("X", "Y") ~ copyNumberAdjusted / 2,  # DFT2 XXYY
#                     TRUE ~ copyNumberAdjusted))
# 
# schema(genotypedNoChanges)
# 
# plan(multisession, workers = availableCores() - 1)
# 
# chunk_ids <- genotypedNoChanges %>%
#   distinct(chunk_id) %>%
#   collect() %>%
#   pull()
# 
# for (cid in chunk_ids) {
# 
#   cat("Processing chunk", cid, "\n")
# 
#   chunk <- genotypedNoChanges %>%
#     filter(chunk_id == cid)  %>%
#     collect()
# 
#   if (nrow(chunk) == 0) next
# 
#   results <- future_pmap(
#     .l = list(
#       chunk$nHostTotal,
#       chunk$nTumorTotal,
#       chunk$hostAlt,
#       chunk$hostTotal,
#       chunk$tumorAlt,
#       chunk$tumorTotal,
#       chunk$purity
#     ),
#     .f = function(hp, tp, ha, ht, ta, tt, p) {
# 
#       g <- genotyper(hp, tp, ha, ht, ta, tt, p)
# 
#       c(g_host = g$E$g_host,
#         g_tumour = g$E$g_tumour,
#         map_g_host = g$map[1, "g_host"],
#         map_g_tumour = g$map[1, "g_tumour"])
#     }
#   )
# 
#   res <- do.call(rbind, results)
# 
#   # chunk$nHostAltGtyr <- res[,1]
#   # chunk$nHostAltGtyrRounded  <- round(res[,1])
#   # chunk$nHostRefGtyrRounded <- chunk$nHostTotal - chunk$nHostAltGtyrRounded
#   # 
#   # chunk$nTumorAltGtyr <- res[,2]
#   # chunk$nTumorAltGtyrRounded  <- round(res[,2])
#   # chunk$nTumorRefGtyrRounded <- chunk$nTumorTotal - chunk$nTumorAltGtyrRounded
# 
#   chunk$nHostAltGtyrMapped <- res[,3]
#   chunk$nHostRefGtyrMapped <- chunk$nHostTotal - chunk$nHostAltGtyrMapped
# 
#   chunk$nTumorAltGtyrMapped <- res[,4]
#   chunk$nTumorRefGtyrMapped <- chunk$nTumorTotal - chunk$nTumorAltGtyrMapped
# 
#   write_parquet(
#     chunk,
#     paste0(resDir, "germlineGenotyperNoChanges/genotyped_chunk_", cid, ".parquet")
#   )
# 
#   rm(chunk, results, res)
#   gc()
# }
# 
# plan(sequential)
# 
# rm(genotypedNoChanges)
# gc()
```


### 3.1.2 Single change
- If host sample is a male, total CN at sex chromosomes == 1
- If tumor sample is DFT2, total CN at sex chromosomes == 1 or 2 depending on original ploidy.
- Don't filter out any of the tumors! Only do this at the stage where we merge with RNAseq data.
```{r}
# genotypedSingleChange <- open_dataset(paste0(resDir,
#                                       "dft2CnvJoinSingleChange/")) %>%
#   left_join(germlinePuritiesPloidy) %>%
#   filter(sampleTumor %in% dft2$sampleTumor) %>%
#   #### Filter germlinePaired on singlet positions ####
#   left_join(germlinePaired %>% filter(tumorLocationRefAlt %in% singlets) %>% dplyr::select(CHROM, tumorLocationRefAlt)) %>%
#   ####
#   mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"),
#                              1,
#                              2),
#        nTumorTotal = case_when(
#                     !sampleTumor %in% dft2$sampleTumor & CHROM == "Y" ~ 0,  # DFT1 female, no Y
#                     sampleTumor %in% dft2$sampleTumor & CHROM %in% c("X", "Y") ~ copyNumberAdjusted / 2,  # DFT2 XXYY
#                     TRUE ~ copyNumberAdjusted))
# 
# schema(genotypedSingleChange)
# 
# plan(multisession, workers = availableCores() - 1)
# 
# chunk_ids <- genotypedSingleChange %>%
#   distinct(chunk_id) %>%
#   collect() %>%
#   pull()
# 
# for (cid in chunk_ids) {
# 
#   cat("Processing chunk", cid, "\n")
# 
#   chunk <- genotypedSingleChange %>%
#     filter(chunk_id == cid)  %>%
#     collect()
# 
#   if (nrow(chunk) == 0) next
# 
#   results <- future_pmap(
#     .l = list(
#       chunk$nHostTotal,
#       chunk$nTumorTotal,
#       chunk$hostAlt,
#       chunk$hostTotal,
#       chunk$tumorAlt,
#       chunk$tumorTotal,
#       chunk$purity
#     ),
#     .f = function(hp, tp, ha, ht, ta, tt, p) {
# 
#       g <- genotyper(hp, tp, ha, ht, ta, tt, p)
# 
#       c(g_host = g$E$g_host,
#         g_tumour = g$E$g_tumour,
#         map_g_host = g$map[1, "g_host"],
#         map_g_tumour = g$map[1, "g_tumour"])
#     }
#   )
# 
#   res <- do.call(rbind, results)
# 
#   # chunk$nHostAltGtyr <- res[,1]
#   # chunk$nHostAltGtyrRounded  <- round(res[,1])
#   # chunk$nHostRefGtyrRounded <- chunk$nHostTotal - chunk$nHostAltGtyrRounded
#   # 
#   # chunk$nTumorAltGtyr <- res[,2]
#   # chunk$nTumorAltGtyrRounded  <- round(res[,2])
#   # chunk$nTumorRefGtyrRounded <- chunk$nTumorTotal - chunk$nTumorAltGtyrRounded
# 
#   chunk$nHostAltGtyrMapped <- res[,3]
#   chunk$nHostRefGtyrMapped <- chunk$nHostTotal - chunk$nHostAltGtyrMapped
# 
#   chunk$nTumorAltGtyrMapped <- res[,4]
#   chunk$nTumorRefGtyrMapped <- chunk$nTumorTotal - chunk$nTumorAltGtyrMapped
# 
#   write_parquet(
#     chunk,
#     paste0(resDir, "germlineGenotyperSingleChange/genotyped_chunk_", cid, ".parquet")
#   )
# 
#   rm(chunk, results, res)
#   gc()
# }
# 
# plan(sequential)
# 
# rm(genotypedSingleChange, dft2CnvJoinSingleChange)
# gc()
```

### 3.1.3 Multiple changes
- If host sample is a male, total CN at sex chromosomes == 1
- If tumor sample is DFT2, total CN at sex chromosomes == 1 or 2 depending on original ploidy.
- Don't filter out any of the tumors! Only do this at the stage where we merge with RNAseq data.
```{r}
# rm(dft2CnvJoinMultipleChanges, dft2CnvJoinMultipleChangesFiltered)
# 
# genotypedMultipleChanges <- open_dataset(paste0(resDir,
#                                       "dft2CnvJoinMultipleChangesFiltered/")) %>%
#   left_join(germlinePuritiesPloidy) %>%
#   filter(sampleTumor %in% dft2$sampleTumor) %>%
#   #### Filter germlinePaired on multiplet positions ####
#   left_join(germlinePaired %>% filter(tumorLocationRefAlt %in% multiplets) %>% 
#               dplyr::select(CHROM, tumorLocationRefAlt)) %>%
#   ####
#   mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"),
#                              1,
#                              2),
#        nTumorTotal = case_when(
#                     !sampleTumor %in% dft2$sampleTumor & CHROM == "Y" ~ 0,  # DFT1 female, no Y
#                     sampleTumor %in% dft2$sampleTumor & CHROM %in% c("X", "Y") ~ copyNumberAdjusted / 2,  # DFT2 XXYY
#                     TRUE ~ copyNumberAdjusted)) 
# 
# # schema(genotypedMultipleChanges)
# 
# plan(multisession, workers = availableCores() - 1)
# 
# chunk_ids <- genotypedMultipleChanges %>%
#   distinct(chunk_id) %>%
#   collect() %>%
#   pull()
# 
# for (cid in chunk_ids) {
# 
#   cat("Processing chunk", cid, "\n")
# 
#   chunk <- genotypedMultipleChanges %>%
#     filter(chunk_id == cid)  %>%
#     collect()
# 
#   if (nrow(chunk) == 0) next
# 
#   results <- future_pmap(
#     .l = list(
#       chunk$nHostTotal,
#       chunk$nTumorTotal,
#       chunk$hostAlt,
#       chunk$hostTotal,
#       chunk$tumorAlt,
#       chunk$tumorTotal,
#       chunk$purity
#     ),
#     .f = function(hp, tp, ha, ht, ta, tt, p) {
# 
#       g <- genotyper(hp, tp, ha, ht, ta, tt, p)
# 
#       c(g_host = g$E$g_host,
#         g_tumour = g$E$g_tumour,
#         map_g_host = g$map[1, "g_host"],
#         map_g_tumour = g$map[1, "g_tumour"])
#     }
#   )
# 
#   res <- do.call(rbind, results)
# 
#   # chunk$nHostAltGtyr <- res[,1]
#   # chunk$nHostAltGtyrRounded  <- round(res[,1])
#   # chunk$nHostRefGtyrRounded <- chunk$nHostTotal - chunk$nHostAltGtyrRounded
#   # 
#   # chunk$nTumorAltGtyr <- res[,2]
#   # chunk$nTumorAltGtyrRounded  <- round(res[,2])
#   # chunk$nTumorRefGtyrRounded <- chunk$nTumorTotal - chunk$nTumorAltGtyrRounded
# 
#   chunk$nHostAltGtyrMapped <- res[,3]
#   chunk$nHostRefGtyrMapped <- chunk$nHostTotal - chunk$nHostAltGtyrMapped
# 
#   chunk$nTumorAltGtyrMapped <- res[,4]
#   chunk$nTumorRefGtyrMapped <- chunk$nTumorTotal - chunk$nTumorAltGtyrMapped
# 
#   write_parquet(
#     chunk,
#     paste0(resDir, "germlineGenotyperMultipleChanges/genotyped_chunk_", cid, ".parquet")
#   )
# 
#   rm(chunk, results, res)
#   gc()
# }
# 
# plan(sequential)
# 
# rm(genotypedMultipleChanges, dft2CnvJoinMultipleChanges)
# gc()
```
### 3.1.4 Merge
```{r}
# noChanges <- open_dataset(paste0(resDir, "germlineGenotyperNoChanges")) 
# singleChange <- open_dataset(paste0(resDir, "germlineGenotyperSingleChange"))
# multipleChanges <- open_dataset(paste0(resDir, "germlineGenotyperMultipleChanges"))
# 
# cols <- intersect(colnames(schema(noChanges) %>% as.data.frame()),
#                             colnames(schema(multipleChanges) %>% as.data.frame()))
# 
# colsOfInterest <- setdiff(cols,
#                           c("tumorLocation", 
#                             "tumorLocationAlt", 
#                             "chunk_id", 
#                             "rowId")
#                           )
# 
# noChanges <- open_dataset(paste0(resDir, "germlineGenotyperNoChanges")) %>% dplyr::select(all_of(colsOfInterest)) %>% collect()
# singleChange <- open_dataset(paste0(resDir, "germlineGenotyperSingleChange")) %>% dplyr::select(all_of(colsOfInterest)) %>% collect()
# multipleChanges <- open_dataset(paste0(resDir, "germlineGenotyperMultipleChanges")) %>% dplyr::select(all_of(colsOfInterest)) %>% collect()
# 
# chunk_size <- 5e4  # adjust to your RAM
# 
# germlineGenotyper <- bind_rows(
#   noChanges,
#   singleChange,
#   multipleChanges
# ) %>%
#   mutate(rowId = row_number(),
#          chunk_id = (rowId - 1) %/% chunk_size) 
# 
# write_dataset(germlineGenotyper,
#               paste0(resDir, "germlineGenotyperDft2/"),
#               partitioning = "chunk_id")


# germlinePairedCheckRows <- germlinePaired %>%
#   dplyr::select(tumorLocationRefAlt) %>%
#   collect()

# nrow(germlinePairedCheckRows$tumorLocationRefAlt)==nrow(germlineGenotyper$tumorLocationRefAlt)
```

# --- BREAK HERE ---
# 4 Majority genotype - Chr 1 to 6
```{r}
chroms <- c(paste0(1:6), "X", "Y")
```

## 4.1 Assign majority genotype 
### 4.1.1 Find multiallelic SNPs
```{r}
germlineGenotyper <- open_dataset(paste0(resDir, "germlineGenotyperDft2/"))
 
# nAltAlleles <- germlineGenotyper %>%
#   dplyr::select(locationRef, ALT, sampleTumor) %>%
#   distinct() %>%
#   collect() %>%
#   group_by(locationRef) %>%
#   summarise(
#     nAltAlleles = n_distinct(ALT),
#     altAlleles = paste(sort(unique(ALT)), collapse = ","),
#     baseA = ifelse("A" %in% ALT, paste(sampleTumor[ALT == "A"], collapse = ","), NA),
#     baseC = ifelse("C" %in% ALT, paste(sampleTumor[ALT == "C"], collapse = ","), NA),
#     baseG = ifelse("G" %in% ALT, paste(sampleTumor[ALT == "G"], collapse = ","), NA),
#     baseT = ifelse("T" %in% ALT, paste(sampleTumor[ALT == "T"], collapse = ","), NA),
#     .groups = "drop"
#   )
# 
# write_parquet(nAltAlleles,
#               paste0(resDir, "dft2Cn2_nAltAlleles.parquet"))

nAltAlleles <- read_parquet(paste0(resDir,  "dft2Cn2_nAltAlleles.parquet")) 

# Are hom ref/alt or het skewed by multiallelic sites?
biallelicData <- nAltAlleles %>%
  filter(nAltAlleles == 1) %>%
  distinct() %>%
  collect()

biallelic <- biallelicData %>%
  pull(locationRef)

# 134 sites. --> explore in IGV
multiallelicData <- nAltAlleles %>%
  filter(nAltAlleles > 1) %>%
   distinct() %>%
  collect()

multiallelic <- multiallelicData %>%
  pull(locationRef)

# Why do all tumors get repeated when a multiallelic site?. All have very low read counts supporting alt allele. Using genotyper, none of these positions exceed 0 in germline table. Do all tumors have both copies? Probably, all repeated in the summary table. 
# testTwo <- germlinePairedGenotypedStatus %>%
#   filter(locationRef %in% multiallelic) %>%
#   collect()

rm(nAltAlleles, biallelicData, multiallelicData)

# germlinePairedMultiallelic <- open_dataset(paste0(resDir, "germlineGenotyperDft2/")) %>%
#   filter(CHROM %in% chroms) %>%
#   mutate(locationRef = paste0(CHROM, ":", POS, "_", REF)) %>%
#   filter(locationRef %in% multiallelic) %>%
#   collect()
# 
# write.xlsx(germlinePairedMultiallelic, 
#            paste0(resDir, fprefix, "_germlinePairedMultiallelic.xlsx"))
# germlinePairedMultiallelicCheck <- germlinePairedMultiallelic %>%
#   filter(obsVafTumor > .20)
# 
# length(unique(germlinePairedMultiallelicCheck$locationRef))
```

### 4.1.2 CN >= 2, all chromosomes, in all tumors. 
- Autosomal
- Positions are biallelic
- Filter for SNPs positions where CN >= 2 in all tumors: set as het, homAlt, or homRef. Using genotyper(), but will eventually use vaf windows defined in Kevin's supp methods. 
- Unclassified are SNPs falling in segments of CN 1 or 0. 
#### 4.1.2.1 Annotate
- Annotate each location with the majority (>= 50% status) flipping status
```{r}
# # # Find positions where all tumors have CN >= 2
# # positionsCn2All <- germlineGenotyper %>%
# #   filter(sampleTumor %in% dft2$sampleTumor & CHROM %in% chroms & locationRef %in% biallelic) %>%
# #   group_by(locationRef) %>%
# #   summarise(minTumorCn = min(nTumorTotal, na.rm = TRUE),
# #             .groups = "drop") %>%
# #   filter(minTumorCn >= 2) %>%
# #   dplyr::select(locationRef, minTumorCn) %>%
# #   collect()
# #
# # gc()
# #
# # length(unique(positionsCn2All$locationRef))
# #
# # gc()
# 
# genotypeDft2Cn2All <- germlineGenotyper %>%
#   ######
#   filter(sampleTumor %in% dft2$sampleTumor) %>%
#   ######
#   mutate(
#     status = case_when(
#       nTumorTotal >= 2 & nTumorRefGtyrMapped == nTumorTotal & nTumorAltGtyrMapped == 0 ~ "homRef",
#       nTumorAltGtyrMapped == nTumorTotal & nTumorRefGtyrMapped == 0                   ~ "homAlt",
#       nTumorTotal >= 2 & nTumorRefGtyrMapped > 0 & nTumorRefGtyrMapped < nTumorTotal &
#         nTumorAltGtyrMapped > 0 & nTumorAltGtyrMapped < nTumorTotal                  ~ "het",
#       nTumorTotal < 2                                                                  ~ "unclassified",
#       TRUE                                                                             ~ "toCheck"
#     ),
#     statusBroad = case_when(
#       status %in% c("homRef", "homAlt") ~ "hom",
#       TRUE                              ~ status
#     )
#   ) %>%
#   dplyr::select(locationRef,
#                 sampleTumor,
#                 status,
#                 statusBroad,
#                 CHROM,
#                 POS,
#                 ploidy,
#                 nTumorTotal) %>%
#   collect()
# 
# # checkStatus <- genotypeDft2Cn2All %>%
# #   filter(is.na(status)==T) # No rows.
# # checkStatus <- genotypeDft2Cn2All %>%
# #    filter(status == "toCheck") # No rows.
# # checkStatusBroad <- genotypeDft2Cn2All %>%
# #   filter(is.na(statusBroad)==T) # No rows.
# # checkStatusBroad <- genotypeDft2Cn2All %>%
# #   filter((statusBroad)=="toCheck")
# 
# # table(genotypeDft2Cn2All$nTumorTotal,
# #       genotypeDft2Cn2All$status)
#   #       het   homAlt   homRef unclassified
#   # 0        0        0        0        22947
#   # 1        0        0        0      2111527
#   # 2  9527753  4840215 18126541            0
#   # 3   365648   184319   629631            0
#   # 4   497817   169053   731135            0
#   # 5    14473     6714    25107            0
#   # 6       29      110      437            0
#   # 7       46       94      436            0
# # 
# # table(genotypeDft2Cn2All$nTumorTotal,
# #       genotypeDft2Cn2All$statusBroad)
# 
# nTumorsTotal = length(unique(genotypeDft2Cn2All$sampleTumor))
# 
# majorityDft2Status <- genotypeDft2Cn2All %>%
#   # # Classify het using purity-corrected VAF intervals
#   # mutate(
#   #   status = case_when(
#   #     nTumorTotal < 2                                                            ~ "unclassified",
#   #     nTumorTotal == 2 & x > 0.25  & x < 0.75  ~ "het",
#   #     nTumorTotal == 3 & x > 0.2   & x < 0.8   ~ "het",
#   #     nTumorTotal == 4 & x > 0.175 & x < 0.825 ~ "het",
#   #     nTumorTotal == 5 & x > 0.15  & x < 0.85  ~ "het",
#   #     nTumorTotal == 6 & x > 0.12  & x < 0.88  ~ "het",
#   #     nTumorTotal >= 7 & x > 0.1   & x < 0.9   ~ "het",
#   #     TRUE                                                                       ~ "hom"
#   #   ),
#   #   statusBroad = case_when(
#   #     status == "het" ~ "het",
#   #     status == "hom" ~ "hom",
#   #     TRUE            ~ status
#   #   )
#   # ) %>%
#   # Remove unclassified SNPs
#   filter(locationRef %in% biallelic & statusBroad != "unclassified") %>%
#   # Calculate majority status per SNP
#   group_by(locationRef) %>%
#   mutate(
#     nHet       = sum(statusBroad == "het"),
#     nHom       = sum(statusBroad == "hom"),
#     nTumoursTotal = nTumorsTotal,
#     majorityStatus = case_when(
#       nHet > nHom ~ "het",
#       nHom > nHet ~ "hom",
#       TRUE        ~ "tied"
#     )
#   ) %>%
#   ungroup()
# 
# table(majorityDft2Status$majorityStatus)
# 
# # > table(majorityDft2Status$majorityStatus)
# #
# #      het      hom     tied
# # 10305740 26911891     5126
# 
# # Step 3: Annotate differences. Rejoin original table, because not all SNP's are classified.
# genotypeDft2Cn2AnnoAll <- genotypeDft2Cn2All %>%
#   filter(CHROM %in% c(paste0(1:6), "X", "Y") & locationRef %in% biallelic) %>%
#   left_join(majorityDft2Status) %>%
#   mutate(majorityCheck = case_when(statusBroad == "unclassified" ~ "unclassified",
#                                    majorityStatus == "tied" ~ "tied",
#                                    # when majority matches genotype of SNP, == 1
#                                    majorityStatus == statusBroad ~ "match",
#                                    # when majority does not match genotype of SNP, flipping snp
#                                    majorityStatus != statusBroad ~ "flip",
#                                    TRUE ~ "toCheck"
#     ))
# 
# table(genotypeDft2Cn2AnnoAll$majorityCheck)
# 
# rm(genotypeDft2Cn2All)
# gc()
# 
# table(genotypeDft2Cn2AnnoAll$CHROM,
#       genotypeDft2Cn2AnnoAll$majorityCheck)
# 
# write_parquet(genotypeDft2Cn2AnnoAll,
#               paste0(resDir, fprefix, "_genotypeDft2Cn2AnnoAll.parquet"))
# 
# genotypeDft2Cn2AnnoAll <- open_dataset(paste0(resDir, 
#                                               fprefix,
#                                         "_genotypeDft2Cn2AnnoAll.parquet"))
```

#### 4.1.2.2 Plot - biallelic non-flagged
```{r}
# cladeOrder <- genotypeDft2Cn2AnnoAll %>%
#   left_join(metadata %>% 
#             dplyr::rename(sampleTumor=tumorId) %>%
#               dplyr::select(sampleTumor, 
#                             HOST.SEX, 
#                             X10_CLADE)) %>%
#   distinct(sampleTumor, X10_CLADE) %>%
#   dplyr::select(sampleTumor, X10_CLADE) %>%
#   arrange(X10_CLADE, sampleTumor) %>%
#   collect() %>%
#   pull(sampleTumor)
```
##### 4.1.2.2.1 Histograms
- X axis = # non-majority SNPs
- What does the distribution look like? I.e., are there many positions with only 1 or 2 flipping SNPs?
```{r}
# Count the number of tumors where status was not equal to majorityStatus
dfPlotAll <- genotypeDft2Cn2AnnoAll %>%
  filter(nTumorTotal >= 2 & majorityCheck == "flip" & locationRef %in% biallelic) %>%
  # CHROM is redundant but useful for plotting.
  group_by(locationRef,
           CHROM) %>%
  summarise(nTumors = n_distinct(sampleTumor),
            .groups = "drop") 

histoData <- dfPlotAll %>%
  ungroup()

# max(dfPlotAll$nTumors)
# 
# histoAll <- ggplot(dfPlotAll,
#                    aes(x = nTumors)) +
#   geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
#   labs(
#     x = "Number of DFT2 tumors",
#     y = "Unique non-majority positions (locationRef)",
#     title = "Biallelic Positions with CN >= 2 (All DFT2 Tumors)"
#   ) +
#   theme_minimal()
# 
# histoAllChrom <- ggplot(dfPlotAll,
#                    aes(x = nTumors,
#                        fill = CHROM)) +
#   geom_histogram(binwidth = 1,
#                  alpha = 0.75) +
#   labs(
#     x = "Number of DFT2 tumors",
#     y = "Unique non-majority positions (locationRef)",
#     title = "Biallelic Positions with CN >= 2 (All DFT2 Tumors)"
#   ) +
#   theme_minimal()
# 
# histoSplitChromFacet <- ggplot(dfPlotAll,
#                    aes(x = nTumors,
#                        fill = CHROM)) +
#   geom_histogram(binwidth = 1,
#                  alpha = 0.75) +
#   facet_wrap(~CHROM,
#              scales = "fixed",
#              ncol = 1) +
#   labs(
#     x = "Number of DFT2 tumors",
#     y = "Unique non-majority positions (locationRef)",
#     title = "Biallelic Positions with CN >= 2 (All DFT2 Tumors)"
#   ) +
#   theme_minimal()
# 
# 
# histoAllChromZoom <- ggplot(dfPlotAll %>% filter(nTumors > 3),
#                    aes(x = nTumors,
#                        fill = CHROM)) +
#   geom_histogram(binwidth = 1,
#                  alpha = 0.75) +
#   labs(
#     x = "Number of DFT2 tumors (n > 3)",
#     y = "Unique non-majority positions (locationRef)",
#     title = "Biallelic Positions with CN >= 2 (All DFT2 Tumors)"
#   ) +
#   theme_minimal()
# 
# 
# pdf(paste0(histoDir, fprefix, "_distributionFlippedSnps_biallelicCN2AllDft2.pdf"),
#     height = 4,
#     width = 8)
# 
# print(histoAll)
# print(histoAllChrom)
# print(histoAllChromZoom)
# 
# dev.off()
# 
# pdf(paste0(histoDir, fprefix, "_distributionFlippedSnps_biallelicCN2AllDft2_facet.pdf"),
#     height = 10,
#     width = 8)
# 
# print(histoSplitChromFacet)
# 
# dev.off()
# 
# write.xlsx(histoAll$data %>% as.data.frame() %>% mutate(varType = "biallelicCN2AllDft2"),
#            paste0(fprefix, "_distributionFlippedSnps_biallelicCN2AllDft2.xlsx"))
```


##### 4.1.2.2.2 Density
- Method from Kevin's paper. 
- SNPs with CN >= 2
- Density of flipping SNPs (proportion of flip/non-flipped, excluding ties & unclassified. )
```{r}
# # Calculate flipping SNP density for each tumour
# flippingDensityDft2 <- genotypeDft2Cn2AnnoAll %>%
#   # Consider only SNPs at CN >= 2
#   filter(nTumorTotal >= 2 & !majorityCheck %in% c("tied", "toCheck")) %>%
#   # Work per tumour
#   group_by(sampleTumor) %>%
#   arrange(CHROM, POS, .by_group = TRUE) %>%
#   mutate(
#     snpIndex = row_number(),
#     # For each SNP, calculate proportion of flipping SNPs in window of 500 preceding + 500 following + current (1001 total)
#     flippingDensity = slider::slide_dbl(
#       majorityCheck == "flip",
#       mean,
#       .before = 500,
#       .after  = 500,
#       .complete = FALSE
#     )
#   ) %>%
#   # Calculate Cook's distance for each SNP within each tumour
#   mutate(
#     # Fit linear model of flipping density ~ snp index to get Cook's distances
#     cooksD = {
#       df <- data.frame(y = flippingDensity, x = snpIndex)
#       mod <- lm(y ~ x, data = df)
#       cooks.distance(mod)
#     },
#     meanCooksD = mean(cooksD, na.rm = TRUE),
#     # Peak selection: Cook's distance > 30x mean Cook's distance
#     cooksPeak = cooksD > 30 * meanCooksD,
#     # Additional filter: density > median + 5*MAD
#     medianDensity = median(flippingDensity, na.rm = TRUE),
#     madDensity    = mad(flippingDensity, na.rm = TRUE),
#     madPeak       = flippingDensity > medianDensity + 5 * madDensity,
#     # Final peak: must pass both filters
#     isPeak = cooksPeak & madPeak
#   ) %>%
#   ungroup()
# 
# # Extract peak SNPs
# peaksDft2 <- flippingDensityDft2 %>%
#   filter(isPeak)
# 
# for (chr in unique(flippingDensityDft2$CHROM)) {
#   
#   tempDf <- flippingDensityDft2 %>%
#     filter(CHROM == chr, nTumorTotal >= 2) %>%
#     mutate(sampleTumor = factor(sampleTumor, levels = cladeOrder), 
#            POS = as.numeric(POS)) %>%
#     left_join(metadata %>%
#                 dplyr::select(X10_CLADE, tumorId) %>%
#                 dplyr::rename(sampleTumor = tumorId))
#   
#   tempPeaks <- peaksDft2 %>%
#     filter(CHROM == chr) %>%
#     mutate(sampleTumor = factor(sampleTumor, levels = cladeOrder), 
#            POS = as.numeric(POS)) %>%
#     left_join(metadata %>%
#                 dplyr::select(X10_CLADE, tumorId) %>%
#                 dplyr::rename(sampleTumor = tumorId))
#   
#   xmax <- max(tempDf$POS, na.rm = TRUE)
#   
#   tempPlot <- ggplot(tempDf,
#       aes(x = POS, 
#           y = flippingDensity, 
#           group = sampleTumor, 
#           colour = X10_CLADE)) +
#     # ── sliding-window density lines ─────────────────────────────────────────
#     geom_line(alpha = 0.6, linewidth = 0.4) +
#     # ── peak points ──────────────────────────────────────────────────────────
#     geom_point(
#       data        = tempPeaks,
#       aes(x = POS, y = flippingDensity),
#       inherit.aes = FALSE,
#       colour      = "firebrick",
#       shape       = 25,
#       size        = 1.2,
#       stroke      = 0.3
#     ) +
#     facet_grid(sampleTumor ~ ., scales = "fixed") +
#     labs(
#       x        = "Genomic position",
#       y        = "Flipping SNP density (proportion in 1001 SNP window)",
#       title    = "Flipping SNPs — DFT2 Tumours",
#       subtitle = paste0("Chromosome: ", chr)
#     ) +
#     scale_x_continuous(
#       breaks = seq(0, xmax, by = 1e7),
#       labels = scales::label_number(scale = 1e-6, suffix = " Mb")
#     ) +
#     scale_y_continuous(limits = c(0, 1)) +
#     theme_bw() +
#     theme(
#       strip.text.y     = element_text(size = 6, angle = 0),
#       axis.text.x      = element_text(angle = 45, hjust = 1, size = 6),
#       axis.text.y      = element_text(size = 5),
#       panel.spacing    = unit(0.2, "lines")
#     )
#   
#   pdf(
#     paste0(barDir, 
#            fprefix, 
#            "_chr", 
#            chr,
#            "_flippedSnps_biallelicCN2AllDft2_density_peaks.pdf"),
#     height = 40, width = 15
#   )
#   print(tempPlot)
#   dev.off()
#   
#   message("Plotted chromosome: ", chr)
# }
```


##### 4.1.2.2.3 Sliding window - idx = snp position. 
```{r}
# # # Step 1: Prepare the data - flag flipping SNPs and get position
# # allSNPs <- genotypeDft2Cn2AnnoAll %>%
# #   filter(majorityStatus != "flag") %>%
# #   mutate(POS = as.numeric(POS)) %>%
# #   arrange(sampleTumor,
# #           CHROM,
# #           POS) %>%
# #   left_join(metadata %>%
# #             dplyr::rename(sampleTumor=tumorId) %>%
# #               dplyr::select(sampleTumor,
# #                             HOST.SEX,
# #                             X10_CLADE))
# # 
# # write_parquet(allSNPs,
# #               paste0(resDir, fprefix, "_dft2Cn2_allSNPs.parquet"))
# # 
# # # Step 2: Sliding window per tumor per chromosome
# # windowedFlips <- allSNPs %>%
# #   group_by(sampleTumor, CHROM) %>%
# #   arrange(POS) %>%
# #   mutate(
# #     # Slide index allows for irregular SNP positions (i.e., as they cluster in non-uniform manner), indexing by position rather than a set number of Mb.
# #     windowCount = slide_index_dbl(
# #       majorityCheck,           # counts
# #       .i = POS,                # index by position
# #       .f = ~ sum(!.x),         # count majority status == F (number of flipping SNPs) per window
# #       .before = 50000,         # 50kb each side = 100kb window
# #       .after  = 50000
# #     )) %>%
# #   ungroup()
# # 
# # dim(windowedFlips)
# # 
# # write_parquet(windowedFlips,
# #               paste0(resDir, fprefix, "_dft2Cn2_windowedFlips.parquet"))
# 
# allSNPS <- read_parquet(paste0(resDir, fprefix, "_dft2Cn2_allSNPs.parquet"))
# 
# windowedFlips <- read_parquet(paste0(resDir, fprefix, "_dft2Cn2_allSNPs.parquet"))
# 
# gc()
# 
# windowedFlipsFiltered <- windowedFlips %>%
#   filter(windowCount > 0)
# 
# rm(windowedFlips)
# 
# dim(windowedFlipsFiltered)  
```

##### 4.1.2.2.4 Plot sliding window
```{r}
library(pracma)
library(zoo)

# ── 1. Extract peaks per tumor per chromosome ─────────────────────────────────
peaks <- windowedFlipsFiltered %>%
  group_by(sampleTumor, CHROM) %>%
  arrange(POS) %>%
  group_modify(~ {
    p <- findpeaks(
      .x$windowCount,
      minpeakheight   = mean(.x$windowCount, na.rm = TRUE) + 2 * sd(.x$windowCount, na.rm = TRUE),
      minpeakdistance = 10
    )
    if (is.null(p)) return(tibble())
    tibble(
      peak_idx    = p[, 2],
      windowCount = p[, 1]
    ) %>%
      mutate(POS = .x$POS[peak_idx])
  }) %>%
  ungroup()


# ── 2. Bin peaks into 500kb windows, count tumors sharing each bin ────────────
bin_size <- 500000  # adjust as needed

sharedPeaks <- peaks %>%
  mutate(bin = floor(POS / bin_size) * bin_size) %>%
  group_by(CHROM, bin) %>%
  summarise(
    n_tumors    = n_distinct(sampleTumor),
    tumor_list  = list(unique(sampleTumor)),
    mean_count  = mean(windowCount),
    .groups     = "drop"
  ) %>%
  filter(n_tumors > 1)   # only keep bins shared by 2+ tumors


# ── 3. Join clade info back onto peaks for plotting ───────────────────────────
# (assumes windowedFlips has X10_CLADE and sampleTumor columns)
cladeKey <- windowedFlipsFiltered %>%
  distinct(sampleTumor, X10_CLADE)

peaksWithClade <- peaks %>%
  left_join(cladeKey, by = "sampleTumor") %>%
  mutate(sampleTumor = factor(sampleTumor, levels = cladeOrder))

sharedBins <- sharedPeaks %>%
  mutate(
    bin_start = bin,
    bin_end   = bin + bin_size
  )

# ── 4. Plot per chromosome ────────────────────────────────────────────────────
for (chr in unique(windowedFlipsFiltered$CHROM)) {

  tempDf <- windowedFlipsFiltered %>%
    filter(CHROM == chr) %>%
    # left_join(cladeKey, by = "sampleTumor") %>%
    mutate(sampleTumor = factor(sampleTumor, 
                                levels = cladeOrder))

  tempPeaks  <- peaksWithClade  %>% filter(CHROM == chr)
  tempShared <- sharedBins      %>% filter(CHROM == chr)

  xmax <- max(tempDf$POS, na.rm = TRUE)

  tempPlotClade <- ggplot(tempDf,
      aes(x = POS, y = windowCount, colour = X10_CLADE)) +
    # ── shared-peak shading ──────────────────────────────────────────────────
    geom_rect(
      data        = tempShared,
      aes(xmin = bin_start, xmax = bin_end,
          ymin = -Inf,      ymax = Inf,
          alpha = n_tumors),
      inherit.aes = FALSE,
      fill        = "grey50"
    ) +
    scale_alpha_continuous(
      name   = "Tumors sharing peak",
      range  = c(0.10, 0.35)
    ) +
    # ── sliding-window lines ─────────────────────────────────────────────────
    geom_line(alpha = 0.6, linewidth = 0.4) +
    # ── individual peak points ───────────────────────────────────────────────
    geom_point(
      data        = tempPeaks,
      aes(x = POS, 
          y = windowCount, 
          colour = X10_CLADE),
      inherit.aes = FALSE,
      shape = 25, size = 1.2, stroke = 0.3
    ) +
    facet_grid(sampleTumor ~ CHROM, scales = "fixed") +
    labs(
      x        = "Genomic position",
      y        = "Flipping SNPs in 100kb window",
      colour   = "Clade",
      title    = "Flipping SNPs — DFT2 Tumors",
      subtitle = paste0("Chromosome: ", chr,
                        "  |  Shading = bins shared by ≥2 tumors")
    ) +
    scale_x_continuous(
  breaks = seq(0, xmax, by = 1e7),
  labels = scales::label_number(scale = 1e-3, suffix = " kb")
)+
    theme(
      legend.position  = "right",
      strip.text.x     = element_text(size = 7),
      axis.text.x      = element_text(angle = 45, hjust = 1, size = 6),
      axis.text.y      = element_text(size = 4)
    )

  pdf(
    paste0(barDir, "germlineDft2/biallelicCn2_countFlips/",
           fprefix, "_chr", chr,
           "_flippedSnps_biallelicCN2AllDft2_slidingWindow100Kb_sharedPeaks.pdf"),
    height = 40, width = 35
  )
  print(tempPlotClade)
  dev.off()
}
```

## 4.2 Correct
- Approach with Liz
  - For non-majority SNP's (CN >= 2), correct genotype to majority. Keep info on what was the originally assigned genotype. Also keep info on how many tumors had this flip. 
  - For the tied SNPs, exclude?
  - For the unclassified SNPs, keep as is? These could be genuine variants?
```{r}
genotypeDft2Cn2AnnoAll <- open_dataset(paste0(resDir, 
                                              fprefix,
                                              "_genotypeDft2Cn2AnnoAll.parquet")) %>%
  left_join(histoData %>%
              dplyr::select(locationRef, 
                            nTumors)) %>%
  collect()
```



# --- BREAK HERE ---
# --- SEE GERMLINE SCRIPTS ---
### 3.1.6 Load genotyper results 
- Filter on conditions of interest
```{r}
germlinePairedInformative<- open_dataset(paste0(resDir, "germlineGenotyper/")) %>%
  filter(
    (nTumorAltGtyrMapped > 0 & nHostAltGtyrMapped == 0) |
    (nTumorRefGtyrMapped > 0 & nHostRefGtyrMapped == 0) |
    (nHostAltGtyrMapped > 0 & nTumorAltGtyrMapped == 0) |
    (nHostRefGtyrMapped > 0 & nTumorRefGtyrMapped == 0)
  ) 

# > nrow(germlinePairedInformative)
# [1] 21168207

germlinePairedInformative <- write_dataset(germlinePairedInformative,
                                           path = paste0(resDir,
                                                         "germlineInformative/"),
  format = "parquet",
  partitioning = "sampleTumor")


gc()

germlinePairedInformative <- open_dataset(paste0(resDir, "germlineInformative/")) %>%
  collect()
```

## 3.3 Merge tumor AlCo with germline vcf information.
- Join filtered vcf file with AlCo results from tumor RNAseq.
- Of note, NOT all samples have an associated RNA-seq file! See metadata script.
- Filtering AlCoDf for any rows where expression > 0. If an informative position has 0 RNAseq reads assigned, the expression of the informative allele is 0. 

dim(joined) --> i.e., 2 million germline variants with an informative allele for which we also have RNAseq data.
[1] 21164054       33
```{r}
germlinePairedInformativeMatched <- open_dataset(paste0(resDir,
                                                        "germlineInformative/")) %>%
  filter(sampleTumor %in% metadata$tumorId)

# 58,966,002 rows.
germlineAlCoDfFiltered <- open_dataset(paste0(resDir, "germlineAlCoDfFiltered/")) %>%
  dplyr::select(c(tumorLocationAlt,
                  tumorIdRnaSeq, 
                  TOTAL,
                  NREF, 
                  NALT)) 

joined <- germlinePairedInformativeMatched %>%
  left_join(germlineAlCoDfFiltered, 
            by = "tumorLocationAlt") %>%
  mutate(POS = as.numeric(POS))

write_dataset(joined, 
              paste0(resDir, "germlineAlCoVarsJoined/"),
              format = "parquet",
              partitioning = "sampleTumor")

# [1] 1346513
# joinedInform <- open_dataset(paste0(resDir, "germlineAlCoVarsJoined/")) %>%
#   filter(TOTAL > 0) %>%
#   collect()
# 
# nrow(joinedInform)
```

## 3.3 GTF annotations
### 3.3.1 Annotate CHROM:POS
- From informative file, get all CHROM and POS. 
- Merge with gtf file on informative positions (exons)

```{r}
# > nrow(joinedChromPos)
# [1] 1018330

joinedChromPos <- joined %>%
  dplyr::select(CHROM, POS) %>%
  distinct() %>%
  collect()

gc()

gtfFile <- "~/gitClones/01_Deconvolution/02_Deconvolution_DFT1/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf"

gtf <- import(gtfFile)

genes <- gtf[gtf$type == "gene"]

exons <- gtf[gtf$type == "exon"]

genes <- genes[genes$gene_biotype == "protein_coding"]

variants <- GRanges(
  seqnames = joinedChromPos$CHROM,
  ranges = IRanges(start = joinedChromPos$POS, 
                   end = joinedChromPos$POS)
)

# Which variants fall in an exon
exonHits <- findOverlaps(variants, exons)

# Map exon hits to gene IDs
exonGeneIds <- exons$gene_id[subjectHits(exonHits)]
exonGeneNames <- exons$gene_name[subjectHits(exonHits)]
exonBiotypes <- exons$gene_biotype[subjectHits(exonHits)]

hitsDf <- data.frame(
  query = queryHits(exonHits),
  gene_id = exonGeneIds,
  gene_name = exonGeneNames,
  biotype = exonBiotypes
)

collapseFun <- function(x) paste(unique(x), collapse = ";")

collapsed <- aggregate(. ~ query, 
                       data = hitsDf, 
                       FUN = collapseFun)

collapsed$CHROM <- joinedChromPos$CHROM[collapsed$query]
collapsed$POS   <- joinedChromPos$POS[collapsed$query]
```

### 3.3.2 Join with AlCo + Germline Info
```{r}
joinedAnno <- open_dataset(paste0(resDir, "germlineAlCoVarsJoined/")) %>%
  left_join(collapsed)

write_dataset(joinedAnno, 
              paste0(resDir, "germlineAlCoVarsJoinedAnno/"),
              format = "parquet",
              partitioning = "sampleTumor")

# Checking. All positions should have been queried, thus should return 0 rows? No, returns 20815775 rows, because exon hits only contains variant hits. 
# nrow(joinedAnnoFiltered %>% filter(is.na(query)==T))
# [1] 348,279 rows in an informative regions!
# joinedAnnoFiltered <- joinedAnno %>%
#   collect() %>%
#   filter(is.na(query)==F)
```

# 4. Normalize nRef and nAlt
## 4.1 Calculate size factors
- Of note, sample 1538T1 was not filtered out properly, we don't have matched RNAseq data for this sample. Excluding here but need to go back and filter this. 
```{r}
joinedAnnoFiltered <- open_dataset(paste0(resDir, "germlineAlCoVarsJoinedAnno/")) %>%
  mutate(locationAlt = paste0(CHROM, ":", POS, "_", REF, ALT)) %>%
  filter(is.na(query)==F) %>%
  filter(is.na(TOTAL)==F) %>%
  collect()

# nrow(joinedAnnoFiltered)
# [1] 341547
```

### 4.1.1 Calculate size factors for all variant positions
- Using TOTAL at each variant position, compute size factors per sample. 
```{r}
dt <- copy(as.data.table(joinedAnnoFiltered))

dt[, samplename := sampleTumor]

# remove variants with zero coverage everywhere
dt[, empty := sum(TOTAL, na.rm = TRUE) == 0, by = locationAlt]
variants <- dt[empty == FALSE]

# geometric means per variant
gmeans <- variants[, .(gmean = geomean(TOTAL + 1)), by = locationAlt]

setkey(gmeans, locationAlt)
setkey(variants, locationAlt)

variants[gmeans, gmean := i.gmean]

# ratios
variants[, scaled_total := TOTAL / gmean]

# size factors
size_factors <- variants[, .(
  size_factor = median(scaled_total, na.rm = TRUE)
), by = samplename]

setkey(size_factors, samplename)

dt[size_factors, on = "samplename", size_factor := i.size_factor]

dt[, `:=`(
  NREF_norm = NREF / size_factor,
  NALT_norm = NALT / size_factor,
  TOTAL_norm = TOTAL / size_factor
)]

dt <- dt %>%
  mutate(variantType = "germline")


write_dataset(dt, 
              paste0(resDir, "germlineAlCoVarsJoinedAnnoNorm/"),
              format = "parquet",
              partitioning = "sampleTumor")
```

### 4.1.2 Calculate size factors for variant positions present in all tumors only
- No variants are shared by all tumors (nrow(chosen) = 0), thus no NREF_norm_all/NALT_norm_all.
- Not compatible with DFT1/DFT2 so leaving this out for now, potentially revisit!
```{r}
# n_samples <- uniqueN(dt$samplename)
# 
# chosen <- dt[TOTAL > 0,
#              .N, by = locationAlt][N == n_samples]
# 
# variants_all <- dt[locationAlt %in% chosen$locationAlt]
# 
# gmeans_all <- variants_all[, .(gmean = geomean(TOTAL + 1)),
#                            by = locationAlt]
# 
# variants_all[gmeans_all, on = "locationAlt", gmean := i.gmean]
# variants_all[, scaled_total := TOTAL / gmean]
# 
# size_factors_all <- variants_all[, .(
#   size_factor_all = median(scaled_total, na.rm = TRUE)
# ), by = samplename]
# 
# dt[size_factors_all, on = "samplename",
#    size_factor_all := i.size_factor_all]
# 
# dt[, `:=`(
#   NREF_norm_all = NREF / size_factor_all,
#   NALT_norm_all = NALT / size_factor_all
# )]
```

## 4.2 Normalize expression by copy number of informative variant

## 4.3 Compute eT and eH


