---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(stringdist)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_metadata"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
```

## 0.3 Custom functions

# 1 Data
## 1.1 VCF files
### 1.1.1 DFT1 somatic variants
- VCF file
```{r}
dft1Som <- read_table(paste0(dataDir, "Data_S1A_DFT1_somatic_variants.txt"), 
                      col_types = cols(.default = "c"))

dft1SomLong <- dft1Som %>%
  pivot_longer(
    cols = -c(CHROM, POS, REF, ALT),
    names_to = "tumor",
    values_to = "ratio"
  ) %>%
  mutate(chrom = as.character(CHROM), 
         location = paste0(CHROM, ":", POS), 
         position = as.numeric(POS)) %>%
  filter(nchar(REF) == 1, nchar(ALT) == 1) 

rm(dft1Som)
gc()

dft1Tumors <- dft1SomLong %>%
  dplyr::select(tumor) %>%
  distinct() %>%
  mutate(VCF = "1", 
         dftType = "dft1")
```

### 1.1.2 DFT2 somatic variants
```{r}
dft2Som <- read_table(paste0(dataDir, "Data_S1B_DFT2_somatic_variants.txt"),
                      col_types = cols(.default = "c"))
dft2SomLong <- dft2Som %>%
  pivot_longer(
    cols = -c(CHROM, POS, REF, ALT),
    names_to = "tumor",
    values_to = "ratio"
  ) %>%
  mutate(chrom = as.character(CHROM),
         location = paste0(CHROM, ":", POS),
         position = as.numeric(POS)) %>%
  filter(nchar(REF) == 1, nchar(ALT) == 1)

rm(dft2Som)
gc()

dft2Tumors <-  dft2SomLong %>%
  dplyr::select(tumor) %>%
  distinct() %>%
  mutate(VCF = "1", 
         dftType = "dft2")
```

### 1.1.3 Germline variants
```{r}
# germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"),
#                       col_types = cols(.default = "c"))
# 
# germlineLong <- head(germline) %>%
#   as.data.frame() %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "sample",
#     values_to = "ratio"
#   ) %>%
#   mutate(chrom = as.character(CHROM),
#          location = paste0(CHROM, ":", POS),
#          position = as.numeric(POS)) %>%
#   filter(nchar(REF) == 1, nchar(ALT) == 1)
# 
# rm(germline)
# gc()
```


## 1.2 CNVs
### 1.2.1 DFT1
```{r}
dft1Cnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx", 
                     sheet = "Table S6B", 
                     startRow = 3)

dft1CnvLong <- dft1Cnv %>%
  pivot_longer(-c(1:12),
              names_to = "tumor", 
              values_to = "presence") %>%
  mutate(presence = as.numeric(presence)) %>%
  mutate(
    START = as.numeric(START),
    END = as.numeric(END),
    CHROM = as.character(CHROM),
    tumor = as.character(tumor)
  )
```


## 1.3 Purities
### 1.3.1 All samples
```{r}
purities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(purities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(purities),
  paste0(colnames(purities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(purities) <- newNames
purities <- purities[-1, ]


# All tumors in vcf files in supplementary metadata.
# setdiff(dft1Tumors$tumor, purities$`TUMOUR.INFORMATION_TUMOUR ID`)
# setdiff(dft2Tumors$tumor, purities$`TUMOUR.INFORMATION_TUMOUR ID`)
```
## 1.4 Allele counter data
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- ALL RNA seq files were genotyped at every position, regardless of DFT1 and DFT2 status. 
- Here checking how many of these RNAseq files were DFT1 versus DFT2. 
### 1.4.1 DFT1
- 115 RNA seq files
```{r}
dft1SomAlCo <- list.files(paste0(resDir, "/063_Results/06_deconvTable_alCo_dft1SomRefAltExport"), 
                          recursive = T)

# Read all files and combine
dft1SomAlCoCombined <- rbindlist(lapply(dft1SomAlCo, function(f) {
  # Extract tumor ID between "dft1SomRefAltExport_" and "_realigned"
  tumorId <- sub(".*dft1SomRefAltExport_(.+)_realigned.*", "\\1", basename(f))
  
  # Read file
  dt <- fread(paste0(resDir, "/063_Results/06_deconvTable_alCo_dft1SomRefAltExport/", f),
              select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
  
  # Add tumor ID column
  dt[, tumorId := tumorId]
  
  return(dt)
})) 
# %>%
# mutate(tumorLocation = paste0())

length(unique(dft1SomAlCoCombined$tumorId))
```

### 1.4.2 DFT2
- 115 RNA seq files
```{r}
dft2SomAlCo <- list.files(paste0(resDir, "/063_Results/06_deconvTable_alCo_dft2SomRefAltExport"), 
                          recursive = T)

# Read all files and combine
dft2SomAlCoCombined <- rbindlist(lapply(dft2SomAlCo, function(f) {
  # Extract tumor ID between "dft2SomRefAltExport_" and "_realigned"
  tumorId <- sub(".*dft2SomRefAltExport_(.+)_realigned.*", "\\1", basename(f))
  
  # Read file
  dt <- fread(paste0(resDir, "/063_Results/06_deconvTable_alCo_dft2SomRefAltExport/", f),
              select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
  
  # Add tumor ID column
  dt[, tumorId := tumorId]
  
  return(dt)
})) 
# %>%
# mutate(tumorLocation = paste0())

length(unique(dft2SomAlCoCombined$tumorId))
```

### 1.4.3 All IRODS files
- Projects 7511 and 5368
- Fairly certain all 5368 files should have been used in Max's study??
- At first filter only 35/121 tumors have a matched entry in Max's metadata.
```{r}
irodsMerged <- read.xlsx(paste0("~/gitClones/01_Deconvolution/02_Deconvolution_DFT1/020_Data/053_devilRna_irods/irodsMerged.xlsx")) %>%
  dplyr::select(-c(collection, 
                   data_object, 
                   total_reads)) %>%
  distinct() 

write.xlsx(irodsMerged, 
           paste0(resDir, fprefix,  "_irodsMergedDistinct.xlsx"))
```

# 2 Filter and join
## 2.1 Add RNAseq column to metadata
```{r}
puritiesIrodsMerged <- purities %>%
  mutate(
    rnaSeq = ifelse(`TUMOUR.INFORMATION_TUMOUR ID` %in% irodsMerged$supplier,
                    `TUMOUR.INFORMATION_TUMOUR ID`,
                    "check"),
    
    rnaSeqPartial = sapply(`TUMOUR.INFORMATION_TUMOUR ID`, function(id) {
      if (id %in% irodsMerged$supplier) {
        return(id)
      }
      
      # compute distances to all supplier names
      dists <- stringdist(id, irodsMerged$supplier, method = "lv")
      
      # find matches with distance 1. Ex. is T1a RNAseq file versus T1 in Max's metadata.
      matches <- irodsMerged$supplier[dists == 1]
      
      if (length(matches) > 0) {
        return(matches[1])  # take first match (or adjust if needed)
      } else {
        "check"
      }
    })
  )
  

write.xlsx(puritiesIrodsMerged, 
           paste0(resDir, 
                  fprefix, 
                  "_purityIrodsMerged.xlsx"))
```



