# 0 Environment
## 0.1 Packages
```{r}
# install.packages("slider")
```

### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(arrow)
library(furrr)
library(patchwork)
library(slider)
library(duckdb)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_dft1Som"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
source("~/gitClones/08_deconvTable/082_Scripts/genotype_germline_variants.R")
histoDir <- paste0(plotDir, 
                   "histo/")
dir.create(histoDir)
barDir <- paste0(plotDir, 
                 "bar/")
dir.create(barDir)


```

## 0.3 Custom functions
### 0.3.1 Compute tAlt
```{r}
computeTAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         hTotal) {
  
    # print("Solving for tAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    tAlt = (obsVaf * denominator)/purityTumor
  return(tAlt)
}
```
### 0.3.2 Compute hAlt
```{r}
computeHAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         tAlt,
                         hTotal) {
  
  # Cases where hAlt = 0, solve for tAlt
 
    print("Solving for hAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    hAlt = (obsVaf*denominator)/(1-purityTumor)
  return(hAlt)
}

computeHAlt(obsVaf = 0.5, 
            purityTumor = 0.8, 
            tTotal = 4, 
            hTotal = 2, 
            tAlt=0)



```
### 0.3.3 Geometric mean
- Kevin's scripts
  - To calculate size factors: 
  - Compute the geom.mean of totals for each variant (optional: add pseudocount)
  - Divide each variant's read total by the geometric mean for each variant
  - Then take the median of these ratios for each sample
  - (Optional) restrict to variants with presence in all tumors
```{r}
geomean <- function(x) {
  if (all(x == 0)) {
    NA_real_
  } else {
    exp(mean(log(x[x > 0])))
  }
}
```

### 0.3.4 Summarize dft1Som variants
```{r}
library(dplyr)
library(tidyr)

library(data.table)

summarise_variants_dt <- function(df) {
  dt <- as.data.table(df)
  
  # ensure uniqueness per tumor per location
  dt <- unique(dt[, .(locationAlt, sampleTumor, statusUpdate)])
  
  # count per location + status
  counts <- dt[, .N, by = .(locationAlt, statusUpdate)]
  
  # total tumors per location
  totals <- dt[, .(total_tumors = uniqueN(sampleTumor)), by = locationAlt]
  
  # merge counts with totals
  counts <- counts[totals, on = "locationAlt"]
  
  # compute percentage
  counts[, pct := N / total_tumors]
  
  # reshape wide (one column per statusUpdate)
  wide <- dcast(
    counts,
    locationAlt + total_tumors ~ statusUpdate,
    value.var = "pct",
    fill = 0
  )
  
  # compute majority
  status_cols <- setdiff(names(wide), c("locationAlt", "total_tumors"))
  
  wide[, majority := {
    vals <- unlist(.SD)
    max_val <- max(vals)
    winners <- names(vals)[vals == max_val]
    
    if (length(winners) > 1) "flag" else winners
  }, by = locationAlt, .SDcols = status_cols]
  
  return(wide)
}
```

### 0.3.4 Copy number state sequence
```{r}
possible_cn_absorbing <- function(start_cn, nGain, nLoss) {
  # Each state is a pair: CN, remaining events (gains and losses)
  states <- list(list(cn = start_cn, g = nGain, l = nLoss))
  
  final_cns <- c()
  
  while (length(states) > 0) {
    new_states <- list()
    
    for (s in states) {
      cn <- s$cn
      g  <- s$g
      l  <- s$l
      
      if (cn == 0 || (g == 0 && l == 0)) {
        # CN is 0 (absorbing) or no events left → record final CN
        final_cns <- c(final_cns, cn)
      } else {
        # Apply a gain if available
        if (g > 0 && cn > 0) {
          new_states <- append(new_states, list(list(cn = cn + 1, g = g - 1, l = l)))
        }
        # Apply a loss if available
        if (l > 0) {
          new_cn <- max(0, cn - 1)
          new_states <- append(new_states, list(list(cn = new_cn, g = g, l = l - 1)))
        }
      }
    }
    
    states <- new_states
  }
  sort(unique((final_cns)))
}

possible_cn_absorbing(3, 2, 3)
```
# 1 Data
## 1.1 Purities and ploidy
```{r}
dft1SomPurities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(dft1SomPurities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(dft1SomPurities),
  paste0(colnames(dft1SomPurities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(dft1SomPurities) <- newNames
dft1SomPurities <- dft1SomPurities[-1, ]

dft1SomPuritiesPloidy <- dft1SomPurities %>%
  mutate(ploidy = ifelse(X15_PLOIDY == "diploid",
                 2, 
                 ifelse(X15_PLOIDY == "tetraploid", 
                        4, 
                        "toCheck") 
                 )) %>%
  filter(ploidy != "toCheck") %>%
  dplyr::rename("sampleTumor"="TUMOUR.INFORMATION_TUMOUR ID", 
                     "purity"="X14_PURITY (VAF-based)", 
                "dftType"="X9_LINEAGE") %>%
              mutate(purity = as.numeric(purity), 
                     ploidy = as.numeric(ploidy), 
                     sex=ifelse(HOST.SEX == "♂", 
                                "male", 
                                ifelse(HOST.SEX == "♀", 
                                       "female", 
                                       "toCheck"))) %>%
              dplyr::select(sampleTumor, 
                            ploidy, 
                            purity, 
                            sex, 
                            dftType, 
                            X15_PLOIDY) 
  
table(dft1SomPuritiesPloidy$sex, 
      dft1SomPuritiesPloidy$dftType)

males <- dft1SomPuritiesPloidy %>%
  filter(sex == "male") 

dft1 <- dft1SomPuritiesPloidy %>%
  filter(dftType == "DFT1") 

dft2 <- dft1SomPuritiesPloidy %>%
  filter(dftType == "DFT2") # 41 samples. 
```
## 1.2 VCF files
- Read in dft1Som vcf file, all variants. 
- Pivot longer so that each variant and each sample is a single row with the ALT/TOTAL ratio.
```{r}
# dft1Som <- read_table(paste0(dataDir, "Data_S1A_DFT1_somatic_variants.txt"),
#                       col_types = cols(.default = "c"))
# 
# dft1SomLong <- dft1Som %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "tumor",
#     values_to = "ratio"
#   ) %>%
#   mutate(chrom = as.character(CHROM),
#          location = paste0(CHROM, ":", POS),
#          position = as.numeric(POS)) %>%
#   filter(nchar(REF) == 1, nchar(ALT) == 1) %>%
#   dplyr::rename("sampleTumor"="tumor") 
# 
# rm(dft1Som)
# gc()
```

## 1.3 Pair tumor/host samples
- Using shared ID's pivot wider so that the ratio for matched tumor/host samples is side by side.
- Filter on variants where alt allele is present in only the host or the tumor sample within a pair! --> so that we have info. on tAlt or hAlt at that position.
- Tumor samples = 112
- Host samples = 81
- Paired = 103 --> 100 if I filter Max's table? Maybe the non-transmissible cases. 
- Add multiallelic annotation, ALL SAMPLES!!
```{r}
# chroms <- c(paste0(1:6),
#             "X",
#             "Y")
# 
# dft1Som <- dft1SomLong %>%
#   mutate(
#     tumorAlt = as.integer(sub("/.*", "", ratio)),
#     tumorTotal = as.integer(sub(".*/", "", ratio))
#   ) %>%
#   mutate(obsVafTumor = tumorAlt/tumorTotal)%>%
#  mutate(locationRef = paste0(CHROM, ":", POS, "_", REF),
#         locationRefAlt = paste0(CHROM, ":", POS, "_", REF, "_", ALT),
#         tumorLocation = paste0(sampleTumor, "_", CHROM, ":", POS),
#         tumorLocationRef = paste0(sampleTumor,
#                                   "_", CHROM, ":", POS, "_", REF),
#         tumorLocationRefAlt = paste0(sampleTumor,
#                                   "_", CHROM, ":", POS, "_", REF, "_", ALT),
#         position = as.numeric(POS)) %>%
#   filter(CHROM %in% chroms)
# 
# # Overwrite data.
# write_dataset(dft1Som,
#               format = "parquet",
#               paste0(resDir,
#                      "dft1Som/08_deconvTable_dft1Som_computedVafs/"),
#               partitioning = "sampleTumor")
```

## 1.4 Annotate multiallelic variants. 
- Somatic variants are not present in any of the host samples. 
- That could mean it has 0 VAF in a given tumor sample, but this was not observed in all samples and is therefore not somatic. 
- How to account for this when calling "true" informative variants?
- 30 multiallelic positions!!!
```{r}
# multiallelicSites <- open_dataset(paste0(resDir, "dft1Som/08_deconvTable_dft1Som_computedVafs/")) %>%
#   dplyr::select(tumorLocationRef, ALT, sampleTumor) %>%
#   distinct() %>%
#   group_by(tumorLocationRef) %>%
#   summarise(nAltAlleles = n_distinct(ALT), .groups = "drop") %>%
#   filter(nAltAlleles > 1) %>%
#   collect()
# 
# nAltAlleles <- open_dataset(paste0(resDir, "dft1Som/08_deconvTable_dft1Som_computedVafs/")) %>%
#   dplyr::select(tumorLocationRef, ALT, sampleTumor) %>%
#   filter(tumorLocationRef %in% multiallelicSites$tumorLocationRef) %>%  # Arrow predicate pushdown
#   distinct() %>%
#   collect() %>%
#   group_by(tumorLocationRef) %>%
#   summarise(
#     nAltAlleles     = n_distinct(ALT),
#     altAlleles      = paste(sort(unique(ALT)), collapse = ","),
#     baseA = ifelse("A" %in% ALT, paste(sampleTumor[ALT == "A"], collapse = ","), NA),
#     baseC = ifelse("C" %in% ALT, paste(sampleTumor[ALT == "C"], collapse = ","), NA),
#     baseG = ifelse("G" %in% ALT, paste(sampleTumor[ALT == "G"], collapse = ","), NA),
#     baseT = ifelse("T" %in% ALT, paste(sampleTumor[ALT == "T"], collapse = ","), NA),
#     .groups = "drop"
#   )
# 
# gc()
# 
# # All tumors have a count at these variant, just flag to exclude from majority calculation??
# test <- open_dataset(paste0(resDir, "dft1Som/08_deconvTable_dft1Som_computedVafs/")) %>%
#   filter(tumorLocationRef %in% multiallelicSites$tumorLocationRef) %>%
#   collect()
# 
# gc()
# 
# length(unique(test$locationRefAlt)) # 30 
# length(unique(test$locationRef)) # 15 
# 
# write.xlsx(test,
#            paste0(resDir, "/dft1Som/08_deconvTable_dft1SomVariants_multiallelicData.xlsx"))
# 
# # Join dft1Som with multiallelic annotation.
# multiallelicSites <- multiallelicSites %>%
#   mutate(multiallelic = 1L)
# 
# dft1Som <- open_dataset(paste0(resDir, "dft1Som/08_deconvTable_dft1Som_computedVafs/")) %>%
#   left_join(multiallelicSites %>%
#               dplyr::select(tumorLocationRef,
#                             multiallelic)) %>%
#   mutate(multiallelic = ifelse(is.na(multiallelic), 0L, multiallelic))
# 
# 
# write_dataset(dft1Som,
#                                 paste0(resDir,
#                                       "dft1Som/08_deconvTable_dft1Som_computedVafs_multiAnno/"),
#                                 partitioning = "sampleTumor",
#                                 format = "parquet")
# 
# 
# dft1Som <- open_dataset(paste0(resDir,
#                                "dft1Som/08_deconvTable_dft1Som_computedVafs_multiAnno/")) 
```

## 1.5 Filter tumors of interest 
```{r}
# dft1Som <- open_dataset(paste0(resDir,
#                             "/dft1Som/08_deconvTable_dft1Som_computedVafs_multiAnno/")) %>%
#   left_join(dft1SomPuritiesPloidy %>% dplyr::select(sampleTumor, X15_PLOIDY))
```

# 2 CNVs - DFT1
## 2.1 Read in CNV data, join with VAF's
- Run selection as background job (Code/scroll down), copy output to global env. 
```{r}
dft1Cnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx",
                     sheet = "Table S6B",
                     startRow = 3)

dft1CnvLong <- dft1Cnv %>%
  pivot_longer(-c(1:12),
              names_to = "tumor",
              values_to = "presence") %>%
  mutate(presence = as.numeric(presence)) %>%
  filter(presence > 0) %>% # If 0, variant not in CNV segment, exclude to reduce size of dataframe.
  mutate(
    START = as.numeric(START),
    END = as.numeric(END),
    CHROM = as.character(CHROM),
    tumor = as.character(tumor)
  ) %>%
  # write as camel case for sql with duck db.
  mutate(eventType = EVENT.TYPE,
         cnvEventId = CNV.EVENT.ID,
         sampleTumorChrom = paste0(tumor, "_", CHROM)) %>%
  dplyr::select(-c("CNV.EVENT.ID",
                   "WIDTH",
                   "NCBI.ANNOTATION",
                   "ENSEMBL.ANNOTATION",
                   "COSMIC",
                   "N",
                   "EVENT.TYPE",
                   "MARKER.5",
                   "DOUBLE.MINUTE"))


rm(dft1Cnv)
gc()

# # check <- dft1CnvLong$tumor
# # check2 <- dft1Som %>%
# #   distinct(sampleTumor) %>%
# #   collect()
# # 
# # length(intersect(check, check2$sampleTumor))
# # > length(intersect(check, check2$sampleTumor))
# # [1] 78
# 
# con <- dbConnect(duckdb(), config = list(threads = availableCores() - 1))
# 
# duckdb_register_arrow(con, "dft1Som", dft1Som)  # no collect() needed
# duckdb_register(con, "cnv", dft1CnvLong)
# 
# # Write parquet directly from DuckDB — never fully loads result into RAM
# dbExecute(con, paste0("
#   COPY (
#     SELECT g.*, c.presence, c.CHROM, c.START, c.END, c.cnvEventId
#     FROM dft1Som g
#     LEFT JOIN cnv c
#            ON  g.CHROM    = c.CHROM
#            AND g.sampleTumor    = c.tumor
#            AND g.position >= c.START
#            AND g.position <= c.END
#   ) TO '", resDir, "/dft1Som/", fprefix, "_dft1CnvJoin/'
#   (FORMAT PARQUET, PARTITION_BY (sampleTumor), OVERWRITE)
# "))
# 
# dbDisconnect(con, shutdown = TRUE)
# 
# # Which variants have > 0 changes?
# presence <- open_dataset(paste0(resDir,
#                                 "/dft1Som/",
#                                 fprefix,
#                                 "_dft1CnvJoin/")) %>%
#   dplyr::select(tumorLocationRefAlt) %>%
#   collect()
# # 16235199 variants. 
```

## 2.3 Starting CN
- Adjust this for all variants. 
- Important to define for running genotyper(). 
- Case when need to keep input type the same (i.e., character string or numeric). Revert back to numeric after checking there are no "toCheck" rows. 
```{r}
# autosomes <- paste0(1:6)
# 
# dft1CnvJoin <- open_dataset(paste0(resDir,
#                                    "/dft1Som/",
#                                    fprefix,
#                                    "_dft1CnvJoin/")) %>%
#   # We left out event type to run DuckDb. Merge eventType with dft1CnvJoin.
#   left_join(dft1CnvLong %>%
#               dplyr::select(cnvEventId,
#                             eventType) %>%
#               distinct()) %>%
#   left_join(dft1SomPuritiesPloidy %>%
#               dplyr::select(X15_PLOIDY,
#                             sampleTumor)) %>%
#   dplyr::select(-c(chrom_1, CHROM_2)) %>%
#   # Set the chrom type. 
#   mutate(chromType = ifelse(CHROM %in% autosomes, 
#                             "autosome", 
#                             CHROM)) %>%
#   # Define CN by combination of dft type, ploidy, and chromType
#   mutate(postWgdCn = case_when(
#     ############################
#     ## Diploid DFT1s ##
#     ############################
#     # Diploid, DFT1, autosome (2 copies)
#     X15_PLOIDY == "diploid" & sampleTumor %in% dft1$sampleTumor & chromType == "autosome" ~ "2",
#     # Diploid, DFT1, sex (XX)
#     X15_PLOIDY == "diploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ "2",
#     # Diploid, DFT1, no Y
#     X15_PLOIDY == "diploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ "0",
#     ############################
#     ## Tetraploid DFT1s ##
#     ############################
#     # Tetraploid, DFT1, autosome
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & chromType == "autosome" ~ "4",
#     # Tetraploid, DFT1, sex (XXXX)
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ "4",
#     # Tetraploid, DFT1, no Y chromosome
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ "0",
#     # ############################
#     # ## Diploid DFT2s ##
#     # ############################
#     # # Diploid, DFT2, autosome
#     # X15_PLOIDY == "diploid" & sampleTumor %in% dft2$sampleTumor & chromType == "autosome" ~ 2,
#     # # Diploid, DFT2, sex (XY)
#     # X15_PLOIDY == "diploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "X" ~ 1,
#     # X15_PLOIDY == "diploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "Y" ~ 1,
#     # ############################
#     # ## Tetraploid DFT2s ##
#     # ############################
#     # # Tetraploid, DFT2, autosome
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & chromType == "autosome" ~ 4,
#     # # Tetraploid, DFT2, sex (XXYY)
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ 2,
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ 2,
#     TRUE ~ "toCheck"
#   )) %>%
#     mutate(preWgdCn = case_when(
#     ############################
#     ## Tetraploid DFT1s ##
#     ############################
#     # Tetraploid, DFT1, autosome
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & chromType == "autosome" ~ "2",
#     # Tetraploid, DFT1, sex (XX)
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ "2",
#     # Tetraploid, DFT1, no Y chromosome
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ "0",
#     # ############################
#     # ## Tetraploid DFT2s ##
#     # ############################
#     # # Tetraploid, DFT2, autosome, preWgd = 2 copies
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & chromType == "autosome" ~ 2,
#     # # Tetraploid, DFT2, sex (XXYY), preWgd = XY (1 copy)
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ 1,
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ 1,
#     TRUE ~ postWgdCn
#     ))
# 
# 
# 
# # toCheck <- dft1CnvJoin %>%
# #   filter(preWgdCn == "toCheck") %>%
# #   collect() # No rows!!
# rm(dft1CnvLong)
# gc()
# 
# dft1CnvJoin <- dft1CnvJoin %>%
#   mutate(preWgdCn = as.numeric(preWgdCn),
#          postWgdCn = as.numeric(postWgdCn)) %>%
#   collect()
# table(dft1CnvJoin$CHROM, dft1CnvJoin$preWgdCn)
  #         2
  # 1 3734355
  # 2 3323590
  # 3 3099301
  # 4 2630401
  # 5 1391415
  # 6 1458734
  # X  597403


# table(dft1CnvJoin$CHROM, dft1CnvJoin$postWgdCn)
  #         2       4
  # 1 2964084  770271
  # 2 2643445  680145
  # 3 2462392  636909
  # 4 2098588  531813
  # 5 1103439  287976
  # 6 1144768  313966
  # X  471609  125794
# write_dataset(dft1CnvJoin, 
#               paste0(resDir, "/dft1Som/dft1CnvJoin_prePostAdjustedCn/"), 
#               partitioning = "sampleTumor", 
#               format = "parquet")
```

## 2.2 Adjust CN
```{r}
# dft1CnvJoin <- open_dataset(paste0(resDir,
#                                    "dft1Som/dft1CnvJoin_prePostAdjustedCn/"))
# 
# # --- 1. Find duplicate keys via aggregation. These are rows where there is more than one CNV event ID spanning a given variant (i.e., a multiplet). This does not overestimate the number of multiplets because of multiallelic variants, because I am counting by ref and alt.
# dup_keys_refalt <- dft1CnvJoin %>%
#   filter(presence > 0) %>%
#   group_by(tumorLocationRefAlt) %>%
#   summarise(nPos = n_distinct(tumorLocationRefAlt), .groups = "drop") %>%
#   filter(nPos > 1) %>%
#   pull(tumorLocationRefAlt)
# 
# # length(dup_keys_refalt)
# # [1] 160299
# 
# # --- 2. Find single keys via aggregation. These are rows where there one CNV event ID spanning a given variant (i.e., a singlet). This does not overestimate the number of multiplets because of multiallelic variants, because I am counting by ref and alt.
# sing_keys_refalt <- dft1CnvJoin %>%
#   filter(presence > 0) %>%
#   group_by(tumorLocationRefAlt) %>%
#   summarise(nPos = n_distinct(tumorLocationRefAlt), .groups = "drop") %>%
#   filter(nPos == 1) %>%
#   pull(tumorLocationRefAlt)
# 
# # length(sing_keys_refalt)
# # [1] 1611830
# 
# noChange <- dft1CnvJoin %>%
#   filter(is.na(presence)) %>%
#   distinct(tumorLocationRefAlt, cnvEventId) %>%
#   pull(tumorLocationRefAlt)
# 
# # [1] 14280661
# 
# dft1CnvJoinMultiplets <- dft1CnvJoin %>%
#   mutate(multipletsRefAlt = ifelse(tumorLocationRefAlt %in% dup_keys_refalt,
#                 1,
#                 0),
#          singletsRefAlt = ifelse(tumorLocationRefAlt %in% sing_keys_refalt,
#                            1,
#                            0),
#          cnvType = ifelse(tumorLocationRefAlt %in% dup_keys_refalt,
#                           "multiplet",
#                           ifelse(tumorLocationRefAlt %in% sing_keys_refalt,
#                                  "singlet",
#                                  "noChange"))) %>% # None should be assigned "noChange" at this point, because we have prefiltered on variants where presence > 0.
#   mutate(variantType = "dft1Som") %>%
#   collect()
# 
# write_dataset(dft1CnvJoinMultiplets,
#               format = "parquet",
#               paste0(resDir,
#                      "/dft1Som/dft1CnvJoinMultiplets/"),
#               partitioning = "sampleTumor")
```

### 2.2.1 Variants with no change
- copyNumberAdjusted = postWgdCn (i.e., no change depending on pre/post.)
```{r}
# dft1CnvJoin <- open_dataset(paste0(resDir,
#                                    "/dft1Som/dft1CnvJoinMultiplets/"))
# 
# dft1CnvJoinNoChanges <- dft1CnvJoin %>%
#   filter(is.na(presence)) %>%
#          mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   # Make copy number the adjusted CN for WGD (in tetraploid tumors, in diploid, no change)
#   mutate(copyNumberAdjusted = postWgdCn) %>%
#   group_by(tumorLocationRefAlt) %>%
#   mutate(
#     nGain = 0,
#     nLoss = 0,
#     nGainPost = 0,
#     nGainPre = 0,
#     nLossPost = 0,
#     nLossPre = 0,
#     sumPresence = 0
#   ) %>%
#   # remove CNV table columns
#   mutate(cnvAdjAnno = "noChanges",
#          multipletsRefAlt = 0,
#          singletsRefAlt = 0,
#          variantType = "dft1Som") %>%
#    # remove CNV table columns, because this has been aggregated by summing rows.
#   dplyr::select(-c(START,
#                    END,
#                    cnvEventId,
#                    eventType,
#                    presence))
# 
# gc()
# 
# write_dataset(dft1CnvJoinNoChanges,
#               paste0(resDir, "/dft1Som/dft1CnvJoinNoChanges/"),
#               partitioning = "sampleTumorChrom",
#               format = "parquet")
# 
# gc()
# 
# rm(dft1CnvJoinNoChanges)
# 
# gc()
```

### 2.2.2 Variants with single change
- change ploidy according to single step
- number of unique variants = number of rows, because only 1 change to cnv per variant. 5 tumorLocation where more than one variant with a single CNV change.
- All presences should be 1 because single change
- If CN goes to 0, it should stay at 0. (Applies for multistep, not here.) Running same code as multiplets as a sanity check, none of the maxCn should differ from minCn. 
```{r}
# dft1CnvJoin <- open_dataset(paste0(resDir,
#                                    "/dft1Som/dft1CnvJoinMultiplets/"))
# 
# # dft1CnvJoinSingleChange <- dft1CnvJoin %>%
# #   filter(singletsRefAlt == 1 & multipletsRefAlt == 0 ) %>%
# #   collect()
# # 
# # min(dft1CnvJoinSingleChange$presence)
# # max(dft1CnvJoinSingleChange$presence)
# # # table(dft1CnvJoinSingleChange$eventType, dft1CnvJoinSingleChange$X15_PLOIDY)
# # # 
# # #                                 diploid tetraploid
# # #   GAIN                           869472     202754
# # #   GAIN (post-tetraploidisation)       0      59025
# # #   GAIN (pre-tetraploidisation)        0       3797
# # #   LOSS                           313631      71323
# # #   LOSS (post-tetraploidisation)       0      87178
# # #   LOSS (pre-tetraploidisation)        0       4650
# 
# # Do we count GAIN as post or pre tetra?
# dft1CnvJoinSingleChange <- dft1CnvJoin %>%
#   filter(singletsRefAlt == 1 & multipletsRefAlt == 0) %>%
#   group_by(tumorLocationRefAlt) %>%
#   mutate(
#     nGain = sum(eventType == "GAIN"),
#     nLoss = sum(eventType == "LOSS"),
#     nGainPost = sum(eventType == "GAIN (post-tetraploidisation)"),
#     nGainPre = sum(eventType == "GAIN (pre-tetraploidisation)"),
#     nLossPost = sum(eventType == "LOSS (post-tetraploidisation)"),
#     nLossPre = sum(eventType == "LOSS (pre-tetraploidisation)"),
#     sumPresence = sum(presence)
#   ) %>%
#    # remove CNV table columns, because this has been aggregated by summing rows.
#   dplyr::select(-c(START,
#                    END,
#                    cnvEventId,
#                    eventType,
#                    presence))  %>%
#   mutate(across(c(nGain, nLoss, nGainPre, nGainPost, 
#                 nLossPre, nLossPost), as.double)) %>%
#   mutate(
#     copyNumberAdjusted = pmax(0,
#     case_when(
#       ####################
#       ### DFT1 diploid ###
#       X15_PLOIDY == "diploid" ~ postWgdCn + nGain - nLoss, # 2 or 0 for DFT1 samples (XX).
#       #######################
#       ### DFT1 tetraploid ###
#       #######################
#       ### Not 100% confident, but Liz/Kevin suggest adding any ambiguous GAIN/LOSS (i.e., not pre- or post-) before tetraploidization. These were samples where it was not 100% certain they were tetraploid in the first place. ###
#       X15_PLOIDY == "tetraploid" ~
#         # Add nGain(Pre) - nLoss(Pre) to starting copy number (before WGD). Then duplicate and adjust for post events.
#         ((preWgdCn + (nGainPre + nGain) - (nLossPre + nLoss)) * 2) + (nGainPost) - (nLossPost),
#       TRUE ~ NA_real_
#     )
#    )
#   ) %>%
#   mutate(cnvAdjAnno = "singleChange")
# 
# 
# # chck <- dft1CnvJoinSingleChange %>%
# #   collect()
# # 
# # 
# # table(chck$CHROM, chck$copyNumberAdjusted)
# # > table(chck$CHROM, chck$copyNumberAdjusted)
# #    
# #          1      2      3      5      6
# #   1  94925  28454  99655     25   4054
# #   2  57434  15385  14551    648   1388
# #   3 125642  20641   4014  38304   3318
# #   4   1622    166 795993      0 195400
# #   5  11600   1940  36244  10979   1333
# #   6   2912    327    370   8892      5
# #   X  19496   9060   5823    177   1053
# # 
# # rm(chck)
# # gc()
# 
# write_dataset(dft1CnvJoinSingleChange,
#               paste0(resDir, "/dft1Som/dft1CnvJoinSingleChange/"),
#               partitioning = "sampleTumor")
# 
# rm(singlets)
# rm(dft1CnvJoinSingleChange)
# 
# gc()
```

### 2.2.3 Variants with multiple changes
- Aggregate steps using same logic as before, only this time ploidy will change using sum of all steps rather than single change of CNV status.
- Issue: we don't know the order of events. Depending on order of losses and gains, the final CN could differ. Adding a permutation function which returns how many times a different final CN state could occur. 
- None of the tetraploid tumors have nLossPre > 1, therefore cannot go to 0.
- The only variable cases are diploid tumors whose variants have two losses and one gain. 
- Filter out the five variants that could have been CN 0 or 1!
  - Check these variants against the counts + gtf file.
```{r}
# dft1CnvJoin <- open_dataset(paste0(resDir,
#                                    "/dft1Som/dft1CnvJoinMultiplets/"))
# 
# multiplets <- dft1CnvJoin %>%
#   filter(multipletsRefAlt == 1) %>%
#   group_by(tumorLocationRefAlt) %>%
#   mutate(
#     nGain = sum(eventType == "GAIN"),
#     nLoss = sum(eventType == "LOSS"),
#     nGainPost = sum(eventType == "GAIN (post-tetraploidisation)"),
#     nGainPre = sum(eventType == "GAIN (pre-tetraploidisation)"),
#     nLossPost = sum(eventType == "LOSS (post-tetraploidisation)"),
#     nLossPre = sum(eventType == "LOSS (pre-tetraploidisation)"),
#     sumPresence = sum(presence)
#   ) %>%
# 
#   # remove CNV table columns, because this has been aggregated by summing rows.
#   dplyr::select(-c(cnvEventId,
#                    eventType,
#                    presence,
#                    START,
#                    END)) %>%
#   distinct() %>%
#   ungroup() %>%
#   collect()
# 
# # > max(multiplets$sumPresence)
# # [1] 17
# 
# dft1CnvJoinMultipleChanges <- multiplets %>%
#   group_by(tumorLocationRefAlt) %>%
#   mutate(
#     #####################
#     #### DFT1 diploid ###
#     #####################
#     possibleCn = list(
#       if (X15_PLOIDY == "diploid") {
#         possible_cn_absorbing(postWgdCn, nGain, nLoss)
# 
#     #####################
#     #### DFT1 tetraploid ###
#     #####################
#       } else if (X15_PLOIDY == "tetraploid") {
# 
#         pre_cn <- possible_cn_absorbing(preWgdCn, nGainPre + nGain, nLossPre + nLoss)
#         dup <- pre_cn * 2
#         unique(unlist(
#           lapply(dup, function(cn) {
#             possible_cn_absorbing(cn, nGainPost, nLossPost)
#           })
#         ))
#       } else {
#         NA_real_
#       }
#     ),
#     minCn = sapply(possibleCn, min),
#     maxCn = sapply(possibleCn, max)
#   )
# 
# gc()
# 
# # 116 conflicting rows, filter these out.
# multipleCns <- dft1CnvJoinMultipleChanges %>%
#   filter(minCn != maxCn)
# 
# dft1CnvJoinMultipleChangesFiltered <- dft1CnvJoinMultipleChanges %>%
#   filter(minCn == maxCn)
# # [1] 160183
# 
# write_dataset(dft1CnvJoinMultipleChangesFiltered,
#               paste0(resDir, "/dft1Som/dft1CnvJoinMultipleChanges/"),
#               partitioning = "sampleTumor")
```






# 3 Compute genotypes
This code joins the CNV adjusted tables from Chunks in Section 1.5.2. Then adds in host copy number info before calculating most likely genotype at each position. 

- In a pure sample, vafH = 0 + (1)*Halt/(0 + 1*Htotal). Therefore, Halt = vafH*Htotal. 

- If host sample is a male, total CN at sex chromosomes == 1 !!!

- If tumor sample is DFT2, total CN was adjusted at sex chromosomes == 1 or 2 depending on original ploidy. Depending on whether a SNP fell in a CNV, this value may have changed, and in reflected in copyNumberAdjusted. 

- If tumor sample is DFT1, total CN was adjusted at Y chromosomes as 0 is changed here. This is because, the variant cannot undergo copy number changes if it's starting point is 0.

- Of note, this will not filter out any of the tumors! Only do this at the stage where we merge with RNAseq data (see Chunk XYZ). 
```{r}
# singleChange <- open_dataset(paste0(resDir, 
#                                     "/dft1Som/dft1CnvJoinSingleChange/"))  %>%
#    mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   collect()
# 
# noChanges <- open_dataset(paste0(resDir, 
#                                  "/dft1Som/dft1CnvJoinNoChanges/")) %>%
#   mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   dplyr::select(colnames(singleChange)) %>%
#   collect()
# 
# 
# multipleChanges <- open_dataset(paste0(resDir, "/dft1Som/dft1CnvJoinMultipleChanges/"))  %>%
#   mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   mutate(copyNumberAdjusted = minCn,
#          cnvAdjAnno = "multipleChanges") %>%
# 
#   dplyr::select(colnames(singleChange)) %>%
#   collect()
# 
# dft1SomGenotyperDft1 <- bind_rows(noChanges,
#                               singleChange,
#                               multipleChanges)
# 
# colnames(dft1SomGenotyperDft1)
# 
# write_dataset(dft1SomGenotyperDft1,
#               paste0(resDir, "/dft1Som/dft1SomCnvCorrected/"),
#               partitioning = "sampleTumorChrom")
# # con <- dbConnect(duckdb())
# # 
# # duckdb_register_arrow(con, "singleChange", singleChange)  # no collect() needed
# # duckdb_register_arrow(con, "multipleChanges", multipleChanges)
# # duckdb_register_arrow(con, "noChanges", noChanges)
# # 
# # duckdb_register(con, "males", males %>% distinct(sampleTumor))
# # duckdb_register(con, "dft1",  dft1  %>% distinct(sampleTumor))
# # 
# # 
# # dbExecute(con, paste0("
# #   COPY (
# #     SELECT
# #       * EXCLUDE (copyNumberAdjusted),
# #       copyNumberAdjusted                              AS nTumorTotal,
# #       CASE
# #         WHEN sampleTumor IN (SELECT sampleTumor FROM males)
# #           AND CHROM IN ('X', 'Y')                    THEN 1
# #         ELSE 2
# #       END                                             AS nHostTotal
# #     FROM read_parquet([
# #         '", paste0(resDir, "/dft1Som/dft1CnvJoinNoChanges/**/*.parquet"), "',
# #         '", paste0(resDir, "/dft1Som/dft1CnvJoinSingleChange/**/*.parquet"), "',
# #         '", paste0(resDir, "/dft1Som/dft1CnvJoinMultipleChanges/**/*.parquet"), "'
# #       ], union_by_name = true) AS unioned
# #   ) TO '", paste0(resDir, "/dft1Som/dft1SomCnvCorrected/"), "'
# #   (FORMAT parquet, PARTITION_BY sampleTumor, OVERWRITE)
# # "))
# # 
# # dbDisconnect(con)
# # gc()
# # # 
# # # [1] 16052674 
# # 
# # # 
# # # dft1SomGenotyperDft1_small <- open_dataset(paste0(resDir, "/dft1Som/", fprefix, "_dft1SomCnvCorrected")) %>%
# # #   mutate(purity = as.numeric(purityTumor)) %>%
# # #   filter(CHROM == "1" & tumor == "477T3" & POS == 587960251) %>%
# # #   collect()
# # # 
# # # dft1SomGenotyperDft1 <- open_dataset(paste0(resDir, 
# # #                                              "/dft1Som/dft1SomCnvCorrected")) %>%
# # #   dplyr::select(-c(START, 
# # #                    END, 
# # #                    cnvType, 
# # #                    CHROM_1))
# # # 
# # # gc()
# # # 
# 
# dft1SomGenotyperDft1 <-open_dataset(paste0(resDir, 
#                                            "/dft1Som/dft1SomCnvCorrected")) %>%
#   # filter(minCn == maxCn) %>%
#   # dplyr::select(-c(possibleCn)) %>%
#   mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   ### Host CN in somatic variants. ### 
#   mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"), 
#                          1, 
#                          2)) %>% ## XX in females, or CN = 2 at autosomes.  
#   left_join(dft1SomPuritiesPloidy %>% dplyr::select(sampleTumor, 
#                                                     purity)) %>%
#   dplyr::rename("purityTumor"="purity") 
# 
# schema(dft1SomGenotyperDft1)
# 
# plan(multisession, workers = availableCores() - 1)
# 
# chunk_ids <- dft1SomGenotyperDft1 %>%
#   distinct(sampleTumorChrom) %>%
#   pull(sampleTumorChrom) 
# 
# # > length(chunk_ids)
# # [1] 546
# 
# for (cid in chunk_ids) {
# 
#   cat("Processing chunk", 
#       cid, 
#       "\n")
# 
#   chunk <- dft1SomGenotyperDft1 %>%
#     filter(sampleTumorChrom == cid) %>%
#     collect()
# 
#   if (nrow(chunk) == 0) next
# 
#   results <- future_pmap(
#     .l = list(
#       chunk$nHostTotal,
#       chunk$copyNumberAdjusted,
#       chunk$tumorAlt,
#       chunk$tumorTotal,
#       chunk$purityTumor
#     ),
#     .f = function(hp, tp, ta, tt, p) {
#       
#       g <- somatic_genotyper(host_ploidy=hp,
#                               tumour_ploidy=tp,
#                               tumour_observed_alt_reads=ta,
#                               tumour_observed_total_reads=tt,
#                               tumour_purity=p,
#                               eps = 1e-3)
# 
#       c(g_host = g$E$g_host,
#         g_tumour = g$E$g_tumour,
#         map_g_host = g$map[1, "g_host"],
#         map_g_tumour = g$map[1, "g_tumour"])
#     }
#   )
# 
#   res <- do.call(rbind, results)
#   
#   chunk$nHostAltGtyrMapped <- res[,3]
#   chunk$nHostRefGtyrMapped <- chunk$nHostTotal - chunk$nHostAltGtyrMapped
# 
#   chunk$nTumorAltGtyrMapped <- res[,4]
#   chunk$nTumorRefGtyrMapped <- chunk$copyNumberAdjusted - chunk$nTumorAltGtyrMapped
#   
#   # chunk$tAltEst <- res[,5]
#   # chunk$tAltEstRounded <- round(res[,5])
#   # 
#   # chunk$tRefEst <- chunk$copyNumberAdjusted - chunk$tAltEstRounded
#   
#   chunk <- chunk %>%
#   mutate(tAltTest = computeTAlt(obsVafTumor, 
#                                 purityTumor, 
#                                 tTotal = copyNumberAdjusted, 
#                                 hTotal = nHostTotal), 
#          nTumorAltTest = round(tAltTest),
#          nTumorRefTest = copyNumberAdjusted-nTumorAltTest
#          ) 
# 
#   write_parquet(
#     chunk,
#     paste0(resDir, 
#            "/dft1Som/dft1SomGenotyperDft1/", 
#            cid, 
#            ".parquet")
#   )
# 
#   rm(chunk, results, res)
#   gc()
# }
# 
# plan(sequential)
```

# 4 Join AlCo data
## 4.1 Metadata
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- Adjusted names manually using metadata! 
- Mapping AlCo results to VCF file names
- FOR now, don't filter where mapping = 1. There are sample names that I am certain match the name in Max's supplementary table. --> do this at the stage where we merge with AlCo results so that everything is genotyped
  - Filter on ! NA in DFT status column, as these we likely highly host contaminated or non-DFT tumors.
```{r}
metadata <- read.xlsx(paste0(resDir,
                          "20260529_08_deconvTable_metadata_purityIrodsMergedManualEdit.xlsx"), 
                      sheet = "withPolyRiboDups") %>%
              dplyr::rename("tumorId"="TUMOUR.INFORMATION_TUMOUR.ID",
                     "tumorIdRnaSeq"="rnaSeqManual") %>%
  filter(X9_LINEAGE %in% c("DFT1"))

# length(intersect(metadata$tumorId, dft1$sampleTumor)) # [1] 78

metadataPairs <- metadata %>%
  filter(mappingUpdate == 1) %>%
  dplyr::select(tumorId, 
                tumorIdRnaSeq) %>%
  distinct() %>%
  # For now, filter out duplicate. 
  filter(!tumorId %in% c("155T", "63T3")) %>%
  filter(tumorIdRnaSeq != "check")

# which(table(metadataPairs$tumorId)>1)
# which(table(metadataPairs$tumorIdRnaSeq)>1)
# 155Ta check 
#    17    55 
# length(intersect(metadataPairs$tumorId, dft1$sampleTumor)) # [1] 57
```

## 4.2 Read in AlCo results
- Here, tumorId = the name of the DNA file. Need to join this with the metadata table and switch out for the VCF tumor name (Max's supp data.)
- We have AlCo (i.e., RNAseq data )
```{r}
# dft1SomAlCo <- list.files(paste0("~/gitClones/08_deconvTable/083_Results/fromFarm/06_alCo_realignedEnsembl_dft1Som/"),
#                           recursive = T)
# 
# # Read all files and combine
# dft1SomAlCoDf <- rbindlist(lapply(dft1SomAlCo, function(f) {
#   # Extract tumor ID between "dft1SomRefAltExport_" and "_realigned"
#   tumorIdRnaSeq <- sub(".*colnames_(.+).txt*", "\\1", basename(f))
# 
#   # Read file
#   dt <- fread(paste0("~/gitClones/08_deconvTable/083_Results/fromFarm/06_alCo_realignedEnsembl_dft1Som/", f),
#               select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
# 
#   # Add tumor ID column
#   dt[, tumorIdRnaSeq := tumorIdRnaSeq]
# 
#   return(dt)
# }))
# 
# # length(unique(dft1SomAlCoDf$tumorIdRnaSeq)) # 115 RNAseq samples were genotyped.
# #
# # check <- unique(dft1SomAlCoDf$tumorIdRnaSeq)
# #
# # length(intersect(dft1SomAlCoDf$tumorIdRnaSeq, metadataPairs$tumorIdRnaSeq)) # 53 DFT2 samples for which we have a paired RNAseq-DNAseq file. --> 4 that don't?
# #
# # setdiff(metadataPairs$tumorIdRnaSeq, intersect(dft1SomAlCoDf$tumorIdRnaSeq, metadataPairs$tumorIdRnaSeq))
# 
# # Left join metadata to map RNAseq ID to DNAseq ID.
# write_dataset(
#   dft1SomAlCoDf,
#   path = paste0(resDir, "/dft1SomAlCoDf/"),
#   format = "parquet",
#   partitioning = "CHROM"
# )
```

# 5 Filter gtf
## 5.1 Overlapping positions
```{r}
# dft1SomGenotyper <- open_dataset(paste0(resDir,
#                            "/dft1Som/dft1SomGenotyperDft1/")) 
# 
# 
# 
# joinedChromPos <- dft1SomGenotyper %>%
#   dplyr::select(CHROM,
#                 POS) %>%
#   distinct() %>%
#   mutate(POS = as.numeric(POS)) %>%
#   collect()
# 
# gc()
# 
# gtfFile <- paste0(dataDir,
#                   "/Sarcophilus_harrisii.mSarHar1.11.115.gtf")
# 
# gtf <- import(gtfFile)
# 
# genes <- gtf[gtf$type == "gene"]
# 
# exons <- gtf[gtf$type == "exon"]
# 
# pcGenes <- genes[genes$gene_biotype == "protein_coding"]
# 
# variants <- GRanges(
#   seqnames = joinedChromPos$CHROM,
#   ranges = IRanges(start = joinedChromPos$POS,
#                    end = joinedChromPos$POS)
# )
# 
# # Which variants fall in an exon
# exonHits <- findOverlaps(variants, exons)
# 
# # Map exon hits to gene IDs
# exonGeneIds <- exons$gene_id[subjectHits(exonHits)]
# exonGeneNames <- exons$gene_name[subjectHits(exonHits)]
# exonBiotypes <- exons$gene_biotype[subjectHits(exonHits)]
# 
# hitsDf <- data.frame(
#   query = queryHits(exonHits),
#   gene_id = exonGeneIds,
#   gene_name = exonGeneNames,
#   biotype = exonBiotypes
# )
# 
# collapseFun <- function(x) paste(unique(x), collapse = ";")
# 
# collapsed <- aggregate(. ~ query,
#                        data = hitsDf,
#                        FUN = collapseFun)
# 
# collapsed$CHROM <- joinedChromPos$CHROM[collapsed$query]
# collapsed$POS   <- joinedChromPos$POS[collapsed$query]
# 
# write.xlsx(collapsed %>%
#            mutate(POS = as.character(POS)),
#            paste0(resDir, "/dft1Som/dft1SomDft1CollapsedGtf.xlsx"))

collapsed <- read.xlsx(paste0(resDir, "/dft1Som/dft1SomDft1CollapsedGtf.xlsx")) %>%
  mutate(location = paste0(CHROM, ":", POS)) # 3,579
```

# 6 Full merge
- Per sampleTumorChrom, merge:
  - Genotypes
  - Majority status based on genotypes
  - AlCo
  - GTF positions
```{r}
chroms <- c(paste0(1:6), 
            "X", 
            "Y")
```

## 6.1 Assign majority genotype 
- Approach with Liz
  - For non-majority SNP's (CN >= 2), correct genotype to majority. Keep info on what was the originally assigned genotype. Also keep info on how many tumors had this flip. 
  - For the tied SNPs, exclude?
  - For the unclassified SNPs, keep as is? These could be genuine variants?
  
### 6.1.1 TEST
- Filter for SNPs positions where CN >= 2 (in any number of tumors): set as het, homAlt, or homRef. Using genotyper(), but will eventually use vaf windows defined in Kevin's supp methods. 
- Unclassified are SNPs falling in segments of CN 1 or 0. 
- Annotate each location with the majority (>= 50% status) flipping status
```{r}
# genotypeDft1Cn2All <- dft1SomGenotyper %>%
#   ######
#   filter(sampleTumor %in% dft1$sampleTumor) %>%
#   ######
#   mutate(
#     status = case_when(
#       ## HOM REF ##
#       nTumorTotal >= 2 & nTumorRefGtyrMapped == nTumorTotal & nTumorAltGtyrMapped == 0 ~ "homRef",
#       ## HOM ALT ##
#       nTumorTotal >= 2 & nTumorAltGtyrMapped == nTumorTotal & nTumorRefGtyrMapped == 0                   ~ "homAlt",
#       
#       nTumorTotal >= 2 & (nTumorRefGtyrMapped > 0 & nTumorAltGtyrMapped > 0)    ~ "het",
#       nTumorTotal < 2                                                                  ~ "unclassified",
#       TRUE                                                                             ~ "toCheck"
#     ),
#     statusBroad = case_when(
#       status %in% c("homRef",
#                     "homAlt") ~ "hom",
#       TRUE                              ~ status
#     )
#   ) %>%
#   dplyr::select(locationRefAlt,
#                 sampleTumor,
#                 status,
#                 statusBroad,
#                 CHROM,
#                 POS,
#                 ploidy,
#                 nTumorTotal,
#                 multiallelic) %>%
#   collect()
# 
# gc()
# 
# # hets <- dft1SomGenotyper %>%
# #   ######
# #   filter(sampleTumor %in% dft1$sampleTumor) %>%
# #   ######
# #   mutate(
# #     status = case_when(
# #       ## HOM REF ##
# #       nTumorTotal >= 2 & nTumorRefGtyrMapped == nTumorTotal & nTumorAltGtyrMapped == 0 ~ "homRef",
# #       ## HOM ALT ##
# #       nTumorTotal >= 2 & nTumorAltGtyrMapped == nTumorTotal & nTumorRefGtyrMapped == 0                   ~ "homAlt",
# #       
# #       nTumorTotal >= 2 & (nTumorRefGtyrMapped > 0 & nTumorAltGtyrMapped > 0)    ~ "het",
# #       nTumorTotal < 2                                                                  ~ "unclassified",
# #       TRUE                                                                             ~ "toCheck"
# #     ),
# #     statusBroad = case_when(
# #       status %in% c("homRef",
# #                     "homAlt") ~ "hom",
# #       TRUE                              ~ status
# #     )
# #   )  %>%
# #   filter(status == "het")  %>%
# #   collect() # [1] 16344391 rows
# # 
# # gc()
# # 
# # nrow(hets) # [1] 16344391
# # min(hets$nTumorTotal) # 2
# # min(hets$nTumorAltGtyrMapped) # 1
# # min(hets$tumorAlt) # 1
# # min(hets$nTumorRefGtyrMapped) # 1
# # 
# # 
# # check <- genotypeDft1Cn2All %>%
# #   filter(status == "toCheck") # [1] 0 rows
# # 
# # gc()
# # 
# # > table(genotypeDft1Cn2All$nTumorTotal,
# # +       genotypeDft1Cn2All$status)
# #     
# #           het   homAlt   homRef unclassified
# #   0         0        0        0         9474
# #   1         0        0        0      1484358
# #   2  11985315  6833191 31103677            0
# #   3    726166   518714  2411845            0
# #   4   3388279  1496718  7446698            0
# #   5    236306   118006   631279            0
# #   6      7479     2168    11241            0
# #   7        31      136      279            0
# #   11      815      500     3500            0
# #   19        0        1        0            0
# 
# 
# majorityDft1Status <- genotypeDft1Cn2All %>%
#   filter(statusBroad != "unclassified") %>% # Exclude variants where CN < 2 when determining majority status. 
#   # Calculate majority status per SNP
#   group_by(locationRefAlt) %>%
#   mutate(
#     nHet       = sum(status == "het"),
#     nHomRef       = sum(status == "homRef"),
#     nHomAlt       = sum(status == "homAlt"),
#     nTumoursTotal = n_distinct(sampleTumor),
#     majorityStatus = case_when(
#       nHet > nHomRef & nHet > nHomAlt ~ "het",
#       nHomRef > nHomAlt & nHomRef > nHet ~ 'homRef',
#       nHomAlt > nHomRef & nHomAlt > nHet ~ 'homAlt',
#       TRUE        ~ "tied"
#     )
#   ) %>%
#   ungroup() 
# 
# # table(majorityDft1Status$majorityStatus)
# # 
# # > table(majorityDft1Status$majorityStatus)
# # 
# #      het   homAlt   homRef     tied 
# # 15938973  9134362 41845930     3079 
# 
# 
# 
# write_dataset(majorityDft1Status,
#               paste0(resDir, "/dft1Som/majorityDft1Status/"),
#               partitioning = "sampleTumor")
# 
# rm(genotypeDft1Cn2All, majorityDft1Status)
# 
# gc()
# # One row per sampleTumorLocationRefAlt
# majorityDft1Status <- open_dataset(paste0(resDir, 
#                                           "/dft1Som/majorityDft1Status")) %>%
#   mutate(tumorLocationRefAlt = paste0(sampleTumor, "_", locationRefAlt)) %>%
#   dplyr::select(locationRefAlt,
#                 CHROM,
#                 sampleTumor,
#                 status,
#                 statusBroad,
#                 nHet,
#                 nHomRef,
#                 nHomAlt,
#                 nTumoursTotal,
#                 majorityStatus, 
#                 tumorLocationRefAlt) %>%
#   distinct() %>%
#   dplyr::rename("nTumorsCn2" = "nTumoursTotal") %>%
#   mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   collect()
# 
# # Stage 1: test join, for single sampleTumorChrom only
# test <- dft1SomGenotyper %>%
#   filter(sampleTumorChrom == "18T_X") %>%
#   left_join(majorityDft1Status %>% dplyr::select(-c(CHROM, 
#                                                     sampleTumor, 
#                                                     locationRefAlt, 
#                                                     sampleTumorChrom)),
#             by = c("tumorLocationRefAlt")) %>%
#     mutate(
#     majorityCheck = case_when(
#       status == "unclassified" ~ "unclassified",
#       majorityStatus == "tied" ~ "tied",
#       majorityStatus == status ~ "match",
#       TRUE ~ "flip"
#     ),
#     transition = paste(status, majorityStatus, sep = "→"),
#     # -------------------------
#     # ALT correction logic
#     # -------------------------
#     nTumorAltCorrected = case_when(
#       # homAlt flipped to: ALT = total, REF = 0
#       majorityCheck == "flip" & majorityStatus == "homAlt" ~ nTumorTotal,
#       # homRef flipped to: ALT = 0, REF = total
#       majorityCheck == "flip" & majorityStatus == "homRef" ~ 0L,
#       # het → homAlt or homRef: split evenly, ALT = ceiling(total / 2)
#       majorityCheck == "flip" &
#         status == "het" &
#         majorityStatus %in% c("homAlt", "homRef") ~ as.integer(ceiling(nTumorTotal / 2)),
#       # default: keep original
#       TRUE ~ nTumorAltGtyrMapped
#     ),
#     # -------------------------
#     # REF correction logic
#     # -------------------------
#     nTumorRefCorrected = case_when(
#       # homRef flipped to: REF = total, ALT = 0
#       majorityCheck == "flip" & majorityStatus == "homRef" ~ nTumorTotal,
#       # homAlt flipped to: REF = 0, ALT = total
#       majorityCheck == "flip" & majorityStatus == "homAlt" ~ 0L,
#       # het → homAlt or homRef: split evenly, REF = floor(total / 2)
#       majorityCheck == "flip" &
#         status == "het" &
#         majorityStatus %in% c("homAlt", 
#                               "homRef") ~ as.integer(floor(nTumorTotal / 2)),
#       # default
#       TRUE ~ nTumorRefGtyrMapped
#     )
#   ) %>%
#   collect()
# 
# write_parquet(test, 
#               paste0(resDir, "dft1Som/18T_3_dft1SomGenotyperMajorityStatus.parquet"))
```


### 6.1.2 FINAL
- Filter dft1SomGenotyper with DNAseq samples of interest
```{r}
# dft1SomGenotyper <- open_dataset(paste0(resDir, "/dft1Som/dft1SomGenotyperDft1/")) %>%
#   mutate(locationRefAlt = paste0(locationRef, "_", ALT)) %>%
#   filter(sampleTumor %in% metadataPairs$tumorId)
# 
# # conflicting <- dft1SomGenotyper %>%
# #   filter(nTumorAltGtyrMapped != nTumorAltTest | nTumorRefGtyrMapped != nTumorRefTest) %>%
# #   collect() --> 9078 rows
# 
# dft1SomAlCoDf <- open_dataset(paste0(resDir, "/dft1SomAlCoDf/")) %>%
#   # filter(tumorIdRnaSeq %in% metadataPairs$tumorIdRnaSeq) %>%
#   left_join(metadataPairs %>% dplyr::select(tumorIdRnaSeq,
#                                             tumorId)) %>%
#   mutate(tumorLocationRefAlt = paste0(tumorId, "_", CHROM, ":", POS, "_", REF, "_", ALT),
#          sampleTumor = tumorId,
#          sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   dplyr::select(-c(tumorId)) %>%
#   dplyr::select(TOTAL,
#                 NREF,
#                 NALT,
#                 tumorIdRnaSeq,
#                 tumorLocationRefAlt)
# 
# dft1Merged <- dft1SomGenotyper %>%
#     ### Start with baseline genotyping results ###
#     ### Correct with majority status ###
#       mutate(
#         status = "somatic",
#         majorityStatus = "somatic",
#       majorityCheck = "somatic",
#       transition = "none",
#       # -------------------------
#       # ALT correction logic
#       # -------------------------
#       nTumorAltCorrected = nTumorAltGtyrMapped,
#       # -------------------------
#       # REF correction logic
#       # -------------------------
#       nTumorRefCorrected = nTumorRefGtyrMapped,
#     ) %>%
#     ### Merge RNAseq counts ###
#     left_join(dft1SomAlCoDf,
#               by = c("tumorLocationRefAlt")) %>%
#     mutate(location = paste0(CHROM, ":", POS)) %>%
#     ### Merge GTF info. Any rows that do not intersect an exon are NA ###
#     left_join(collapsed %>%
#                 dplyr::select(-c(CHROM,
#                                  POS)),
#               by = "location")
# 
# dft1MergedCol <- dft1Merged %>%
#   filter(is.na(query)==F) %>%
#   collect() ## 193320 variants.
# 
# # > length(unique(dft1MergedCol$gene_name))
# # [1] 2938
# 
# write_dataset(
#     dft1Merged,
#     paste0(resDir,
#            "/dft1Som/dft1SomAnnoStatusDft1/"),
#     partitioning = "sampleTumor"
#   )
```






# 4. Normalize nRef and nAlt
## 4.1 Calculate size factors
- Of note, sample 1538T1 was not filtered out properly, we don't have matched RNAseq data for this sample. Excluding here but need to go back and filter this. 
```{r}
joinedAnnoFiltered <- open_dataset(paste0(resDir, 
                             "/dft1Som/dft1SomAnnoStatus/")) %>%
    filter(
    (nTumorAltGtyrMapped > 0 & nHostAltGtyrMapped == 0) | # Tumor-inf
    (nTumorRefGtyrMapped > 0 & nHostRefGtyrMapped == 0) | # Tumor-inf
    (nHostAltGtyrMapped > 0 & nTumorAltGtyrMapped == 0) | # Host-inf
    (nHostRefGtyrMapped > 0 & nTumorRefGtyrMapped == 0)) %>% # Host-inf

  filter(is.na(query)==F) %>%
  collect()

# > nrow(dft1MergedColInformative)
# [1] 5497

# > length(unique(dft1MergedColInformative$gene_name))
# a <- unique(dft1MergedColInformative$gene_name) %>% as.data.frame()
# [1] 1853
```

### 4.1.1 Calculate size factors for all variant positions
- Using TOTAL at each variant position, compute size factors per sample. 
```{r}
dt <- copy(as.data.table(joinedAnnoFiltered))

# rm(joinedAnnoFiltered)

gc()

dt[, samplename := sampleTumor]

variants <- dt

# geometric means per variant
geomean <- function(x) {
  x <- x[!is.na(x) & x > 0]
  if (length(x) == 0) return(NA_real_)
  exp(mean(log(x)))
}

# geometric means per variant
gmeans <- variants[, .(gmean = geomean(TOTAL + 1)), by = locationRefAlt]

# na <- gmeans %>% filter(is.na(gmean))
# 
# dtNa <- dt %>%
#   filter(locationRefAlt %in% na$locationRefAlt)

setkey(gmeans, locationRefAlt)
setkey(variants, locationRefAlt)

variants[gmeans, gmean := i.gmean]

# ratios
variants[, scaled_total := TOTAL / gmean]

# size factors
size_factors <- variants[, .(
  size_factor = median(scaled_total, na.rm = TRUE)
), by = sampleTumor]

setkey(size_factors, sampleTumor)

dt[size_factors, on = "sampleTumor", size_factor := i.size_factor]

dt[, `:=`(
  NREF_norm = NREF / size_factor,
  NALT_norm = NALT / size_factor,
  TOTAL_norm = TOTAL / size_factor
)]

write_dataset(dt, 
              paste0(resDir, "dft1Som/dft1SomAnnoStatusNorm/"),
              format = "parquet",
              partitioning = "sampleTumor")
```

### 4.1.2 Calculate size factors for variant positions present in all tumors only
- No variants are shared by all tumors (nrow(chosen) = 0), thus no NREF_norm_all/NALT_norm_all.
- Not compatible with DFT1/DFT2 so leaving this out for now, potentially revisit!
```{r}
# n_samples <- uniqueN(dt$samplename)
# 
# chosen <- dt[TOTAL > 0,
#              .N, by = locationAlt][N == n_samples]
# 
# variants_all <- dt[locationAlt %in% chosen$locationAlt]
# 
# gmeans_all <- variants_all[, .(gmean = geomean(TOTAL + 1)),
#                            by = locationAlt]
# 
# variants_all[gmeans_all, on = "locationAlt", gmean := i.gmean]
# variants_all[, scaled_total := TOTAL / gmean]
# 
# size_factors_all <- variants_all[, .(
#   size_factor_all = median(scaled_total, na.rm = TRUE)
# ), by = samplename]
# 
# dt[size_factors_all, on = "samplename",
#    size_factor_all := i.size_factor_all]
# 
# dt[, `:=`(
#   NREF_norm_all = NREF / size_factor_all,
#   NALT_norm_all = NALT / size_factor_all
# )]
```

## 4.2 Normalize expression by copy number of informative variant
```{r}

```

## 4.3 Compute eT and eH
```{r}

```



