---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
```{r}
# install.packages("slider")
```

### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(arrow)
library(furrr)
library(patchwork)
library(slider)
library(duckdb)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_germline_dft1"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
source("~/gitClones/08_deconvTable/082_Scripts/genotype_germline_variants.R")
histoDir <- paste0(plotDir, 
                   "histo/")
dir.create(histoDir)
barDir <- paste0(plotDir, 
                 "bar/")
dir.create(barDir)


```

## 0.3 Custom functions
### 0.3.1 Compute tAlt
```{r}
computeTAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         hTotal, 
                         hAlt) {
  
    print("Solving for tAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    tAlt = ((obsVaf*denominator)-(1-purityTumor)*hAlt)/purityTumor
  return(tAlt)
}

assign_nAlt_from_vaf <- function(vaf, total_cn) {
  
  if (is.na(total_cn) || is.na(vaf)) return(NA_integer_)
  
  if (total_cn == 0) return(0)
  
  if (total_cn == 1) {
    return(ifelse(vaf < 0.5, 0, 1))
  }
  
  if (total_cn == 2) {
    if (vaf < 0.25) return(0)
    else if (vaf < 0.75) return(1)
    else return(2)
  }
  
  if (total_cn == 3) {
    if (vaf < 0.15) return(0)
    else if (vaf < 0.5) return(1)
    else if (vaf < 0.82) return(2)
    else return(3)
  }
  
  if (total_cn == 4) {
    if (vaf < 0.12) return(0)
    else if (vaf < 0.39) return(1)
    else if (vaf < 0.64) return(2)
    else if (vaf < 0.86) return(3)
    else return(4)
  }
  
  return(NA_integer_)  # fallback
}
```
### 0.3.2 Compute hAlt
```{r}
computeHAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         tAlt,
                         hTotal) {
  
  # Cases where hAlt = 0, solve for tAlt
 
    print("Solving for hAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    hAlt = (obsVaf*denominator)/(1-purityTumor)
  return(hAlt)
}

computeHAlt(obsVaf = 0.5, 
            purityTumor = 0.8, 
            tTotal = 4, 
            hTotal = 2, 
            tAlt=0)



```
### 0.3.3 Geometric mean
- Kevin's scripts
  - To calculate size factors: 
  - Compute the geom.mean of totals for each variant (optional: add pseudocount)
  - Divide each variant's read total by the geometric mean for each variant
  - Then take the median of these ratios for each sample
  - (Optional) restrict to variants with presence in all tumors
```{r}
geomean <- function(x) {
  if (all(x == 0)) {
    NA_real_
  } else {
    exp(mean(log(x[x > 0])))
  }
}
```

### 0.3.4 Summarize germline variants
```{r}
library(dplyr)
library(tidyr)

library(data.table)

summarise_variants_dt <- function(df) {
  dt <- as.data.table(df)
  
  # ensure uniqueness per tumor per location
  dt <- unique(dt[, .(locationAlt, sampleTumor, statusUpdate)])
  
  # count per location + status
  counts <- dt[, .N, by = .(locationAlt, statusUpdate)]
  
  # total tumors per location
  totals <- dt[, .(total_tumors = uniqueN(sampleTumor)), by = locationAlt]
  
  # merge counts with totals
  counts <- counts[totals, on = "locationAlt"]
  
  # compute percentage
  counts[, pct := N / total_tumors]
  
  # reshape wide (one column per statusUpdate)
  wide <- dcast(
    counts,
    locationAlt + total_tumors ~ statusUpdate,
    value.var = "pct",
    fill = 0
  )
  
  # compute majority
  status_cols <- setdiff(names(wide), c("locationAlt", "total_tumors"))
  
  wide[, majority := {
    vals <- unlist(.SD)
    max_val <- max(vals)
    winners <- names(vals)[vals == max_val]
    
    if (length(winners) > 1) "flag" else winners
  }, by = locationAlt, .SDcols = status_cols]
  
  return(wide)
}
```

### 0.3.4 Copy number state sequence
```{r}
possible_cn_absorbing <- function(start_cn, nGain, nLoss) {
  # Each state is a pair: CN, remaining events (gains and losses)
  states <- list(list(cn = start_cn, g = nGain, l = nLoss))
  
  final_cns <- c()
  
  while (length(states) > 0) {
    new_states <- list()
    
    for (s in states) {
      cn <- s$cn
      g  <- s$g
      l  <- s$l
      
      if (cn == 0 || (g == 0 && l == 0)) {
        # CN is 0 (absorbing) or no events left → record final CN
        final_cns <- c(final_cns, cn)
      } else {
        # Apply a gain if available
        if (g > 0 && cn > 0) {
          new_states <- append(new_states, list(list(cn = cn + 1, g = g - 1, l = l)))
        }
        # Apply a loss if available
        if (l > 0) {
          new_cn <- max(0, cn - 1)
          new_states <- append(new_states, list(list(cn = new_cn, g = g, l = l - 1)))
        }
      }
    }
    
    states <- new_states
  }
  sort(unique((final_cns)))
}

possible_cn_absorbing(3, 2, 3)
```
# 1 Data
## 1.1 Purities and ploidy
```{r}
germlinePurities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(germlinePurities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(germlinePurities),
  paste0(colnames(germlinePurities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(germlinePurities) <- newNames
germlinePurities <- germlinePurities[-1, ]

germlinePuritiesPloidy <- germlinePurities %>%
  mutate(ploidy = ifelse(X15_PLOIDY == "diploid",
                 2, 
                 ifelse(X15_PLOIDY == "tetraploid", 
                        4, 
                        "toCheck") 
                 )) %>%
  filter(ploidy != "toCheck") %>%
  dplyr::rename("sampleTumor"="TUMOUR.INFORMATION_TUMOUR ID", 
                     "purity"="X14_PURITY (VAF-based)", 
                "dftType"="X9_LINEAGE") %>%
              mutate(purity = as.numeric(purity), 
                     ploidy = as.numeric(ploidy), 
                     sex=ifelse(HOST.SEX == "♂", 
                                "male", 
                                ifelse(HOST.SEX == "♀", 
                                       "female", 
                                       "toCheck"))) %>%
              dplyr::select(sampleTumor, 
                            ploidy, 
                            purity, 
                            sex, 
                            dftType, 
                            X15_PLOIDY) 
  
table(germlinePuritiesPloidy$sex, 
      germlinePuritiesPloidy$dftType)

males <- germlinePuritiesPloidy %>%
  filter(sex == "male") 

dft1 <- germlinePuritiesPloidy %>%
  filter(dftType == "DFT1") 
```
## 1.2 VCF files
- Read in germline vcf file, all variants. 
- Pivot longer so that each variant and each sample is a single row with the ALT/TOTAL ratio.
```{r}
# germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"),
#                       col_types = cols(.default = "c"))
# # 
# # # Starting with autosomal germline and somatic substitution variants, we identified genomic
# # # intervals containing germline absolute copy number polymorphisms (see section 7.1) and removed
# # # germline substitutions occurring either within these intervals, or within 10 kb on either side of
# # # these intervals.
# # # This alignment included 1,175,235 substitution variants
# # # (1,070,436 germline, 104,799 somatic).
# # # > 1175235-1070436
# # # [1] 104799 --> fall into a CNV region??
# # # # 
# # 
# sampleInfo <- tibble(sample = colnames(germline)[-(1:4)]) %>%
#   mutate(
#     sampleType = str_extract(sample, "T|H"),
#     tcgId = str_extract(sample, "^[^TH]+")
#   )
# 
# 
# germlineLong <- germline %>%
#   filter(nchar(REF) == 1,
#          nchar(ALT) == 1) %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "sample",
#     values_to = "ratio"
#   ) %>%
#   left_join(sampleInfo, by = "sample") %>%
#   transmute(
#     CHROM,
#     POS,
#     REF,
#     ALT,
#     tcgId,
#     sampleType,
#     ratio,
#     sample
#   )
# rm(germline)
# saveRDS(germlineLong,
#         paste0(resDir,
#                "08_deconvTable_germline_germlineLong.rds"))
# 
# 
# germlineLong <- readRDS(paste0(resDir,
#                "08_deconvTable_germline_germlineLong.rds"))
```

## 1.3 Pair tumor/host samples
- Using shared ID's pivot wider so that the ratio for matched tumor/host samples is side by side.
- Filter on variants where alt allele is present in only the host or the tumor sample within a pair! --> so that we have info. on tAlt or hAlt at that position.
- Tumor samples = 112
- Host samples = 81
- Paired = 103 --> 100 if I filter Max's table? Maybe the non-transmissible cases. 
- Add multiallelic annotation, ALL SAMPLES!!
```{r}
# tumorSamples <- germlineLong %>%
#   filter(sampleType == "T") %>%
#   dplyr::select(c("tcgId", "sample")) %>%
#   distinct()
# 
# normalSamples <- germlineLong %>%
#   filter(sampleType == "H") %>%
#   dplyr::select(c("tcgId", "sample")) %>%
#   distinct()
# 
# pairedSamples <- tumorSamples %>%
#   inner_join(normalSamples,
#             by= c("tcgId"))
# 
# # > setdiff(tumorSamples$tcgId, pairedSamples$tcgId). Checked all manually in Max's S2 table. None have a matched host. 
# #  [1] "49" --> no host.   "86" --> cell line + no host  "88" --> cell line + no host "102" --> no host "140" --> no host  "208"--> no host  "367"  "368"  "378"  "384"  "837"  "876"  "1071" "1515" "1524" "1553" "2692" "2695".
# 
# write.xlsx(pairedSamples,
#            paste0(paste0(resDir,
#                "08_deconvTable_germline_germlinePaired.xlsx")))
# 
# normal <- germlineLong %>%
#   filter(sampleType == "H") %>%
#   dplyr::rename(
#     vafHost = ratio,
#     sampleHost = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# gc()
# 
# tumor <- germlineLong %>%
#   filter(sampleType == "T") %>%
#   dplyr::rename(
#     vafTumor = ratio,
#     sampleTumor = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# gc()
# 
# paired <- tumor %>%
#   inner_join(normal,
#              by = c("CHROM",
#                     "POS", 
#                     "REF", 
#                     "ALT", 
#                     "tcgId"))
# 
# gc()
# 
# rm(tumor, normal, germlineLong)
# gc()
# 
# pairedNas <- paired %>%
#   filter(is.na(vafTumor) | is.na(vafHost)) # --> 0 rows!
# 
# write_dataset(paired,
#               format = "parquet",
#         paste0(resDir,
#                "08_deconvTable_germline_germlinePaired/"),
#         partitioning = "tcgId")
# 
# 
# germlinePaired <- open_dataset(paste0(resDir,
#                "08_deconvTable_germline_germlinePaired/")) %>%
#   mutate(
#     tumorAlt = as.integer(sub("/.*", "", vafTumor)),
#     tumorTotal = as.integer(sub(".*/", "", vafTumor)),
#     hostAlt = as.integer(sub("/.*", "", vafHost)),
#     hostTotal = as.integer(sub(".*/", "", vafHost))
#   ) %>%
#   mutate(obsVafTumor = tumorAlt/tumorTotal,
#          obsVafHost = hostAlt/hostTotal)%>%
#  mutate(locationRef = paste0(CHROM, ":", POS, "_", REF, "_", ALT),
#         tumorLocationRef = paste0(sampleTumor, 
#                                   "_", CHROM, ":", POS, "_", REF),
#         tumorLocationRefAlt = paste0(sampleTumor, 
#                                   "_", CHROM, ":", POS, "_", REF, "_", ALT), 
#         position = as.numeric(POS), 
#         tumor = as.character(sampleTumor))
# 
# # Overwrite data. 
# write_dataset(germlinePaired,
#               format = "parquet",
#               paste0(resDir,
#                      "08_deconvTable_germline_germlinePaired_computedVafs/"),
#               partitioning = "tcgId")
# chroms <- c(paste0(1:6), 
#             "X", 
#             "Y")
# 
# germlinePaired <- open_dataset(paste0(resDir,
#                                       "08_deconvTable_germline_germlinePaired_computedVafs/")) %>%
#   mutate(locationRef = paste0(CHROM, ":", POS, "_", REF),
#          tumorLocation = paste0(sampleTumor, "_", CHROM, ":", POS), 
#          tumorLocationAlt =paste0(sampleTumor, "_", CHROM, ":", POS, "_", ALT)) %>%
#   filter(CHROM %in% chroms)
```

## 1.4 Annotate multiallelic variants. 
- Germline (host+tumour) variants are defined as those with VAF >0.25 in any host.
- That could mean it has 0 VAF in a given tumor sample, but this was not observed in all samples and is therefore not somatic. 
- How to account for this when calling "true" informative variants?
```{r}
# multiallelicSites <- germlinePaired %>%
#   dplyr::select(tumorLocationRef, ALT, sampleTumor) %>%
#   distinct() %>%
#   group_by(tumorLocationRef) %>%
#   summarise(nAltAlleles = n_distinct(ALT), .groups = "drop") %>%
#   filter(nAltAlleles > 1) %>%
#   collect()
# 
# nAltAlleles <- germlinePaired %>%
#   dplyr::select(tumorLocationRef, ALT, sampleTumor) %>%
#   distinct() %>%
#   filter(tumorLocationRef %in% multiallelicSites$tumorLocationRef) %>%  # Arrow predicate pushdown
#   collect() %>%
#   group_by(tumorLocationRef) %>%
#   summarise(
#     nAltAlleles     = n_distinct(ALT),
#     altAlleles      = paste(sort(unique(ALT)), collapse = ","),
#     baseA = ifelse("A" %in% ALT, paste(sampleTumor[ALT == "A"], collapse = ","), NA),
#     baseC = ifelse("C" %in% ALT, paste(sampleTumor[ALT == "C"], collapse = ","), NA),
#     baseG = ifelse("G" %in% ALT, paste(sampleTumor[ALT == "G"], collapse = ","), NA),
#     baseT = ifelse("T" %in% ALT, paste(sampleTumor[ALT == "T"], collapse = ","), NA),
#     .groups = "drop"
#   )
# 
# All tumors have a count at these variant, just flag to exclude from majority calculation?? 
# test <- germlinePaired %>%
#   filter(tumorLocationRef %in% multiallelicSites$tumorLocationRef) %>%
#   collect()
# 
# write.xlsx(test,
#            paste0(resDir, "/dft1Germline/08_deconvTable_germlineVariants_multiallelicData.xlsx"))
# 
# # Join germlinePaired with multiallelic annotation. 
# multiallelicSites <- multiallelicSites %>%
#   mutate(multiallelic = 1L)
# 
# germlinePaired <- germlinePaired %>%
#   left_join(multiallelicSites %>%
#               dplyr::select(tumorLocationRef, 
#                             multiallelic)) %>%
#   mutate(multiallelic = ifelse(is.na(multiallelic), 0L, multiallelic))
# 
# 
# germlinePaired <- write_dataset(germlinePaired,
#                                 paste0(resDir,
#                                       "08_deconvTable_germline_germlinePaired_computedVafs_multiAnno/"),
#                                 partitioning = "tcgId",
#                                 format = "parquet")
# 

```

## 1.5 Filter tumors of interest 
```{r}
# germlinePaired <- open_dataset(paste0(resDir,
#                                   "08_deconvTable_germline_germlinePaired_computedVafs_multiAnno/")) %>%
#   left_join(germlinePuritiesPloidy %>%
#               dplyr::select(sampleTumor, X15_PLOIDY)) %>%
#   filter(sampleTumor %in% dft1$sampleTumor) %>%
#   mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM))
```


# 2 CNVs - DFT1
## 2.1 Read in CNV data, join with VAF's
- Run selection as background job (Code/scroll down), copy output to global env. 
```{r}
dft1Cnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx",
                     sheet = "Table S6B",
                     startRow = 3)

dft1CnvLong <- dft1Cnv %>%
  pivot_longer(-c(1:12),
              names_to = "tumor",
              values_to = "presence") %>%
  mutate(presence = as.numeric(presence)) %>%
  filter(presence > 0) %>%
  mutate(
    START = as.numeric(START),
    END = as.numeric(END),
    CHROM = as.character(CHROM),
    tumor = as.character(tumor)
  ) %>%
  # write as camel case for sql with duck db.
  mutate(eventType = EVENT.TYPE,
         cnvEventId = CNV.EVENT.ID,
         sampleTumorChrom = paste0(tumor, "_", CHROM)) %>%
  dplyr::select(-c("CNV.EVENT.ID",
                   "WIDTH",
                   "NCBI.ANNOTATION",
                   "ENSEMBL.ANNOTATION",
                   "COSMIC",
                   "N",
                   "EVENT.TYPE",
                   "MARKER.5",
                   "DOUBLE.MINUTE"))


rm(dft1Cnv)
gc()
# 
# con <- dbConnect(duckdb(), config = list(threads = availableCores() - 1))
# 
# duckdb_register_arrow(con, "germlinePaired", germlinePaired)  # no collect() needed
# duckdb_register(con, "cnv", dft1CnvLong)
# 
# # Write parquet directly from DuckDB — never fully loads result into RAM
# dbExecute(con, paste0("
#   COPY (
#     SELECT g.*, c.presence, c.CHROM, c.START, c.END, c.cnvEventId
#     FROM germlinePaired g
#     LEFT JOIN cnv c
#            ON  g.CHROM    = c.CHROM
#            AND g.tumor    = c.tumor
#            AND g.position >= c.START
#            AND g.position <= c.END
#   ) TO '", resDir, "/dft1Germline/dft1CnvJoin/'
#   (FORMAT PARQUET, PARTITION_BY (sampleTumorChrom), OVERWRITE)
# "))
# 
# dbDisconnect(con, shutdown = TRUE)
# 
# # [1] 60328849  
# 
# 
# # # Which variants have > 0 changes?
# # presence <- open_dataset(paste0(resDir,
# #                                 "/dft1Germline/dft1CnvJoin/")) %>%
# #   filter(presence > 0) %>%
# #   dplyr::select(tumorLocationRefAlt) %>%
# #   collect()
# # 
# # # > nrow(presence)
# # # [1] 6556035
# # 
# # # Which variants have > 0 changes?
# # absence <- open_dataset(paste0(resDir,
# #                                 "/dft1Germline/dft1CnvJoin/")) %>%
# #   filter(presence == 0 | is.na(presence)) %>%
# #   dplyr::select(tumorLocationRefAlt) %>%
# #   collect()
# # 
# # [1] 53772814
```

## 2.3 Starting CN
- Adjust this for all variants. 
- Important to define for running genotyper(). 
- Case when need to keep input type the same (i.e., character string or numeric). Revert back to numeric after checking there are no "toCheck" rows. 
```{r}
autosomes <- paste0(1:6)

dft1CnvJoin <- open_dataset(paste0(resDir,
                                   "/dft1Germline/dft1CnvJoin/")) %>%
  # We left out event type to run DuckDb. Merge eventType with dft1CnvJoin.
  left_join(dft1CnvLong %>%
              dplyr::select(cnvEventId,
                            eventType) %>%
              distinct()) %>%
  left_join(germlinePuritiesPloidy %>%
              dplyr::select(X15_PLOIDY,
                            sampleTumor)) %>%
  dplyr::select(-c(CHROM_1)) %>%
  # Set the chrom type.
  mutate(chromType = ifelse(CHROM %in% autosomes,
                            "autosome",
                            CHROM)) %>%
  # Define CN by combination of dft type, ploidy, and chromType
  mutate(postWgdCn = case_when(
    ############################
    ## Diploid DFT1s ##
    ############################
    # Diploid, DFT1, autosome (2 copies)
    X15_PLOIDY == "diploid" & sampleTumor %in% dft1$sampleTumor & chromType == "autosome" ~ "2",
    # Diploid, DFT1, sex (XX)
    X15_PLOIDY == "diploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ "2",
    # Diploid, DFT1, no Y
    X15_PLOIDY == "diploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ "0",
    ############################
    ## Tetraploid DFT1s ##
    ############################
    # Tetraploid, DFT1, autosome
    X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & chromType == "autosome" ~ "4",
    # Tetraploid, DFT1, sex (XXXX)
    X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ "4",
    # Tetraploid, DFT1, no Y chromosome
    X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ "0",
    # ############################
    # ## Diploid DFT2s ##
    # ############################
    # # Diploid, DFT2, autosome
    # X15_PLOIDY == "diploid" & sampleTumor %in% dft2$sampleTumor & chromType == "autosome" ~ "2",
    # # Diploid, DFT2, sex (XY)
    # X15_PLOIDY == "diploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "X" ~ "1",
    # X15_PLOIDY == "diploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "Y" ~ "1",
    # ############################
    # ## Tetraploid DFT2s ##
    # ############################
    # # Tetraploid, DFT2, autosome
    # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & chromType == "autosome" ~ "4",
    # # Tetraploid, DFT2, sex (XXYY)
    # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "X" ~ "2",
    # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "Y" ~ "2",
    TRUE ~ "toCheck"
  )) %>%
    mutate(preWgdCn = case_when(
    ############################
    ## Tetraploid DFT1s ##
    ############################
    # Tetraploid, DFT1, autosome
    X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & chromType == "autosome" ~ "2",
    # Tetraploid, DFT1, sex (XX)
    X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ "2",
    # Tetraploid, DFT1, no Y chromosome
    X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ "0",
    # ############################
    # ## Tetraploid DFT2s ##
    # ############################
    # # Tetraploid, DFT2, autosome, preWgd = 2 copies
    # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & chromType == "autosome" ~ "2",
    # # Tetraploid, DFT2, sex (XXYY), preWgd = XY (1 copy)
    # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "X" ~ "1",
    # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "Y" ~ "1",
    TRUE ~ postWgdCn
    ))



# toCheck <- dft1CnvJoin %>%
#   filter(preWgdCn == "toCheck") %>%
#   collect() # No rows!!

rm(dft1CnvLong, toCheck)
gc()

dft1CnvJoin <- dft1CnvJoin %>%
  mutate(preWgdCn = as.numeric(preWgdCn),
         postWgdCn = as.numeric(postWgdCn)) %>%
  collect() 

# > table(dft1CnvJoin$CHROM, dft1CnvJoin$preWgdCn)
#    
#            0        2
#   1        0 13973698
#   2        0 12834342
#   3        0 11755027
#   4        0  8564237
#   5        0  5859511
#   6        0  6898568
#   X        0   443411
#   Y       55        0 --> female devil, ignore Y vars.

# > table(dft1CnvJoin$CHROM, dft1CnvJoin$postWgdCn)
#    
#            0        2        4
#   1        0 10662580  3311118
#   2        0  9812765  3021577
#   3        0  8970318  2784709
#   4        0  6562442  2001795
#   5        0  4463151  1396360
#   6        0  5173140  1725428
#   X        0   335227   108184
#   Y       55        0        0

write_dataset(dft1CnvJoin,
              paste0(resDir, "/dft1Som/dft1CnvJoin_prePostAdjustedCn/"),
              partitioning = "sampleTumorChrom",
              format = "parquet")

rm(dft1CnvJoin)
gc()
```



## 2.2 Adjust CN
```{r}
# dft1CnvJoin <- open_dataset(paste0(resDir,
#                                    "/dft1Germline/",
#                                    fprefix,
#                                    "_dft1CnvJoin/")) %>%
#   # We left out event type to run DuckDb. Merge eventType with dft1CnvJoin.
#   left_join(dft1CnvLong %>%
#               dplyr::select(cnvEventId,
#                             eventType) %>%
#               distinct()) %>%
#   left_join(germlinePuritiesPloidy %>% 
#               dplyr::select(X15_PLOIDY, 
#                             sampleTumor))
# 
# 
# # --- 1. Find duplicate keys via aggregation. These are rows where there is more than one CNV event ID spanning a given variant (i.e., a multiplet). This does not overestimate the number of multiplets because of multiallelic variants, because I am counting by ref and alt.
# dup_keys_refalt <- dft1CnvJoin %>%
#   filter(presence > 0) %>%
#   count(tumorLocationRefAlt) %>%
#   collect() %>%
#   filter(n > 1) %>%
#   pull(tumorLocationRefAlt)
# 
# # > length(dup_keys_refalt)
# # [1] 520010
# 
# # --- 2. Find single keys via aggregation. These are rows where there one CNV event ID spanning a given variant (i.e., a singlet). This does not overestimate the number of multiplets because of multiallelic variants, because I am counting by ref and alt.
# sing_keys_refalt <- dft1CnvJoin %>%
#   filter(presence > 0) %>%
#   count(tumorLocationRefAlt) %>%
#   collect() %>%
#   filter(n == 1) %>%
#   pull(tumorLocationRefAlt)
# 
# > length(sing_keys_refalt)
# [1] 6081404
# 
# dft1CnvJoinMultiplets <- dft1CnvJoin %>%
#   filter(tumorLocationRefAlt %in% dup_keys_refalt) %>%
#   mutate(multipletsRefAlt = ifelse(tumorLocationRefAlt %in% dup_keys_refalt,
#                 1,
#                 0),
#          singletsRefAlt = ifelse(tumorLocationRefAlt %in% sing_keys_refalt,
#                            1,
#                            0),
#          cnvType = ifelse(tumorLocationRefAlt %in% dup_keys_refalt,
#                           "multiplet",
#                           ifelse(tumorLocationRefAlt %in% sing_keys_refalt,
#                                  "singlet",
#                                  "noChange")))%>% # None should be assigned "noChange" at this point, because we have prefiltered on variants where presence > 0.
#   mutate(variantType = "dft1Germline") %>%
#   left_join(germlinePuritiesPloidy) %>%
#   ### Special condition for CN of DFT2 tumors only
#   ### Adjust this before merging with CNV information, so that starting point (before copy number changes) is accurate.
#   mutate(ploidy = ifelse(CHROM %in% c("X", "Y") & dftType == "DFT2",
#                          ploidy / 2, ## In DFT2 tumors, either XY or XXYY
#                          ploidy)) %>%
# 
#   ### Set ploidy to 0 at Y chrom of DFT1 tumors.
#   mutate(ploidy = ifelse(CHROM %in% c("Y") & dftType == "DFT1",
#                          0, ## In DFT1 tumors, either XX or XXXX, so set Y to 0.
#                          ploidy),
#          sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) ## X starting ploidy = same as autosome CN (i.e., 2 or 4)
# 
# write_dataset(dft1CnvJoinMultiplets,
#               format = "parquet",
#               paste0(resDir,
#                      "/dft1Germline/",
#                      fprefix,
#                      "_dft1CnvJoinMultiplets/"),
#               partitioning = "sampleTumorChrom")
# 
# 
# dft1CnvJoinSinglets <- dft1CnvJoin %>%
#   filter(tumorLocationRefAlt %in% sing_keys_refalt) %>%
#   mutate(multipletsRefAlt = ifelse(tumorLocationRefAlt %in% dup_keys_refalt,
#                 1,
#                 0),
#          singletsRefAlt = ifelse(tumorLocationRefAlt %in% sing_keys_refalt,
#                            1,
#                            0),
#          cnvType = ifelse(tumorLocationRefAlt %in% dup_keys_refalt,
#                           "multiplet",
#                           ifelse(tumorLocationRefAlt %in% sing_keys_refalt,
#                                  "singlet",
#                                  "noChange")))%>% # None should be assigned "noChange" at this point, because we have prefiltered on variants where presence > 0.
#   mutate(variantType = "dft1Germline") %>%
#   left_join(germlinePuritiesPloidy) %>%
#   ### Special condition for CN of DFT2 tumors only
#   ### Adjust this before merging with CNV information, so that starting point (before copy number changes) is accurate.
#   mutate(ploidy = ifelse(CHROM %in% c("X", "Y") & dftType == "DFT2",
#                          ploidy / 2, ## In DFT2 tumors, either XY or XXYY
#                          ploidy)) %>%
# 
#   ### Set ploidy to 0 at Y chrom of DFT1 tumors.
#   mutate(ploidy = ifelse(CHROM %in% c("Y") & dftType == "DFT1",
#                          0, ## In DFT1 tumors, either XX or XXXX, so set Y to 0.
#                          ploidy),
#          sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) ## X starting ploidy = same as autosome CN (i.e., 2 or 4)
# 
# write_dataset(dft1CnvJoinSinglets,
#               format = "parquet",
#               paste0(resDir,
#                      "/dft1Germline/",
#                      fprefix,
#                      "_dft1CnvJoinSinglets/"),
#               partitioning = "sampleTumorChrom")
# 
# 
# rm(dft1CnvJoin)
# gc()
# 
# # 60 variants.
# dft1Test <- open_dataset(paste0(resDir,
#                      "/dft1Germline/",
#                      fprefix,
#                      "_dft1CnvJoinMultiplets/")) %>%
# 
#   group_by(tumorLocationRefAlt) %>%
# 
#   filter(multipletsRefAlt == 1 & presence > 0) %>%
# 
#   collect() %>%
# 
#   mutate(
#     nGain = sum(eventType == "GAIN"),
#     nLoss = sum(eventType == "LOSS"),
#     nGainPost = sum(eventType == "GAIN (post-tetraploidisation)"),
#     nGainPre = sum(eventType == "GAIN (pre-tetraploidisation)"),
#     nLossPost = sum(eventType == "LOSS (post-tetraploidisation)"),
#     nLossPre = sum(eventType == "LOSS (pre-tetraploidisation)")
#   ) %>%
# 
#   dplyr::select(-c(setdiff(colnames(dft1CnvLong), "tumor"))) %>%
#   distinct() %>%
#   rowwise()  %>%
#   filter(ploidy == 2 & nGain == 1 & nLoss == 2)
# 
# 
# 
# # Test functions
# dft1TestFunction <- dft1Test %>%
# 
#   mutate(
#     # ---- CASE 1: diploid ----
#     possibleCN = list(
#       if (ploidy == 2) {
#         possible_cn_absorbing(ploidy, nGain, nLoss)
# 
#       # ---- CASE 2: tetraploid with pre-events ----
#       } else if (ploidy == 4 & (nGainPre > 0 | nLossPre > 0)) {
# 
#         pre_cn <- possible_cn_absorbing(2, nGainPre, nLossPre)
#         pre_cn <- pre_cn * 2
# 
#         # apply post events to each possible pre state
#         unique(unlist(
#           lapply(pre_cn, function(cn) {
#             possible_cn_absorbing(
#               cn,
#               nGain + nGainPost,
#               nLoss + nLossPost
#             )
#           })
#         ))
# 
#       # ---- CASE 3: tetraploid without pre-events ----
#       } else if (ploidy == 4) {
# 
#         possible_cn_absorbing(
#           ploidy,
#           nGain + nGainPost,
#           nLoss + nLossPost
#         )
# 
#       } else {
#         NA_real_
#       }
#     ),
# 
#     minCN = min(possibleCN),
#     maxCN = max(possibleCN)
# 
#   ) %>%
#   ungroup() %>%
#   mutate(cnvAdjAnno = "multipleChangesTest")
# 
# dft1TestFunctionConflicting <- dft1TestFunction %>%
#   filter(minCN != maxCN)
# dft1CnvJoinSinglets <- open_dataset(paste0(resDir, 
#                      "/dft1Germline/",
#                      fprefix,
#                      "_dft1CnvJoinSinglets/"))
# 
# singletPositions <- dft1CnvJoinSinglets %>%
#   dplyr::select(tumorLocationRefAlt) %>%
#   distinct() %>%
#   collect()
# 
# dft1CnvJoinMultiplets<- open_dataset(paste0(resDir, 
#                      "/dft1Germline/",
#                      fprefix,
#                      "_dft1CnvJoinMultiplets/"))
# 
# multipletPositions <- dft1CnvJoinMultiplets %>%
#   dplyr::select(tumorLocationRefAlt) %>%
#   distinct() %>%
#   collect()
# 
# allPositions <- germlinePaired %>%
#   dplyr::select(tumorLocationRefAlt) %>%
#   distinct() %>%
#   collect()
# 
# 
# res <- list("allPositions"= allPositions$tumorLocationRefAlt,
#             "multipletPositions"=multipletPositions$tumorLocationRefAlt,
#             "singletPositions"=singletPositions$tumorLocationRefAlt)
# 
# saveRDS(res, 
#         paste0(resDir, 
#                "/dft1Germline/", 
#                fprefix, 
#                "positions.rds"
#                ))
# 
# noChangePositions <- readRDS(paste0(resDir,
#                              "/dft1Germline/",
#                              fprefix,
#                              "noChangePositions.rds"))
# 
# saveRDS(noChangePositions,
#         paste0(resDir,
#                "/dft1Germline/",
#                fprefix,
#                "noChangePositions.rds"
#                ))
# rm(res)
# gc()
```

### 2.2.1 Variants with no change
- copyNumberAdjusted = ploidy
- read in multiplets table, these are variants where presence > 0 intersected a CNV. 
- if tumorLocationRefAlt not in multiplets table, no change!
- number of unique variants (taking into account alt) = number of rows, because no changes to cnv. YES? --> multiallelic variants possible (i.e., subclones, because these variants don't occur in any of the host samples.)
```{r}
# dft1CnvJoinNoChanges <- germlinePaired %>%
#   left_join(germlinePuritiesPloidy) %>%
#   filter(tumorLocationRefAlt %in% noChangePositions) %>%
#   mutate(ploidy = ifelse(CHROM %in% c("X", "Y") & dftType == "DFT2",
#                          ploidy / 2, ## In DFT2 tumors, either XY or XXYY
#                          ploidy)) %>% # In DFT1 tumors, either XX or XXXX
#   mutate(ploidy = ifelse(CHROM %in% c("Y") & dftType == "DFT1",
#                          0, ## In DFT1 tumors, either XX or XXXX, so set Y to 0.
#                          ploidy),
#          sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>% ## X starting ploidy = same as autosome CN (i.e., 2 or 4)
#   mutate(copyNumberAdjusted = ploidy) %>%
#   group_by(tumorLocationRefAlt) %>%
#   mutate(
#     nGain = 0,
#     nLoss = 0,
#     nGainPost = 0,
#     nGainPre = 0,
#     nLossPost = 0,
#     nLossPre = 0,
#     sumPresence = 0
#   ) %>%
#   # remove CNV table columns
#   mutate(cnvAdjAnno = "noChanges", 
#          multipletsRefAlt = 0, 
#          singletsRefAlt = 0, 
#          variantType = "dft1Germline", 
#          CHROM_1 = CHROM)
# 
# gc()
# 
# write_dataset(dft1CnvJoinNoChanges,
#               paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinNoChanges/"),
#               partitioning = "sampleTumorChrom",
#               format = "parquet")
# 
# gc()
```

### 2.2.2 Variants with single change
- change ploidy according to single step
- number of unique variants = number of rows, because only 1 change to cnv per variant. 5 tumorLocation where more than one variant with a single CNV change.
- All presences should be 1 because single change
- If CN goes to 0, it should stay at 0. (Applies for multistep, not here.) Running same code as multiplets as a sanity check, none of the maxCn should differ from minCn. 
```{r}
# # dft1CnvJoinSingleChange <- open_dataset(paste0(resDir, "/dft1Germline/",
# #                                             fprefix,
# #                                             "_dft1CnvJoinMultiplets/")) %>%
# #   # has a CNV, but a singlet!!
# #   filter(presence > 0 & multipletsRefAlt == 0) %>%
# #   group_by(tumorLocationRefAlt) %>%
# #   mutate(
# #     nGain = sum(eventType == "GAIN"),
# #     nLoss = sum(eventType == "LOSS"),
# #     nGainPost = sum(eventType == "GAIN (post-tetraploidisation)"),
# #     nGainPre = sum(eventType == "GAIN (pre-tetraploidisation)"),
# #     nLossPost = sum(eventType == "LOSS (post-tetraploidisation)"),
# #     nLossPre = sum(eventType == "LOSS (pre-tetraploidisation)"),
# #     sumPresence = sum(presence)
# #   ) %>%
# #    # remove CNV table columns, because this has been aggregated by summing rows.
# #   dplyr::select(-c(cnvEventId,
# #                    eventType,
# #                    presence)) %>%
# #   mutate(
# #     copyNumberAdjusted = pmax(0,
# #     case_when(
# #       # Case 1: At sex chromosomes, starting CN of tumor is 1 because DFT has a male founder devil.
# #       ploidy == 1 ~ ploidy + nGain - nLoss,
# #
# #       # Case 1: diploid
# #       ploidy == 2 ~ ploidy + nGain - nLoss,
# #
# #       # Case 2: tetraploid with pre-tetra events present
# #       ploidy == 4 & (nGainPre > 0 | nLossPre > 0) ~
# #         (2 + nGainPre - nLossPre) * 2 +
# #         (nGain + nGainPost) - (nLoss + nLossPost),
# #
# #       # Case 3: tetraploid without pre-tetra events present
# #       ploidy == 4 & nGainPre == 0 & nLossPre == 0 ~
# #         ploidy + (nGain + nGainPost) - (nLoss + nLossPost),
# #
# #       TRUE ~ NA_real_
# #     )
# #    )
# #   ) %>%
# #   mutate(cnvAdjAnno = "singleChange")
# #
# # write_dataset(dft1CnvJoinSingleChange,
# #               paste0(resDir, "/dft1Germline/",
# #                                             fprefix,
# #                                             "_dft1CnvJoinSingleChange/"),
# #               partitioning = "tcgId")
# 
# con <- dbConnect(duckdb())
# 
# duckdb_register_arrow(con, "singlets",
#   open_dataset(paste0(resDir,
#                       "/dft1Germline/", 
#                       fprefix, 
#                       "_dft1CnvJoinSinglets/")))
# 
# dbExecute(con, paste0("
#   COPY (
#     SELECT
#       * EXCLUDE (cnvEventId, eventType, presence),
#       'singleChange' AS cnvAdjAnno,
#       -- window aggregations
#       SUM(CASE WHEN eventType = 'GAIN'                          THEN 1 ELSE 0 END)
#         OVER (PARTITION BY tumorLocationRefAlt) AS nGain,
#       SUM(CASE WHEN eventType = 'LOSS'                          THEN 1 ELSE 0 END)
#         OVER (PARTITION BY tumorLocationRefAlt) AS nLoss,
#       SUM(CASE WHEN eventType = 'GAIN (post-tetraploidisation)' THEN 1 ELSE 0 END)
#         OVER (PARTITION BY tumorLocationRefAlt) AS nGainPost,
#       SUM(CASE WHEN eventType = 'GAIN (pre-tetraploidisation)'  THEN 1 ELSE 0 END)
#         OVER (PARTITION BY tumorLocationRefAlt) AS nGainPre,
#       SUM(CASE WHEN eventType = 'LOSS (post-tetraploidisation)' THEN 1 ELSE 0 END)
#         OVER (PARTITION BY tumorLocationRefAlt) AS nLossPost,
#       SUM(CASE WHEN eventType = 'LOSS (pre-tetraploidisation)'  THEN 1 ELSE 0 END)
#         OVER (PARTITION BY tumorLocationRefAlt) AS nLossPre,
#       SUM(presence)
#         OVER (PARTITION BY tumorLocationRefAlt) AS sumPresence,
#       GREATEST(0,
#         CASE
#           -- 1) Diploid tumors
#           WHEN X15_PLOIDY != 'tetraploid'
#             THEN ploidy + nGain - nLoss
#           -- 2) Tetraploid tumors, sex chromosomes (ploidy == 2, base pre-WGD = 1)
#           WHEN X15_PLOIDY = 'tetraploid' AND chrom IN ('X', 'Y')
#             THEN CASE
#                    WHEN nGainPre > 0 OR nLossPre > 0
#                      THEN (1 + nGainPre - nLossPre) * 2 + (nGain + nGainPost) - (nLoss + nLossPost)
#                    ELSE
#                      ploidy + (nGain + nGainPost) - (nLoss + nLossPost)
#                  END
#           -- 3) Tetraploid tumors, autosomes (ploidy == 4, base pre-WGD = 2)
#           WHEN X15_PLOIDY = 'tetraploid'
#             THEN CASE
#                    WHEN nGainPre > 0 OR nLossPre > 0
#                      THEN (2 + nGainPre - nLossPre) * 2 + (nGain + nGainPost) - (nLoss + nLossPost)
#                    ELSE
#                      ploidy + (nGain + nGainPost) - (nLoss + nLossPost)
#                  END
#           ELSE NULL
#         END
#       ) AS copyNumberAdjusted
#     FROM singlets
#     WHERE presence > 0
#       AND multipletsRefAlt = 0
#   ) TO '", paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinSingleChange/"), "'
#   (FORMAT parquet, PARTITION_BY sampleTumorChrom, OVERWRITE)
# "))
# 
# dbDisconnect(con)
# gc()
# 
# [1] 6081404
# 
# 
# check <- open_dataset(paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinSingleChange/")) %>%
#   dplyr::select(copyNumberAdjusted, CHROM) %>%
#   collect()
# 
# table(check$CHROM, check$copyNumberAdjusted)
# 
#           0       1       2       3       4       5       6
#   1       0  420164     319  532505       0   12480       0
#   2       0  325346     537  178427       0   16497       0
#   3       0  610411      51  114104       0  205646   13723
#   4       0    1155     256 2634301       0  603265       6
#   5       0   85248     207  121721       0   79571       3
#   6       0   20870     823    5101       0   64159       0
#   X    4156   20760       0    8749     677     166       0
# 
# dft1CnvJoinSingleChange <- open_dataset(paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinSingleChange/"))
# 
# names(schema(dft1CnvJoinSingleChange))
```

### 2.2.3 Variants with multiple changes
- Aggregate steps using same logic as before, only this time ploidy will change using sum of all steps rather than single change of CNV status.
- Issue: we don't know the order of events. Depending on order of losses and gains, the final CN could differ. Adding a permutation function which returns how many times a different final CN state could occur. 
- None of the tetraploid tumors have nLossPre > 1, therefore cannot go to 0.
- The only variable cases are diploid tumors whose variants have two losses and one gain. 
- Filter out the five variants that could have been CN 0 or 1!
  - Check these variants against the counts + gtf file.
```{r}
# dft1CnvJoinMultipleChanges <- open_dataset(paste0(resDir, 
#                                                   "/dft1Germline/",
#                                                   fprefix,
#                                                   "_dft1CnvJoinMultiplets/")) %>%
# 
#   group_by(tumorLocationRefAlt) %>%
# 
#   filter(multipletsRefAlt > 0 & presence > 0) %>%
# 
#   mutate(
#     nGain = sum(eventType == "GAIN"),
#     nLoss = sum(eventType == "LOSS"),
#     nGainPost = sum(eventType == "GAIN (post-tetraploidisation)"),
#     nGainPre = sum(eventType == "GAIN (pre-tetraploidisation)"),
#     nLossPost = sum(eventType == "LOSS (post-tetraploidisation)"),
#     nLossPre = sum(eventType == "LOSS (pre-tetraploidisation)"),
#     sumPresence = sum(presence)
#   ) %>%
# 
#   # remove CNV table columns, because this has been aggregated by summing rows.
#   dplyr::select(-c(cnvEventId,
#                    eventType,
#                    presence)) %>%
#   distinct() %>%
#   # Collect to compute adjusted copy number at each SNP.
#   collect() %>%
#   rowwise() %>%
# 
#   mutate(
# 
#     # ---- CASE 1: diploid ----
#     possibleCn = list(
#       if (ploidy == 1) {
#         possible_cn_absorbing(ploidy, nGain, nLoss)
# 
#       # ---- CASE 2: tetraploid with pre-events ----
#       }
#       else if (ploidy == 2) {
#         possible_cn_absorbing(ploidy, nGain, nLoss)
# 
#       # ---- CASE 2: tetraploid with pre-events ----
#       } else if (ploidy == 4 & (nGainPre > 0 | nLossPre > 0)) {
# 
#         pre_cn <- possible_cn_absorbing(2, nGainPre, nLossPre)
#         pre_cn <- pre_cn * 2
# 
#         # apply post events to each possible pre state
#         unique(unlist(
#           lapply(pre_cn, function(cn) {
#             possible_cn_absorbing(
#               cn,
#               nGain + nGainPost,
#               nLoss + nLossPost
#             )
#           })
#         ))
# 
#       # ---- CASE 3: tetraploid without pre-events ----
#       } else if (ploidy == 4) {
# 
#         possible_cn_absorbing(
#           ploidy,
#           nGain + nGainPost,
#           nLoss + nLossPost
#         )
# 
#       } else {
#         NA_real_
#       }
#     ),
# 
#     minCn = min(possibleCn),
#     maxCn = max(possibleCn)
# 
#   ) %>%
#   ungroup()
# 
# gc()
# 
# # 60 conflicting rows, filter these out.
# multipleCns <- dft1CnvJoinMultipleChanges %>%
#   filter(minCn != maxCn)
# 
# dft1CnvJoinMultipleChanges <- open_dataset(paste0(resDir, 
#                                                   "/dft1Germline/",
#                                                   fprefix,
#                                                   "_dft1CnvJoinMultiplets/")) %>%
#   group_by(tumorLocationRefAlt) %>%
#   filter(multipletsRefAlt > 0 & presence > 0) %>%
#   mutate(
#     nGain     = sum(eventType == "GAIN"),
#     nLoss     = sum(eventType == "LOSS"),
#     nGainPost = sum(eventType == "GAIN (post-tetraploidisation)"),
#     nGainPre  = sum(eventType == "GAIN (pre-tetraploidisation)"),
#     nLossPost = sum(eventType == "LOSS (post-tetraploidisation)"),
#     nLossPre  = sum(eventType == "LOSS (pre-tetraploidisation)"),
#     sumPresence = sum(presence)
#   ) %>%
#   dplyr::select(-c(cnvEventId, eventType, presence)) %>%
#   distinct() %>%
#   collect() %>%
#   rowwise() %>%
#   mutate(
#     possibleCn = list(
#       # ---- CASE 1: Diploid tumors ----
#       if (X15_PLOIDY != "tetraploid") {
#         possible_cn_absorbing(ploidy, nGain, nLoss)
# 
#       # ---- CASE 2: Tetraploid, sex chromosomes (pre-WGD base = 1) ----
#       } else if (X15_PLOIDY == "tetraploid" & CHROM %in% c("X", "Y")) {
#         if (nGainPre > 0 | nLossPre > 0) {
#           pre_cn <- possible_cn_absorbing(1, nGainPre, nLossPre)
#           pre_cn <- pre_cn * 2
#           unique(unlist(
#             lapply(pre_cn, function(cn) {
#               possible_cn_absorbing(cn, nGain + nGainPost, nLoss + nLossPost)
#             })
#           ))
#         } else {
#           possible_cn_absorbing(ploidy, nGain + nGainPost, nLoss + nLossPost)
#         }
# 
#       # ---- CASE 3: Tetraploid, autosomes (pre-WGD base = 2) ----
#       } else if (X15_PLOIDY == "tetraploid") {
#         if (nGainPre > 0 | nLossPre > 0) {
#           pre_cn <- possible_cn_absorbing(2, nGainPre, nLossPre)
#           pre_cn <- pre_cn * 2
#           unique(unlist(
#             lapply(pre_cn, function(cn) {
#               possible_cn_absorbing(cn, nGain + nGainPost, nLoss + nLossPost)
#             })
#           ))
#         } else {
#           possible_cn_absorbing(ploidy, nGain + nGainPost, nLoss + nLossPost)
#         }
# 
#       } else {
#         NA_real_
#       }
#     ),
#     minCn = min(possibleCn),
#     maxCn = max(possibleCn)
#   ) %>%
#   ungroup()
# 
# gc()
# 
# # 60 conflicting rows, filter these out. --> 120 
# multipleCns <- dft1CnvJoinMultipleChanges %>%
#   distinct() %>%
#   filter(minCn != maxCn)
# 
# dft1CnvJoinMultipleChanges <- dft1CnvJoinMultipleChanges %>%
#   filter(minCn == maxCn) %>%
#   mutate(copyNumberAdjusted = minCn,
#          cnvAdjAnno = "multipleChanges") %>%
#   dplyr::select(names(schema(dft1CnvJoinSingleChange)))
# # 
# # write_dataset(dft1CnvJoinMultipleChanges,
# #               paste0(resDir, 
# #                      "/dft1Germline/",
# #                      fprefix,
# #                      "_dft1CnvJoinMultipleChanges/"),
# #               partitioning = "sampleTumorChrom")
# # gc()
# 
# checkNa <- dft1CnvJoinMultipleChanges %>%
#   filter(is.na(copyNumberAdjusted)) %>%
#   collect()
# 
# check <- dft1CnvJoinMultipleChanges %>%
#   dplyr::select(copyNumberAdjusted, CHROM) %>%
#   collect()
# 
# table(check$CHROM, 
#       check$copyNumberAdjusted)
# 
# dft1CnvJoinMultipleChanges <- dft1CnvJoinMultipleChanges %>%
#   # Filter any computation columns not needed for singleChange and noChange dataframes.
#   dplyr::select(setdiff(names(schema(dft1CnvJoinSingleChange)), c("START", 
#                                                                    "END", 
#                                                                    "cnvType", 
#                                                                    "cnvEventId"))) %>%
#   distinct()
# 
# write_dataset(dft1CnvJoinMultipleChanges,
#               paste0(resDir, "/dft1Germline/",
#                                             fprefix,
#                                             "_dft1CnvJoinMultipleChanges/"),
#               partitioning = "sampleTumorChrom")
# gc()
# 
# checkNa <- dft1CnvJoinMultipleChanges %>%
#   filter(is.na(copyNumberAdjusted)) %>%
#   collect()
# 
# check <- dft1CnvJoinMultipleChanges %>%
#   dplyr::select(copyNumberAdjusted, CHROM) %>%
#   collect()
# 
# table(check$CHROM, check$copyNumberAdjusted) ### Multiplets not affecting sex chromosomes.
```






# 3 Compute genotypes
This code joins the CNV adjusted tables from Chunks in Section 1.5.2. Then adds in host copy number info before calculating most likely genotype at each position. 

- In a pure sample, vafH = 0 + (1)*Halt/(0 + 1*Htotal). Therefore, Halt = vafH*Htotal. 

- If host sample is a male, total CN at sex chromosomes == 1 !!!

- If tumor sample is DFT2, total CN was adjusted at sex chromosomes == 1 or 2 depending on original ploidy. Depending on whether a SNP fell in a CNV, this value may have changed, and in reflected in copyNumberAdjusted. 

- If tumor sample is DFT1, total CN was adjusted at Y chromosomes as 0 is changed here. This is because, the variant cannot undergo copy number changes if it's starting point is 0.

- Of note, this will not filter out any of the tumors! Only do this at the stage where we merge with RNAseq data (see Chunk XYZ). 
```{r}
# noChanges <- open_dataset(paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinNoChanges/")) %>%
#   dplyr::select(tumorLocationRef) %>%
#   collect()
# singleChange <- open_dataset(paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinSingleChange/"))  %>%
#   dplyr::select(tumorLocationRef) %>%
#   collect()
# multipleChanges <- open_dataset(paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinMultipleChanges/"))  %>%
#   dplyr::select(tumorLocationRef) %>%
#   collect()
# 
# nrow(noChanges) + nrow(singleChange) + nrow(multipleChanges)
# [1] 68416176
# 
# con <- dbConnect(duckdb())
# 
# duckdb_register(con, "males", males %>% distinct(sampleTumor))
# duckdb_register(con, "dft1",  dft1  %>% distinct(sampleTumor))
# 
# dbExecute(con, paste0("
#   COPY (
#     SELECT
#       * EXCLUDE (purity, copyNumberAdjusted),
#       purity                                          AS purityTumor,
#       copyNumberAdjusted                              AS nTumorTotal,
#       CASE
#         WHEN sampleTumor IN (SELECT sampleTumor FROM males)
#           AND CHROM IN ('X', 'Y')                    THEN 1
#         ELSE 2
#       END                                             AS nHostTotal
#     FROM read_parquet([
#         '", paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinNoChanges/**/*.parquet"), "',
#         '", paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinSingleChange/**/*.parquet"), "',
#         '", paste0(resDir, "/dft1Germline/", fprefix, "_dft1CnvJoinMultipleChanges/**/*.parquet"), "'
#       ], union_by_name = true) AS unioned
#   ) TO '", paste0(resDir, "/dft1Germline/", fprefix, "_dft1SomCnvCorrected/"), "'
#   (FORMAT parquet, PARTITION_BY sampleTumorChrom, OVERWRITE)
# "))
# 
# dbDisconnect(con)
# gc()
# 
# [1] 68,416,176 

# # [1] 159,407,245      
# # 
# # 
# # # 
# # # dft1SomCnvCorrected <-  %>%
# # #   dplyr::rename("purityTumor"="purity") %>%
# # #   mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"),
# # #                              1,
# # #                              2),
# # #        nTumorTotal = case_when(
# # #                     sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ 0,          # DFT1 female founder, no Y.
# # #                     TRUE ~ copyNumberAdjusted)) # Copy number adjusted takes into account starting copy number at sex chromosomes in tumors already, so no need to specify.
# # # 
# # # 
# # # chunk_size <- 5e4  # adjust to your RAM
# # # 
# # # dft1SomCnvCorrected <- dft1SomCnvCorrected %>%
# # #   mutate(rowId = row_number(),
# # #          chunk_id = (rowId - 1) %/% chunk_size)
# # # 
# # # gc()
# # # 
# # # write_dataset(dft1SomCnvCorrected,
# # #               paste0(resDir, "/dft1Germline/", fprefix, "_dft1SomCnvCorrected/"),
# # #               partitioning = "chunk_id")
# # # 
# # # rm(dft1SomCnvCorrected)
# 
# 
# germlineGenotyperDft1_small <- open_dataset(paste0(resDir, "/dft1Germline/", fprefix, "_dft1SomCnvCorrected")) %>%
#   mutate(purity = as.numeric(purityTumor)) %>%
#   filter(CHROM == "1" & tumor == "477T3" & POS == 587960251) %>%
#   collect()
# 
# germlineGenotyperDft1 <- open_dataset(paste0(resDir, 
#                                              "/dft1Germline/", 
#                                              fprefix, 
#                                              "_dft1SomCnvCorrected")) %>%
#   dplyr::select(-c(START, 
#                    END, 
#                    cnvType, 
#                    CHROM_1))
# 
# gc()
# 
# plan(multisession, workers = availableCores() - 1)
# 
# chunk_ids <- germlineGenotyperDft1 %>%
#   distinct(sampleTumorChrom) %>%
#   collect() %>%
#   pull()
# 
# # > length(chunk_ids)
# # [1] 504
# 
# for (cid in chunk_ids) {
# 
#   cat("Processing chunk", cid, "\n")
# 
#   chunk <- germlineGenotyperDft1 %>%
#     filter(sampleTumorChrom == cid) %>%
#     collect()
# 
#   if (nrow(chunk) == 0) next
# 
#   results <- future_pmap(
#     .l = list(
#       chunk$nHostTotal,
#       chunk$nTumorTotal,
#       chunk$hostAlt,
#       chunk$hostTotal,
#       chunk$tumorAlt,
#       chunk$tumorTotal,
#       chunk$purityTumor
#     ),
#     .f = function(hp, tp, ha, ht, ta, tt, p) {
# 
#       g <- genotyper(hp, tp, ha, ht, ta, tt, p)
# 
#       c(g_host = g$E$g_host,
#         g_tumour = g$E$g_tumour,
#         map_g_host = g$map[1, "g_host"],
#         map_g_tumour = g$map[1, "g_tumour"])
#     }
#   )
# 
#   res <- do.call(rbind, results)
# 
#   # chunk$nHostAltGtyr <- res[,1]
#   # chunk$nHostAltGtyrRounded  <- round(res[,1])
#   # chunk$nHostRefGtyrRounded <- chunk$nHostTotal - chunk$nHostAltGtyrRounded
#   #
#   # chunk$nTumorAltGtyr <- res[,2]
#   # chunk$nTumorAltGtyrRounded  <- round(res[,2])
#   # chunk$nTumorRefGtyrRounded <- chunk$nTumorTotal - chunk$nTumorAltGtyrRounded
# 
#   chunk$nHostAltGtyrMapped <- res[,3]
#   chunk$nHostRefGtyrMapped <- chunk$nHostTotal - chunk$nHostAltGtyrMapped
# 
#   chunk$nTumorAltGtyrMapped <- res[,4]
#   chunk$nTumorRefGtyrMapped <- chunk$nTumorTotal - chunk$nTumorAltGtyrMapped
# 
#   write_parquet(
#     chunk,
#     paste0(resDir, "/dft1Germline/", "germlineGenotyperDft1/", cid, ".parquet")
#   )
# 
#   rm(chunk, results, res)
#   gc()
# }
# 
# plan(sequential)
# 
# rm(germlineGenotyperDft1)
# gc()
# 
# dft1SomCorrectedCheck <- open_dataset(paste0(resDir, "/dft1Germline/", "germlineGenotyperDft1/"))%>%
#   filter(CHROM == "1" & tumor == "477T3" & POS == 587960251) %>%
#   collect()

germlineGenotyper <- open_dataset(paste0(resDir, "/dft1Germline/germlineGenotyperDft1/")) %>%
  mutate(locationRefAlt = paste0(locationRef, "_", ALT))
```

# 4 Join AlCo data
## 4.1 Metadata
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- Adjusted names manually using metadata! 
- Mapping AlCo results to VCF file names
- FOR now, don't filter where mapping = 1. There are sample names that I am certain match the name in Max's supplementary table. --> do this at the stage where we merge with AlCo results so that everything is genotyped
  - Filter on ! NA in DFT status column, as these we likely highly host contaminated or non-DFT tumors.
```{r}
metadata <- read.xlsx(paste0(resDir,
                          "20260529_08_deconvTable_metadata_purityIrodsMergedManualEdit.xlsx"), 
                      sheet = "withPolyRiboDups") %>%
              dplyr::rename("tumorId"="TUMOUR.INFORMATION_TUMOUR.ID",
                     "tumorIdRnaSeq"="rnaSeqManual") %>%
  filter(X9_LINEAGE %in% c("DFT1"))

# length(intersect(metadata$tumorId, dft1$sampleTumor)) # [1] 78

metadataPairs <- metadata %>%
  filter(mappingUpdate == 1) %>%
  dplyr::select(tumorId, 
                tumorIdRnaSeq) %>%
  distinct() %>%
  # For now, filter out duplicate. 
  filter(tumorId != "155T") %>%
  filter(tumorIdRnaSeq != "check")

# which(table(metadataPairs$tumorId)>1)
# which(table(metadataPairs$tumorIdRnaSeq)>1)
# 155Ta check 
#    17    55 
# length(intersect(metadataPairs$tumorId, dft1$sampleTumor)) # [1] 57
```

## 4.2 Read in AlCo results
- Here, tumorId = the name of the DNA file. Need to join this with the metadata table and switch out for the VCF tumor name (Max's supp data.)
- We have AlCo (i.e., RNAseq data )
```{r}
# germlineAlCo <- list.files(paste0("~/gitClones/08_deconvTable/083_Results/fromFarm/06_alCo_realignedEnsembl_germline/"),
#                           recursive = T)
# 
# # Read all files and combine
# germlineAlCoDf <- rbindlist(lapply(germlineAlCo, function(f) {
#   # Extract tumor ID between "germlineRefAltExport_" and "_realigned"
#   tumorIdRnaSeq <- sub(".*colnames_(.+).txt*", "\\1", basename(f))
# 
#   # Read file
#   dt <- fread(paste0("~/gitClones/08_deconvTable/083_Results/fromFarm/06_alCo_realignedEnsembl_germline/", f),
#               select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
# 
#   # Add tumor ID column
#   dt[, tumorIdRnaSeq := tumorIdRnaSeq]
# 
#   return(dt)
# }))
# 
# # length(unique(germlineAlCoDf$tumorIdRnaSeq)) # 115 RNAseq samples were genotyped.
# # 
# # check <- unique(germlineAlCoDf$tumorIdRnaSeq)
# # 
# # length(intersect(germlineAlCoDf$tumorIdRnaSeq, metadataPairs$tumorIdRnaSeq)) # 53 DFT2 samples for which we have a paired RNAseq-DNAseq file. --> 4 that don't?
# # 
# # setdiff(metadataPairs$tumorIdRnaSeq, intersect(germlineAlCoDf$tumorIdRnaSeq, metadataPairs$tumorIdRnaSeq))
# 
# # Left join metadata to map RNAseq ID to DNAseq ID. 
# write_dataset(
#   germlineAlCoDf,
#   path = paste0(resDir, "/germlineAlCoDf/"),
#   format = "parquet",
#   partitioning = "CHROM"
# )
# 
# germlineAlCoDf <- open_dataset(paste0(resDir, "/germlineAlCoDf/")) %>%
#   filter(tumorIdRnaSeq %in% metadataPairs$tumorIdRnaSeq) %>%
#   left_join(metadataPairs %>% dplyr::select(tumorIdRnaSeq, 
#                                             tumorId)) %>%
#   mutate(tumorLocationRefAlt = paste0(tumorId, "_", CHROM, ":", POS, "_", REF, "_", ALT),
#          sampleTumor = tumorId, 
#          sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   dplyr::select(-c(tumorId)) %>%
#   dplyr::select(TOTAL, 
#                 NREF, 
#                 NALT, 
#                 tumorIdRnaSeq, 
#                 tumorLocationRefAlt, 
#                 sampleTumorChrom)
```

# 5 Filter gtf
## 5.1 Overlapping positions
```{r}
# joinedChromPos <- germlinePaired %>% ## Crashes my session if I use the table with merged genotype info and AlCo results. 
#   dplyr::select(CHROM, 
#                 POS) %>%
#   distinct() %>%
#   mutate(POS = as.numeric(POS)) %>%
#   collect()
# 
# gc()
# 
# gtfFile <- paste0(dataDir, 
#                   "/Sarcophilus_harrisii.mSarHar1.11.115.gtf")
# 
# gtf <- import(gtfFile)
# 
# genes <- gtf[gtf$type == "gene"]
# 
# exons <- gtf[gtf$type == "exon"]
# 
# pcGenes <- genes[genes$gene_biotype == "protein_coding"]
# 
# variants <- GRanges(
#   seqnames = joinedChromPos$CHROM,
#   ranges = IRanges(start = joinedChromPos$POS, 
#                    end = joinedChromPos$POS)
# )
# 
# # Which variants fall in an exon
# exonHits <- findOverlaps(variants, exons)
# 
# # Map exon hits to gene IDs
# exonGeneIds <- exons$gene_id[subjectHits(exonHits)]
# exonGeneNames <- exons$gene_name[subjectHits(exonHits)]
# exonBiotypes <- exons$gene_biotype[subjectHits(exonHits)]
# 
# hitsDf <- data.frame(
#   query = queryHits(exonHits),
#   gene_id = exonGeneIds,
#   gene_name = exonGeneNames,
#   biotype = exonBiotypes
# )
# 
# collapseFun <- function(x) paste(unique(x), collapse = ";")
# 
# collapsed <- aggregate(. ~ query, 
#                        data = hitsDf, 
#                        FUN = collapseFun)
# 
# collapsed$CHROM <- joinedChromPos$CHROM[collapsed$query]
# collapsed$POS   <- joinedChromPos$POS[collapsed$query]
# 
# write.xlsx(collapsed %>% 
#            mutate(POS = as.character(POS)), 
#            paste0(resDir, "/dft1Germline/germlineDft1CollapsedGtf.xlsx"))

collapsed <- read.xlsx(paste0(resDir, "/dft1Germline/germlineDft1CollapsedGtf.xlsx")) %>%
  mutate(location = paste0(CHROM, ":", POS))
```

# 6 Full merge
- Per sampleTumorChrom, merge:
  - Genotypes
  - Majority status based on genotypes
  - AlCo
  - GTF positions
```{r}
chroms <- c(paste0(1:6), 
            "X", 
            "Y")
```

## 6.1 Assign majority genotype 
- Approach with Liz
  - For non-majority SNP's (CN >= 2), correct genotype to majority. Keep info on what was the originally assigned genotype. Also keep info on how many tumors had this flip. 
  - For the tied SNPs, exclude?
  - For the unclassified SNPs, keep as is? These could be genuine variants?
  
### 6.1.1 TEST
- Filter for SNPs positions where CN >= 2 (in any number of tumors): set as het, homAlt, or homRef. Using genotyper(), but will eventually use vaf windows defined in Kevin's supp methods. 
- Unclassified are SNPs falling in segments of CN 1 or 0. 
- Annotate each location with the majority (>= 50% status) flipping status
```{r}
# genotypeDft1Cn2All <- germlineGenotyper %>%
#   ######
#   filter(sampleTumor %in% dft1$sampleTumor) %>%
#   ######
#   mutate(
#     status = case_when(
#       ## HOM REF ##
#       nTumorTotal >= 2 & nTumorRefGtyrMapped == nTumorTotal & nTumorAltGtyrMapped == 0 ~ "homRef",
#       ## HOM ALT ##
#       nTumorTotal >= 2 & nTumorAltGtyrMapped == nTumorTotal & nTumorRefGtyrMapped == 0                   ~ "homAlt",
#       
#       nTumorTotal >= 2 & (nTumorRefGtyrMapped > 0 & nTumorAltGtyrMapped > 0)    ~ "het",
#       nTumorTotal < 2                                                                  ~ "unclassified",
#       TRUE                                                                             ~ "toCheck"
#     ),
#     statusBroad = case_when(
#       status %in% c("homRef",
#                     "homAlt") ~ "hom",
#       TRUE                              ~ status
#     )
#   ) %>%
#   dplyr::select(locationRefAlt,
#                 sampleTumor,
#                 status,
#                 statusBroad,
#                 CHROM,
#                 POS,
#                 ploidy,
#                 nTumorTotal,
#                 multiallelic) %>%
#   collect()
# 
# gc()
# 
# # hets <- germlineGenotyper %>%
# #   ######
# #   filter(sampleTumor %in% dft1$sampleTumor) %>%
# #   ######
# #   mutate(
# #     status = case_when(
# #       ## HOM REF ##
# #       nTumorTotal >= 2 & nTumorRefGtyrMapped == nTumorTotal & nTumorAltGtyrMapped == 0 ~ "homRef",
# #       ## HOM ALT ##
# #       nTumorTotal >= 2 & nTumorAltGtyrMapped == nTumorTotal & nTumorRefGtyrMapped == 0                   ~ "homAlt",
# #       
# #       nTumorTotal >= 2 & (nTumorRefGtyrMapped > 0 & nTumorAltGtyrMapped > 0)    ~ "het",
# #       nTumorTotal < 2                                                                  ~ "unclassified",
# #       TRUE                                                                             ~ "toCheck"
# #     ),
# #     statusBroad = case_when(
# #       status %in% c("homRef",
# #                     "homAlt") ~ "hom",
# #       TRUE                              ~ status
# #     )
# #   )  %>%
# #   filter(status == "het")  %>%
# #   collect() # [1] 16344391 rows
# # 
# # gc()
# # 
# # nrow(hets) # [1] 16344391
# # min(hets$nTumorTotal) # 2
# # min(hets$nTumorAltGtyrMapped) # 1
# # min(hets$tumorAlt) # 1
# # min(hets$nTumorRefGtyrMapped) # 1
# # 
# # 
# # check <- genotypeDft1Cn2All %>%
# #   filter(status == "toCheck") # [1] 0 rows
# # 
# # gc()
# # 
# # > table(genotypeDft1Cn2All$nTumorTotal,
# # +       genotypeDft1Cn2All$status)
# #     
# #           het   homAlt   homRef unclassified
# #   0         0        0        0         9474
# #   1         0        0        0      1484358
# #   2  11985315  6833191 31103677            0
# #   3    726166   518714  2411845            0
# #   4   3388279  1496718  7446698            0
# #   5    236306   118006   631279            0
# #   6      7479     2168    11241            0
# #   7        31      136      279            0
# #   11      815      500     3500            0
# #   19        0        1        0            0
# 
# 
# majorityDft1Status <- genotypeDft1Cn2All %>%
#   filter(statusBroad != "unclassified") %>% # Exclude variants where CN < 2 when determining majority status. 
#   # Calculate majority status per SNP
#   group_by(locationRefAlt) %>%
#   mutate(
#     nHet       = sum(status == "het"),
#     nHomRef       = sum(status == "homRef"),
#     nHomAlt       = sum(status == "homAlt"),
#     nTumoursTotal = n_distinct(sampleTumor),
#     majorityStatus = case_when(
#       nHet > nHomRef & nHet > nHomAlt ~ "het",
#       nHomRef > nHomAlt & nHomRef > nHet ~ 'homRef',
#       nHomAlt > nHomRef & nHomAlt > nHet ~ 'homAlt',
#       TRUE        ~ "tied"
#     )
#   ) %>%
#   ungroup() 
# 
# # table(majorityDft1Status$majorityStatus)
# # 
# # > table(majorityDft1Status$majorityStatus)
# # 
# #      het   homAlt   homRef     tied 
# # 15938973  9134362 41845930     3079 
# 
# 
# 
# write_dataset(majorityDft1Status,
#               paste0(resDir, "/dft1Germline/majorityDft1Status/"),
#               partitioning = "sampleTumor")
# 
# rm(genotypeDft1Cn2All, majorityDft1Status)
# 
# gc()
# # One row per sampleTumorLocationRefAlt
# majorityDft1Status <- open_dataset(paste0(resDir, 
#                                           "/dft1Germline/majorityDft1Status")) %>%
#   mutate(tumorLocationRefAlt = paste0(sampleTumor, "_", locationRefAlt)) %>%
#   dplyr::select(locationRefAlt,
#                 CHROM,
#                 sampleTumor,
#                 status,
#                 statusBroad,
#                 nHet,
#                 nHomRef,
#                 nHomAlt,
#                 nTumoursTotal,
#                 majorityStatus, 
#                 tumorLocationRefAlt) %>%
#   distinct() %>%
#   dplyr::rename("nTumorsCn2" = "nTumoursTotal") %>%
#   mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   collect()
# 
# # Stage 1: test join, for single sampleTumorChrom only
# test <- germlineGenotyper %>%
#   filter(sampleTumorChrom == "18T_X") %>%
#   left_join(majorityDft1Status %>% dplyr::select(-c(CHROM, 
#                                                     sampleTumor, 
#                                                     locationRefAlt, 
#                                                     sampleTumorChrom)),
#             by = c("tumorLocationRefAlt")) %>%
#     mutate(
#     majorityCheck = case_when(
#       status == "unclassified" ~ "unclassified",
#       majorityStatus == "tied" ~ "tied",
#       majorityStatus == status ~ "match",
#       TRUE ~ "flip"
#     ),
#     transition = paste(status, majorityStatus, sep = "→"),
#     # -------------------------
#     # ALT correction logic
#     # -------------------------
#     nTumorAltCorrected = case_when(
#       # homAlt flipped to: ALT = total, REF = 0
#       majorityCheck == "flip" & majorityStatus == "homAlt" ~ nTumorTotal,
#       # homRef flipped to: ALT = 0, REF = total
#       majorityCheck == "flip" & majorityStatus == "homRef" ~ 0L,
#       # het → homAlt or homRef: split evenly, ALT = ceiling(total / 2)
#       majorityCheck == "flip" &
#         status == "het" &
#         majorityStatus %in% c("homAlt", "homRef") ~ as.integer(ceiling(nTumorTotal / 2)),
#       # default: keep original
#       TRUE ~ nTumorAltGtyrMapped
#     ),
#     # -------------------------
#     # REF correction logic
#     # -------------------------
#     nTumorRefCorrected = case_when(
#       # homRef flipped to: REF = total, ALT = 0
#       majorityCheck == "flip" & majorityStatus == "homRef" ~ nTumorTotal,
#       # homAlt flipped to: REF = 0, ALT = total
#       majorityCheck == "flip" & majorityStatus == "homAlt" ~ 0L,
#       # het → homAlt or homRef: split evenly, REF = floor(total / 2)
#       majorityCheck == "flip" &
#         status == "het" &
#         majorityStatus %in% c("homAlt", 
#                               "homRef") ~ as.integer(floor(nTumorTotal / 2)),
#       # default
#       TRUE ~ nTumorRefGtyrMapped
#     )
#   ) %>%
#   collect()
# 
# write_parquet(test, 
#               paste0(resDir, "dft1Germline/18T_3_germlineGenotyperMajorityStatus.parquet"))
```


### 6.1.2 FINAL
- Filter germlineGenotyper with DNAseq samples of interest
```{r}
# germlineGenotyper <- open_dataset(paste0(resDir, "/dft1Germline/germlineGenotyperDft1/")) %>%
#   mutate(locationRefAlt = paste0(locationRef, "_", ALT)) %>%
#   filter(sampleTumor %in% metadataPairs$tumorId)
# 
# sampleTumorChroms <- open_dataset(paste0(resDir, 
#                                           "/dft1Germline/majorityDft1Status")) %>%
#   filter(sampleTumor %in% metadataPairs$tumorId) %>%
#   mutate(tumorLocationRefAlt = paste0(sampleTumor, "_", locationRefAlt)) %>%
#   dplyr::select(locationRefAlt,
#                 CHROM,
#                 sampleTumor,
#                 status,
#                 statusBroad,
#                 nHet,
#                 nHomRef,
#                 nHomAlt,
#                 nTumoursTotal,
#                 majorityStatus, 
#                 tumorLocationRefAlt) %>%
#   distinct() %>%
#   dplyr::rename("nTumorsCn2" = "nTumoursTotal") %>%
#   mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   dplyr::select(sampleTumorChrom) %>%
#   distinct() %>%
#   collect() %>%
#   pull(sampleTumorChrom) 
# 
# majorityDft1Status <- open_dataset(paste0(resDir,
#                                             "/dft1Germline/majorityDft1Status")) %>%
#   filter(sampleTumor %in% metadataPairs$tumorId) %>%
#     mutate(tumorLocationRefAlt = paste0(sampleTumor, "_", locationRefAlt)) %>%
#     dplyr::select(locationRefAlt,
#                   CHROM,
#                   sampleTumor,
#                   status,
#                   statusBroad,
#                   nHet,
#                   nHomRef,
#                   nHomAlt,
#                   nTumoursTotal,
#                   majorityStatus,
#                   tumorLocationRefAlt) %>%
#     distinct() %>%
#     dplyr::rename("nTumorsCn2" = "nTumoursTotal") %>%
#     mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM))
# 
# gc()
# 
# for(schr in sampleTumorChroms){
#   
#   print(paste0("Running ", schr))
#   
#   tempStat <- majorityDft1Status %>%
#     filter(sampleTumorChrom == schr) %>% 
#     dplyr::select(-c(CHROM,
#                                  sampleTumor,
#                                  locationRefAlt,
#                                  sampleTumorChrom))
#     
#   
#   tempAlCo <- germlineAlCoDf %>%
#     filter(sampleTumorChrom == schr) %>%
#     dplyr::select(-c(sampleTumorChrom))
#   
#   tempMerge <- germlineGenotyper %>%
#     ### Start with baseline genotyping results ###
#     filter(sampleTumorChrom == schr) %>%
#     ### Correct with majority status ###
#     left_join(tempStat,
#               by = c("tumorLocationRefAlt")) %>%
#       mutate(
#       majorityCheck = case_when(
#         status == "unclassified" ~ "unclassified",
#         majorityStatus == "tied" ~ "tied",
#         majorityStatus == status ~ "match",
#         TRUE ~ "flip"
#       ),
#       transition = paste(status, majorityStatus, sep = "→"),
#       # -------------------------
#       # ALT correction logic
#       # -------------------------
#       nTumorAltCorrected = case_when(
#         # homAlt flipped to: ALT = total, REF = 0
#         majorityCheck == "flip" & majorityStatus == "homAlt" ~ nTumorTotal,
#         # homRef flipped to: ALT = 0, REF = total
#         majorityCheck == "flip" & majorityStatus == "homRef" ~ 0L,
#         # het → homAlt or homRef: split evenly, ALT = ceiling(total / 2)
#         majorityCheck == "flip" &
#           status == "het" &
#           majorityStatus %in% c("homAlt", "homRef") ~ as.integer(ceiling(nTumorTotal / 2)),
#         # default: keep original
#         TRUE ~ nTumorAltGtyrMapped
#       ),
#       # -------------------------
#       # REF correction logic
#       # -------------------------
#       nTumorRefCorrected = case_when(
#         # homRef flipped to: REF = total, ALT = 0
#         majorityCheck == "flip" & majorityStatus == "homRef" ~ nTumorTotal,
#         # homAlt flipped to: REF = 0, ALT = total
#         majorityCheck == "flip" & majorityStatus == "homAlt" ~ 0L,
#         # het → homAlt or homRef: split evenly, REF = floor(total / 2)
#         majorityCheck == "flip" &
#           status == "het" &
#           majorityStatus %in% c("homAlt",
#                                 "homRef") ~ as.integer(floor(nTumorTotal / 2)),
#         # default
#         TRUE ~ nTumorRefGtyrMapped
#       )
#     ) %>%
#     ### Merge RNAseq counts ###
#     left_join(tempAlCo, 
#               by = c("tumorLocationRefAlt")) %>%
#     mutate(location = paste0(CHROM, ":", POS)) %>%
#     ### Merge GTF info. Any rows that do not intersect an exon are NA ###
#     left_join(collapsed %>%
#                 dplyr::select(-c(CHROM, 
#                                  POS)),
#               by = "location")
#   ### 
#   write_parquet(
#     tempMerge,
#     paste0(resDir, 
#            "/dft1Germline/germlineAnnoStatusDft1/", 
#            schr, 
#            ".parquet")
#   )
#   ###
# }
```
### 6.1.3 Check all variants are accounted for in final table. 
- Only 60 positions lost between allPositions and mergedPositions (for matched RNA/DNA seq data). These were multiplet positions where maxCN != minCN. 
```{r}
# allPositions <- open_dataset(paste0(resDir,
#                                   "08_deconvTable_germline_germlinePaired_computedVafs_multiAnno/")) %>%
#   filter(sampleTumor %in% metadataPairs$tumorId) %>%
#   distinct(tumorLocationRefAlt) %>%
#   dplyr::pull(tumorLocationRefAlt)
# 
# 
# mergedPositions <- open_dataset( paste0(resDir, 
#            "/dft1Germline/germlineAnnoStatusDft1/")) %>%
#   distinct(tumorLocationRefAlt) %>%
#   pull(tumorLocationRefAlt)  
# 
# length(allPositions)==length(mergedPositions)
# # 60 positions differ, because these are those where minCN != maxCN (multiplets.)
```

## 6.2 Informative positions
```{r}
mergedInformative <- open_dataset( paste0(resDir,
           "/dft1Germline/germlineAnnoStatusDft1/")) %>%
  filter(is.na(query)==F) %>%
  
```







