---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(arrow)
library(furrr)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_germline"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
source("~/gitClones/08_deconvTable/082_Scripts/genotype_germline_variants.R")
```

## 0.3 Custom functions
### 0.3.1 Compute tAlt
```{r}
computeTAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         hTotal, 
                         hAlt) {
  
    print("Solving for tAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    tAlt = ((obsVaf*denominator)-(1-purityTumor)*hAlt)/purityTumor
  return(tAlt)
}

assign_nAlt_from_vaf <- function(vaf, total_cn) {
  
  if (is.na(total_cn) || is.na(vaf)) return(NA_integer_)
  
  if (total_cn == 0) return(0)
  
  if (total_cn == 1) {
    return(ifelse(vaf < 0.5, 0, 1))
  }
  
  if (total_cn == 2) {
    if (vaf < 0.25) return(0)
    else if (vaf < 0.75) return(1)
    else return(2)
  }
  
  if (total_cn == 3) {
    if (vaf < 0.15) return(0)
    else if (vaf < 0.5) return(1)
    else if (vaf < 0.82) return(2)
    else return(3)
  }
  
  if (total_cn == 4) {
    if (vaf < 0.12) return(0)
    else if (vaf < 0.39) return(1)
    else if (vaf < 0.64) return(2)
    else if (vaf < 0.86) return(3)
    else return(4)
  }
  
  return(NA_integer_)  # fallback
}
```
### 0.3.2 Compute hAlt
```{r}
computeHAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         tAlt,
                         hTotal) {
  
  # Cases where hAlt = 0, solve for tAlt
 
    print("Solving for hAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    hAlt = (obsVaf*denominator)/(1-purityTumor)
  return(hAlt)
}

computeHAlt(obsVaf = 0.5, 
            purityTumor = 0.8, 
            tTotal = 4, 
            hTotal = 2, 
            tAlt=0)



```
### 0.3.3 Geometric mean
- Kevin's scripts
  - To calculate size factors: 
  - Compute the geom.mean of totals for each variant (optional: add pseudocount)
  - Divide each variant's read total by the geometric mean for each variant
  - Then take the median of these ratios for each sample
  - (Optional) restrict to variants with presence in all tumors
```{r}
geomean <- function(x) {
  if (all(x == 0)) {
    NA_real_
  } else {
    exp(mean(log(x[x > 0])))
  }
}
```

### 0.3.4 Summarize germline variants
```{r}
library(dplyr)
library(tidyr)

library(data.table)

summarise_variants_dt <- function(df) {
  dt <- as.data.table(df)
  
  # ensure uniqueness per tumor per location
  dt <- unique(dt[, .(locationAlt, sampleTumor, statusUpdate)])
  
  # count per location + status
  counts <- dt[, .N, by = .(locationAlt, statusUpdate)]
  
  # total tumors per location
  totals <- dt[, .(total_tumors = uniqueN(sampleTumor)), by = locationAlt]
  
  # merge counts with totals
  counts <- counts[totals, on = "locationAlt"]
  
  # compute percentage
  counts[, pct := N / total_tumors]
  
  # reshape wide (one column per statusUpdate)
  wide <- dcast(
    counts,
    locationAlt + total_tumors ~ statusUpdate,
    value.var = "pct",
    fill = 0
  )
  
  # compute majority
  status_cols <- setdiff(names(wide), c("locationAlt", "total_tumors"))
  
  wide[, majority := {
    vals <- unlist(.SD)
    max_val <- max(vals)
    winners <- names(vals)[vals == max_val]
    
    if (length(winners) > 1) "flag" else winners
  }, by = locationAlt, .SDcols = status_cols]
  
  return(wide)
}
```


# 1 Data
## 1.1 VCF files
- Read in germline vcf file
- Pivot longer so that each variant and each sample is a single row with the ALT/TOTAL ratio.
```{r}
# germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"),
#                       col_types = cols(.default = "c"))
# 
# sampleInfo <- tibble(sample = colnames(germline)[-(1:4)]) %>%
#   mutate(
#     sampleType = str_extract(sample, "T|H"),
#     tcgId = str_extract(sample, "^[^TH]+")
#   )
# 
# 
# germlineLong <- germline %>%
#   filter(nchar(REF) == 1, 
#          nchar(ALT) == 1) %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "sample",
#     values_to = "ratio"
#   ) %>%
#   left_join(sampleInfo, by = "sample") %>%
#   transmute(
#     CHROM,
#     POS,
#     REF,
#     ALT,
#     tcgId,
#     sampleType,
#     ratio,
#     sample
#   )
# 
# saveRDS(germlineLong, 
#         paste0(resDir, 
#                fprefix, 
#                "_germlineLong.rds"))
```

## 1.2 Pair tumor/host samples
- Using shared ID's pivot wider so that the ratio for matched tumor/host samples is side by side.
- Filter on variants where alt allele is present in only the host or the tumor sample within a pair! --> so that we have info. on tAlt or hAlt at that position.
  - 75/121 tumors in the VCF file have an informative variant, where one of the alt alleles is 0.
  - Filter for tumor id's for which we have matched RNAseq data. 
```{r}
# tumor <- germlineLong %>%
#   filter(sampleType == "T") %>%
#   dplyr::rename(
#     vafTumor = ratio,
#     sampleTumor = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# normal <- germlineLong %>%
#   filter(sampleType == "H") %>%
#   dplyr::rename(
#     vafHost = ratio,
#     sampleHost = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# paired <- tumor %>%
#   inner_join(normal,
#              by = c("CHROM", 
#                     "POS", "REF", "ALT", "tcgId"))
# 
# saveRDS(paired, 
#         paste0(resDir, 
#                fprefix, 
#                "_germlinePaired.rds"))
# 
# germlinePairedFiltered <- paired %>%
#   filter(
#     str_starts(vafHost, "0/") != str_starts(vafTumor, "0/")
# germlinePairedFiltered$tumorLocationAlt <- paste0(germlinePairedFiltered$sampleTumor, 
#                                                    "_", germlinePairedFiltered$CHROM, 
#                                                   ":", germlinePairedFiltered$POS, 
#                                                   "_", germlinePairedFiltered$ALT)
#   )
# 
# saveRDS(germlinePairedFiltered,
#         paste0(resDir,
#                fprefix,
#                "_germlinePairedFiltered.rds"))
# rm(germlineLong, 
#    germline, 
#    paired, 
#    normal, 
#    tumor)
# 
# gc()
# germlinePairedFiltered <- readRDS(paste0(resDir,
#                fprefix,
#                "_germlinePairedFiltered.rds"))
# germlinePairedTest <- germlinePairedFiltered[1:1000,] %>%
#   mutate(
#     tumorAlt = as.integer(sub("/.*", "", vafTumor)),
#     tumorTotal = as.integer(sub(".*/", "", vafTumor)),
#     hostAlt = as.integer(sub("/.*", "", vafHost)), 
#     hostTotal = as.integer(sub(".*/", "", vafHost))
#   ) %>%
#   mutate(vafTumor = tumorAlt/tumorTotal, 
#          vafHost = hostAlt/hostTotal) %>%
#   mutate(hAlt = ifelse(tumorAlt == 0, 
#                        vafHost*2, 
#                        ifelse(hostAlt == 0, 
#                               0, 
#                               NA)), 
#          nHostAlt = round(hAlt), 
#          nHostRef = 2-nHostAlt)
# 
# %>%
#   mutate(tumorLocationAlt = paste(sampleTumor, "_", CHROM, ":", POS, "_", ALT)
# germlinePaired <- readRDS(paste0(resDir,
#                fprefix,
#                "_germlinePaired.rds")) %>%
#   mutate(
#     tumorAlt = as.integer(sub("/.*", "", vafTumor)),
#     tumorTotal = as.integer(sub(".*/", "", vafTumor)),
#     hostAlt = as.integer(sub("/.*", "", vafHost)),
#     hostTotal = as.integer(sub(".*/", "", vafHost))
#   ) %>%
#   mutate(obsVafTumor = tumorAlt/tumorTotal,
#          obsVafHost = hostAlt/hostTotal)%>%
#  mutate(tumorLocationAlt = paste0(sampleTumor, "_", CHROM, ":", POS, "_", ALT))
# 
# saveRDS(germlinePaired,
#         paste0(resDir,
#                fprefix,
#                "_germlinePaired.rds") )
# set.seed(123)
# test <- germlinePaired[sample(nrow(germlinePaired), 5000),]
# saveRDS(test, 
#         paste0(resDir, 
#                fprefix, 
#                "_germlinePairedTest.rds"))
# 

# test <- readRDS(paste0(resDir,
#                fprefix,
#                "_germlinePairedTest.rds")) %>%
#   mutate(
#     tumorAlt = as.integer(sub("/.*", "", vafTumor)),
#     tumorTotal = as.integer(sub(".*/", "", vafTumor)),
#     hostAlt = as.integer(sub("/.*", "", vafHost)),
#     hostTotal = as.integer(sub(".*/", "", vafHost))
#   ) %>%
#   mutate(obsVafTumor = tumorAlt/tumorTotal,
#          obsVafHost = hostAlt/hostTotal)%>%
#  mutate(tumorLocationAlt = paste0(sampleTumor, "_", CHROM, ":", POS, "_", ALT))

# library(arrow)
# 
# write_parquet(germlinePaired, paste0(resDir,
#                fprefix,
#                "_germlinePaired.parquet"))

# germlinePaired <- open_dataset(paste0(resDir, 
#                                       fprefix,
#                                       "_germlinePaired.parquet")) 
```

## 1.3 Subset germline vcfs by host, tumor, or shared vaf's.
- Germline (host+tumour) variants are defined as those with VAF >0.25 in any host.
- That could mean it has 0 VAF in a given tumor sample, but this was not observed in all samples and is therefore not somatic. 
- How to account for this when calling "true" informative variants?
```{r}
# germlinePairedFiltered <- germlinePaired 
```

## 1.2 CNVs
- Any germline variants falling within a CNV event were filtered out, so no need to adjust ploidy! 
```{r}
# germlineCnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx", 
#                      sheet = "Table S6C", 
#                      startRow = 3)
# 
# germlineCnvLong <- germlineCnv %>%
#   pivot_longer(-c(1:10),
#               names_to = "tumor", 
#               values_to = "presence") %>%
#   mutate(presence = as.numeric(presence)) %>%
#   mutate(
#     START = as.numeric(START),
#     END = as.numeric(END),
#     CHROM = as.character(CHROM),
#     tumor = as.character(tumor)
#   )
# 
# rm(germlineCnv)
# gc()
```


## 1.3 Purities and ploidy
```{r}
germlinePurities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(germlinePurities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(germlinePurities),
  paste0(colnames(germlinePurities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(germlinePurities) <- newNames
germlinePurities <- germlinePurities[-1, ]

germlinePuritiesPloidy <- germlinePurities %>%
  mutate(ploidy = ifelse(X15_PLOIDY == "diploid",
                 2, 
                 ifelse(X15_PLOIDY == "tetraploid", 
                        4, 
                        "toCheck") 
                 )) %>%
  filter(ploidy != "toCheck") %>%
  dplyr::rename("sampleTumor"="TUMOUR.INFORMATION_TUMOUR ID", 
                     "purity"="X14_PURITY (VAF-based)", 
                "dftType"="X9_LINEAGE") %>%
              mutate(purity = as.numeric(purity), 
                     ploidy = as.numeric(ploidy), 
                     sex=ifelse(HOST.SEX == "♂", 
                                "male", 
                                ifelse(HOST.SEX == "♀", 
                                       "female", 
                                       "toCheck"))) %>%
              dplyr::select(sampleTumor, 
                            ploidy, 
                            purity, 
                            sex, 
                            dftType
                            ) 
  
table(germlinePuritiesPloidy$sex, germlinePuritiesPloidy$dftType)

males <- germlinePuritiesPloidy %>%
  filter(sex == "male") 

dft2 <- germlinePuritiesPloidy %>%
  filter(dftType == "DFT2") 
```

## 1.4 Allele counter data
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- Adjust names using metadata! 

### 1.4.1 Metadata
- Mapping AlCo results to VCF file names
- FOR now, filter where mapping = 1. There are sample names that I am certain match the name in Max's supplementary table. 
  - 46/75 tumors in Max's VCF file WITH an informative germline variant have a matched RNAseq file!
- Go back and check once workflow is done
```{r}
metadata <- read.xlsx(paste0(resDir,
                          "08_deconvTable_metadata_purityIrodsMergedManualEdit.xlsx")) %>%
  filter(mapping==1)  %>%
              dplyr::rename("tumorId"="TUMOUR.INFORMATION_TUMOUR.ID",
                     "tumorIdRnaSeq"="rnaSeqManual")

# length(intersect(germlinePaired$sampleTumor, metadata$tumorId))
## 46  

dft2 <- metadata %>% 
  filter(X9_LINEAGE == "DFT2") %>%
  pull(tumorId)
```

### 1.4.2 Read in AlCo results
- Here, tumorId = the name of the RNAseq file. Need to join this with the metadata table and switch out for the VCF tumor name (Max's supp data.)
- We have AlCo data from 115 tumors, calling this the tumorRnaSeqId. 
  - This can be left joined with the paired tumorId from the metadata.

```{r}
# germlineAlCo <- list.files(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport"),
#                           recursive = T)
# 
# # Read all files and combine
# germlineAlCoDf <- rbindlist(lapply(germlineAlCo, function(f) {
#   # Extract tumor ID between "germlineRefAltExport_" and "_realigned"
#   tumorIdRnaSeq <- sub(".*germlineRefAltExport_(.+)_realigned.*", "\\1", basename(f))
# 
#   # Read file
#   dt <- fread(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport/", f),
#               select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
# 
#   # Add tumor ID column
#   dt[, tumorIdRnaSeq := tumorIdRnaSeq]
# 
#   return(dt)
# }))
# 
# length(unique(germlineAlCoDf$tumorIdRnaSeq)) # 115 RNAseq samples were genotyped.
# 
# length(intersect(germlineAlCoDf$tumorIdRnaSeq, metadata$tumorIdRnaSeq)) # 53 samples for which we have a paired RNAseq-DNAseq file.
# 
chunk_size <- 5e4  # adjust to your RAM
# 
# ## Left join metadata
# germlineAlCoDf <- germlineAlCoDf %>%
#   mutate(rowId = row_number(),
#          chunk_id = (rowId - 1) %/% chunk_size) 
# 
# write_dataset(
#   germlineAlCoDf,
#   path = paste0(resDir, "germlineAlCoDf/"),
#   format = "parquet",
#   partitioning = "chunk_id"
# )

# Perform filtering and mutate on arrow, less memory intensive.
# Rows left after filter informative samples = 58966002
# germlineAlCoDfFiltered <- open_dataset(paste0(resDir, "germlineAlCoDf/")) %>%
#   filter(tumorIdRnaSeq %in% metadata$tumorIdRnaSeq) %>%
#   left_join(metadata) %>%
#   # Tumor ID = sample ID from germline vcf.
#   mutate(tumorLocationAlt = paste0(tumorId, "_", CHROM, ":", POS, "_", ALT), 
#          sampleTumor = tumorId) %>%
#   dplyr::select(-c(tumorId))
# 
# write_dataset(
#   germlineAlCoDfFiltered,
#   path = paste0(resDir, "germlineAlCoDfFiltered/"),
#   format = "parquet", 
#   partitioning = "sampleTumor"
#   
# )
```

### 1.4.3 Adjust tumorId of RNAseq data to match Max's metadata
- Filter to keep only samples for which we have RNAseq data.
  - Filter tumorId in AlCo for those in metadata$tumorIdRnaSeq
- Left join with metadata table, to get the matched tumorId
- Create new column, tumorId_CHROM_POS_ALT
- Filter for informative AlCo positions in germlinePaired, otherwise there are too many values in the table (slow to work with).
  - 53 / 115 tumor samples have matched RNA (need to check this.)
  - 46/75 samples have an informative variant in the germline vcf file AND a matched RNAseq position. 
  - /46 have at least one variant with TOTAL AlCo > 0. 
```{r}
# setDT(germlineAlCoDf)
# 
# # fast filtering
# germlineAlCoCombined <- germlineAlCoDf[tumorIdRnaSeq %in% metadata$tumorIdRnaSeq] %>%
#   left_join(metadata %>%
#               dplyr::select(tumorId, tumorIdRnaSeq))
# 
# germlineAlCoCombined[, c("tumorLocation", "tumorLocationAlt") := .(
#   paste0(tumorId, "_", CHROM, ":", POS),
#   paste0(tumorId, "_", CHROM, ":", POS, "_", ALT)
# )]
# 
# saveRDS(germlineAlCoCombined,
#         paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombined.rds"))
# 
# germlineAlCoCombined <- readRDS(paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombined.rds"))
# 
# 
# germlineAlCoCombinedFiltered <- germlineAlCoCombined[tumorLocationAlt %in% germlinePaired$tumorLocationAlt]
# 
# 
# saveRDS(germlineAlCoCombinedFiltered,
#         paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombinedFiltered.rds"))
# 
# rm(germlineAlCoCombined,
#    germlineAlCoDf,
#    germlinePairedFiltered)
# 
# gc()
# 
# germlineAlCoCombinedFiltered <- readRDS(paste0(resDir, 
#                fprefix, 
#                "_germlineAlCoCombinedFiltered.rds")) 
```

# 3 Construct final table
## 3.1 Compute genotypes
- Adjust total CN for X and Y chromosomes in male samples
- In a pure sample, vafH = 0 + (1)*Halt/(0 + 1*Htotal). Therefore, Halt = vafH*Htotal. 
- How do deal with ploidy for X and Y in male samples in 
### 3.1.1 Test dataset
- 5000 random rows
- Compare basic implementation to Kevin's formula, which should only affect flagged cases.
```{r}
# testGenotyped <- test %>%
#   left_join(germlinePuritiesPloidy) %>%
#   filter(sampleTumor %in% metadata$tumorId) %>%
#   filter(!sampleTumor %in% c("212T",
#                             "818T1")) %>%
#   mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"), 
#                              1, 
#                              2), 
#          nTumorTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"), 
#                              ploidy/2,  # tetraploid tumors have XXYY.
#                              ploidy))  %>%
#   # Using a pure host sample (i.e., p = 0)
#   mutate(hAlt = obsVafHost * nHostTotal,
#          nHostAlt = round(hAlt),
#          nHostAltFlagged = pmin(nHostAlt, nHostTotal),
#          flag = nHostAlt > nHostTotal,
#          nHostRef = nHostTotal - nHostAltFlagged) %>%
#   
#   # Estimate tAlt
#   mutate(tAlt = computeTAlt(obsVaf = obsVafTumor,
#                             purityTumor = purity,
#                             tTotal = nTumorTotal,
#                             hTotal = nHostTotal,
#                             hAlt = nHostAltFlagged)) %>%
#   ########
#   # Round for nTumorAlt
#   mutate(nTumorAlt = round(tAlt),
#          flag = nTumorAlt > nTumorTotal, 
#          nTumorAltFlagged = pmin(nTumorAlt, nTumorTotal)) %>%
#   # No conditions where using estimate of NAlt changes nTumorAlt.
#   # filter(nTumorAltFromHAlt!=nTumorAltFromNHAlt) %>%
#   mutate(nTumorRef = nTumorTotal-nTumorAltFlagged,
#          flag = nTumorAlt > nTumorTotal) 
#   
# 
# ### Now test with Kevin's function. Run genotyper in parallel. 
# plan(multisession, workers = availableCores() - 1)
# 
# # Run genotyper in parallel — both outputs extracted in a single pass per row.
# results <- furrr::future_pmap(
#   .l = list(
#     hostPloidy = testGenotyped$nHostTotal,
#     tumorPloidy = testGenotyped$nTumorTotal,
#     hostObsAlt = testGenotyped$hostAlt,
#     hostObsTotal = testGenotyped$hostTotal,
#     tumorObsAlt = testGenotyped$tumorAlt,
#     tumorObsTotal = testGenotyped$tumorTotal,
#     purity  = testGenotyped$purity
#   ),
#   .f = function(hostPloidy,
#                 tumorPloidy,
#                 hostObsAlt,
#                 hostObsTotal,
#                 tumorObsAlt,
#                 tumorObsTotal,
#                 purity) {
#     # g <- genotyper(hp, tp, ha, ht, ta, tt, p)$E
#     g <- genotyper(host_ploidy = hostPloidy,
#                    tumour_ploidy = tumorPloidy,
#                    host_observed_alt_reads = hostObsAlt,
#                    host_observed_total_reads = hostObsTotal,
#                    tumour_observed_alt_reads = tumorObsAlt ,
#                    tumour_observed_total_reads = tumorObsTotal,
#                    tumour_purity = purity)
#     list(g_host = g$E$g_host, 
#          g_tumour = g$E$g_tumour, 
#          g=g)
#   },
#   .options = furrr_options(seed = NULL)  # no RNG used, avoids overhead
# )
# 
# testGenotyped$nHostAltGenotyper  <- vapply(results, `[[`, numeric(1), "g_host")
# testGenotyped$nTumorAltGenotyper <- vapply(results, `[[`, numeric(1), "g_tumour")
# 
# # Reset to sequential when done
# plan(sequential)
# 
# saveRDS(testGenotyped, 
#         paste0(resDir, fprefix, "_testGenotyped.rds"))
```

### 3.1.2 Full dataset
- If host sample is a male, total CN at sex chromosomes == 1
- If tumor sample is DFT2, total CN at sex chromosomes == 1 or 2 depending on original ploidy.
```{r}
germlinePaired <- read_parquet(paste0(resDir,
                                      "08_deconvTable_germline_germlinePaired.parquet"))


germlinePairedGenotyped <- arrow_table(germlinePaired) %>%
  left_join(germlinePuritiesPloidy) %>%
  filter(sampleTumor %in% metadata$tumorId) %>%
  filter(!sampleTumor %in% c("212T",
                            "818T1")) %>%
  mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"),
                             1,
                             2),
         nTumorTotal = ifelse(sampleTumor %in% dft2$sampleTumor & CHROM %in% c("X", "Y"),
                             ploidy/2,  # tetraploid tumors have XXYY.
                             ploidy)) %>%
  collect()

germlinePairedGenotyped$rowId <- 1:nrow(germlinePairedGenotyped)

chunk_size <- 5e4  # adjust to your RAM

germlinePairedGenotyped <- germlinePairedGenotyped %>%
  mutate(chunk_id = (rowId - 1) %/% chunk_size)

gc()

write_dataset(
  germlinePairedGenotyped,
  path = paste0(resDir, "germlinePartitioned/"),
  format = "parquet",
  partitioning = "chunk_id"
)
```

### 3.1.3 Run genotyper in chunks
```{r}
ds <- open_dataset(paste0(resDir, "germlinePartitioned/"))

plan(multisession, workers = availableCores() - 1)

chunk_ids <- ds %>%
  distinct(chunk_id) %>%
  collect() %>%
  pull()

for (cid in chunk_ids) {

  cat("Processing chunk", cid, "\n")

  chunk <- ds %>%
    filter(chunk_id == cid)  %>%
    collect()

  if (nrow(chunk) == 0) next

  results <- future_pmap(
    .l = list(
      chunk$nHostTotal,
      chunk$nTumorTotal,
      chunk$hostAlt,
      chunk$hostTotal,
      chunk$tumorAlt,
      chunk$tumorTotal,
      chunk$purity
    ),
    .f = function(hp, tp, ha, ht, ta, tt, p) {

      g <- genotyper(hp, tp, ha, ht, ta, tt, p)

      c(g_host = g$E$g_host,
        g_tumour = g$E$g_tumour, 
        map_g_host = g$map[1, "g_host"],
        map_g_tumour = g$map[1, "g_tumour"])
    }
  )

  res <- do.call(rbind, results)

  chunk$nHostAltGtyr <- res[,1]
  chunk$nHostAltGtyrRounded  <- round(res[,1])
  chunk$nHostRefGtyrRounded <- chunk$nHostTotal - chunk$nHostAltGtyrRounded

  chunk$nTumorAltGtyr <- res[,2]
  chunk$nTumorAltGtyrRounded  <- round(res[,2])
  chunk$nTumorRefGtyrRounded <- chunk$nTumorTotal - chunk$nTumorAltGtyrRounded
  
  chunk$nHostAltGtyrMapped <- res[,3]
  chunk$nHostRefGtyrMapped <- chunk$nHostTotal - chunk$nHostAltGtyrMapped

  chunk$nTumorAltGtyrMapped <- res[,4]
  chunk$nTumorRefGtyrMapped <- chunk$nTumorTotal - chunk$nTumorAltGtyrMapped

  write_parquet(
    chunk,
    paste0(resDir, "germlineGenotyper/genotyped_chunk_", cid, ".parquet")
  )

  rm(chunk, results, res)
  gc()
}

plan(sequential)
```
### 3.1.4 Plot genotyper() results
- Annotate germline status as hom alt, hom ref, or het
- Check what the majority status is at each position (e.g., are >50% of tumors hom alt?)
  - Add % column
  - Add majority assignment
- Create 10KB bins along entire genome
- Per tumor, plo
```{r}
# Diploid tumors
germlinePairedGenotypedStatus <- open_dataset(paste0(resDir, "germlineGenotyper/")) %>%
  mutate(status = ifelse(nTumorRefGtyrMapped == ploidy & nTumorAltGtyrMapped == 0, 
                "homRef", 
                ifelse(nTumorAltGtyrMapped == ploidy & nTumorRefGtyrMapped == 0, 
                       "homAlt", 
                       ifelse(nTumorRefGtyrMapped > 0 & nTumorRefGtyrMapped <  ploidy & nTumorAltGtyrMapped > 0 & nTumorAltGtyrMapped < ploidy, 
                              "het", 
                              "toCheck"))), 
         statusUpdate = ifelse(status == "toCheck" & nTumorRefGtyrMapped > 0 & nTumorAltGtyrMapped == 0, 
                "Ref", 
                ifelse(status == "toCheck" & nTumorAltGtyrMapped > 0 & nTumorRefGtyrMapped == 0, 
                       "Alt", 
                       status))) %>%
  mutate(locationAlt = paste0(CHROM, ":", POS, "_", REF, ALT)) 


# table(germlinePairedGenotypedStatus$sampleTumor, 
#       germlinePairedGenotypedStatus$statusUpdate)
# germlinePairedGenotypedStatusNormal <- germlinePairedGenotypedStatus %>%
#   filter(status != "toCheck")
# 
# germlinePairedGenotypedStatusCheck <- germlinePairedGenotypedStatus %>%
#   filter(status == "toCheck") %>%
#   mutate(status = ifelse(nTumorRefGtyrMapped > 0 & nTumorAltGtyrMapped == 0,
#                 "Ref",
#                 ifelse(nTumorAltGtyrMapped > 0 & nTumorRefGtyrMapped == 0,
#                        "Alt",
#                        "toCheck"))) %>%
#   collect()
# 
# table(germlinePairedGenotypedStatusCheck$status)
# table(germlinePairedGenotypedStatusCheck$status)
# > table(germlinePairedGenotypedStatusCheck$CHROM, germlinePairedGenotypedStatusCheck$sex)
#       male
#   X 227124
#   Y     26
# table(germlinePairedGenotypedStatusCheck$statusUpdate)
# 
#     Ref toCheck 
#  158950   68200 


germlinePairedGenotypedStatusDft2 <- germlinePairedGenotypedStatus  %>%
  filter(sampleTumor %in% dft2$sampleTumor) %>%
  distinct(locationAlt, sampleTumor, statusUpdate) %>%
  group_by(locationAlt, statusUpdate) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(locationAlt) %>%
  mutate(total_tumors = sum(n),
         pct = n / total_tumors) %>% 
  collect()


dt <- as.data.table(germlinePairedGenotypedStatusDft2)

all_status <- c("homRef", "homAlt", "het", "Ref", "Alt", "toCheck")

resultDft2 <- dt[, {
  # initialize all statuses to 0
  pct_vec <- setNames(rep(0, length(all_status)), all_status)
  
  # fill observed values
  pct_vec[statusUpdate] <- pct
  
  # majority logic
  max_val <- max(pct_vec)
  winners <- names(pct_vec)[pct_vec == max_val]
  majority <- if (length(winners) > 1) "flag" else winners
  
  c(
    list(total_tumors = unique(total_tumors)),
    as.list(pct_vec),
    list(majority = majority)
  )
}, by = locationAlt]


checkFlagsDft2 <- resultDft2 %>%
  filter(majority == "flag")


germlinePairedGenotypedStatusDft1 <- germlinePairedGenotypedStatus  %>%
  filter(!sampleTumor %in% dft2$sampleTumor) %>%
  distinct(locationAlt, sampleTumor, statusUpdate) %>%
  group_by(locationAlt, statusUpdate) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(locationAlt) %>%
  mutate(total_tumors = sum(n),
         pct = n / total_tumors) %>% 
  collect()

dt <- as.data.table(germlinePairedGenotypedStatusDft1)

all_status <- c("homRef", "homAlt", "het", "Ref", "Alt", "toCheck")

resultDft1 <- dt[, {
  # initialize all statuses to 0
  pct_vec <- setNames(rep(0, length(all_status)), all_status)
  
  # fill observed values
  pct_vec[statusUpdate] <- pct
  
  # majority logic
  max_val <- max(pct_vec)
  winners <- names(pct_vec)[pct_vec == max_val]
  majority <- if (length(winners) > 1) "flag" else winners
  
  c(
    list(total_tumors = unique(total_tumors)),
    as.list(pct_vec),
    list(majority = majority)
  )
}, by = locationAlt]

table(resultDft1$majority)

checkFlagsDft1 <- resultDft1 %>%
  filter(majority == "flag")
```

### 3.1.5 Load genotyper results 
- Filter on conditions of interest
```{r}
germlinePairedInformative<- open_dataset(paste0(resDir, "germlineGenotyper/")) %>%
  filter(
    (nTumorAltGtyrMapped > 0 & nHostAltGtyrMapped == 0) |
    (nTumorRefGtyrMapped > 0 & nHostRefGtyrMapped == 0) |
    (nHostAltGtyrMapped > 0 & nTumorAltGtyrMapped == 0) |
    (nHostRefGtyrMapped > 0 & nTumorRefGtyrMapped == 0)
  ) 

# > nrow(germlinePairedInformative)
# [1] 21168207

germlinePairedInformative <- write_dataset(germlinePairedInformative,
                                           path = paste0(resDir,
                                                         "germlineInformative/"),
  format = "parquet",
  partitioning = "sampleTumor")


gc()

germlinePairedInformative <- open_dataset(paste0(resDir, "germlineInformative/")) %>%
  collect()
```

## 3.3 Merge tumor AlCo with germline vcf information.
- Join filtered vcf file with AlCo results from tumor RNAseq.
- Of note, NOT all samples have an associated RNA-seq file! See metadata script.
- Filtering AlCoDf for any rows where expression > 0. If an informative position has 0 RNAseq reads assigned, the expression of the informative allele is 0. 

dim(joined) --> i.e., 2 million germline variants with an informative allele for which we also have RNAseq data.
[1] 21164054       33
```{r}
germlinePairedInformativeMatched <- open_dataset(paste0(resDir,
                                                        "germlineInformative/")) %>%
  filter(sampleTumor %in% metadata$tumorId)

# 58,966,002 rows.
germlineAlCoDfFiltered <- open_dataset(paste0(resDir, "germlineAlCoDfFiltered/")) %>%
  dplyr::select(c(tumorLocationAlt,
                  tumorIdRnaSeq, 
                  TOTAL,
                  NREF, 
                  NALT)) 

joined <- germlinePairedInformativeMatched %>%
  left_join(germlineAlCoDfFiltered, 
            by = "tumorLocationAlt") %>%
  mutate(POS = as.numeric(POS))

write_dataset(joined, 
              paste0(resDir, "germlineAlCoVarsJoined/"),
              format = "parquet",
              partitioning = "sampleTumor")

# [1] 1346513
# joinedInform <- open_dataset(paste0(resDir, "germlineAlCoVarsJoined/")) %>%
#   filter(TOTAL > 0) %>%
#   collect()
# 
# nrow(joinedInform)
```

## 3.3 GTF annotations
### 3.3.1 Annotate CHROM:POS
- From informative file, get all CHROM and POS. 
- Merge with gtf file on informative positions (exons)

```{r}
# > nrow(joinedChromPos)
# [1] 1018330

joinedChromPos <- joined %>%
  dplyr::select(CHROM, POS) %>%
  distinct() %>%
  collect()

gc()

gtfFile <- "~/gitClones/01_Deconvolution/02_Deconvolution_DFT1/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf"

gtf <- import(gtfFile)

genes <- gtf[gtf$type == "gene"]

exons <- gtf[gtf$type == "exon"]

genes <- genes[genes$gene_biotype == "protein_coding"]

variants <- GRanges(
  seqnames = joinedChromPos$CHROM,
  ranges = IRanges(start = joinedChromPos$POS, 
                   end = joinedChromPos$POS)
)

# Which variants fall in an exon
exonHits <- findOverlaps(variants, exons)

# Map exon hits to gene IDs
exonGeneIds <- exons$gene_id[subjectHits(exonHits)]
exonGeneNames <- exons$gene_name[subjectHits(exonHits)]
exonBiotypes <- exons$gene_biotype[subjectHits(exonHits)]

hitsDf <- data.frame(
  query = queryHits(exonHits),
  gene_id = exonGeneIds,
  gene_name = exonGeneNames,
  biotype = exonBiotypes
)

collapseFun <- function(x) paste(unique(x), collapse = ";")

collapsed <- aggregate(. ~ query, 
                       data = hitsDf, 
                       FUN = collapseFun)

collapsed$CHROM <- joinedChromPos$CHROM[collapsed$query]
collapsed$POS   <- joinedChromPos$POS[collapsed$query]
```

### 3.3.2 Join with AlCo + Germline Info
```{r}
joinedAnno <- open_dataset(paste0(resDir, "germlineAlCoVarsJoined/")) %>%
  left_join(collapsed)

write_dataset(joinedAnno, 
              paste0(resDir, "germlineAlCoVarsJoinedAnno/"),
              format = "parquet",
              partitioning = "sampleTumor")

# Checking. All positions should have been queried, thus should return 0 rows? No, returns 20815775 rows, because exon hits only contains variant hits. 
# nrow(joinedAnnoFiltered %>% filter(is.na(query)==T))
# [1] 348,279 rows in an informative regions!
# joinedAnnoFiltered <- joinedAnno %>%
#   collect() %>%
#   filter(is.na(query)==F)
```

# 4. Normalize nRef and nAlt
## 4.1 Calculate size factors
- Of note, sample 1538T1 was not filtered out properly, we don't have matched RNAseq data for this sample. Excluding here but need to go back and filter this. 
```{r}
joinedAnnoFiltered <- open_dataset(paste0(resDir, "germlineAlCoVarsJoinedAnno/")) %>%
  mutate(locationAlt = paste0(CHROM, ":", POS, "_", REF, ALT)) %>%
  filter(is.na(query)==F) %>%
  filter(is.na(TOTAL)==F) %>%
  collect()

# nrow(joinedAnnoFiltered)
# [1] 341547
```

### 4.1.1 Calculate size factors for all variant positions
- Using TOTAL at each variant position, compute size factors per sample. 
```{r}
dt <- copy(as.data.table(joinedAnnoFiltered))

dt[, samplename := sampleTumor]

# remove variants with zero coverage everywhere
dt[, empty := sum(TOTAL, na.rm = TRUE) == 0, by = locationAlt]
variants <- dt[empty == FALSE]

# geometric means per variant
gmeans <- variants[, .(gmean = geomean(TOTAL + 1)), by = locationAlt]

setkey(gmeans, locationAlt)
setkey(variants, locationAlt)

variants[gmeans, gmean := i.gmean]

# ratios
variants[, scaled_total := TOTAL / gmean]

# size factors
size_factors <- variants[, .(
  size_factor = median(scaled_total, na.rm = TRUE)
), by = samplename]

setkey(size_factors, samplename)

dt[size_factors, on = "samplename", size_factor := i.size_factor]

dt[, `:=`(
  NREF_norm = NREF / size_factor,
  NALT_norm = NALT / size_factor,
  TOTAL_norm = TOTAL / size_factor
)]

dt <- dt %>%
  mutate(variantType = "germline")


write_dataset(dt, 
              paste0(resDir, "germlineAlCoVarsJoinedAnnoNorm/"),
              format = "parquet",
              partitioning = "sampleTumor")
```

### 4.1.2 Calculate size factors for variant positions present in all tumors only
- No variants are shared by all tumors (nrow(chosen) = 0), thus no NREF_norm_all/NALT_norm_all.
- Not compatible with DFT1/DFT2 so leaving this out for now, potentially revisit!
```{r}
# n_samples <- uniqueN(dt$samplename)
# 
# chosen <- dt[TOTAL > 0,
#              .N, by = locationAlt][N == n_samples]
# 
# variants_all <- dt[locationAlt %in% chosen$locationAlt]
# 
# gmeans_all <- variants_all[, .(gmean = geomean(TOTAL + 1)),
#                            by = locationAlt]
# 
# variants_all[gmeans_all, on = "locationAlt", gmean := i.gmean]
# variants_all[, scaled_total := TOTAL / gmean]
# 
# size_factors_all <- variants_all[, .(
#   size_factor_all = median(scaled_total, na.rm = TRUE)
# ), by = samplename]
# 
# dt[size_factors_all, on = "samplename",
#    size_factor_all := i.size_factor_all]
# 
# dt[, `:=`(
#   NREF_norm_all = NREF / size_factor_all,
#   NALT_norm_all = NALT / size_factor_all
# )]
```

## 4.2 Normalize expression by copy number of informative variant

## 4.3 Compute eT and eH


