# 0 Environment
## 0.1 Packages
```{r}
# install.packages("slider")
```

### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(arrow)
library(furrr)
library(patchwork)
library(slider)
library(duckdb)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_dft1Som"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
source("~/gitClones/08_deconvTable/082_Scripts/genotype_germline_variants.R")
histoDir <- paste0(plotDir, 
                   "histo/")
dir.create(histoDir)
barDir <- paste0(plotDir, 
                 "bar/")
dir.create(barDir)
```

## 0.3 Custom functions
### 0.3.1 Compute tAlt
```{r}
computeTAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         hTotal) {
  
    # print("Solving for tAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    tAlt = (obsVaf * denominator)/purityTumor
  return(tAlt)
}
```
### 0.3.2 Compute hAlt
```{r}
computeHAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         tAlt,
                         hTotal) {
  
  # Cases where hAlt = 0, solve for tAlt
 
    print("Solving for hAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    hAlt = (obsVaf*denominator)/(1-purityTumor)
  return(hAlt)
}

computeHAlt(obsVaf = 0.5, 
            purityTumor = 0.8, 
            tTotal = 4, 
            hTotal = 2, 
            tAlt=0)
```
### 0.3.3 Geometric mean
- Kevin's scripts
  - To calculate size factors: 
  - Compute the geom.mean of totals for each variant (optional: add pseudocount)
  - Divide each variant's read total by the geometric mean for each variant
  - Then take the median of these ratios for each sample
  - (Optional) restrict to variants with presence in all tumors
```{r}
geomean <- function(x) {
  if (all(x == 0)) {
    NA_real_
  } else {
    exp(mean(log(x[x > 0])))
  }
}
```

### 0.3.4 Summarize dft1Som variants
```{r}
library(dplyr)
library(tidyr)

library(data.table)

summarise_variants_dt <- function(df) {
  dt <- as.data.table(df)
  
  # ensure uniqueness per tumor per location
  dt <- unique(dt[, .(locationAlt, sampleTumor, statusUpdate)])
  
  # count per location + status
  counts <- dt[, .N, by = .(locationAlt, statusUpdate)]
  
  # total tumors per location
  totals <- dt[, .(total_tumors = uniqueN(sampleTumor)), by = locationAlt]
  
  # merge counts with totals
  counts <- counts[totals, on = "locationAlt"]
  
  # compute percentage
  counts[, pct := N / total_tumors]
  
  # reshape wide (one column per statusUpdate)
  wide <- dcast(
    counts,
    locationAlt + total_tumors ~ statusUpdate,
    value.var = "pct",
    fill = 0
  )
  
  # compute majority
  status_cols <- setdiff(names(wide), c("locationAlt", "total_tumors"))
  
  wide[, majority := {
    vals <- unlist(.SD)
    max_val <- max(vals)
    winners <- names(vals)[vals == max_val]
    
    if (length(winners) > 1) "flag" else winners
  }, by = locationAlt, .SDcols = status_cols]
  
  return(wide)
}
```

### 0.3.4 Copy number state sequence
```{r}
possible_cn_absorbing <- function(start_cn, nGain, nLoss) {
  # Each state is a pair: CN, remaining events (gains and losses)
  states <- list(list(cn = start_cn, g = nGain, l = nLoss))
  
  final_cns <- c()
  
  while (length(states) > 0) {
    new_states <- list()
    
    for (s in states) {
      cn <- s$cn
      g  <- s$g
      l  <- s$l
      
      if (cn == 0 || (g == 0 && l == 0)) {
        # CN is 0 (absorbing) or no events left → record final CN
        final_cns <- c(final_cns, cn)
      } else {
        # Apply a gain if available
        if (g > 0 && cn > 0) {
          new_states <- append(new_states, list(list(cn = cn + 1, g = g - 1, l = l)))
        }
        # Apply a loss if available
        if (l > 0) {
          new_cn <- max(0, cn - 1)
          new_states <- append(new_states, list(list(cn = new_cn, g = g, l = l - 1)))
        }
      }
    }
    
    states <- new_states
  }
  sort(unique((final_cns)))
}

possible_cn_absorbing(3, 2, 3)
```
# 1 Data
## 1.1 Purities and ploidy
```{r}
dft1SomPurities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(dft1SomPurities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(dft1SomPurities),
  paste0(colnames(dft1SomPurities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(dft1SomPurities) <- newNames
dft1SomPurities <- dft1SomPurities[-1, ]

dft1SomPuritiesPloidy <- dft1SomPurities %>%
  mutate(ploidy = ifelse(X15_PLOIDY == "diploid",
                 2, 
                 ifelse(X15_PLOIDY == "tetraploid", 
                        4, 
                        "toCheck") 
                 )) %>%
  filter(ploidy != "toCheck") %>%
  dplyr::rename("sampleTumor"="TUMOUR.INFORMATION_TUMOUR ID", 
                     "purity"="X14_PURITY (VAF-based)", 
                "dftType"="X9_LINEAGE") %>%
              mutate(purity = as.numeric(purity), 
                     ploidy = as.numeric(ploidy), 
                     sex=ifelse(HOST.SEX == "♂", 
                                "male", 
                                ifelse(HOST.SEX == "♀", 
                                       "female", 
                                       "toCheck"))) %>%
              dplyr::select(sampleTumor, 
                            ploidy, 
                            purity, 
                            sex, 
                            dftType, 
                            X15_PLOIDY) 
  
table(dft1SomPuritiesPloidy$sex, 
      dft1SomPuritiesPloidy$dftType)

males <- dft1SomPuritiesPloidy %>%
  filter(sex == "male") 

dft1 <- dft1SomPuritiesPloidy %>%
  filter(dftType == "DFT1") 

dft2 <- dft1SomPuritiesPloidy %>%
  filter(dftType == "DFT2") # 41 samples. 
```
## 1.2 VCF files
- Read in dft1Som vcf file, all variants. 
- Pivot longer so that each variant and each sample is a single row with the ALT/TOTAL ratio.
```{r}
# dft1Som <- read_table(paste0(dataDir, "Data_S1A_DFT1_somatic_variants.txt"),
#                       col_types = cols(.default = "c"))
# 
# dft1SomLong <- dft1Som %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "tumor",
#     values_to = "ratio"
#   ) %>%
#   mutate(chrom = as.character(CHROM),
#          location = paste0(CHROM, ":", POS),
#          position = as.numeric(POS)) %>%
#   filter(nchar(REF) == 1, nchar(ALT) == 1) %>%
#   dplyr::rename("sampleTumor"="tumor") 
# 
# rm(dft1Som)
# gc()
```

## 1.3 Pair tumor/host samples
- Using shared ID's pivot wider so that the ratio for matched tumor/host samples is side by side.
- Filter on variants where alt allele is present in only the host or the tumor sample within a pair! --> so that we have info. on tAlt or hAlt at that position.
- Tumor samples = 112
- Host samples = 81
- Paired = 103 --> 100 if I filter Max's table? Maybe the non-transmissible cases. 
- Add multiallelic annotation, ALL SAMPLES!!
```{r}
# chroms <- c(paste0(1:6),
#             "X",
#             "Y")
# 
# dft1Som <- dft1SomLong %>%
#   mutate(
#     tumorAlt = as.integer(sub("/.*", "", ratio)),
#     tumorTotal = as.integer(sub(".*/", "", ratio))
#   ) %>%
#   mutate(obsVafTumor = tumorAlt/tumorTotal)%>%
#  mutate(locationRef = paste0(CHROM, ":", POS, "_", REF),
#         locationRefAlt = paste0(CHROM, ":", POS, "_", REF, "_", ALT),
#         tumorLocation = paste0(sampleTumor, "_", CHROM, ":", POS),
#         tumorLocationRef = paste0(sampleTumor,
#                                   "_", CHROM, ":", POS, "_", REF),
#         tumorLocationRefAlt = paste0(sampleTumor,
#                                   "_", CHROM, ":", POS, "_", REF, "_", ALT),
#         position = as.numeric(POS)) %>%
#   filter(CHROM %in% chroms)
# 
# # Overwrite data.
# write_dataset(dft1Som,
#               format = "parquet",
#               paste0(resDir,
#                      "dft1Som/08_deconvTable_dft1Som_computedVafs/"),
#               partitioning = "sampleTumor")
```

## 1.4 Annotate multiallelic variants. 
- Somatic variants are not present in any of the host samples. 
- That could mean it has 0 VAF in a given tumor sample, but this was not observed in all samples and is therefore not somatic. 
- How to account for this when calling "true" informative variants?
- 30 multiallelic positions!!!
```{r}
# multiallelicSites <- open_dataset(paste0(resDir, "dft1Som/08_deconvTable_dft1Som_computedVafs/")) %>%
#   dplyr::select(tumorLocationRef, ALT, sampleTumor) %>%
#   distinct() %>%
#   group_by(tumorLocationRef) %>%
#   summarise(nAltAlleles = n_distinct(ALT), .groups = "drop") %>%
#   filter(nAltAlleles > 1) %>%
#   collect()
# 
# nAltAlleles <- open_dataset(paste0(resDir, "dft1Som/08_deconvTable_dft1Som_computedVafs/")) %>%
#   dplyr::select(tumorLocationRef, ALT, sampleTumor) %>%
#   filter(tumorLocationRef %in% multiallelicSites$tumorLocationRef) %>%  # Arrow predicate pushdown
#   distinct() %>%
#   collect() %>%
#   group_by(tumorLocationRef) %>%
#   summarise(
#     nAltAlleles     = n_distinct(ALT),
#     altAlleles      = paste(sort(unique(ALT)), collapse = ","),
#     baseA = ifelse("A" %in% ALT, paste(sampleTumor[ALT == "A"], collapse = ","), NA),
#     baseC = ifelse("C" %in% ALT, paste(sampleTumor[ALT == "C"], collapse = ","), NA),
#     baseG = ifelse("G" %in% ALT, paste(sampleTumor[ALT == "G"], collapse = ","), NA),
#     baseT = ifelse("T" %in% ALT, paste(sampleTumor[ALT == "T"], collapse = ","), NA),
#     .groups = "drop"
#   )
# 
# gc()
# 
# # All tumors have a count at these variant, just flag to exclude from majority calculation??
# test <- open_dataset(paste0(resDir, "dft1Som/08_deconvTable_dft1Som_computedVafs/")) %>%
#   filter(tumorLocationRef %in% multiallelicSites$tumorLocationRef) %>%
#   collect()
# 
# gc()
# 
# length(unique(test$locationRefAlt)) # 30 
# length(unique(test$locationRef)) # 15 
# 
# write.xlsx(test,
#            paste0(resDir, "/dft1Som/08_deconvTable_dft1SomVariants_multiallelicData.xlsx"))
# 
# # Join dft1Som with multiallelic annotation.
# multiallelicSites <- multiallelicSites %>%
#   mutate(multiallelic = 1L)
# 
# dft1Som <- open_dataset(paste0(resDir, "dft1Som/08_deconvTable_dft1Som_computedVafs/")) %>%
#   left_join(multiallelicSites %>%
#               dplyr::select(tumorLocationRef,
#                             multiallelic)) %>%
#   mutate(multiallelic = ifelse(is.na(multiallelic), 0L, multiallelic))
# 
# 
# write_dataset(dft1Som,
#                                 paste0(resDir,
#                                       "dft1Som/08_deconvTable_dft1Som_computedVafs_multiAnno/"),
#                                 partitioning = "sampleTumor",
#                                 format = "parquet")
# 
# 
# dft1Som <- open_dataset(paste0(resDir,
#                                "dft1Som/08_deconvTable_dft1Som_computedVafs_multiAnno/")) 
```

## 1.5 Filter tumors of interest 
```{r}
# dft1Som <- open_dataset(paste0(resDir,
#                             "/dft1Som/08_deconvTable_dft1Som_computedVafs_multiAnno/")) %>%
#   left_join(dft1SomPuritiesPloidy %>% dplyr::select(sampleTumor, X15_PLOIDY))
```


# 2 CNVs - DFT1
## 2.1 Read in CNV data, join with VAF's
- Run selection as background job (Code/scroll down), copy output to global env. 
```{r}
dft1Cnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx",
                     sheet = "Table S6B",
                     startRow = 3)

dft1CnvLong <- dft1Cnv %>%
  pivot_longer(-c(1:12),
              names_to = "tumor",
              values_to = "presence") %>%
  mutate(presence = as.numeric(presence)) %>%
  filter(presence > 0) %>% # If 0, variant not in CNV segment, exclude to reduce size of dataframe.
  mutate(
    START = as.numeric(START),
    END = as.numeric(END),
    CHROM = as.character(CHROM),
    tumor = as.character(tumor)
  ) %>%
  # write as camel case for sql with duck db.
  mutate(eventType = EVENT.TYPE,
         cnvEventId = CNV.EVENT.ID,
         sampleTumorChrom = paste0(tumor, "_", CHROM)) %>%
  dplyr::select(-c("CNV.EVENT.ID",
                   "WIDTH",
                   "NCBI.ANNOTATION",
                   "ENSEMBL.ANNOTATION",
                   "COSMIC",
                   "N",
                   "EVENT.TYPE",
                   "MARKER.5",
                   "DOUBLE.MINUTE"))


rm(dft1Cnv)
gc()

# # check <- dft1CnvLong$tumor
# # check2 <- dft1Som %>%
# #   distinct(sampleTumor) %>%
# #   collect()
# # 
# # length(intersect(check, check2$sampleTumor))
# # > length(intersect(check, check2$sampleTumor))
# # [1] 78
# 
# con <- dbConnect(duckdb(), config = list(threads = availableCores() - 1))
# 
# duckdb_register_arrow(con, "dft1Som", dft1Som)  # no collect() needed
# duckdb_register(con, "cnv", dft1CnvLong)
# 
# # Write parquet directly from DuckDB — never fully loads result into RAM
# dbExecute(con, paste0("
#   COPY (
#     SELECT g.*, c.presence, c.CHROM, c.START, c.END, c.cnvEventId
#     FROM dft1Som g
#     LEFT JOIN cnv c
#            ON  g.CHROM    = c.CHROM
#            AND g.sampleTumor    = c.tumor
#            AND g.position >= c.START
#            AND g.position <= c.END
#   ) TO '", resDir, "/dft1Som/", fprefix, "_dft1CnvJoin/'
#   (FORMAT PARQUET, PARTITION_BY (sampleTumor), OVERWRITE)
# "))
# 
# dbDisconnect(con, shutdown = TRUE)
# 
# # Which variants have > 0 changes?
# presence <- open_dataset(paste0(resDir,
#                                 "/dft1Som/",
#                                 fprefix,
#                                 "_dft1CnvJoin/")) %>%
#   dplyr::select(tumorLocationRefAlt) %>%
#   collect()
# # 16235199 variants. 
```

## 2.3 Starting CN
- Adjust this for all variants. 
- Important to define for running genotyper(). 
- Case when need to keep input type the same (i.e., character string or numeric). Revert back to numeric after checking there are no "toCheck" rows. 
```{r}
# autosomes <- paste0(1:6)
# 
# dft1CnvJoin <- open_dataset(paste0(resDir,
#                                    "/dft1Som/",
#                                    fprefix,
#                                    "_dft1CnvJoin/")) %>%
#   # We left out event type to run DuckDb. Merge eventType with dft1CnvJoin.
#   left_join(dft1CnvLong %>%
#               dplyr::select(cnvEventId,
#                             eventType) %>%
#               distinct()) %>%
#   left_join(dft1SomPuritiesPloidy %>%
#               dplyr::select(X15_PLOIDY,
#                             sampleTumor)) %>%
#   dplyr::select(-c(chrom_1, CHROM_2)) %>%
#   # Set the chrom type. 
#   mutate(chromType = ifelse(CHROM %in% autosomes, 
#                             "autosome", 
#                             CHROM)) %>%
#   # Define CN by combination of dft type, ploidy, and chromType
#   mutate(postWgdCn = case_when(
#     ############################
#     ## Diploid DFT1s ##
#     ############################
#     # Diploid, DFT1, autosome (2 copies)
#     X15_PLOIDY == "diploid" & sampleTumor %in% dft1$sampleTumor & chromType == "autosome" ~ "2",
#     # Diploid, DFT1, sex (XX)
#     X15_PLOIDY == "diploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ "2",
#     # Diploid, DFT1, no Y
#     X15_PLOIDY == "diploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ "0",
#     ############################
#     ## Tetraploid DFT1s ##
#     ############################
#     # Tetraploid, DFT1, autosome
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & chromType == "autosome" ~ "4",
#     # Tetraploid, DFT1, sex (XXXX)
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ "4",
#     # Tetraploid, DFT1, no Y chromosome
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ "0",
#     # ############################
#     # ## Diploid DFT2s ##
#     # ############################
#     # # Diploid, DFT2, autosome
#     # X15_PLOIDY == "diploid" & sampleTumor %in% dft2$sampleTumor & chromType == "autosome" ~ 2,
#     # # Diploid, DFT2, sex (XY)
#     # X15_PLOIDY == "diploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "X" ~ 1,
#     # X15_PLOIDY == "diploid" & sampleTumor %in% dft2$sampleTumor & CHROM == "Y" ~ 1,
#     # ############################
#     # ## Tetraploid DFT2s ##
#     # ############################
#     # # Tetraploid, DFT2, autosome
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & chromType == "autosome" ~ 4,
#     # # Tetraploid, DFT2, sex (XXYY)
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ 2,
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ 2,
#     TRUE ~ "toCheck"
#   )) %>%
#     mutate(preWgdCn = case_when(
#     ############################
#     ## Tetraploid DFT1s ##
#     ############################
#     # Tetraploid, DFT1, autosome
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & chromType == "autosome" ~ "2",
#     # Tetraploid, DFT1, sex (XX)
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ "2",
#     # Tetraploid, DFT1, no Y chromosome
#     X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ "0",
#     # ############################
#     # ## Tetraploid DFT2s ##
#     # ############################
#     # # Tetraploid, DFT2, autosome, preWgd = 2 copies
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft2$sampleTumor & chromType == "autosome" ~ 2,
#     # # Tetraploid, DFT2, sex (XXYY), preWgd = XY (1 copy)
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "X" ~ 1,
#     # X15_PLOIDY == "tetraploid" & sampleTumor %in% dft1$sampleTumor & CHROM == "Y" ~ 1,
#     TRUE ~ postWgdCn
#     ))
# 
# 
# 
# # toCheck <- dft1CnvJoin %>%
# #   filter(preWgdCn == "toCheck") %>%
# #   collect() # No rows!!
# rm(dft1CnvLong)
# gc()
# 
# dft1CnvJoin <- dft1CnvJoin %>%
#   mutate(preWgdCn = as.numeric(preWgdCn),
#          postWgdCn = as.numeric(postWgdCn)) %>%
#   collect()
# table(dft1CnvJoin$CHROM, dft1CnvJoin$preWgdCn)
  #         2
  # 1 3734355
  # 2 3323590
  # 3 3099301
  # 4 2630401
  # 5 1391415
  # 6 1458734
  # X  597403


# table(dft1CnvJoin$CHROM, dft1CnvJoin$postWgdCn)
  #         2       4
  # 1 2964084  770271
  # 2 2643445  680145
  # 3 2462392  636909
  # 4 2098588  531813
  # 5 1103439  287976
  # 6 1144768  313966
  # X  471609  125794
# write_dataset(dft1CnvJoin, 
#               paste0(resDir, "/dft1Som/dft1CnvJoin_prePostAdjustedCn/"), 
#               partitioning = "sampleTumor", 
#               format = "parquet")


```

## 2.2 Adjust CN
```{r}
# dft1CnvJoin <- open_dataset(paste0(resDir,
#                                    "dft1Som/dft1CnvJoin_prePostAdjustedCn/"))
# 
# # --- 1. Find duplicate keys via aggregation. These are rows where there is more than one CNV event ID spanning a given variant (i.e., a multiplet). This does not overestimate the number of multiplets because of multiallelic variants, because I am counting by ref and alt.
# dup_keys_refalt <- dft1CnvJoin %>%
#   filter(presence > 0) %>%
#   distinct(tumorLocationRefAlt, cnvEventId) %>%
#   count(tumorLocationRefAlt) %>%
#   collect() %>%
#   filter(n > 1) %>%
#   pull(tumorLocationRefAlt)
# 
# # length(dup_keys_refalt)
# # [1] 160299
# 
# # --- 2. Find single keys via aggregation. These are rows where there one CNV event ID spanning a given variant (i.e., a singlet). This does not overestimate the number of multiplets because of multiallelic variants, because I am counting by ref and alt.
# sing_keys_refalt <- dft1CnvJoin %>%
#   filter(presence > 0) %>%
#   distinct(tumorLocationRefAlt, cnvEventId) %>%
#   count(tumorLocationRefAlt) %>%
#   collect() %>%
#   filter(n == 1) %>%
#   pull(tumorLocationRefAlt)
# 
# # length(sing_keys_refalt)
# # [1] 1611830
# 
# noChange <- dft1CnvJoin %>%
#   filter(is.na(presence)) %>%
#   distinct(tumorLocationRefAlt, cnvEventId) %>%
#   pull(tumorLocationRefAlt)
# 
# # [1] 14280661
# 
# dft1CnvJoinMultiplets <- dft1CnvJoin %>%
#   mutate(multipletsRefAlt = ifelse(tumorLocationRefAlt %in% dup_keys_refalt,
#                 1,
#                 0),
#          singletsRefAlt = ifelse(tumorLocationRefAlt %in% sing_keys_refalt,
#                            1,
#                            0),
#          cnvType = ifelse(tumorLocationRefAlt %in% dup_keys_refalt,
#                           "multiplet",
#                           ifelse(tumorLocationRefAlt %in% sing_keys_refalt,
#                                  "singlet",
#                                  "noChange"))) %>% # None should be assigned "noChange" at this point, because we have prefiltered on variants where presence > 0.
#   mutate(variantType = "dft1Som") %>%
#   collect()
# 
# write_dataset(dft1CnvJoinMultiplets,
#               format = "parquet",
#               paste0(resDir,
#                      "/dft1Som/dft1CnvJoinMultiplets/"),
#               partitioning = "sampleTumor")
```

### 2.2.1 Variants with no change
- copyNumberAdjusted = postWgdCn (i.e., no change depending on pre/post.)
```{r}
# dft1CnvJoin <- open_dataset(paste0(resDir,
#                      "/dft1Som/dft1CnvJoinMultiplets/"))
# 
# dft1CnvJoinNoChanges <- dft1CnvJoin %>%
#   filter(is.na(presence)) %>%
#          mutate(sampleTumorChrom = paste0(sampleTumor, "_", CHROM)) %>%
#   # Make copy number the adjusted CN for WGD (in tetraploid tumors, in diploid, no change)
#   mutate(copyNumberAdjusted = postWgdCn) %>%
#   group_by(tumorLocationRefAlt) %>%
#   mutate(
#     nGain = 0,
#     nLoss = 0,
#     nGainPost = 0,
#     nGainPre = 0,
#     nLossPost = 0,
#     nLossPre = 0,
#     sumPresence = 0
#   ) %>%
#   # remove CNV table columns
#   mutate(cnvAdjAnno = "noChanges",
#          multipletsRefAlt = 0,
#          singletsRefAlt = 0,
#          variantType = "dft1Som") %>%
#    # remove CNV table columns, because this has been aggregated by summing rows.
#   dplyr::select(-c(START, 
#                    END, 
#                    cnvEventId,
#                    eventType,
#                    presence))
# 
# gc()
# 
# write_dataset(dft1CnvJoinNoChanges,
#               paste0(resDir, "/dft1Som/dft1CnvJoinNoChanges/"),
#               partitioning = "sampleTumorChrom",
#               format = "parquet")
# 
# gc()
```

### 2.2.2 Variants with single change
- change ploidy according to single step
- number of unique variants = number of rows, because only 1 change to cnv per variant. 5 tumorLocation where more than one variant with a single CNV change.
- All presences should be 1 because single change
- If CN goes to 0, it should stay at 0. (Applies for multistep, not here.) Running same code as multiplets as a sanity check, none of the maxCn should differ from minCn. 

- I have subset regions from the bam files of 18T and 528T2. 
- Manually inspect coverage to check what the CN is at these "conflicting positions"?
```{r}
dft1CnvJoin <- open_dataset(paste0(resDir,
                     "/dft1Som/dft1CnvJoinMultiplets/"))

# dft1CnvJoinSingleChange <- dft1CnvJoin %>%
#   filter(singletsRefAlt == 1 & multipletsRefAlt == 0 ) %>%
#   collect()

dft1CnvJoinSingleChangeTetraploid  <- dft1CnvJoin %>%
  filter(singletsRefAlt == 1 & multipletsRefAlt == 0 ) %>%
  filter(sampleTumor %in% c("18T", "528T2", "1011T1")) %>%
  filter(CHROM == "1" & position >= 530000000 & position <= 600000000 ) %>%
  collect()

table(dft1CnvJoinSingleChangeTetraploid$sampleTumor, 
      dft1CnvJoinSingleChangeTetraploid$eventType)

  #        LOSS
  # 1011T1  931
  # 18T     931
  # 528T2     4

dft1CnvJoinSingleChangeTetraploid  <- dft1CnvJoin %>%
  filter(singletsRefAlt == 1 & multipletsRefAlt == 0 ) %>%
  filter(sampleTumor %in% c("18T", "528T2", "1011T1")) %>%
  filter(CHROM == "2" & position >= 240000000 & position <= 250000000 ) %>%
  collect()

table(dft1CnvJoinSingleChangeTetraploid$sampleTumor, 
      dft1CnvJoinSingleChangeTetraploid$eventType)

  #        LOSS
  # 1011T1  274 --> should be CN 1
  # 18T     274 --> either 3 or 2?
  # 528T2   274 --> either 3 or 2?

dft1CnvJoinSingleChangeTetraploid  <- dft1CnvJoin %>%
  filter(singletsRefAlt == 1 & multipletsRefAlt == 0 ) %>%
  filter(sampleTumor %in% c("18T", "528T2", "1011T1")) %>%
  filter(CHROM == "3" & position >= 115000000 & position <= 135000000) %>%
  collect()

table(dft1CnvJoinSingleChangeTetraploid$sampleTumor, 
      dft1CnvJoinSingleChangeTetraploid$eventType)

  #       GAIN (post-tetraploidisation)
  # 528T2                          1747 --> should be 5

# CHR4_305000000_310000000 
dft1CnvJoinSingleChangeTetraploid  <- dft1CnvJoin %>%
  filter(singletsRefAlt == 1 & multipletsRefAlt == 0 ) %>%
  filter(sampleTumor %in% c("18T", "528T2", "1011T1")) %>%
  filter(CHROM == "4" & position >= 305000000 & position <= 310000000 ) %>%
  collect()

# table(dft1CnvJoinSingleChangeTetraploid$sampleTumor, 
#       dft1CnvJoinSingleChangeTetraploid$eventType)
# 
#        GAIN
#   1011T1  301 --> CN = 3
#   18T     301 --> CN = 6 or 5
#   528T2   301 --> CN = 6 or 5

# CHR5_240000000_250000000
dft1CnvJoinSingleChangeTetraploid  <- dft1CnvJoin %>%
  filter(singletsRefAlt == 1 & multipletsRefAlt == 0 ) %>%
  filter(sampleTumor %in% c("18T", "528T2", "1011T1")) %>%
  filter(CHROM == "5" & position >= 240000000 & position <= 250000000) %>%
  collect()

# table(dft1CnvJoinSingleChangeTetraploid$sampleTumor,
#       dft1CnvJoinSingleChangeTetraploid$eventType)
# < table of extent 0 x 0 >

# CHR6_135000000_140000000
dft1CnvJoinSingleChangeTetraploid  <- dft1CnvJoin %>%
  filter(singletsRefAlt == 1 & multipletsRefAlt == 0 ) %>%
  filter(sampleTumor %in% c("18T", "528T2", "1011T1")) %>%
  filter(CHROM == "6" & position >= 135000000 & position <= 140000000) %>%
  collect()

# table(dft1CnvJoinSingleChangeTetraploid$sampleTumor,
#       dft1CnvJoinSingleChangeTetraploid$eventType)
# 
#          GAIN (post-tetraploidisation) LOSS
#   1011T1                             0   17 --> should be 1 copy
#   528T2                            372    0 --> should be 5

# CHR6_230000000_240000000

dft1CnvJoinSingleChangeTetraploid  <- dft1CnvJoin %>%
  filter(singletsRefAlt == 1 & multipletsRefAlt == 0 ) %>%
  filter(sampleTumor %in% c("18T", "528T2", "1011T1")) %>%
  filter(CHROM == "6" & position >= 230000000 & position <= 240000000) %>%
  collect()

# table(dft1CnvJoinSingleChangeTetraploid$sampleTumor,
#       dft1CnvJoinSingleChangeTetraploid$eventType)
# 
#         GAIN (post-tetraploidisation)
#   528T2                           576 --> likely 5?

```

### 2.2.3 Variants with multiple changes
- Aggregate steps using same logic as before, only this time ploidy will change using sum of all steps rather than single change of CNV status.
- Issue: we don't know the order of events. Depending on order of losses and gains, the final CN could differ. Adding a permutation function which returns how many times a different final CN state could occur. 
- None of the tetraploid tumors have nLossPre > 1, therefore cannot go to 0.
- The only variable cases are diploid tumors whose variants have two losses and one gain. 
- Filter out the five variants that could have been CN 0 or 1!
  - Check these variants against the counts + gtf file.
```{r}
dft1CnvJoin <- open_dataset(paste0(resDir,
                     "/dft1Som/dft1CnvJoinMultiplets/"))

# CHR6_135000000_140000000
dft1CnvJoinSingleChangeTetraploid  <- dft1CnvJoin %>%
  filter(singletsRefAlt == 0 & multipletsRefAlt == 1 ) %>%
  filter(sampleTumor %in% c("18T", "528T2", "1011T1")) %>%
  filter(CHROM == "6" & position >= 135000000 & position <= 140000000) %>%
  collect()

table(dft1CnvJoinSingleChangeTetraploid$sampleTumor,
      dft1CnvJoinSingleChangeTetraploid$eventType)

```

