---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(txdbmaker)
library(AnnotationDbi)
library(GenomicRanges)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
scriptDir <-  paste0(wDir, "/082_Scripts/")
dir.create(scriptDir)
presDir <-  paste0(wDir, "/085_Presentations/")
dir.create(presDir)


hmDir <- paste0(plotDir, "/heatmap/")
barDir <- paste0(plotDir, "/bar/")
corrDir <- paste0(plotDir, "/corr/")
upsetDir <- paste0(plotDir, "/upset/")
dotDir <- paste0(plotDir, "/dot/")

matDir <- paste0(resDir, "/matrix/")


# dir.create(hmDir)
# dir.create(corrDir)
# dir.create(barDir)
# dir.create(upsetDir)
# dir.create(matDir)
# dir.create(dotDir)
```

## 0.3 Custom functions

# 1 Data
## 1.1 Supplementary files
### 1.1.1 DFT1 somatic variants
- For alleleCounter, exclude indels and export version with column names (Kevin's version) and without (alleleCounter)
```{r}
dft1Som <- read_table(paste0(dataDir, "Data_S1A_DFT1_somatic_variants.txt"), 
                      col_types = cols(.default = "c"))

dft1SomLong <- dft1Som %>%
  pivot_longer(
    cols = -c(CHROM, POS, REF, ALT),
    names_to = "tumor",
    values_to = "ratio"
  ) %>%
  mutate(chrom = as.character(CHROM), 
         location = paste0(CHROM, ":", POS), 
         position = as.numeric(POS)) %>%
  filter(nchar(REF) == 1, nchar(ALT) == 1)

nas <- dft1SomLong %>%
  filter(is.na(CHROM)==T)

dft1SomExport <- dft1SomLong %>%
  dplyr::select(1:4) %>%
  distinct()

write_tsv(dft1SomExport, 
          paste0(resDir, fprefix, "_dft1SomRefAltExport_colnames.tsv"),
          col_names = T, 
          quote = "none")

write_tsv(dft1SomExport, 
          paste0(resDir, fprefix, "_dft1SomRefAltExport_withoutColnames.tsv"),
          col_names = F, 
          quote = "none")

dft1SomExport <- dft1SomLong %>%
  dplyr::select(1:2) %>%
  distinct()

write_tsv(dft1SomExport, 
          paste0(resDir, fprefix, "_dft1SomExport_colnames.tsv"),
          col_names = T, 
          quote = "none")

write_tsv(dft1SomExport, 
          paste0(resDir, fprefix, "_dft1SomExport_withoutColnames.tsv"),
          col_names = F, 
          quote = "none")
```

### 1.1.2 DFT2 somatic variants
```{r}
dft2Som <- read_table(paste0(dataDir, "Data_S1B_DFT2_somatic_variants.txt"), 
                      col_types = cols(.default = "c"))

dft2SomLong <- dft2Som %>%
  pivot_longer(
    cols = -c(CHROM, POS, REF, ALT),
    names_to = "tumor",
    values_to = "ratio"
  ) %>%
  mutate(chrom = as.character(CHROM), 
         location = paste0(CHROM, ":", POS), 
         position = as.numeric(POS)) %>%
  filter(nchar(REF) == 1, nchar(ALT) == 1) 

nas <- dft2SomLong %>%
  filter(is.na(CHROM)==T)

dft2SomExport <- dft2SomLong %>%
  dplyr::select(1:4) %>%
  distinct()

write_tsv(dft2SomExport, 
          paste0(resDir, fprefix, "_dft2SomRefAltExport_colnames.tsv"),
          col_names = T, 
          quote = "none")

write_tsv(dft2SomExport, 
          paste0(resDir, fprefix, "_dft2SomRefAltExport_withoutColnames.tsv"),
          col_names = F, 
          quote = "none")

dft2SomExport <- dft2SomLong %>%
  dplyr::select(1:2) %>%
  distinct()

write_tsv(dft2SomExport, 
          paste0(resDir, fprefix, "_dft2SomExport_colnames.tsv"),
          col_names = T, 
          quote = "none")

write_tsv(dft2SomExport, 
          paste0(resDir, fprefix, "_dft2SomExport_withoutColnames.tsv"),
          col_names = F, 
          quote = "none")
```

### 1.1.3 Germline variants
```{r}
germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"), 
                      col_types = cols(.default = "c"))

germlineLong <- germline %>%
  pivot_longer(
    cols = -c(CHROM, POS, REF, ALT),
    names_to = "tumor",
    values_to = "ratio"
  ) %>%
  mutate(chrom = as.character(CHROM), 
         location = paste0(CHROM, ":", POS), 
         position = as.numeric(POS)) %>%
  filter(nchar(REF) == 1, nchar(ALT) == 1)

nas <- germlineLong %>%
  filter(is.na(CHROM)==T)

germlineExport <- germlineLong %>%
  dplyr::select(1:4) %>%
  distinct()

write_tsv(germlineExport, 
          paste0(resDir, fprefix, "_germlineRefAltExport_withoutColnames.tsv"),
          col_names = F, 
          quote = "none")

write_tsv(germlineExport, 
          paste0(resDir, fprefix, "_germlineRefAltExport_colnames.tsv"),
          col_names = T, 
          quote = "none")
```

## 1.2 GTF file
```{r}
exons <- rtracklayer::import(paste0("~/gitClones/01_Deconvolution/02_Deconvolution_DFT1/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf")) %>%
  as.data.frame() %>%
  filter(type == "exon")
```

## 1.4 Adrian's table (example)
```{r}
adrianDf <- read_csv(paste0("~/gitClones/01_Deconvolution/03_Deconvolution_CTVT/031_Data/ctvt_rnaseq_expression-2.csv"))
head(adrianDf)

length(unique((adrianDf$Tumour))) # [1] 42 - 9 from 
length(unique(adrianDf$Symbol)) # [1] 17243
length(unique(adrianDf$GeneID)) # [1] 17243
```

# 2 Filtering
## 2.1 DFT1
Per tumor-host pair, filter variants where alternative allele is exclusively in tumor or host!
```{r}

```


