---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicRanges)

```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_germline"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
```

## 0.3 Custom functions

# 1 Data
## 1.1 VCF files
- Read in germline vcf file
- Pivot longer so that each variant and each sample is a single row with the ALT/TOTAL ratio.
```{r}
germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"),
                      col_types = cols(.default = "c"))

sampleInfo <- tibble(sample = colnames(germline)[-(1:4)]) %>%
  mutate(
    sampleType = str_extract(sample, "T|H"),
    tcgId = str_extract(sample, "^[^TH]+")
  )


germlineLong <- germline %>%
  filter(nchar(REF) == 1, 
         nchar(ALT) == 1) %>%
  pivot_longer(
    cols = -c(CHROM, POS, REF, ALT),
    names_to = "sample",
    values_to = "ratio"
  ) %>%
  left_join(sampleInfo, by = "sample") %>%
  transmute(
    CHROM,
    POS,
    REF,
    ALT,
    tcgId,
    sampleType,
    ratio,
    sample
  )
```

## 1.2 Pair tumor/host samples
- Using shared ID's pivot wider so that the ratio for matched tumor/host samples is side by side.
- Filter on variants where alt allele is present in only the host or the tumor sample! 
```{r}
tumor <- germlineLong %>%
  filter(sampleType == "T") %>%
  rename(
    ratio_T = ratio,
    sample_T = sample
  ) %>%
  select(-sampleType)

normal <- germlineLong %>%
  filter(sampleType == "H") %>%
  rename(
    ratio_H = ratio,
    sample_H = sample
  ) %>%
  select(-sampleType)

paired <- tumor %>%
  inner_join(normal,
             by = c("CHROM", "POS", "REF", "ALT", "tcgId"))

```


## 1.2 CNVs
- Any germline variants falling within a CNV event were filtered out, so no need to adjust ploidy! 
```{r}
# germlineCnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx", 
#                      sheet = "Table S6C", 
#                      startRow = 3)
# 
# germlineCnvLong <- germlineCnv %>%
#   pivot_longer(-c(1:10),
#               names_to = "tumor", 
#               values_to = "presence") %>%
#   mutate(presence = as.numeric(presence)) %>%
#   mutate(
#     START = as.numeric(START),
#     END = as.numeric(END),
#     CHROM = as.character(CHROM),
#     tumor = as.character(tumor)
#   )
# 
# rm(germlineCnv)
# gc()
```


## 1.3 Purities
```{r}
germlinePurities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(germlinePurities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(germlinePurities),
  paste0(colnames(germlinePurities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(germlinePurities) <- newNames
germlinePurities <- germlinePurities[-1, ]
```

## 1.4 Allele counter data
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- Adjust names using metadata! 

### 1.4.1 Metadata
- Mapping AlCo results to VCF file names
- FOR now, filter where mapping = 1. There are sample names that I am certain match the name in Max's supplementary table. 
- Go back and check once workflow is done
```{r}
metadata <- read.xlsx(paste0(resDir,
                          "08_deconvTable_metadata_purityIrodsMergedManualEdit.xlsx")) %>%
  filter(mapping==1)  %>%
              dplyr::rename("tumorId"="TUMOUR.INFORMATION_TUMOUR.ID", 
                     "tumorIdRnaSeq"="rnaSeqManual")
```

### 1.4.2 DFT2
#### 1.4.2.1 Read in AlCo results
```{r}
germlineAlCo <- list.files(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport"), 
                          recursive = T)

# Read all files and combine
germlineAlCoDf <- rbindlist(lapply(germlineAlCo, function(f) {
  # Extract tumor ID between "germlineRefAltExport_" and "_realigned"
  tumorId <- sub(".*germlineRefAltExport_(.+)_realigned.*", "\\1", basename(f))
  
  # Read file
  dt <- fread(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport/", f),
              select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
  
  # Add tumor ID column
  dt[, tumorId := tumorId]
  
  return(dt)
})) 

length(unique(germlineAlCoDf$tumorId))
```
#### 1.4.2.2 Adjust tumorId of RNAseq data to match Max's metadata
- Make tumorId the metadata id.
- Save RNAseq data name.
- Adjust tumorLocationAlt
```{r}
germlineAlCoCombined <- germlineAlCoDf %>%
  filter(tumorId %in% metadata$tumorIdRnaSeq) %>%
  left_join(metadata %>%
              dplyr::select(tumorId,
                            tumorIdRnaSeq)) %>%
  mutate(tumorLocation = paste0(tumorId, "_", CHROM, ":", POS), 
         tumorLocationAlt = paste0(tumorId, "_", CHROM, ":", POS, "_", ALT))

length(unique(germlineAlCoCombined$tumorId))
```
