---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(arrow)
library(furrr)
library(patchwork)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_germline"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
source("~/gitClones/08_deconvTable/082_Scripts/genotype_germline_variants.R")
```

## 0.3 Custom functions
### 0.3.1 Compute tAlt
```{r}
computeTAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         hTotal, 
                         hAlt) {
  
    print("Solving for tAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    tAlt = ((obsVaf*denominator)-(1-purityTumor)*hAlt)/purityTumor
  return(tAlt)
}

assign_nAlt_from_vaf <- function(vaf, total_cn) {
  
  if (is.na(total_cn) || is.na(vaf)) return(NA_integer_)
  
  if (total_cn == 0) return(0)
  
  if (total_cn == 1) {
    return(ifelse(vaf < 0.5, 0, 1))
  }
  
  if (total_cn == 2) {
    if (vaf < 0.25) return(0)
    else if (vaf < 0.75) return(1)
    else return(2)
  }
  
  if (total_cn == 3) {
    if (vaf < 0.15) return(0)
    else if (vaf < 0.5) return(1)
    else if (vaf < 0.82) return(2)
    else return(3)
  }
  
  if (total_cn == 4) {
    if (vaf < 0.12) return(0)
    else if (vaf < 0.39) return(1)
    else if (vaf < 0.64) return(2)
    else if (vaf < 0.86) return(3)
    else return(4)
  }
  
  return(NA_integer_)  # fallback
}
```
### 0.3.2 Compute hAlt
```{r}
computeHAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         tAlt,
                         hTotal) {
  
  # Cases where hAlt = 0, solve for tAlt
 
    print("Solving for hAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    hAlt = (obsVaf*denominator)/(1-purityTumor)
  return(hAlt)
}

computeHAlt(obsVaf = 0.5, 
            purityTumor = 0.8, 
            tTotal = 4, 
            hTotal = 2, 
            tAlt=0)



```
### 0.3.3 Geometric mean
- Kevin's scripts
  - To calculate size factors: 
  - Compute the geom.mean of totals for each variant (optional: add pseudocount)
  - Divide each variant's read total by the geometric mean for each variant
  - Then take the median of these ratios for each sample
  - (Optional) restrict to variants with presence in all tumors
```{r}
geomean <- function(x) {
  if (all(x == 0)) {
    NA_real_
  } else {
    exp(mean(log(x[x > 0])))
  }
}
```

### 0.3.4 Summarize germline variants
```{r}
library(dplyr)
library(tidyr)

library(data.table)

summarise_variants_dt <- function(df) {
  dt <- as.data.table(df)
  
  # ensure uniqueness per tumor per location
  dt <- unique(dt[, .(locationAlt, sampleTumor, statusUpdate)])
  
  # count per location + status
  counts <- dt[, .N, by = .(locationAlt, statusUpdate)]
  
  # total tumors per location
  totals <- dt[, .(total_tumors = uniqueN(sampleTumor)), by = locationAlt]
  
  # merge counts with totals
  counts <- counts[totals, on = "locationAlt"]
  
  # compute percentage
  counts[, pct := N / total_tumors]
  
  # reshape wide (one column per statusUpdate)
  wide <- dcast(
    counts,
    locationAlt + total_tumors ~ statusUpdate,
    value.var = "pct",
    fill = 0
  )
  
  # compute majority
  status_cols <- setdiff(names(wide), c("locationAlt", "total_tumors"))
  
  wide[, majority := {
    vals <- unlist(.SD)
    max_val <- max(vals)
    winners <- names(vals)[vals == max_val]
    
    if (length(winners) > 1) "flag" else winners
  }, by = locationAlt, .SDcols = status_cols]
  
  return(wide)
}
```


# 1 Data
## 1.1 VCF files
- Read in germline vcf file
- Pivot longer so that each variant and each sample is a single row with the ALT/TOTAL ratio.
```{r}
# germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"),
#                       col_types = cols(.default = "c"))
# 
# sampleInfo <- tibble(sample = colnames(germline)[-(1:4)]) %>%
#   mutate(
#     sampleType = str_extract(sample, "T|H"),
#     tcgId = str_extract(sample, "^[^TH]+")
#   )
# 
# 
# germlineLong <- germline %>%
#   filter(nchar(REF) == 1, 
#          nchar(ALT) == 1) %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "sample",
#     values_to = "ratio"
#   ) %>%
#   left_join(sampleInfo, by = "sample") %>%
#   transmute(
#     CHROM,
#     POS,
#     REF,
#     ALT,
#     tcgId,
#     sampleType,
#     ratio,
#     sample
#   )
# 
# saveRDS(germlineLong,
#         paste0(resDir,
#                fprefix,
#                "_germlineLong.rds"))
```

## 1.2 Pair tumor/host samples
- Using shared ID's pivot wider so that the ratio for matched tumor/host samples is side by side.
- Filter on variants where alt allele is present in only the host or the tumor sample within a pair! --> so that we have info. on tAlt or hAlt at that position.
- 
```{r}
# germlineLong <- readRDS(paste0(resDir,
#                fprefix,
#                "_germlineLong.rds"))
# 
# tumorSamples <- germlineLong %>%
#   filter(sampleType == "T") %>%
#   dplyr::select(c("tcgId", "sample")) %>%
#   distinct()
# 
# 
# normalSamples <- germlineLong %>%
#   filter(sampleType == "H") %>%
#   dplyr::select(c("tcgId", "sample")) %>%
#   distinct()
# 
# pairedSamples <- tumorSamples %>%
#   inner_join(normalSamples,
#             by= c("tcgId"))
# 
# write.xlsx(pairedSamples, 
#            paste0(paste0(resDir,
#                fprefix,
#                "_germlinePaired.xlsx")))
# 
# normal <- germlineLong %>%
#   filter(sampleType == "H") %>%
#   dplyr::rename(
#     vafHost = ratio,
#     sampleHost = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# tumor <- germlineLong %>%
#   filter(sampleType == "T") %>%
#   dplyr::rename(
#     vafTumor = ratio,
#     sampleTumor = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# paired <- tumor %>%
#   inner_join(normal,
#              by = c("CHROM",
#                     "POS", "REF", "ALT", "tcgId")) 
# 
# pairedNas <- paired %>%
#   filter(is.na(vafTumor) | is.na(vafHost)) --> 0 rows!
# 
# write_dataset(paired,
#               format = "parquet",
#         paste0(resDir,
#                fprefix,
#                "_germlinePaired/"), 
#         partitioning = "tcgId")
# germlinePaired <- open_dataset(paste0(resDir,
#                                       "_germlinePaired/")) %>%
#   mutate(
#     tumorAlt = as.integer(sub("/.*", "", vafTumor)),
#     tumorTotal = as.integer(sub(".*/", "", vafTumor)),
#     hostAlt = as.integer(sub("/.*", "", vafHost)),
#     hostTotal = as.integer(sub(".*/", "", vafHost))
#   ) %>%
#   mutate(obsVafTumor = tumorAlt/tumorTotal,
#          obsVafHost = hostAlt/hostTotal)%>%
#  mutate(tumorLocationAlt = paste0(sampleTumor, "_", CHROM, ":", POS, "_", ALT))
# 
# write_dataset(germlinePaired,
#               format = "parquet",
#               paste0(resDir,
#                      "germlinePaired/"), 
#               partitioning = "tcgId")
# germlinePaired <- open_dataset(paste0(resDir,
#                                       fprefix,
#                                       "_germlinePaired/"))
```

## 1.3 Subset germline vcfs by host, tumor, or shared vaf's.
- Germline (host+tumour) variants are defined as those with VAF >0.25 in any host.
- That could mean it has 0 VAF in a given tumor sample, but this was not observed in all samples and is therefore not somatic. 
- How to account for this when calling "true" informative variants?
```{r}
# germlinePairedFiltered <- germlinePaired 
```

## 1.2 CNVs
- Any germline variants falling within a CNV event were filtered out, so no need to adjust ploidy! 
```{r}
# germlineCnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx", 
#                      sheet = "Table S6C", 
#                      startRow = 3)
# 
# germlineCnvLong <- germlineCnv %>%
#   pivot_longer(-c(1:10),
#               names_to = "tumor", 
#               values_to = "presence") %>%
#   mutate(presence = as.numeric(presence)) %>%
#   mutate(
#     START = as.numeric(START),
#     END = as.numeric(END),
#     CHROM = as.character(CHROM),
#     tumor = as.character(tumor)
#   )
# 
# rm(germlineCnv)
# gc()
```


## 1.3 Purities and ploidy
```{r}
germlinePurities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(germlinePurities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(germlinePurities),
  paste0(colnames(germlinePurities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(germlinePurities) <- newNames
germlinePurities <- germlinePurities[-1, ]

germlinePuritiesPloidy <- germlinePurities %>%
  mutate(ploidy = ifelse(X15_PLOIDY == "diploid",
                 2, 
                 ifelse(X15_PLOIDY == "tetraploid", 
                        4, 
                        "toCheck") 
                 )) %>%
  filter(ploidy != "toCheck") %>%
  dplyr::rename("sampleTumor"="TUMOUR.INFORMATION_TUMOUR ID", 
                     "purity"="X14_PURITY (VAF-based)", 
                "dftType"="X9_LINEAGE") %>%
              mutate(purity = as.numeric(purity), 
                     ploidy = as.numeric(ploidy), 
                     sex=ifelse(HOST.SEX == "♂", 
                                "male", 
                                ifelse(HOST.SEX == "♀", 
                                       "female", 
                                       "toCheck"))) %>%
              dplyr::select(sampleTumor, 
                            ploidy, 
                            purity, 
                            sex, 
                            dftType
                            ) 
  
table(germlinePuritiesPloidy$sex, germlinePuritiesPloidy$dftType)

males <- germlinePuritiesPloidy %>%
  filter(sex == "male") 

dft2 <- germlinePuritiesPloidy %>%
  filter(dftType == "DFT2") 
```

## 1.4 Allele counter data
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- Adjust names using metadata! 

### 1.4.1 Metadata
- Mapping AlCo results to VCF file names
- FOR now, don't filter where mapping = 1. There are sample names that I am certain match the name in Max's supplementary table. --> do this at the stage where we merge with AlCo results so that everything is genotyped
  - Filter on na in DFT status column, as these we likely highly host contaminated or non-DFT tumors.

- 
```{r}
metadata <- read.xlsx(paste0(resDir,
                          "08_deconvTable_metadata_purityIrodsMergedManualEdit.xlsx")) %>%
              dplyr::rename("tumorId"="TUMOUR.INFORMATION_TUMOUR.ID",
                     "tumorIdRnaSeq"="rnaSeqManual") %>%
  filter(X9_LINEAGE %in% c("DFT1", 
                           "DFT2"))

# table(metadata$X9_LINEAGE)
# 
# DFT1 DFT2 
#   78   41 

metadataPairs <- metadata %>%
  filter(HOST.INFORMATION_HOST.ID!="N/A")

# DFT1 DFT2 
#   63   37 
```

### 1.4.2 Read in AlCo results
- Here, tumorId = the name of the RNAseq file. Need to join this with the metadata table and switch out for the VCF tumor name (Max's supp data.)
- We have AlCo data from 115 tumors, calling this the tumorRnaSeqId. 
  - This can be left joined with the paired tumorId from the metadata.

```{r}
# germlineAlCo <- list.files(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport"),
#                           recursive = T)
# 
# # Read all files and combine
# germlineAlCoDf <- rbindlist(lapply(germlineAlCo, function(f) {
#   # Extract tumor ID between "germlineRefAltExport_" and "_realigned"
#   tumorIdRnaSeq <- sub(".*germlineRefAltExport_(.+)_realigned.*", "\\1", basename(f))
# 
#   # Read file
#   dt <- fread(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport/", f),
#               select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
# 
#   # Add tumor ID column
#   dt[, tumorIdRnaSeq := tumorIdRnaSeq]
# 
#   return(dt)
# }))
# 
# length(unique(germlineAlCoDf$tumorIdRnaSeq)) # 115 RNAseq samples were genotyped.
# 
# length(intersect(germlineAlCoDf$tumorIdRnaSeq, metadata$tumorIdRnaSeq)) # 53 samples for which we have a paired RNAseq-DNAseq file.
# 
# chunk_size <- 5e4  # adjust to your RAM
# 
# ## Left join metadata
# germlineAlCoDf <- germlineAlCoDf %>%
#   mutate(rowId = row_number(),
#          chunk_id = (rowId - 1) %/% chunk_size) 
# 
# write_dataset(
#   germlineAlCoDf,
#   path = paste0(resDir, "germlineAlCoDf/"),
#   format = "parquet",
#   partitioning = "chunk_id"
# )

# Perform filtering and mutate on arrow, less memory intensive.
# Rows left after filter informative samples = 58966002
# germlineAlCoDfFiltered <- open_dataset(paste0(resDir, "germlineAlCoDf/")) %>%
#   filter(tumorIdRnaSeq %in% metadata$tumorIdRnaSeq) %>%
#   left_join(metadata) %>%
#   # Tumor ID = sample ID from germline vcf.
#   mutate(tumorLocationAlt = paste0(tumorId, "_", CHROM, ":", POS, "_", ALT), 
#          sampleTumor = tumorId) %>%
#   dplyr::select(-c(tumorId))
# 
# write_dataset(
#   germlineAlCoDfFiltered,
#   path = paste0(resDir, "germlineAlCoDfFiltered/"),
#   format = "parquet", 
#   partitioning = "sampleTumor"
#   
# )
```

### 1.4.3 Adjust tumorId of RNAseq data to match Max's metadata
- Filter to keep only samples for which we have RNAseq data.
  - Filter tumorId in AlCo for those in metadata$tumorIdRnaSeq
- Left join with metadata table, to get the matched tumorId
- Create new column, tumorId_CHROM_POS_ALT
- Filter for informative AlCo positions in germlinePaired, otherwise there are too many values in the table (slow to work with).
  - 53 / 115 tumor samples have matched RNA (need to check this.)
  - 46/75 samples have an informative variant in the germline vcf file AND a matched RNAseq position. 
  - /46 have at least one variant with TOTAL AlCo > 0. 
```{r}
# setDT(germlineAlCoDf)
# 
# # fast filtering
# germlineAlCoCombined <- germlineAlCoDf[tumorIdRnaSeq %in% metadata$tumorIdRnaSeq] %>%
#   left_join(metadata %>%
#               dplyr::select(tumorId, tumorIdRnaSeq))
# 
# germlineAlCoCombined[, c("tumorLocation", "tumorLocationAlt") := .(
#   paste0(tumorId, "_", CHROM, ":", POS),
#   paste0(tumorId, "_", CHROM, ":", POS, "_", ALT)
# )]
# 
# saveRDS(germlineAlCoCombined,
#         paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombined.rds"))
# 
# germlineAlCoCombined <- readRDS(paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombined.rds"))
# 
# 
# germlineAlCoCombinedFiltered <- germlineAlCoCombined[tumorLocationAlt %in% germlinePaired$tumorLocationAlt]
# 
# 
# saveRDS(germlineAlCoCombinedFiltered,
#         paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombinedFiltered.rds"))
# 
# rm(germlineAlCoCombined,
#    germlineAlCoDf,
#    germlinePairedFiltered)
# 
# gc()
# 
# germlineAlCoCombinedFiltered <- readRDS(paste0(resDir, 
#                fprefix, 
#                "_germlineAlCoCombinedFiltered.rds")) 
```

# 3 Construct final table
## 3.1 Compute genotypes
- Adjust total CN for X and Y chromosomes in male samples
- In a pure sample, vafH = 0 + (1)*Halt/(0 + 1*Htotal). Therefore, Halt = vafH*Htotal. 
- How do deal with ploidy for X and Y in male samples in 
### 3.1.1 Test dataset
- 5000 random rows
- Compare basic implementation to Kevin's formula, which should only affect flagged cases.
```{r}
# testGenotyped <- test %>%
#   left_join(germlinePuritiesPloidy) %>%
#   filter(sampleTumor %in% metadata$tumorId) %>%
#   filter(!sampleTumor %in% c("212T",
#                             "818T1")) %>%
#   mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"), 
#                              1, 
#                              2), 
#          nTumorTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"), 
#                              ploidy/2,  # tetraploid tumors have XXYY.
#                              ploidy))  %>%
#   # Using a pure host sample (i.e., p = 0)
#   mutate(hAlt = obsVafHost * nHostTotal,
#          nHostAlt = round(hAlt),
#          nHostAltFlagged = pmin(nHostAlt, nHostTotal),
#          flag = nHostAlt > nHostTotal,
#          nHostRef = nHostTotal - nHostAltFlagged) %>%
#   
#   # Estimate tAlt
#   mutate(tAlt = computeTAlt(obsVaf = obsVafTumor,
#                             purityTumor = purity,
#                             tTotal = nTumorTotal,
#                             hTotal = nHostTotal,
#                             hAlt = nHostAltFlagged)) %>%
#   ########
#   # Round for nTumorAlt
#   mutate(nTumorAlt = round(tAlt),
#          flag = nTumorAlt > nTumorTotal, 
#          nTumorAltFlagged = pmin(nTumorAlt, nTumorTotal)) %>%
#   # No conditions where using estimate of NAlt changes nTumorAlt.
#   # filter(nTumorAltFromHAlt!=nTumorAltFromNHAlt) %>%
#   mutate(nTumorRef = nTumorTotal-nTumorAltFlagged,
#          flag = nTumorAlt > nTumorTotal) 
#   
# 
# ### Now test with Kevin's function. Run genotyper in parallel. 
# plan(multisession, workers = availableCores() - 1)
# 
# # Run genotyper in parallel — both outputs extracted in a single pass per row.
# results <- furrr::future_pmap(
#   .l = list(
#     hostPloidy = testGenotyped$nHostTotal,
#     tumorPloidy = testGenotyped$nTumorTotal,
#     hostObsAlt = testGenotyped$hostAlt,
#     hostObsTotal = testGenotyped$hostTotal,
#     tumorObsAlt = testGenotyped$tumorAlt,
#     tumorObsTotal = testGenotyped$tumorTotal,
#     purity  = testGenotyped$purity
#   ),
#   .f = function(hostPloidy,
#                 tumorPloidy,
#                 hostObsAlt,
#                 hostObsTotal,
#                 tumorObsAlt,
#                 tumorObsTotal,
#                 purity) {
#     # g <- genotyper(hp, tp, ha, ht, ta, tt, p)$E
#     g <- genotyper(host_ploidy = hostPloidy,
#                    tumour_ploidy = tumorPloidy,
#                    host_observed_alt_reads = hostObsAlt,
#                    host_observed_total_reads = hostObsTotal,
#                    tumour_observed_alt_reads = tumorObsAlt ,
#                    tumour_observed_total_reads = tumorObsTotal,
#                    tumour_purity = purity)
#     list(g_host = g$E$g_host, 
#          g_tumour = g$E$g_tumour, 
#          g=g)
#   },
#   .options = furrr_options(seed = NULL)  # no RNG used, avoids overhead
# )
# 
# testGenotyped$nHostAltGenotyper  <- vapply(results, `[[`, numeric(1), "g_host")
# testGenotyped$nTumorAltGenotyper <- vapply(results, `[[`, numeric(1), "g_tumour")
# 
# # Reset to sequential when done
# plan(sequential)
# 
# saveRDS(testGenotyped, 
#         paste0(resDir, fprefix, "_testGenotyped.rds"))
```

### 3.1.2 Full dataset
- If host sample is a male, total CN at sex chromosomes == 1
- If tumor sample is DFT2, total CN at sex chromosomes == 1 or 2 depending on original ploidy.
- Don't filter out any of the tumors! Only do this at the stage where we merge with RNAseq data.
```{r}
# germlinePairedCheckPairs <- open_dataset(paste0(resDir,
#                                       "germlinePaired/")) %>%
#   dplyr::select(sampleTumor,
#                 sampleHost) %>%
#   distinct() %>%
#   collect()
# 
# # setdiff(germlinePairedCheckPairs$sampleTumor, metadataPairs$tumorId)
# # > setdiff(germlinePairedCheckPairs$sampleTumor, metadataPairs$tumorId) --> samples that are paired, but in Max's table have no matched host?
# # [1] "212T"  "340T"  "818T1"
# 
# # > setdiff(germlinePairedCheckPairs$sampleHost, metadataPairs$HOST.INFORMATION_HOST.ID)
# # [1] "212H" "340H"
# 
# germlinePairedGenotyped <- open_dataset(paste0(resDir,
#                                       "germlinePaired/")) %>%
#   left_join(germlinePuritiesPloidy) %>%
#   filter(sampleTumor %in% metadataPairs$tumorId) %>%
#   mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"),
#                              1,
#                              2),
         nTumorTotal = ifelse(sampleTumor %in% dft2$sampleTumor & CHROM %in% c("X", "Y"),
                             ploidy/2,  # tetraploid tumors have XXYY.
                             ploidy)) %>%
  collect()
# 
# germlinePairedGenotyped$rowId <- 1:nrow(germlinePairedGenotyped)
# 
# chunk_size <- 5e4  # adjust to your RAM
# 
# germlinePairedGenotyped <- germlinePairedGenotyped %>%
#   mutate(chunk_id = (rowId - 1) %/% chunk_size)
# 
# gc()
# 
# write_dataset(
#   germlinePairedGenotyped,
#   path = paste0(resDir, "germlinePartitioned/"),
#   format = "parquet",
#   partitioning = "chunk_id"
# )
```

### 3.1.3 Run genotyper() in chunks
- Chunk size = 50,000 rows
```{r}
# ds <- open_dataset(paste0(resDir, "germlinePartitioned/"))
# 
# plan(multisession, workers = availableCores() - 1)
# 
# chunk_ids <- ds %>%
#   distinct(chunk_id) %>%
#   collect() %>%
#   pull()
# 
# for (cid in chunk_ids) {
# 
#   cat("Processing chunk", cid, "\n")
# 
#   chunk <- ds %>%
#     filter(chunk_id == cid)  %>%
#     collect()
# 
#   if (nrow(chunk) == 0) next
# 
#   results <- future_pmap(
#     .l = list(
#       chunk$nHostTotal,
#       chunk$nTumorTotal,
#       chunk$hostAlt,
#       chunk$hostTotal,
#       chunk$tumorAlt,
#       chunk$tumorTotal,
#       chunk$purity
#     ),
#     .f = function(hp, tp, ha, ht, ta, tt, p) {
# 
#       g <- genotyper(hp, tp, ha, ht, ta, tt, p)
# 
#       c(g_host = g$E$g_host,
#         g_tumour = g$E$g_tumour, 
#         map_g_host = g$map[1, "g_host"],
#         map_g_tumour = g$map[1, "g_tumour"])
#     }
#   )
# 
#   res <- do.call(rbind, results)
# 
#   chunk$nHostAltGtyr <- res[,1]
#   chunk$nHostAltGtyrRounded  <- round(res[,1])
#   chunk$nHostRefGtyrRounded <- chunk$nHostTotal - chunk$nHostAltGtyrRounded
# 
#   chunk$nTumorAltGtyr <- res[,2]
#   chunk$nTumorAltGtyrRounded  <- round(res[,2])
#   chunk$nTumorRefGtyrRounded <- chunk$nTumorTotal - chunk$nTumorAltGtyrRounded
#   
#   chunk$nHostAltGtyrMapped <- res[,3]
#   chunk$nHostRefGtyrMapped <- chunk$nHostTotal - chunk$nHostAltGtyrMapped
# 
#   chunk$nTumorAltGtyrMapped <- res[,4]
#   chunk$nTumorRefGtyrMapped <- chunk$nTumorTotal - chunk$nTumorAltGtyrMapped
# 
#   write_parquet(
#     chunk,
#     paste0(resDir, "germlineGenotyper/genotyped_chunk_", cid, ".parquet")
#   )
# 
#   rm(chunk, results, res)
#   gc()
# }
# 
# plan(sequential)
```
### 3.1.4 Assign majority genotype 
#### 3.1.3.1 Annotate
- Annotate germline status as hom alt, hom ref, or het
- Check what the majority status is at each position (e.g., are >50% of tumors hom alt?)
  - Add % column
  - Add majority assignment
```{r}
chroms <- c(paste0(seq(1,6)), "X", "Y")

germlinePairedGenotypedStatus <- open_dataset(paste0(resDir, "germlineGenotyper/")) %>%
  filter(CHROM %in% chroms) %>%
  mutate(status = ifelse(nTumorRefGtyrMapped == ploidy & nTumorAltGtyrMapped == 0, 
                "homRef", 
                ifelse(nTumorAltGtyrMapped == ploidy & nTumorRefGtyrMapped == 0, 
                       "homAlt", 
                       ifelse(nTumorRefGtyrMapped > 0 & nTumorRefGtyrMapped <  ploidy & nTumorAltGtyrMapped > 0 & nTumorAltGtyrMapped < ploidy, 
                              "het", 
                              "toCheck")))) %>%
  mutate(statusUpdate = ifelse(status == "toCheck" & nTumorRefGtyrMapped > 0 & nTumorAltGtyrMapped == 0,
                "Ref",
                ifelse(status == "toCheck" & nTumorAltGtyrMapped > 0 & nTumorRefGtyrMapped == 0,
                       "Alt",
                       status))) %>%
  mutate(locationAlt = paste0(CHROM, ":", POS, "_", REF, ALT)) 
         
# toCheck <- germlinePairedGenotypedStatus %>%
#   filter(status == "toCheck") %>%
#   collect()
# table(toCheck$CHROM, toCheck$sex)
#     female   male
#   X  47682 244403
#   Y      6     29
#   
# table(toCheck$sex, toCheck$dftType)
#         
#            DFT2
#   female  47688
#   male   244432
# 
# toCheckDft1 <- germlinePairedGenotypedStatus %>%
#   filter(dftType == "DFT1" & CHROM %in% c("X", "Y")) %>%
#   collect()
# table(toCheckDft1$status)
 #   het homAlt homRef 
 # 86173  94701 319850 
```
  
#### 3.1.3.2 DFT1
- Compute number of tumors per variant position
- Calculate, per genotype, the percentage of tumors in each state
- Paste the tumors in each state
- Determline the "majority" state if this exists.
- Save using arrow, to merge back into large genotype table 
  - Per tumorLocationAlt annotate as "majority/minority/flagged" and paste the majority or "flag" as the genotype. 
```{r}
# germlinePairedGenotypedStatusDft1 <- germlinePairedGenotypedStatus  %>%
#   filter(!sampleTumor %in% dft2$sampleTumor) %>%
#   distinct(locationAlt, sampleTumor, statusUpdate) %>%
#   group_by(locationAlt, statusUpdate) %>%
#   summarise(n = n(),
#             .groups = "drop") %>%
#   group_by(locationAlt) %>%
#   mutate(total_tumors = sum(n),
#          pct = n / total_tumors) %>%
#   collect()
# 
# tumor_map <- germlinePairedGenotypedStatusCheck %>%
#   filter(!sampleTumor %in% dft2$sampleTumor) %>%
#   distinct(locationAlt, sampleTumor, statusUpdate) %>%
#   collect() %>%
#   group_by(locationAlt, statusUpdate) %>%
#   summarise(tumors = paste(sampleTumor, collapse = ","), .groups = "drop")
# 
# germlinePairedGenotypedStatusDft1TumorAnno <- germlinePairedGenotypedStatusDft1 %>%
#   left_join(tumor_map, by = c("locationAlt", "statusUpdate")) %>%
#   group_by(locationAlt) %>%
#   mutate(
#     total_tumors = sum(n),
#     pct = n / total_tumors
#   )
# write_dataset(germlinePairedGenotypedStatusDft1TumorAnno,
#               paste0(resDir, "/germlineStatusDft1/"),
#               partitioning = "statusUpdate")
# 
# 
# germlineGenotypedMajorityAnno <- open_dataset(paste0(resDir, "/germlineStatusDft1/")) %>%
#   collect()

# ============================================================
# STEP 1: Build resultDft1 with tumor strings in wide format
# ============================================================

# dt <- as.data.table(germlineGenotypedMajorityAnno)
# 
# all_status <- c("homRef", "homAlt", "het", "Ref", "Alt", "toCheck")
# 
# resultDft1 <- dt[, {
#   pct_vec   <- setNames(rep(0,  length(all_status)), all_status)
#   tumor_vec <- setNames(rep("", length(all_status)), all_status)
#   
#   pct_vec[statusUpdate]   <- pct
#   tumor_vec[statusUpdate] <- tumors          # <-- tumor strings kept
#   
#   max_val <- max(pct_vec)
#   winners <- names(pct_vec)[pct_vec == max_val]
#   majority <- if (length(winners) > 1) "flag" else winners
#   
#   c(
#     list(total_tumors = unique(total_tumors)),
#     as.list(pct_vec),
#     as.list(setNames(tumor_vec, paste0(names(tumor_vec), "_tumors"))),
#     list(majority = majority)
#   )
# }, by = locationAlt] %>%
#   mutate(result = "dft1")


# write_parquet(resultDft1, 
#               paste0(resDir, "resultDft1.parquet"))

# # ============================================================
# # STEP 2: Build per-tumor table and join majority info
# # ============================================================
# 
# dt_full <- germlinePairedGenotypedStatus %>%
#   filter(!sampleTumor %in% dft2$sampleTumor) %>%
#   dplyr::select(locationAlt, sampleTumor, statusUpdate) %>%
#   collect() %>%
#   as.data.table()
# 
# # Join majority columns (total_tumors, pct cols, tumor string cols, majority)
# dt_full <- resultDft1[dt_full, on = "locationAlt"]
# 
# # Parse CHROM + POS
# dt_full[, c("CHROM", "POS") := {
#   tmp <- tstrsplit(locationAlt, "[:_]", perl = TRUE)
#   list(tmp[[1]], as.numeric(tmp[[2]]))
# }]
# 
# # Flag whether this tumor's genotype differs from the majority
# dt_full[, flip := fcase(
#   majority == "flag",  NA,          # ambiguous position — keep but mark NA
#   statusUpdate != majority, TRUE,
#   default = FALSE
# )]
# 
# # Convenience: label what the majority was (or "flag")
# dt_full[, majority_label := majority]
# 
# # ============================================================
# # STEP 3: Screen for hotspot regions of repeated flips
# # ============================================================
# # --- 3a. 100 kb bin-level flip counts (per tumor) ---
# dt_full[, bin := (POS %/% 100000L) * 100000L]
# 
# flip_counts <- dt_full[
#   flip == TRUE,
#   .(n_flips = .N),
#   by = .(sampleTumor, CHROM, bin)
# ]
# 
# # Fill all bins so zeros are explicit
# all_bins <- dt_full[
#   ,
#   .(bin = seq(min(bin), max(bin), by = 100000L)),
#   by = .(sampleTumor, CHROM)
# ]
# 
# flip_counts <- merge(all_bins, flip_counts,
#                      by = c("sampleTumor", "CHROM", "bin"),
#                      all.x = TRUE)
# 
# flip_counts[is.na(n_flips), n_flips := 0L]

resultDft1 <- read_parquet(paste0(resDir, "/resultDft1.parquet")) %>%
  collect()
```

#### 3.1.3.3 DFT2
```{r}
# germlinePairedGenotypedStatusDft2 <- germlinePairedGenotypedStatus  %>%
#   filter(sampleTumor %in% dft2$sampleTumor) %>%
#   distinct(locationAlt, sampleTumor, statusUpdate) %>%
#   group_by(locationAlt, statusUpdate) %>%
#   summarise(n = n(),
#             .groups = "drop") %>%
#   group_by(locationAlt) %>%
#   mutate(total_tumors = sum(n),
#          pct = n / total_tumors) %>%
#   collect()
# 
# tumor_map <- germlinePairedGenotypedStatus %>%
#   filter(sampleTumor %in% dft2$sampleTumor) %>%
#   distinct(locationAlt, sampleTumor, statusUpdate) %>%
#   collect() %>%
#   group_by(locationAlt, statusUpdate) %>%
#   summarise(tumors = paste(sampleTumor, collapse = ","), .groups = "drop")
# 
# germlinePairedGenotypedStatusDft2TumorAnno <- germlinePairedGenotypedStatusDft2 %>%
#   left_join(tumor_map, by = c("locationAlt", "statusUpdate")) %>%
#   group_by(locationAlt) %>%
#   mutate(
#     total_tumors = sum(n),
#     pct = n / total_tumors
#   )
# 
# write_dataset(germlinePairedGenotypedStatusDft2TumorAnno,
#               paste0(resDir, "/germlineStatusDft2/"),
#               partitioning = "statusUpdate")
# 
# 
# germlineGenotypedMajorityAnno <- open_dataset(paste0(resDir, "/germlineStatusDft2/")) %>%
#   collect()

# ============================================================
# STEP 1: Build resultDft2 with tumor strings in wide format
# ============================================================

# dt <- as.data.table(germlineGenotypedMajorityAnno)
# 
# all_status <- c("homRef", "homAlt", "het", "Ref", "Alt", "toCheck")
# 
# resultDft2 <- dt[, {
#   pct_vec   <- setNames(rep(0,  length(all_status)), all_status)
#   tumor_vec <- setNames(rep("", length(all_status)), all_status)
# 
#   pct_vec[statusUpdate]   <- pct
#   tumor_vec[statusUpdate] <- tumors          # <-- tumor strings kept
# 
#   max_val <- max(pct_vec)
#   winners <- names(pct_vec)[pct_vec == max_val]
#   majority <- if (length(winners) > 1) "flag" else winners
# 
#   c(
#     list(total_tumors = unique(total_tumors)),
#     as.list(pct_vec),
#     as.list(setNames(tumor_vec, paste0(names(tumor_vec), "_tumors"))),
#     list(majority = majority)
#   )
# }, by = locationAlt] %>%
#   mutate(result = "dft2")
# 
# table(resultDft2$majority)
# 
# write_parquet(resultDft2,
#               paste0(resDir, "resultDft2.parquet"))

resultDft2 <- read_parquet(paste0(resDir, "/resultDft2.parquet")) %>%
  collect()
```


#### 3.1.3.4 Merge with genotyped rows
```{r}
# ============================================================
# DFT1: Merge majority + flip status into germlinePairedGenotypedStatus
# ============================================================

# germlinePairedGenotypedStatusDft1Annotated <- germlinePairedGenotypedStatus %>%
#   filter(!sampleTumor %in% dft2$sampleTumor) %>%
#   dplyr::select(locationAlt, sampleTumor, statusUpdate) %>%
#   collect() %>%
#   as.data.table() %>%
#   # Join majority columns (total_tumors, pct cols, tumor string cols, majority)
#   merge(resultDft1[, .(locationAlt, majority, total_tumors,
#                         homRef, homAlt, het, Ref, Alt, toCheck)],  # keep only needed cols
#         by = "locationAlt",
#         all.x = TRUE) %>%
#   # Flag whether this tumor's genotype differs from the majority
#   .[, flip := fcase(
#     majority == "flag",       NA,     # ambiguous position
#     statusUpdate != majority, TRUE,
#     default =                 FALSE
#   )]

# ============================================================
# DFT2: Merge majority + flip status into germlinePairedGenotypedStatus
# ============================================================

# germlinePairedGenotypedStatusDft2Annotated <- germlinePairedGenotypedStatus %>%
#   filter(sampleTumor %in% dft2$sampleTumor) %>%
#   dplyr::select(locationAlt, sampleTumor, statusUpdate) %>%
#   collect() %>%
#   as.data.table() %>%
#   # Join majority columns
#   merge(resultDft2[, .(locationAlt, total_tumors, majority,
#                         homRef, homAlt, het, Ref, Alt, toCheck)],
#         by = "locationAlt",
#         all.x = TRUE) %>%
#   # Flag whether this tumor's genotype differs from the majority
#   .[, flip := fcase(
#     majority == "flag",       NA,
#     statusUpdate != majority, TRUE,
#     default =                 FALSE
#   )]

# ============================================================
# Combine DFT1 + DFT2 back into a single annotated table
# ============================================================

# germlinePairedGenotypedStatusAnnotated <- rbindlist(
#   list(germlinePairedGenotypedStatusDft1Annotated,
#        germlinePairedGenotypedStatusDft2Annotated),
#   use.names = TRUE
# )
#
# write_dataset(germlinePairedGenotypedStatusAnnotated,
#               paste0(resDir, "germlinePairedGenotypedStatusAnnotated/"),
#               partitioning = "sampleTumor")
# 
# 
# 
# germlinePairedGenotypedStatusAnnotated <- open_dataset(paste0(resDir, "germlinePairedGenotypedStatusAnnotated/"))
# 
# germlinePairedAnnoDft1 <- germlinePairedGenotypedStatus %>%
#   left_join(germlinePairedGenotypedStatusAnnotated) %>%
#   filter(!sampleTumor %in% dft2$sampleTumor) %>%
#   collect()
```

#### 3.1.3.4 Plot DFT1
- Create 10KB bins along entire genome
  - Count number of times a SNP has a different genotype than the majority. 
  - If a tie ("flagged") plot bar in red. 

- Plot
  - Per tumor
  - Per chromosome (all tumors)
  - All tumors all chromosomes
  
##### 3.1.3.4.1 Exclude flagged
```{r}
# dt_full <- germlinePairedGenotypedStatus  %>%
#   filter(!sampleTumor %in% dft2$sampleTumor) %>%
#   dplyr::select(locationAlt, sampleTumor, statusUpdate) %>%
#   collect()
# 
# dt_full <- as.data.table(dt_full)
# 
# # join majority
# dt_full <- resultDft1[dt_full, on = "locationAlt"]
# 
# # extract CHROM + POS
# dt_full[, c("CHROM", "POS") := {
#   tmp <- tstrsplit(locationAlt, "[:_]", perl = TRUE)
#   list(tmp[[1]], as.numeric(tmp[[2]]))
# }]
# 
# write_dataset(dt_full, 
#               paste0(resDir, "dft1DtFull/"), 
#               partitioning = "sampleTumor")

dt_full <- open_dataset(paste0(resDir, "dft1DtFull/")) %>%
  collect()

# # filter out flags early (important for speed)
# dt_full <- dt_full[majority != "flag"]
# 
# # define flipping SNPs
# dt_full[, flip := statusUpdate != majority]
# 
# # create 100kb bins
# dt_full[, bin := (POS %/% 100000) * 100000]
# 
# #---------------------------
# # 2. Count flips
# #---------------------------
# flip_counts <- dt_full[
#   flip == TRUE,
#   .(n_flips = .N),
#   by = .(sampleTumor, CHROM, bin)
# ]
# 
# #---------------------------
# # 3. Optional: fill missing bins (recommended)
# #---------------------------
# all_bins <- dt_full[
#   , .(bin = seq(min(bin), max(bin), by = 100000)),
#   by = .(sampleTumor, CHROM)
# ]
# 
# flip_counts <- merge(
#   all_bins,
#   flip_counts,
#   by = c("sampleTumor", "CHROM", "bin"),
#   all.x = TRUE
# )
# 
# flip_counts[is.na(n_flips), n_flips := 0]
# 
# 
# write_parquet(flip_counts, 
#               paste0(resDir, "dft1_flipCounts.parquet"))
```
###### 3.1.3.4.1.1 Per chromosome per tumor
```{r}
# tumors <- unique(flip_counts$sampleTumor)
# chroms <- unique(flip_counts$CHROM)
# 
# for (tumor in tumors) {
#   dir.create(paste0(plotDir, "/bar/germlineDft1/", tumor, "/"),
#              recursive = T)
# 
#   for (chr in chroms) {
# 
#     dt_plot <- flip_counts[
#       sampleTumor == tumor & CHROM == chr
#     ]
# 
#     if (nrow(dt_plot) == 0) next
# 
#     p <- ggplot(dt_plot, aes(x = bin, y = n_flips, fill = n_flips)) +
#       geom_col() +
#         scale_fill_gradient(
#       low = "lightblue",
#       high = "navy"
#     ) +
#       labs(
#         title = paste("DFT1 Tumor:", tumor, "| Chromosome:", chr),
#         x = "Genomic position (100kb bins)",
#         y = "Flipping SNP count",
#         fill = "Number of Flips"
#       ) +
#       theme_bw()
# 
#     pdf(paste0(plotDir, "/bar/germlineDft1/", tumor, "/", fprefix, "_germlineFlip_", tumor, "_chr", chr, ".pdf"),
#       width = 10,
#       height = 4
#     )
# 
#         print(p)
# 
#     dev.off()
#   }
# }
```


###### 3.1.3.4.1.2 Per chromosome
```{r}
chroms <- unique(flip_counts$CHROM)

for (chr in chroms) {

  dt_plot <- flip_counts[CHROM == chr]

  if (nrow(dt_plot) == 0) next

  p <- ggplot(dt_plot, aes(x = bin, y = n_flips, fill = n_flips)) +
    geom_col() +

    # one tumor per row
    facet_wrap(~ sampleTumor, ncol = 1, scales = "fixed") +

    scale_fill_gradient(
      low = "lightblue",
      high = "navy"
    ) +

    labs(
      title = paste("Chromosome:", chr),
      x = "Genomic position (100kb bins)",
      y = "Flipping SNP count",
      fill = "n_flips"
    ) +

    theme_bw() +
    theme(
      strip.text = element_text(size = 8),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )

  pdf(paste0(plotDir, "/bar/germlineDft1/", fprefix, "_germlineFlip_allTumors_chr", chr, ".pdf"),
      width = 10,
      height = 50
    )

        print(p)

    dev.off()

}
```

###### 3.1.3.4.1.3 All chromosomes, all tumors
```{r}
bin_size <- 100000

# Calculate per-chromosome offsets based on actual bin span
chr_offsets <- flip_counts[
  , .(min_bin = min(bin), max_bin = max(bin)),
  by = CHROM
][order(CHROM)][
  , chr_width := max_bin - min_bin + bin_size
]

# Calculate offsets manually
chr_offsets[, offset := c(0, cumsum(chr_width)[-.N])]

# Add genome-wide position to flip_counts
flip_counts <- merge(flip_counts, chr_offsets[, .(CHROM, offset, min_bin)],
                     by = "CHROM", all.x = TRUE)
flip_counts[, genome_pos := bin - min_bin + offset]  # normalise within chromosome first

# Chromosome label positions (midpoint of each chromosome)
axis_labels <- chr_offsets[
  , .(label_pos = offset + chr_width / 2, CHROM)
]

# Chromosome boundary positions (for vertical lines)
chr_boundaries <- chr_offsets[, .(offset, CHROM)]

# Plot
p <- ggplot(flip_counts, aes(x = genome_pos, y = n_flips, fill = n_flips)) +
  geom_col(width = bin_size) +
  facet_wrap(~ sampleTumor, ncol = 1, scales = "fixed") +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2) +
  scale_fill_gradient(low = "lightblue", high = "navy") +
  scale_x_continuous(
    breaks = axis_labels$label_pos,
    labels = axis_labels$CHROM
  ) +
  labs(
    title = "DFT1 - All Tumors, All Chromosomes",
    x = "Chromosome",
    y = "Flipping SNP count",
    fill = "n_flips"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

n_tumors <- length(unique(flip_counts$sampleTumor))

pdf(paste0(plotDir, "/bar/germlineDft1/", fprefix, "_germlineFlip_allTumors_allChr.pdf"),
    width = 80,
    height = n_tumors * 3)   # scale height by number of tumors
print(p)
dev.off()
# 
# p <- ggplot(flip_counts, aes(x = genome_pos, y = sampleTumor, fill = n_flips)) +
#   geom_tile(width = bin_size, height = 0.9) +
#   geom_vline(data = chr_boundaries,
#              aes(xintercept = offset),
#              color = "grey50", linewidth = 0.2) +
#   scale_fill_gradient(low = "black", high = "white") +
#   scale_x_continuous(
#     breaks = axis_labels$label_pos,
#     labels = axis_labels$CHROM
#   ) +
#   labs(
#     title = "DFT1 - All Tumors, All Chromosomes",
#     x = "Chromosome",
#     y = "Tumor",
#     fill = "n_flips"
#   ) +
#   theme_bw() +
#   theme(
#     axis.text.x = element_text(angle = 45, hjust = 1),
#     axis.text.y = element_text(size = 8)
#   )
# 
# pdf(paste0(plotDir, "/tile/germlineDft1/", fprefix, "_germlineFlip_allTumors_allChr.pdf"),
#     width = 80,
#     height = n_tumors * 2)   # scale height by number of tumors
# print(p)
# dev.off()
```


##### 3.1.3.4.2 Number of tumors with n_flips > X
###### 3.1.3.4.2.1 X = 0

```{r}
# ============================================================
# Compute: number of tumors with n_flips > 0 per bin
# ============================================================

tumors_flipping_per_bin <- flip_counts[
  n_flips > 0,
  .(n_tumors_flipping = uniqueN(sampleTumor)),
  by = .(CHROM, bin)
]

# Fill missing bins with 0
all_bins <- unique(flip_counts[, .(CHROM, bin)])
tumors_flipping_per_bin <- merge(
  all_bins,
  tumors_flipping_per_bin,
  by = c("CHROM", "bin"),
  all.x = TRUE
)
tumors_flipping_per_bin[is.na(n_tumors_flipping), n_tumors_flipping := 0]

# Add genome-wide positions
tumors_flipping_per_bin <- merge(
  tumors_flipping_per_bin,
  chr_offsets[, .(CHROM, offset, min_bin)],
  by = "CHROM", all.x = TRUE
)
tumors_flipping_per_bin[, genome_pos := bin - min_bin + offset]

global_max_tumors <- max(tumors_flipping_per_bin$n_tumors_flipping)

# ============================================================
# All chromosomes - bar plot
# ============================================================

p_bar_all <- ggplot(tumors_flipping_per_bin,
                    aes(x = genome_pos, y = n_tumors_flipping,
                        fill = n_tumors_flipping)) +
  geom_col(width = bin_size) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "lightblue", high = "navy",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  scale_y_continuous(limits = c(0, global_max_tumors)) +
  labs(
    title = "DFT1 - Tumors with n_flips > 0 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = "N tumors with flipping SNPs"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())

# ============================================================
# All chromosomes - tile plot
# ============================================================

p_tile_all <- ggplot(tumors_flipping_per_bin,
                     aes(x = genome_pos, y = 1,
                         fill = n_tumors_flipping)) +
  geom_tile(width = bin_size, height = 0.9) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "white", high = "navy",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  labs(
    title = "DFT1 - Tumors with n_flips > 0 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = NULL
  ) +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid   = element_blank())

pdf(paste0(plotDir, "/bar/germlineDft1/", fprefix,
           "_germlineFlipTumors_allChr.pdf"),
    width = 80, height = 6)
print(p_bar_all)
print(p_tile_all)
dev.off()
```

###### 3.1.3.4.2.2 X = 10

```{r}
# ============================================================
# Compute: number of tumors with n_flips > 0 per bin
# ============================================================

tumors_flipping_per_bin <- flip_counts[
  n_flips > 10,
  .(n_tumors_flipping = uniqueN(sampleTumor)),
  by = .(CHROM, bin)
]

# Fill missing bins with 0
all_bins <- unique(flip_counts[, .(CHROM, bin)])
tumors_flipping_per_bin <- merge(
  all_bins,
  tumors_flipping_per_bin,
  by = c("CHROM", "bin"),
  all.x = TRUE
)
tumors_flipping_per_bin[is.na(n_tumors_flipping), n_tumors_flipping := 0]

# Add genome-wide positions
tumors_flipping_per_bin <- merge(
  tumors_flipping_per_bin,
  chr_offsets[, .(CHROM, offset, min_bin)],
  by = "CHROM", all.x = TRUE
)
tumors_flipping_per_bin[, genome_pos := bin - min_bin + offset]

global_max_tumors <- max(tumors_flipping_per_bin$n_tumors_flipping)

# ============================================================
# All chromosomes - bar plot
# ============================================================

p_bar_all <- ggplot(tumors_flipping_per_bin,
                    aes(x = genome_pos, y = n_tumors_flipping,
                        fill = n_tumors_flipping)) +
  geom_col(width = bin_size) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "lightblue", high = "navy",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  scale_y_continuous(limits = c(0, global_max_tumors)) +
  labs(
    title = "DFT1 - Tumors with n_flips > 10 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = "N tumors with flipping SNPs"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())

# ============================================================
# All chromosomes - tile plot
# ============================================================

p_tile_all <- ggplot(tumors_flipping_per_bin,
                     aes(x = genome_pos, y = 1,
                         fill = n_tumors_flipping)) +
  geom_tile(width = bin_size, height = 0.9) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "white", high = "navy",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  labs(
    title = "DFT1 - Tumors with n_flips > 10 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = NULL
  ) +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid   = element_blank())

pdf(paste0(plotDir, "/bar/germlineDft1/", fprefix,
           "_germlineFlipTumors10_allChr.pdf"),
    width = 80, height = 6)
print(p_bar_all)
print(p_tile_all)
dev.off()
```
###### 3.1.3.4.2.2 X = 25

```{r}
# ============================================================
# Compute: number of tumors with n_flips > 0 per bin
# ============================================================

tumors_flipping_per_bin <- flip_counts[
  n_flips > 25,
  .(n_tumors_flipping = uniqueN(sampleTumor)),
  by = .(CHROM, bin)
]

# Fill missing bins with 0
all_bins <- unique(flip_counts[, .(CHROM, bin)])
tumors_flipping_per_bin <- merge(
  all_bins,
  tumors_flipping_per_bin,
  by = c("CHROM", "bin"),
  all.x = TRUE
)
tumors_flipping_per_bin[is.na(n_tumors_flipping), n_tumors_flipping := 0]

# Add genome-wide positions
tumors_flipping_per_bin <- merge(
  tumors_flipping_per_bin,
  chr_offsets[, .(CHROM, offset, min_bin)],
  by = "CHROM", all.x = TRUE
)
tumors_flipping_per_bin[, genome_pos := bin - min_bin + offset]

global_max_tumors <- max(tumors_flipping_per_bin$n_tumors_flipping)

# ============================================================
# All chromosomes - bar plot
# ============================================================

p_bar_all <- ggplot(tumors_flipping_per_bin,
                    aes(x = genome_pos, y = n_tumors_flipping,
                        fill = n_tumors_flipping)) +
  geom_col(width = bin_size) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "lightblue", high = "navy",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  scale_y_continuous(limits = c(0, global_max_tumors)) +
  labs(
    title = "DFT1 - Tumors with n_flips > 25 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = "N tumors with flipping SNPs"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())

# ============================================================
# All chromosomes - tile plot
# ============================================================

p_tile_all <- ggplot(tumors_flipping_per_bin,
                     aes(x = genome_pos, y = 1,
                         fill = n_tumors_flipping)) +
  geom_tile(width = bin_size, height = 0.9) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "white", high = "navy",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  labs(
    title = "DFT1 - Tumors with n_flips > 25 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = NULL
  ) +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid   = element_blank())

pdf(paste0(plotDir, "/bar/germlineDft1/", fprefix,
           "_germlineFlipTumors25_allChr.pdf"),
    width = 80, height = 6)
print(p_bar_all)
print(p_tile_all)
dev.off()
```

##### 3.1.3.5.4 Double check majorities in bins with > n_flips
```{r}
# ============================================================
# STEP 1: Per-position tumor flip counts
# (not binned - exact SNP positions)
# ============================================================

n_total_tumors <- uniqueN(dt_nonflag$sampleTumor)

pos_flip_check <- dt_nonflag[
  , .(
    n_tumors_flipping    = sum(flip == TRUE,  na.rm = TRUE),
    n_tumors_not_flipping = sum(flip == FALSE, na.rm = TRUE),
    n_tumors_flag        = sum(is.na(flip)),
    tumors_flipping      = paste(sampleTumor[flip == TRUE],  collapse = ","),
    tumors_not_flipping  = paste(sampleTumor[flip == FALSE], collapse = ","),
    majority_genotype    = unique(majority),
    status_observed      = paste(sort(unique(statusUpdate)), collapse = ",")
  ),
  by = .(locationAlt, CHROM, POS)
][, pct_tumors_flipping := n_tumors_flipping / n_total_tumors]

# ============================================================
# STEP 2: Flag suspicious positions
# ============================================================

# Case 1: ALL tumors flip at the same position
# This should not happen if majority is correctly assigned
all_tumors_flip <- pos_flip_check[
  n_tumors_flipping == n_total_tumors
]
cat("Positions where ALL tumors flip (possible majority misassignment):\n")
cat(nrow(all_tumors_flip), "\n\n")

# Case 2: Majority of tumors flip (>50%) - weaker signal but worth checking
majority_tumors_flip <- pos_flip_check[
  pct_tumors_flipping > 0.5
]
cat("Positions where >50% of tumors flip:\n")
cat(nrow(majority_tumors_flip), "\n\n")

# Case 3: Check what the majority genotype is at suspicious positions
# If majority was misassigned, you'd expect the majority genotype to be
# the minority statusUpdate at that position
cat("Majority genotype distribution at positions where ALL tumors flip:\n")
print(table(all_tumors_flip$majority_genotype))

cat("\nObserved status combinations at positions where ALL tumors flip:\n")
print(table(all_tumors_flip$status_observed))

# ============================================================
# STEP 3: Cross-check majority assignment directly
# Compare majority from resultDft1 vs recomputed from dt_nonflag
# ============================================================

majority_recomputed <- dt_nonflag[
  , .(
    n_tumors = uniqueN(sampleTumor)
  ),
  by = .(locationAlt, statusUpdate)
][
  , pct := n_tumors / sum(n_tumors), by = locationAlt
][
  , .SD[which.max(n_tumors)], by = locationAlt
][
  , .(locationAlt, recomputed_majority = statusUpdate, recomputed_pct = pct)
]

# Join with original majority from resultDft1
majority_comparison <- merge(
  majority_recomputed,
  resultDft1[, .(locationAlt, original_majority = majority)],
  by = "locationAlt"
)

# Find mismatches
majority_mismatches <- majority_comparison[
  recomputed_majority != original_majority
]

cat("\nMismatches between original and recomputed majority:\n")
cat(nrow(majority_mismatches), "\n")

if (nrow(majority_mismatches) > 0) {
  cat("\nMismatch breakdown:\n")
  print(table(majority_mismatches$original_majority,
              majority_mismatches$recomputed_majority))
}

# ============================================================
# STEP 4: For bins where n_tumors_flipping == n_total_tumors,
# verify flips are at DIFFERENT positions within the bin
# ============================================================

suspicious_bins <- tumors_flipping_per_bin[
  n_tumors_flipping == n_total_tumors,
  .(CHROM, bin)
]

if (nrow(suspicious_bins) > 0) {

  # Get all flipping SNPs in those bins
  suspicious_snps <- dt_nonflag[
    flip == TRUE
  ][suspicious_bins, on = .(CHROM, bin)]

  # Count how many tumors flip per position within suspicious bins
  pos_counts_in_suspicious_bins <- suspicious_snps[
    , .(n_tumors_at_pos = uniqueN(sampleTumor)),
    by = .(locationAlt, CHROM, bin, POS)
  ]

  cat("\nIn bins where all tumors flip:\n")
  cat("Max tumors flipping at a single position:",
      max(pos_counts_in_suspicious_bins$n_tumors_at_pos), "\n")
  cat("Positions where ALL tumors flip at exactly the same SNP:\n")
  print(pos_counts_in_suspicious_bins[
    n_tumors_at_pos == n_total_tumors
  ])

  # Visualise: for each suspicious bin, how many positions have
  # each number of tumors flipping?
  p_check <- ggplot(pos_counts_in_suspicious_bins,
                    aes(x = n_tumors_at_pos)) +
    geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
    facet_wrap(~ CHROM, scales = "free_y") +
    scale_x_continuous(breaks = seq(1, n_total_tumors, by = 1)) +
    labs(
      title = "Distribution of per-position tumor flip counts in suspicious bins",
      subtitle = "Bins where n_tumors_flipping == total tumors",
      x = "N tumors flipping at same position",
      y = "N positions"
    ) +
    theme_bw()

  pdf(paste0(plotDir, "/bar/germlineDft1/", fprefix,
             "_majorityCheck_suspiciousBins.pdf"),
      width = 12, height = 8)
  print(p_check)
  dev.off()
}
```
##### 3.1.3.4.3 Clear
```{r}
rm(dt_full, 
   flip_counts)

gc()
```

#### 3.1.3.5 Plot DFT2
##### 3.1.3.5.1 Exclude flagged
```{r}
# dt_full <- germlinePairedGenotypedStatus  %>%
#   filter(sampleTumor %in% dft2$sampleTumor) %>%
#   dplyr::select(locationAlt, sampleTumor, statusUpdate) %>%
#   collect()
# 
# dt_full <- as.data.table(dt_full)
# 
# # join majority
# dt_full <- resultDft2[dt_full, on = "locationAlt"]
# 
# # extract CHROM + POS
# dt_full[, c("CHROM", "POS") := {
#   tmp <- tstrsplit(locationAlt, "[:_]", perl = TRUE)
#   list(tmp[[1]], as.numeric(tmp[[2]]))
# }]

write_dataset(dt_full, 
              paste0(resDir, "dft2DtFull/"), 
              partitioning = "sampleTumor")


dt_full <- open_dataset(paste0(resDir, "dft2DtFull/")) %>%
  collect()

# filter out flags early (important for speed)
dt_full <- dt_full[majority != "flag"]

# define flipping SNPs
dt_full[, flip := statusUpdate != majority]

# create 100kb bins
dt_full[, bin := (POS %/% 100000) * 100000]

#---------------------------
# 2. Count flips
#---------------------------
flip_counts <- dt_full[
  flip == TRUE,
  .(n_flips = .N),
  by = .(sampleTumor, CHROM, bin)
]

#---------------------------
# 3. Optional: fill missing bins (recommended)
#---------------------------
all_bins <- dt_full[
  , .(bin = seq(min(bin), max(bin), by = 100000)),
  by = .(sampleTumor, CHROM)
]

flip_counts <- merge(
  all_bins,
  flip_counts,
  by = c("sampleTumor", "CHROM", "bin"),
  all.x = TRUE
)

flip_counts[is.na(n_flips), n_flips := 0]

```
###### 3.1.3.5.1.1 Per chromosome per tumor
```{r}
tumors <- unique(flip_counts$sampleTumor)
chroms <- unique(flip_counts$CHROM)

for (tumor in tumors) {
  dir.create(paste0(plotDir, "/bar/germlineDft2/", tumor, "/"),
             recursive = T)

  for (chr in chroms) {

    dt_plot <- flip_counts[
      sampleTumor == tumor & CHROM == chr
    ]

    if (nrow(dt_plot) == 0) next

    p <- ggplot(dt_plot, aes(x = bin, y = n_flips, fill = n_flips)) +
      geom_col() +
        scale_fill_gradient(
      low = "pink",
      high = "firebrick"
    ) +
      labs(
        title = paste("DFT2 Tumor:", tumor, "| Chromosome:", chr),
        x = "Genomic position (100kb bins)",
        y = "Flipping SNP count",
        fill = "Number of Flips"
      ) +
      theme_bw()

    pdf(paste0(plotDir, "/bar/germlineDft2/", tumor, "/", fprefix, "_germlineFlip_", tumor, "_chr", chr, ".pdf"),
      width = 10,
      height = 4
    )

        print(p)

    dev.off()
  }
}
```


###### 3.1.3.5.1.2 Per chromosome
```{r}
chroms <- unique(flip_counts$CHROM)

for (chr in chroms) {

  dt_plot <- flip_counts[CHROM == chr]

  if (nrow(dt_plot) == 0) next

  p <- ggplot(dt_plot, aes(x = bin, y = n_flips, fill = n_flips)) +
    geom_col() +

    # one tumor per row
    facet_wrap(~ sampleTumor, ncol = 1, scales = "fixed") +

    scale_fill_gradient(
      low = "pink",
      high = "firebrick"
    ) +

    labs(
      title = paste("Chromosome:", chr),
      x = "Genomic position (100kb bins)",
      y = "Flipping SNP count",
      fill = "n_flips"
    ) +

    theme_bw() +
    theme(
      strip.text = element_text(size = 8),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )

  pdf(paste0(plotDir, "/bar/germlineDft2/", fprefix, "_germlineFlip_allTumors_chr", chr, ".pdf"),
      width = 10,
      height = 50
    )

        print(p)

    dev.off()

}
```

###### 3.1.3.5.1.3 All chromosomes, all tumors
```{r}
bin_size <- 100000

# Calculate per-chromosome offsets based on actual bin span
chr_offsets <- flip_counts[
  , .(min_bin = min(bin), max_bin = max(bin)),
  by = CHROM
][order(CHROM)][
  , chr_width := max_bin - min_bin + bin_size
]

# Calculate offsets manually
chr_offsets[, offset := c(0, cumsum(chr_width)[-.N])]

# Add genome-wide position to flip_counts
flip_counts <- merge(flip_counts, chr_offsets[, .(CHROM, offset, min_bin)],
                     by = "CHROM", all.x = TRUE)
flip_counts[, genome_pos := bin - min_bin + offset]  # normalise within chromosome first

# Chromosome label positions (midpoint of each chromosome)
axis_labels <- chr_offsets[
  , .(label_pos = offset + chr_width / 2, CHROM)
]

# Chromosome boundary positions (for vertical lines)
chr_boundaries <- chr_offsets[, .(offset, CHROM)]

# Plot
p <- ggplot(flip_counts, aes(x = genome_pos, y = n_flips, fill = n_flips)) +
  geom_col(width = bin_size) +
  facet_wrap(~ sampleTumor, ncol = 1, scales = "fixed") +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2) +
  scale_fill_gradient(low = "pink",
      high = "firebrick") +
  scale_x_continuous(
    breaks = axis_labels$label_pos,
    labels = axis_labels$CHROM
  ) +
  labs(
    title = "DFT2 - All Tumors, All Chromosomes",
    x = "Chromosome",
    y = "Flipping SNP count",
    fill = "n_flips"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

n_tumors <- length(unique(flip_counts$sampleTumor))

pdf(paste0(plotDir, "/bar/germlineDft2/", fprefix, "_germlineFlip_allTumors_allChr.pdf"),
    width = 80,
    height = n_tumors * 3)   # scale height by number of tumors
print(p)
dev.off()

# p <- ggplot(flip_counts, aes(x = genome_pos, y = sampleTumor, fill = n_flips)) +
#   geom_tile(width = bin_size, height = 0.9) +
#   geom_vline(data = chr_boundaries,
#              aes(xintercept = offset),
#              color = "grey50", linewidth = 0.2) +
#   scale_fill_gradient(low = "black", high = "white") +
#   scale_x_continuous(
#     breaks = axis_labels$label_pos,
#     labels = axis_labels$CHROM
#   ) +
#   labs(
#     title = "DFT2 - All Tumors, All Chromosomes",
#     x = "Chromosome",
#     y = "Tumor",
#     fill = "n_flips"
#   ) +
#   theme_bw() +
#   theme(
#     axis.text.x = element_text(angle = 45, hjust = 1),
#     axis.text.y = element_text(size = 8)
#   )
# 
# pdf(paste0(plotDir, "/tile/germlineDft2/", fprefix, "_germlineFlip_allTumors_allChr.pdf"),
#     width = 80,
#     height = n_tumors * 2)   # scale height by number of tumors
# print(p)
# dev.off()
```

##### 3.1.3.5.2 Number of tumors with n_flips > X
###### 3.1.3.5.2.1 X = 0

```{r}
# ============================================================
# Compute: number of tumors with n_flips > 0 per bin
# ============================================================

tumors_flipping_per_bin <- flip_counts[
  n_flips > 0,
  .(n_tumors_flipping = uniqueN(sampleTumor)),
  by = .(CHROM, bin)
]

# Fill missing bins with 0
all_bins <- unique(flip_counts[, .(CHROM, bin)])
tumors_flipping_per_bin <- merge(
  all_bins,
  tumors_flipping_per_bin,
  by = c("CHROM", "bin"),
  all.x = TRUE
)
tumors_flipping_per_bin[is.na(n_tumors_flipping), n_tumors_flipping := 0]

# Add genome-wide positions
tumors_flipping_per_bin <- merge(
  tumors_flipping_per_bin,
  chr_offsets[, .(CHROM, offset, min_bin)],
  by = "CHROM", all.x = TRUE
)
tumors_flipping_per_bin[, genome_pos := bin - min_bin + offset]

global_max_tumors <- max(tumors_flipping_per_bin$n_tumors_flipping)

# ============================================================
# All chromosomes - bar plot
# ============================================================

p_bar_all <- ggplot(tumors_flipping_per_bin,
                    aes(x = genome_pos, y = n_tumors_flipping,
                        fill = n_tumors_flipping)) +
  geom_col(width = bin_size) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "pink", high = "firebrick",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  scale_y_continuous(limits = c(0, global_max_tumors)) +
  labs(
    title = "DFT2 - Tumors with n_flips > 0 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = "N tumors with flipping SNPs"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())

# ============================================================
# All chromosomes - tile plot
# ============================================================

p_tile_all <- ggplot(tumors_flipping_per_bin,
                     aes(x = genome_pos, y = 1,
                         fill = n_tumors_flipping)) +
  geom_tile(width = bin_size, height = 0.9) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "white", high = "firebrick",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  labs(
    title = "DFT2 - Tumors with n_flips > 0 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = NULL
  ) +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid   = element_blank())

pdf(paste0(plotDir, "/bar/germlineDft2/", fprefix,
           "_germlineFlipTumors_allChr.pdf"),
    width = 80, height = 6)
print(p_bar_all)
print(p_tile_all)
dev.off()

# ============================================================
# Per chromosome
# ============================================================

for (chr in unique(tumors_flipping_per_bin$CHROM)) {

  dt_chr <- tumors_flipping_per_bin[CHROM == chr]

  if (nrow(dt_chr) == 0) next

  p_bar_chr <- ggplot(dt_chr, aes(x = bin, y = n_tumors_flipping,
                                   fill = n_tumors_flipping)) +
    geom_col(width = bin_size) +
    scale_fill_gradient(low = "pink", high = "firebrick",
                        limits = c(0, global_max_tumors),
                        name   = "N tumors") +
    scale_x_continuous(labels = scales::comma) +
    scale_y_continuous(limits = c(0, global_max_tumors)) +
    labs(
      title = paste("DFT2 - Tumors with n_flips > 0 | Chromosome:", chr),
      x     = "Genomic position (100kb bins)",
      y     = "N tumors with flipping SNPs"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid  = element_blank())

  p_tile_chr <- ggplot(dt_chr, aes(x = bin, y = 1,
                                    fill = n_tumors_flipping)) +
    geom_tile(width = bin_size, height = 0.9) +
    scale_fill_gradient(low = "white", high = "firebrick",
                        limits = c(0, global_max_tumors),
                        name   = "N tumors") +
    scale_x_continuous(labels = scales::comma) +
    labs(
      title = paste("DFT2 - Tumors with n_flips > 0 | Chromosome:", chr),
      x     = "Genomic position (100kb bins)",
      y     = NULL
    ) +
    theme_bw() +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1),
          axis.text.y  = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid   = element_blank())

  pdf(paste0(plotDir, "/bar/germlineDft2/", fprefix,
             "_germlineFlipTumors_chr", chr, ".pdf"),
      width = 14, height = 5)
  print(p_bar_chr)
  print(p_tile_chr)
  dev.off()
}
```

###### 3.1.3.5.2.2 X = 10

```{r}
# ============================================================
# Compute: number of tumors with n_flips > 0 per bin
# ============================================================

tumors_flipping_per_bin <- flip_counts[
  n_flips > 10,
  .(n_tumors_flipping = uniqueN(sampleTumor)),
  by = .(CHROM, bin)
]

# Fill missing bins with 0
all_bins <- unique(flip_counts[, .(CHROM, bin)])
tumors_flipping_per_bin <- merge(
  all_bins,
  tumors_flipping_per_bin,
  by = c("CHROM", "bin"),
  all.x = TRUE
)
tumors_flipping_per_bin[is.na(n_tumors_flipping), n_tumors_flipping := 0]

# Add genome-wide positions
tumors_flipping_per_bin <- merge(
  tumors_flipping_per_bin,
  chr_offsets[, .(CHROM, offset, min_bin)],
  by = "CHROM", all.x = TRUE
)
tumors_flipping_per_bin[, genome_pos := bin - min_bin + offset]

global_max_tumors <- max(tumors_flipping_per_bin$n_tumors_flipping)

# ============================================================
# All chromosomes - bar plot
# ============================================================

p_bar_all <- ggplot(tumors_flipping_per_bin,
                    aes(x = genome_pos, y = n_tumors_flipping,
                        fill = n_tumors_flipping)) +
  geom_col(width = bin_size) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "pink", high = "firebrick",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  scale_y_continuous(limits = c(0, global_max_tumors)) +
  labs(
    title = "DFT2 - Tumors with n_flips > 10 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = "N tumors with flipping SNPs"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())

# ============================================================
# All chromosomes - tile plot
# ============================================================

p_tile_all <- ggplot(tumors_flipping_per_bin,
                     aes(x = genome_pos, y = 1,
                         fill = n_tumors_flipping)) +
  geom_tile(width = bin_size, height = 0.9) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "white", high = "firebrick",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  labs(
    title = "DFT2 - Tumors with n_flips > 10 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = NULL
  ) +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid   = element_blank())

pdf(paste0(plotDir, "/bar/germlineDft2/", fprefix,
           "_germlineFlipTumors10_allChr.pdf"),
    width = 80, height = 6)
print(p_bar_all)
print(p_tile_all)
dev.off()
```
###### 3.1.3.5.2.2 X = 25

```{r}
# ============================================================
# Compute: number of tumors with n_flips > 0 per bin
# ============================================================

tumors_flipping_per_bin <- flip_counts[
  n_flips > 25,
  .(n_tumors_flipping = uniqueN(sampleTumor)),
  by = .(CHROM, bin)
]

# Fill missing bins with 0
all_bins <- unique(flip_counts[, .(CHROM, bin)])
tumors_flipping_per_bin <- merge(
  all_bins,
  tumors_flipping_per_bin,
  by = c("CHROM", "bin"),
  all.x = TRUE
)
tumors_flipping_per_bin[is.na(n_tumors_flipping), n_tumors_flipping := 0]

# Add genome-wide positions
tumors_flipping_per_bin <- merge(
  tumors_flipping_per_bin,
  chr_offsets[, .(CHROM, offset, min_bin)],
  by = "CHROM", all.x = TRUE
)
tumors_flipping_per_bin[, genome_pos := bin - min_bin + offset]

global_max_tumors <- max(tumors_flipping_per_bin$n_tumors_flipping)

# ============================================================
# All chromosomes - bar plot
# ============================================================

p_bar_all <- ggplot(tumors_flipping_per_bin,
                    aes(x = genome_pos, y = n_tumors_flipping,
                        fill = n_tumors_flipping)) +
  geom_col(width = bin_size) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "pink", high = "firebrick",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  scale_y_continuous(limits = c(0, global_max_tumors)) +
  labs(
    title = "DFT2 - Tumors with n_flips > 25 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = "N tumors with flipping SNPs"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())

# ============================================================
# All chromosomes - tile plot
# ============================================================

p_tile_all <- ggplot(tumors_flipping_per_bin,
                     aes(x = genome_pos, y = 1,
                         fill = n_tumors_flipping)) +
  geom_tile(width = bin_size, height = 0.9) +
  geom_vline(data = chr_boundaries,
             aes(xintercept = offset),
             color = "grey50", linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_gradient(low = "white", high = "firebrick",
                      limits = c(0, global_max_tumors),
                      name   = "N tumors") +
  scale_x_continuous(breaks = axis_labels$label_pos,
                     labels = axis_labels$CHROM) +
  labs(
    title = "DFT2 - Tumors with n_flips > 25 per bin, All Chromosomes",
    x     = "Chromosome",
    y     = NULL
  ) +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid   = element_blank())

pdf(paste0(plotDir, "/bar/germlineDft2/", fprefix,
           "_germlineFlipTumors25_allChr.pdf"),
    width = 80, height = 6)
print(p_bar_all)
print(p_tile_all)
dev.off()
```


##### 3.1.3.5.4 Double check majorities in bins with > n_flips
```{r}
# ============================================================
# STEP 1: Per-position tumor flip counts
# (not binned - exact SNP positions)
# ============================================================

n_total_tumors <- uniqueN(dt_nonflag$sampleTumor)

pos_flip_check <- dt_nonflag[
  , .(
    n_tumors_flipping    = sum(flip == TRUE,  na.rm = TRUE),
    n_tumors_not_flipping = sum(flip == FALSE, na.rm = TRUE),
    n_tumors_flag        = sum(is.na(flip)),
    tumors_flipping      = paste(sampleTumor[flip == TRUE],  collapse = ","),
    tumors_not_flipping  = paste(sampleTumor[flip == FALSE], collapse = ","),
    majority_genotype    = unique(majority),
    status_observed      = paste(sort(unique(statusUpdate)), collapse = ",")
  ),
  by = .(locationAlt, CHROM, POS)
][, pct_tumors_flipping := n_tumors_flipping / n_total_tumors]

# ============================================================
# STEP 2: Flag suspicious positions
# ============================================================

# Case 1: ALL tumors flip at the same position
# This should not happen if majority is correctly assigned
all_tumors_flip <- pos_flip_check[
  n_tumors_flipping == n_total_tumors
]
cat("Positions where ALL tumors flip (possible majority misassignment):\n")
cat(nrow(all_tumors_flip), "\n\n")

# Case 2: Majority of tumors flip (>50%) - weaker signal but worth checking
majority_tumors_flip <- pos_flip_check[
  pct_tumors_flipping > 0.5
]
cat("Positions where >50% of tumors flip:\n")
cat(nrow(majority_tumors_flip), "\n\n")

# Case 3: Check what the majority genotype is at suspicious positions
# If majority was misassigned, you'd expect the majority genotype to be
# the minority statusUpdate at that position
cat("Majority genotype distribution at positions where ALL tumors flip:\n")
print(table(all_tumors_flip$majority_genotype))

cat("\nObserved status combinations at positions where ALL tumors flip:\n")
print(table(all_tumors_flip$status_observed))

# ============================================================
# STEP 3: Cross-check majority assignment directly
# Compare majority from resultDft1 vs recomputed from dt_nonflag
# ============================================================

majority_recomputed <- dt_nonflag[
  , .(
    n_tumors = uniqueN(sampleTumor)
  ),
  by = .(locationAlt, statusUpdate)
][
  , pct := n_tumors / sum(n_tumors), by = locationAlt
][
  , .SD[which.max(n_tumors)], by = locationAlt
][
  , .(locationAlt, recomputed_majority = statusUpdate, recomputed_pct = pct)
]

# Join with original majority from resultDft1
majority_comparison <- merge(
  majority_recomputed,
  resultDft1[, .(locationAlt, original_majority = majority)],
  by = "locationAlt"
)

# Find mismatches
majority_mismatches <- majority_comparison[
  recomputed_majority != original_majority
]

cat("\nMismatches between original and recomputed majority:\n")
cat(nrow(majority_mismatches), "\n")

if (nrow(majority_mismatches) > 0) {
  cat("\nMismatch breakdown:\n")
  print(table(majority_mismatches$original_majority,
              majority_mismatches$recomputed_majority))
}

# ============================================================
# STEP 4: For bins where n_tumors_flipping == n_total_tumors,
# verify flips are at DIFFERENT positions within the bin
# ============================================================

suspicious_bins <- tumors_flipping_per_bin[
  n_tumors_flipping == n_total_tumors,
  .(CHROM, bin)
]

if (nrow(suspicious_bins) > 0) {

  # Get all flipping SNPs in those bins
  suspicious_snps <- dt_nonflag[
    flip == TRUE
  ][suspicious_bins, on = .(CHROM, bin)]

  # Count how many tumors flip per position within suspicious bins
  pos_counts_in_suspicious_bins <- suspicious_snps[
    , .(n_tumors_at_pos = uniqueN(sampleTumor)),
    by = .(locationAlt, CHROM, bin, POS)
  ]

  cat("\nIn bins where all tumors flip:\n")
  cat("Max tumors flipping at a single position:",
      max(pos_counts_in_suspicious_bins$n_tumors_at_pos), "\n")
  cat("Positions where ALL tumors flip at exactly the same SNP:\n")
  print(pos_counts_in_suspicious_bins[
    n_tumors_at_pos == n_total_tumors
  ])

  # Visualise: for each suspicious bin, how many positions have
  # each number of tumors flipping?
  p_check <- ggplot(pos_counts_in_suspicious_bins,
                    aes(x = n_tumors_at_pos)) +
    geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
    facet_wrap(~ CHROM, scales = "free_y") +
    scale_x_continuous(breaks = seq(1, n_total_tumors, by = 1)) +
    labs(
      title = "Distribution of per-position tumor flip counts in suspicious bins",
      subtitle = "Bins where n_tumors_flipping == total tumors",
      x = "N tumors flipping at same position",
      y = "N positions"
    ) +
    theme_bw()

  pdf(paste0(plotDir, "/bar/germlineDft1/", fprefix,
             "_majorityCheck_suspiciousBins.pdf"),
      width = 12, height = 8)
  print(p_check)
  dev.off()
}
```

##### 3.1.3.5.4 Clear
```{r}
rm(dt_full, 
   flip_counts)

gc()
```
### 3.1.5 Edit genotyper results
- Based on majority status, correct genotype if deemed a flipping SNP
```{r}

```

### 3.1.6 Load genotyper results 
- Filter on conditions of interest
```{r}
germlinePairedInformative<- open_dataset(paste0(resDir, "germlineGenotyper/")) %>%
  filter(
    (nTumorAltGtyrMapped > 0 & nHostAltGtyrMapped == 0) |
    (nTumorRefGtyrMapped > 0 & nHostRefGtyrMapped == 0) |
    (nHostAltGtyrMapped > 0 & nTumorAltGtyrMapped == 0) |
    (nHostRefGtyrMapped > 0 & nTumorRefGtyrMapped == 0)
  ) 

# > nrow(germlinePairedInformative)
# [1] 21168207

germlinePairedInformative <- write_dataset(germlinePairedInformative,
                                           path = paste0(resDir,
                                                         "germlineInformative/"),
  format = "parquet",
  partitioning = "sampleTumor")


gc()

germlinePairedInformative <- open_dataset(paste0(resDir, "germlineInformative/")) %>%
  collect()
```

## 3.3 Merge tumor AlCo with germline vcf information.
- Join filtered vcf file with AlCo results from tumor RNAseq.
- Of note, NOT all samples have an associated RNA-seq file! See metadata script.
- Filtering AlCoDf for any rows where expression > 0. If an informative position has 0 RNAseq reads assigned, the expression of the informative allele is 0. 

dim(joined) --> i.e., 2 million germline variants with an informative allele for which we also have RNAseq data.
[1] 21164054       33
```{r}
germlinePairedInformativeMatched <- open_dataset(paste0(resDir,
                                                        "germlineInformative/")) %>%
  filter(sampleTumor %in% metadata$tumorId)

# 58,966,002 rows.
germlineAlCoDfFiltered <- open_dataset(paste0(resDir, "germlineAlCoDfFiltered/")) %>%
  dplyr::select(c(tumorLocationAlt,
                  tumorIdRnaSeq, 
                  TOTAL,
                  NREF, 
                  NALT)) 

joined <- germlinePairedInformativeMatched %>%
  left_join(germlineAlCoDfFiltered, 
            by = "tumorLocationAlt") %>%
  mutate(POS = as.numeric(POS))

write_dataset(joined, 
              paste0(resDir, "germlineAlCoVarsJoined/"),
              format = "parquet",
              partitioning = "sampleTumor")

# [1] 1346513
# joinedInform <- open_dataset(paste0(resDir, "germlineAlCoVarsJoined/")) %>%
#   filter(TOTAL > 0) %>%
#   collect()
# 
# nrow(joinedInform)
```

## 3.3 GTF annotations
### 3.3.1 Annotate CHROM:POS
- From informative file, get all CHROM and POS. 
- Merge with gtf file on informative positions (exons)

```{r}
# > nrow(joinedChromPos)
# [1] 1018330

joinedChromPos <- joined %>%
  dplyr::select(CHROM, POS) %>%
  distinct() %>%
  collect()

gc()

gtfFile <- "~/gitClones/01_Deconvolution/02_Deconvolution_DFT1/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf"

gtf <- import(gtfFile)

genes <- gtf[gtf$type == "gene"]

exons <- gtf[gtf$type == "exon"]

genes <- genes[genes$gene_biotype == "protein_coding"]

variants <- GRanges(
  seqnames = joinedChromPos$CHROM,
  ranges = IRanges(start = joinedChromPos$POS, 
                   end = joinedChromPos$POS)
)

# Which variants fall in an exon
exonHits <- findOverlaps(variants, exons)

# Map exon hits to gene IDs
exonGeneIds <- exons$gene_id[subjectHits(exonHits)]
exonGeneNames <- exons$gene_name[subjectHits(exonHits)]
exonBiotypes <- exons$gene_biotype[subjectHits(exonHits)]

hitsDf <- data.frame(
  query = queryHits(exonHits),
  gene_id = exonGeneIds,
  gene_name = exonGeneNames,
  biotype = exonBiotypes
)

collapseFun <- function(x) paste(unique(x), collapse = ";")

collapsed <- aggregate(. ~ query, 
                       data = hitsDf, 
                       FUN = collapseFun)

collapsed$CHROM <- joinedChromPos$CHROM[collapsed$query]
collapsed$POS   <- joinedChromPos$POS[collapsed$query]
```

### 3.3.2 Join with AlCo + Germline Info
```{r}
joinedAnno <- open_dataset(paste0(resDir, "germlineAlCoVarsJoined/")) %>%
  left_join(collapsed)

write_dataset(joinedAnno, 
              paste0(resDir, "germlineAlCoVarsJoinedAnno/"),
              format = "parquet",
              partitioning = "sampleTumor")

# Checking. All positions should have been queried, thus should return 0 rows? No, returns 20815775 rows, because exon hits only contains variant hits. 
# nrow(joinedAnnoFiltered %>% filter(is.na(query)==T))
# [1] 348,279 rows in an informative regions!
# joinedAnnoFiltered <- joinedAnno %>%
#   collect() %>%
#   filter(is.na(query)==F)
```

# 4. Normalize nRef and nAlt
## 4.1 Calculate size factors
- Of note, sample 1538T1 was not filtered out properly, we don't have matched RNAseq data for this sample. Excluding here but need to go back and filter this. 
```{r}
joinedAnnoFiltered <- open_dataset(paste0(resDir, "germlineAlCoVarsJoinedAnno/")) %>%
  mutate(locationAlt = paste0(CHROM, ":", POS, "_", REF, ALT)) %>%
  filter(is.na(query)==F) %>%
  filter(is.na(TOTAL)==F) %>%
  collect()

# nrow(joinedAnnoFiltered)
# [1] 341547
```

### 4.1.1 Calculate size factors for all variant positions
- Using TOTAL at each variant position, compute size factors per sample. 
```{r}
dt <- copy(as.data.table(joinedAnnoFiltered))

dt[, samplename := sampleTumor]

# remove variants with zero coverage everywhere
dt[, empty := sum(TOTAL, na.rm = TRUE) == 0, by = locationAlt]
variants <- dt[empty == FALSE]

# geometric means per variant
gmeans <- variants[, .(gmean = geomean(TOTAL + 1)), by = locationAlt]

setkey(gmeans, locationAlt)
setkey(variants, locationAlt)

variants[gmeans, gmean := i.gmean]

# ratios
variants[, scaled_total := TOTAL / gmean]

# size factors
size_factors <- variants[, .(
  size_factor = median(scaled_total, na.rm = TRUE)
), by = samplename]

setkey(size_factors, samplename)

dt[size_factors, on = "samplename", size_factor := i.size_factor]

dt[, `:=`(
  NREF_norm = NREF / size_factor,
  NALT_norm = NALT / size_factor,
  TOTAL_norm = TOTAL / size_factor
)]

dt <- dt %>%
  mutate(variantType = "germline")


write_dataset(dt, 
              paste0(resDir, "germlineAlCoVarsJoinedAnnoNorm/"),
              format = "parquet",
              partitioning = "sampleTumor")
```

### 4.1.2 Calculate size factors for variant positions present in all tumors only
- No variants are shared by all tumors (nrow(chosen) = 0), thus no NREF_norm_all/NALT_norm_all.
- Not compatible with DFT1/DFT2 so leaving this out for now, potentially revisit!
```{r}
# n_samples <- uniqueN(dt$samplename)
# 
# chosen <- dt[TOTAL > 0,
#              .N, by = locationAlt][N == n_samples]
# 
# variants_all <- dt[locationAlt %in% chosen$locationAlt]
# 
# gmeans_all <- variants_all[, .(gmean = geomean(TOTAL + 1)),
#                            by = locationAlt]
# 
# variants_all[gmeans_all, on = "locationAlt", gmean := i.gmean]
# variants_all[, scaled_total := TOTAL / gmean]
# 
# size_factors_all <- variants_all[, .(
#   size_factor_all = median(scaled_total, na.rm = TRUE)
# ), by = samplename]
# 
# dt[size_factors_all, on = "samplename",
#    size_factor_all := i.size_factor_all]
# 
# dt[, `:=`(
#   NREF_norm_all = NREF / size_factor_all,
#   NALT_norm_all = NALT / size_factor_all
# )]
```

## 4.2 Normalize expression by copy number of informative variant

## 4.3 Compute eT and eH


