---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
```{r}
# install.packages("slider")
```

### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(arrow)
library(furrr)
library(patchwork)
library(slider)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_germline"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
source("~/gitClones/08_deconvTable/082_Scripts/genotype_germline_variants.R")
histoDir <- paste0(plotDir, 
                   "histo/")
dir.create(histoDir)
barDir <- paste0(plotDir, 
                 "bar/")
dir.create(barDir)

resDirOut <- paste0(resDir, "/dft1GermlineVariants/")
dir.create(resDirOut)


```

## 0.3 Custom functions
### 0.3.1 Compute tAlt
```{r}
computeTAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         hTotal, 
                         hAlt) {
  
    print("Solving for tAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    tAlt = ((obsVaf*denominator)-(1-purityTumor)*hAlt)/purityTumor
  return(tAlt)
}

assign_nAlt_from_vaf <- function(vaf, total_cn) {
  
  if (is.na(total_cn) || is.na(vaf)) return(NA_integer_)
  
  if (total_cn == 0) return(0)
  
  if (total_cn == 1) {
    return(ifelse(vaf < 0.5, 0, 1))
  }
  
  if (total_cn == 2) {
    if (vaf < 0.25) return(0)
    else if (vaf < 0.75) return(1)
    else return(2)
  }
  
  if (total_cn == 3) {
    if (vaf < 0.15) return(0)
    else if (vaf < 0.5) return(1)
    else if (vaf < 0.82) return(2)
    else return(3)
  }
  
  if (total_cn == 4) {
    if (vaf < 0.12) return(0)
    else if (vaf < 0.39) return(1)
    else if (vaf < 0.64) return(2)
    else if (vaf < 0.86) return(3)
    else return(4)
  }
  
  return(NA_integer_)  # fallback
}
```
### 0.3.2 Compute hAlt
```{r}
computeHAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         tAlt,
                         hTotal) {
  
  # Cases where hAlt = 0, solve for tAlt
 
    print("Solving for hAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    hAlt = (obsVaf*denominator)/(1-purityTumor)
  return(hAlt)
}

computeHAlt(obsVaf = 0.5, 
            purityTumor = 0.8, 
            tTotal = 4, 
            hTotal = 2, 
            tAlt=0)



```
### 0.3.3 Geometric mean
- Kevin's scripts
  - To calculate size factors: 
  - Compute the geom.mean of totals for each variant (optional: add pseudocount)
  - Divide each variant's read total by the geometric mean for each variant
  - Then take the median of these ratios for each sample
  - (Optional) restrict to variants with presence in all tumors
```{r}
geomean <- function(x) {
  if (all(x == 0)) {
    NA_real_
  } else {
    exp(mean(log(x[x > 0])))
  }
}
```

### 0.3.4 Summarize germline variants
```{r}
library(dplyr)
library(tidyr)

library(data.table)

summarise_variants_dt <- function(df) {
  dt <- as.data.table(df)
  
  # ensure uniqueness per tumor per location
  dt <- unique(dt[, .(locationAlt, sampleTumor, statusUpdate)])
  
  # count per location + status
  counts <- dt[, .N, by = .(locationAlt, statusUpdate)]
  
  # total tumors per location
  totals <- dt[, .(total_tumors = uniqueN(sampleTumor)), by = locationAlt]
  
  # merge counts with totals
  counts <- counts[totals, on = "locationAlt"]
  
  # compute percentage
  counts[, pct := N / total_tumors]
  
  # reshape wide (one column per statusUpdate)
  wide <- dcast(
    counts,
    locationAlt + total_tumors ~ statusUpdate,
    value.var = "pct",
    fill = 0
  )
  
  # compute majority
  status_cols <- setdiff(names(wide), c("locationAlt", "total_tumors"))
  
  wide[, majority := {
    vals <- unlist(.SD)
    max_val <- max(vals)
    winners <- names(vals)[vals == max_val]
    
    if (length(winners) > 1) "flag" else winners
  }, by = locationAlt, .SDcols = status_cols]
  
  return(wide)
}
```

### 0.3.5 P and Q arm adjustment
```{r, warning=F}
split_location_ref <- function(location_ref, chrom_lengths_path) {
  
  chrom_lengths <- read_tsv(
    chrom_lengths_path,
    col_names = c("chrom", "length"),
  ) %>%
    dplyr::select(chrom, length)
  
  p_arm_lengths <- chrom_lengths %>%
    filter(grepl("^[0-9]+p$", chrom)) %>%
    mutate(base = sub("p$", "", chrom)) %>%
    deframe()
  
  names(p_arm_lengths) <- str_replace(names(p_arm_lengths), "p", "")
  
  chrom <- sub(":.*", "", location_ref)
  pos   <- as.numeric(sub(".*:(\\d+)_.*", "\\1", location_ref))
  ref   <- sub(".*_", "", location_ref)
  
  if (chrom %in% names(p_arm_lengths)) {
    p_len <- p_arm_lengths[[chrom]]
    if (pos <= p_len) {
      paste0(chrom, "p:", pos, "_", ref)
    } else {
      paste0(chrom, "q:", pos - p_len, "_", ref)
    }
  } else {
    location_ref
  }
}

chrom_lengths <- "~/gitClones/08_deconvTable/083_Results/08_deconvTable_sophiaArchivedSamples/dft1Germline/chrom_lengths.tsv"

split_location_ref(location_ref = "1:450172129_C", 
                   chrom_lengths_path = chrom_lengths)


# split_location_ref(location_ref = "1:56076961_G", 
#                    chrom_lengths_path = chrom_lengths) # A/T
# 1p:56076961_G
# 
# split_location_ref(location_ref = "1:57865046_G", 
#                    chrom_lengths_path = chrom_lengths) # C/T
# 1p:57865046_G
# 
# split_location_ref(location_ref = "1:58159695_C", 
#                    chrom_lengths_path = chrom_lengths) # G/T
# 
# 1p:58159695_C

# split_location_ref(location_ref = "1:593662106_T",
#                    chrom_lengths_path = chrom_lengths) # C/G
# # "1q:306958333_T"


split_location_ref(location_ref = "2:240000000_C", 
                   chrom_lengths_path = chrom_lengths)

# "2p:2.4e+08_C"

split_location_ref(location_ref = "3:135000000_A", 
                   chrom_lengths_path = chrom_lengths)

"3p:1.15e+08_A"
"3p:1.35e+08_A"

3_115000000_135000000

```

# 1 Data
## 1.1 VCF files
- Read in germline vcf file
- Pivot longer so that each variant and each sample is a single row with the ALT/TOTAL ratio.
```{r}
# germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"),
#                       col_types = cols(.default = "c"))
# 
# sampleInfo <- tibble(sample = colnames(germline)[-(1:4)]) %>%
#   mutate(
#     sampleType = str_extract(sample, "T|H"),
#     tcgId = str_extract(sample, "^[^TH]+")
#   )
# 
# 
# germlineLong <- germline %>%
#   filter(nchar(REF) == 1, 
#          nchar(ALT) == 1) %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "sample",
#     values_to = "ratio"
#   ) %>%
#   left_join(sampleInfo, by = "sample") %>%
#   transmute(
#     CHROM,
#     POS,
#     REF,
#     ALT,
#     tcgId,
#     sampleType,
#     ratio,
#     sample
#   )
# 
# saveRDS(germlineLong,
#         paste0(resDir,
#                fprefix,
#                "_germlineLong.rds"))
```

## 1.2 Pair tumor/host samples
- Using shared ID's pivot wider so that the ratio for matched tumor/host samples is side by side.
- Filter on variants where alt allele is present in only the host or the tumor sample within a pair! --> so that we have info. on tAlt or hAlt at that position.
- 
```{r}
# germlineLong <- readRDS(paste0(resDir,
#                fprefix,
#                "_germlineLong.rds"))
# 
# tumorSamples <- germlineLong %>%
#   filter(sampleType == "T") %>%
#   dplyr::select(c("tcgId", "sample")) %>%
#   distinct()
# 
# 
# normalSamples <- germlineLong %>%
#   filter(sampleType == "H") %>%
#   dplyr::select(c("tcgId", "sample")) %>%
#   distinct()
# 
# pairedSamples <- tumorSamples %>%
#   inner_join(normalSamples,
#             by= c("tcgId"))
# 
# write.xlsx(pairedSamples, 
#            paste0(paste0(resDir,
#                fprefix,
#                "_germlinePaired.xlsx")))
# 
# normal <- germlineLong %>%
#   filter(sampleType == "H") %>%
#   dplyr::rename(
#     vafHost = ratio,
#     sampleHost = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# tumor <- germlineLong %>%
#   filter(sampleType == "T") %>%
#   dplyr::rename(
#     vafTumor = ratio,
#     sampleTumor = sample
#   ) %>%
#   dplyr::select(-sampleType)
# 
# paired <- tumor %>%
#   inner_join(normal,
#              by = c("CHROM",
#                     "POS", "REF", "ALT", "tcgId")) 
# 
# pairedNas <- paired %>%
#   filter(is.na(vafTumor) | is.na(vafHost)) --> 0 rows!
# 
# write_dataset(paired,
#               format = "parquet",
#         paste0(resDir,
#                fprefix,
#                "_germlinePaired/"), 
#         partitioning = "tcgId")
# germlinePaired <- open_dataset(paste0(resDir,
#                                       "_germlinePaired/")) %>%
#   mutate(
#     tumorAlt = as.integer(sub("/.*", "", vafTumor)),
#     tumorTotal = as.integer(sub(".*/", "", vafTumor)),
#     hostAlt = as.integer(sub("/.*", "", vafHost)),
#     hostTotal = as.integer(sub(".*/", "", vafHost))
#   ) %>%
#   mutate(obsVafTumor = tumorAlt/tumorTotal,
#          obsVafHost = hostAlt/hostTotal)%>%
#  mutate(tumorLocationRef = paste0(sampleTumor, "_", CHROM, ":", POS, "_", REF))
# 
# write_dataset(germlinePaired,
#               format = "parquet",
#               paste0(resDir,
#                      "germlinePaired/"), 
#               partitioning = "tcgId")
germlinePairedTest <- open_dataset(paste0(resDir,
                                      fprefix,
                                      "_germlinePaired/")) %>%
  filter(CHROM %in% c("6") & sampleTumor %in% c("18T")) %>%
  mutate(POS=as.numeric(POS)) %>%
  filter(255000000 > POS & POS > 240000000) %>%
           collect()

germlinePairedTestChr1 <- open_dataset(paste0(resDir,
                                      fprefix,
                                      "_germlinePaired/")) %>%
  filter(CHROM %in% c("1") & sampleTumor %in% c("18T")) %>%
  mutate(POS=as.numeric(POS)) %>%
  filter(550000000 > POS & POS > 535000000) %>%
           collect()
```

## 1.3 Subset germline vcfs by host, tumor, or shared vaf's.
- Germline (host+tumour) variants are defined as those with VAF >0.25 in any host.
- That could mean it has 0 VAF in a given tumor sample, but this was not observed in all samples and is therefore not somatic. 
- How to account for this when calling "true" informative variants?
```{r}
# germlinePairedFiltered <- germlinePaired 
```

## 1.2 CNVs
- Any germline variants falling within a CNV event were filtered out, so no need to adjust ploidy! 
```{r}
# germlineCnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx", 
#                      sheet = "Table S6C", 
#                      startRow = 3)
# 
# germlineCnvLong <- germlineCnv %>%
#   pivot_longer(-c(1:10),
#               names_to = "tumor", 
#               values_to = "presence") %>%
#   mutate(presence = as.numeric(presence)) %>%
#   mutate(
#     START = as.numeric(START),
#     END = as.numeric(END),
#     CHROM = as.character(CHROM),
#     tumor = as.character(tumor)
#   )
# 
# rm(germlineCnv)
# gc()
```


## 1.3 Purities and ploidy
```{r}
germlinePurities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(germlinePurities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(germlinePurities),
  paste0(colnames(germlinePurities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(germlinePurities) <- newNames
germlinePurities <- germlinePurities[-1, ]

germlinePuritiesPloidy <- germlinePurities %>%
  mutate(ploidy = ifelse(X15_PLOIDY == "diploid",
                 2, 
                 ifelse(X15_PLOIDY == "tetraploid", 
                        4, 
                        "toCheck") 
                 )) %>%
  filter(ploidy != "toCheck") %>%
  dplyr::rename("sampleTumor"="TUMOUR.INFORMATION_TUMOUR ID", 
                     "purity"="X14_PURITY (VAF-based)", 
                "dftType"="X9_LINEAGE") %>%
              mutate(purity = as.numeric(purity), 
                     ploidy = as.numeric(ploidy), 
                     sex=ifelse(HOST.SEX == "♂", 
                                "male", 
                                ifelse(HOST.SEX == "♀", 
                                       "female", 
                                       "toCheck"))) %>%
              dplyr::select(sampleTumor, 
                            ploidy, 
                            purity, 
                            sex, 
                            dftType) 
  
table(germlinePuritiesPloidy$sex, germlinePuritiesPloidy$dftType)

males <- germlinePuritiesPloidy %>%
  filter(sex == "male") 

dft2 <- germlinePuritiesPloidy %>%
  filter(dftType == "DFT2") 
```

## 1.4 Allele counter data
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- Adjust names using metadata! 

### 1.4.1 Metadata
- Mapping AlCo results to VCF file names
- FOR now, don't filter where mapping = 1. There are sample names that I am certain match the name in Max's supplementary table. --> do this at the stage where we merge with AlCo results so that everything is genotyped
  - Filter on na in DFT status column, as these we likely highly host contaminated or non-DFT tumors.

- 
```{r}
metadata <- read.xlsx(paste0(resDir,
                          "08_deconvTable_metadata_purityIrodsMergedManualEdit.xlsx")) %>%
              dplyr::rename("tumorId"="TUMOUR.INFORMATION_TUMOUR.ID",
                     "tumorIdRnaSeq"="rnaSeqManual") %>%
  filter(X9_LINEAGE %in% c("DFT1", 
                           "DFT2"))

# table(metadata$X9_LINEAGE)
# 
# DFT1 DFT2 
#   78   41 

metadataPairs <- metadata %>%
  filter(HOST.INFORMATION_HOST.ID!="N/A")

# DFT1 DFT2 
#   63   37 
```

### 1.4.2 Read in AlCo results
- Here, tumorId = the name of the RNAseq file. Need to join this with the metadata table and switch out for the VCF tumor name (Max's supp data.)
- We have AlCo data from 115 tumors, calling this the tumorRnaSeqId. 
  - This can be left joined with the paired tumorId from the metadata.

```{r}
# germlineAlCo <- list.files(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport"),
#                           recursive = T)
# 
# # Read all files and combine
# germlineAlCoDf <- rbindlist(lapply(germlineAlCo, function(f) {
#   # Extract tumor ID between "germlineRefAltExport_" and "_realigned"
#   tumorIdRnaSeq <- sub(".*germlineRefAltExport_(.+)_realigned.*", "\\1", basename(f))
# 
#   # Read file
#   dt <- fread(paste0(resDir, "/063_Results/06_deconvTable_alCo_germlineRefAltExport/", f),
#               select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
# 
#   # Add tumor ID column
#   dt[, tumorIdRnaSeq := tumorIdRnaSeq]
# 
#   return(dt)
# }))
# 
# length(unique(germlineAlCoDf$tumorIdRnaSeq)) # 115 RNAseq samples were genotyped.
# 
# length(intersect(germlineAlCoDf$tumorIdRnaSeq, metadata$tumorIdRnaSeq)) # 53 samples for which we have a paired RNAseq-DNAseq file.
# 
# chunk_size <- 5e4  # adjust to your RAM
# 
# ## Left join metadata
# germlineAlCoDf <- germlineAlCoDf %>%
#   mutate(rowId = row_number(),
#          chunk_id = (rowId - 1) %/% chunk_size) 
# 
# write_dataset(
#   germlineAlCoDf,
#   path = paste0(resDir, "germlineAlCoDf/"),
#   format = "parquet",
#   partitioning = "chunk_id"
# )

# Perform filtering and mutate on arrow, less memory intensive.
# Rows left after filter informative samples = 58966002
# germlineAlCoDfFiltered <- open_dataset(paste0(resDir, "germlineAlCoDf/")) %>%
#   filter(tumorIdRnaSeq %in% metadata$tumorIdRnaSeq) %>%
#   left_join(metadata) %>%
#   # Tumor ID = sample ID from germline vcf.
#   mutate(tumorLocationRef = paste0(tumorId, "_", CHROM, ":", POS, "_", REF), 
#          sampleTumor = tumorId) %>%
#   dplyr::select(-c(tumorId))
# 
# write_dataset(
#   germlineAlCoDfFiltered,
#   path = paste0(resDir, "germlineAlCoDfFiltered/"),
#   format = "parquet", 
#   partitioning = "sampleTumor"
#   
# )
```

### 1.4.3 Adjust tumorId of RNAseq data to match Max's metadata
- Filter to keep only samples for which we have RNAseq data.
  - Filter tumorId in AlCo for those in metadata$tumorIdRnaSeq
- Left join with metadata table, to get the matched tumorId
- Create new column, tumorId_CHROM_POS_ALT
- Filter for informative AlCo positions in germlinePaired, otherwise there are too many values in the table (slow to work with).
  - 53 / 115 tumor samples have matched RNA (need to check this.)
  - 46/75 samples have an informative variant in the germline vcf file AND a matched RNAseq position. 
  - /46 have at least one variant with TOTAL AlCo > 0. 
```{r}
# setDT(germlineAlCoDf)
# 
# # fast filtering
# germlineAlCoCombined <- germlineAlCoDf[tumorIdRnaSeq %in% metadata$tumorIdRnaSeq] %>%
#   left_join(metadata %>%
#               dplyr::select(tumorId, tumorIdRnaSeq))
# 
# germlineAlCoCombined[, c("tumorLocation", "tumorLocationRef") := .(
#   paste0(tumorId, "_", CHROM, ":", POS),
#   paste0(tumorId, "_", CHROM, ":", POS, "_", REF)
# )]
# 
# saveRDS(germlineAlCoCombined,
#         paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombined.rds"))
# 
# germlineAlCoCombined <- readRDS(paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombined.rds"))
# 
# 
# germlineAlCoCombinedFiltered <- germlineAlCoCombined[tumorLocationRef %in% germlinePaired$tumorLocationRef]
# 
# 
# saveRDS(germlineAlCoCombinedFiltered,
#         paste0(resDir,
#                fprefix,
#                "_germlineAlCoCombinedFiltered.rds"))
# 
# rm(germlineAlCoCombined,
#    germlineAlCoDf,
#    germlinePairedFiltered)
# 
# gc()
# 
# germlineAlCoCombinedFiltered <- readRDS(paste0(resDir, 
#                fprefix, 
#                "_germlineAlCoCombinedFiltered.rds")) 
```

# 3 Construct final table
## 3.1 Compute genotypes
- Adjust total CN for X and Y chromosomes in male samples
- In a pure sample, vafH = 0 + (1)*Halt/(0 + 1*Htotal). Therefore, Halt = vafH*Htotal. 
- How do deal with ploidy for X and Y in male samples in 
### 3.1.1 Full dataset
- If host sample is a male, total CN at sex chromosomes == 1
- If tumor sample is DFT2, total CN at sex chromosomes == 1 or 2 depending on original ploidy.
- Don't filter out any of the tumors! Only do this at the stage where we merge with RNAseq data.
```{r}
# germlinePairedCheckPairs <- open_dataset(paste0(resDir,
#                                       "germlinePaired/")) %>%
#   dplyr::select(sampleTumor,
#                 sampleHost) %>%
#   distinct() %>%
#   collect()
#
# setdiff(germlinePairedCheckPairs$sampleTumor, metadataPairs$tumorId)
# > setdiff(germlinePairedCheckPairs$sampleTumor, metadataPairs$tumorId) --> samples that are paired, but in Max's table have no matched host?
# [1] "212T"  "340T"  "818T1"

# > setdiff(germlinePairedCheckPairs$sampleHost, metadataPairs$HOST.INFORMATION_HOST.ID)
# [1] "212H" "340H"
#
# germlinePairedGenotyped <- open_dataset(paste0(resDir,
#                                       "germlinePaired/")) %>%
#   left_join(germlinePuritiesPloidy) %>%
#   filter(sampleTumor %in% metadataPairs$tumorId) %>%
#   mutate(nHostTotal = ifelse(sampleTumor %in% males$sampleTumor & CHROM %in% c("X", "Y"),
#                              1,
#                              2),
#        nTumorTotal = case_when(
#                     !sampleTumor %in% dft2$sampleTumor & CHROM == "Y" ~ 0,          # DFT1 female, no Y
#                     sampleTumor %in% dft2$sampleTumor & CHROM %in% c("X", "Y") ~ ploidy / 2,  # DFT2 XXYY
#                     TRUE ~ ploidy
#                   ))
#
#
# germlinePairedGenotyped <- germlinePairedGenotyped %>%
#   collect()
#
# gc()
#
# germlinePairedGenotyped$rowId <- 1:nrow(germlinePairedGenotyped)
#
# chunk_size <- 5e4  # adjust to your RAM
#
# germlinePairedGenotyped <- germlinePairedGenotyped %>%
#   mutate(chunk_id = (rowId - 1) %/% chunk_size)
#
# gc()
#
# write_dataset(
#   germlinePairedGenotyped,
#   path = paste0(resDir, "germlinePartitioned/"),
#   format = "parquet",
#   partitioning = "chunk_id"
# )
#
#
# test <- open_dataset(paste0(resDir, "germlinePartitioned/")) %>%
#   filter(!sampleTumor %in% dft2$sampleTumor & CHROM %in% c("Y")) %>%
#   collect()
```

### 3.1.3 Run genotyper() in chunks
- Chunk size = 50,000 rows
```{r}
# ds <- open_dataset(paste0(resDir, "germlinePartitioned/"))
# 
# plan(multisession, workers = availableCores() - 1)
# 
# chunk_ids <- ds %>%
#   distinct(chunk_id) %>%
#   collect() %>%
#   pull()
# 
# for (cid in chunk_ids) {
# 
#   cat("Processing chunk", cid, "\n")
# 
#   chunk <- ds %>%
#     filter(chunk_id == cid)  %>%
#     collect()
# 
#   if (nrow(chunk) == 0) next
# 
#   results <- future_pmap(
#     .l = list(
#       chunk$nHostTotal,
#       chunk$nTumorTotal,
#       chunk$hostAlt,
#       chunk$hostTotal,
#       chunk$tumorAlt,
#       chunk$tumorTotal,
#       chunk$purity
#     ),
#     .f = function(hp, tp, ha, ht, ta, tt, p) {
# 
#       g <- genotyper(hp, tp, ha, ht, ta, tt, p)
# 
#       c(g_host = g$E$g_host,
#         g_tumour = g$E$g_tumour,
#         map_g_host = g$map[1, "g_host"],
#         map_g_tumour = g$map[1, "g_tumour"])
#     }
#   )
# 
#   res <- do.call(rbind, results)
# 
#   chunk$nHostAltGtyr <- res[,1]
#   chunk$nHostAltGtyrRounded  <- round(res[,1])
#   chunk$nHostRefGtyrRounded <- chunk$nHostTotal - chunk$nHostAltGtyrRounded
# 
#   chunk$nTumorAltGtyr <- res[,2]
#   chunk$nTumorAltGtyrRounded  <- round(res[,2])
#   chunk$nTumorRefGtyrRounded <- chunk$nTumorTotal - chunk$nTumorAltGtyrRounded
# 
#   chunk$nHostAltGtyrMapped <- res[,3]
#   chunk$nHostRefGtyrMapped <- chunk$nHostTotal - chunk$nHostAltGtyrMapped
# 
#   chunk$nTumorAltGtyrMapped <- res[,4]
#   chunk$nTumorRefGtyrMapped <- chunk$nTumorTotal - chunk$nTumorAltGtyrMapped
# 
#   write_parquet(
#     chunk,
#     paste0(resDir, "germlineGenotyper/genotyped_chunk_", cid, ".parquet")
#   )
# 
#   rm(chunk, results, res)
#   gc()
# }
# 
# plan(sequential)
```
### 3.1.4 Assign majority genotype 
#### 3.1.3.1 Annotate
- Annotate germline status as hom alt, hom ref, or het
- Check what the majority status is at each position (e.g., are >50% of tumors hom alt?)
  - Add % column
  - Add majority assignment
  - Check, how many ALT alleles 
```{r}
chroms <- c(paste0(seq(1,6)), "X")


germlinePairedGenotypedStatus <- open_dataset(paste0(resDir, "germlineGenotyper/")) %>%
  filter(CHROM %in% chroms) %>%
  mutate(locationRef = paste0(CHROM, ":", POS, "_", REF)) 

germlinePairedGenotypedStatus <- germlinePairedGenotypedStatus %>%
  filter(sampleTumor == "18T") %>%
  collect()

# schema(germlinePairedGenotypedStatus)

# Originally flagged, checking what the raw data looks like
# test <- germlinePairedGenotypedStatus %>%
#   filter(nTumorAltGtyrMapped > 2) %>%
#   collect()
# 
# nAltAlleles <- germlinePairedGenotypedStatus %>%
#   filter(!sampleTumor %in% dft2$sampleTumor) %>%
#   dplyr::select(locationRef, ALT, sampleTumor) %>%
#   distinct() %>%
#   collect() %>%
#   group_by(locationRef) %>%
#   summarise(
#     nAltAlleles = n_distinct(ALT),
#     altAlleles = paste(sort(unique(ALT)), collapse = ","),
#     baseA = ifelse("A" %in% ALT, paste(sampleTumor[ALT == "A"], collapse = ","), NA),
#     baseC = ifelse("C" %in% ALT, paste(sampleTumor[ALT == "C"], collapse = ","), NA),
#     baseG = ifelse("G" %in% ALT, paste(sampleTumor[ALT == "G"], collapse = ","), NA),
#     baseT = ifelse("T" %in% ALT, paste(sampleTumor[ALT == "T"], collapse = ","), NA),
#     .groups = "drop"
#   )
# 
# write_parquet(nAltAlleles,
#               paste0(resDir, fprefix, "_dft1Cn2_nAltAlleles.parquet"))

nAltAlleles <- read_parquet(paste0(resDir, fprefix, "_dft1Cn2_nAltAlleles.parquet"))

# Are hom ref/alt or het skewed by multiallelic sites?
biallelicData <- nAltAlleles %>%
  filter(nAltAlleles == 1) %>%
  distinct()

biallelic <- biallelicData %>%
  pull(locationRef)

# 97 sites. --> explore in IGV
multiallelicData <- nAltAlleles %>%
  filter(nAltAlleles > 1) %>%
  distinct()

multiallelic <- multiallelicData %>%
  pull(locationRef)

# Why do all tumors get repeated when a multiallelic site?. All have very low read counts supporting alt allele. Using genotyper, none of these positions exceed 0 in germline table. Do all tumors have both copies? Probably, all repeated in the summary table. 
# testTwo <- germlinePairedGenotypedStatus %>%
#   filter(locationRef %in% multiallelic) %>%
#   collect()

rm(nAltAlleles)

germlinePairedMultiallelic <- open_dataset(paste0(resDir, "germlineGenotyper/")) %>%
  filter(CHROM %in% chroms) %>%
  mutate(locationRef = paste0(CHROM, ":", POS, "_", REF)) %>%
  filter(locationRef %in% multiallelic) %>%
  collect()

write.xlsx(germlinePairedMultiallelic, 
           paste0(resDir, fprefix, "_germlinePairedMultiallelic.xlsx"))

germlinePairedMultiallelicCheck <- germlinePairedMultiallelic %>%
  filter(obsVafTumor > .20)

length(unique(germlinePairedMultiallelicCheck$locationRef))
```
  
#### 3.1.3.2 DFT1
###### ---
##### 3.1.3.2.2 CN >= 2, all chromosomes, in all tumors. 
- Autosomal
- Positions are biallelic
- Filter for positions where CN >= 2 in all tumors. 
- Compute number of tumors per variant position
- Calculate, per genotype, the percentage of tumors in each state
- Paste the tumors in each state
- Determline the "majority" state if this exists.
- Save using arrow, to merge back into large genotype table 
  - Per tumorLocationRef annotate as "majority/minority/flagged" and paste the majority or "flag" as the genotype. 
  
###### 3.1.3.2.1.1 Annotate
- Annotate each location with the majority (>= 66% status) flipping status
```{r}
# Find positions where all tumors have CN >= 2
positionsCn2All <- germlinePairedGenotypedStatus %>%
  filter(!sampleTumor %in% dft2$sampleTumor & CHROM %in% chroms & locationRef %in% biallelic) %>%
  group_by(locationRef) %>%
  summarise(minTumorCn = min(nTumorTotal, na.rm = TRUE), 
            .groups = "drop") %>%
  filter(minTumorCn >= 2) %>%
  dplyr::select(locationRef, minTumorCn) %>%
  collect()

gc()

length(unique(positionsCn2All$locationRef))

gc()

genotypeDft1Cn2All <- germlinePairedGenotypedStatus %>%
  ######
  filter(!sampleTumor %in% dft2$sampleTumor & locationRef %in% positionsCn2All$locationRef) %>%
  ######
  mutate(status = ifelse(nTumorRefGtyrMapped == nTumorTotal & nTumorAltGtyrMapped == 0, 
                "homRef", 
                ifelse(nTumorAltGtyrMapped == nTumorTotal & nTumorRefGtyrMapped == 0, 
                       "homAlt", 
                         ifelse(nTumorRefGtyrMapped > 0 & nTumorRefGtyrMapped < nTumorTotal & nTumorAltGtyrMapped > 0 & nTumorAltGtyrMapped < nTumorTotal, 
                                "het", 
                                "toCheck")))) %>%
  dplyr::select(locationRef, 
                sampleTumor, 
                status, 
                CHROM, 
                POS, 
                ploidy, 
                nTumorTotal) %>%
  collect()

majorityStatusAll <- genotypeDft1Cn2All %>%
  group_by(locationRef, status) %>%
  summarise(n = n_distinct(sampleTumor), .groups = "drop") %>%
  group_by(locationRef) %>%
  mutate(
    totalTumors = sum(n),        # total tumors with a call at this position
    maxN        = max(n),
    nTied       = sum(n == maxN),
    propMax     = maxN / totalTumors   # proportion of tumors with majority status
  ) %>%
  filter(n == maxN) %>%
  mutate(majorityStatus = case_when(
    propMax <= 0.5         ~ "flag",   # no status exceeds 50%, which would support a tie (2 or 3 way). 
    TRUE                   ~ status
  )) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::rename(majorityN = n) %>%
  dplyr::select(locationRef,
                majorityStatus,
                majorityN,
                totalTumors,
                propMax)           # keep for diagnostics

# table(majorityStatusAll$majorityStatus)
# 
# > range(majorityStatusAll$propMax)
# [1] 0.3650794 1.0000000

# Step 3: Annotate differences
genotypeDft1Cn2AnnoAll <- genotypeDft1Cn2All %>%
  # Filter on informative positions
  filter(locationRef %in% positionsCn2All$locationRef) %>%
  left_join(majorityStatusAll) %>%
  mutate(majorityCheck = status == majorityStatus) 

rm(genotypeDft1Cn2All)

# [1] 23 --> to have a majority must be greater than 63/2, i.e., 32. Why is 23 the majority? Possibly because it is > 1/3 of the samples?? Yes. Because 63/3 = 21, so the min between three genotypes is 21 samples. 
flagsAll <- genotypeDft1Cn2AnnoAll %>%
  filter(majorityStatus %in% c("flag")) %>%
  mutate(tumorLocationRef = paste0(sampleTumor, "_", locationRef)) %>%
  left_join(majorityStatusAll)

length(unique(flagsAll$locationRef))

table(flagsAll$status, 
      flagsAll$locationRef)

germlinePairedFlagged <- germlinePairedGenotypedStatus %>%
  mutate(tumorLocationRef = paste0(sampleTumor, "_", locationRef)) %>%
  filter(tumorLocationRef %in% flagsAll$tumorLocationRef) %>%
  collect()

length(unique(germlinePairedFlagged$locationRef))

write.xlsx(germlinePairedFlagged, 
           paste0(resDirOut, fprefix, "_biallelicCN2AllDft1_germlinePairedFlagged.xlsx"))
```

###### 3.1.3.2.1.2 Plot - biallelic non-flagged
```{r}
cladeOrder <- genotypeDft1Cn2AnnoAll %>%
  left_join(metadata %>% 
            dplyr::rename(sampleTumor=tumorId) %>%
              dplyr::select(sampleTumor, 
                            HOST.SEX, 
                            X10_CLADE)) %>%
  distinct(sampleTumor, X10_CLADE) %>%
  dplyr::select(sampleTumor, X10_CLADE) %>%
  arrange(X10_CLADE, sampleTumor) %>%
  collect() %>%
  pull(sampleTumor)
```

####### 3.1.3.2.1.2.1 Histograms - non flagged
- X axis = # non-majority SNPs
- What does the distribution look like? I.e., are there many positions with only 1 or 2 flipping SNPs?
```{r}
# # Count the number of tumors where status was not equal to majorityStatus
# dfPlotAll <- genotypeDft1Cn2AnnoAll %>%
#   filter(majorityCheck == F & majorityStatus!="flag" & locationRef %in% biallelic) %>%
#   # CHROM is redundant but useful for plotting.
#   group_by(locationRef,
#            CHROM) %>%
#   summarise(nTumors = n_distinct(sampleTumor), .groups = "drop")
# 
# max(dfPlotAll$nTumors) 
# 
# histoAll <- ggplot(dfPlotAll,
#                    aes(x = nTumors)) +
#   geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
#   labs(
#     x = "Number of DFT1 tumors",
#     y = "Unique non-majority positions (locationRef)",
#     title = "Biallelic Positions with CN >= 2 (All DFT1 Tumors)"
#   ) +
#   theme_minimal()
# 
# histoAllChrom <- ggplot(dfPlotAll,
#                    aes(x = nTumors, 
#                        fill = CHROM)) +
#   geom_histogram(binwidth = 1, 
#                  alpha = 0.75) +
#   labs(
#     x = "Number of DFT1 tumors",
#     y = "Unique non-majority positions (locationRef)",
#     title = "Biallelic Positions with CN >= 2 (All DFT1 Tumors)"
#   ) +
#   theme_minimal()
# 
# histoSplitChromFacet <- ggplot(dfPlotAll,
#                    aes(x = nTumors, 
#                        fill = CHROM)) +
#   geom_histogram(binwidth = 1, 
#                  alpha = 0.75) +
#   facet_wrap(~CHROM, 
#              scales = "fixed", 
#              ncol = 1) +
#   labs(
#     x = "Number of DFT1 tumors",
#     y = "Unique non-majority positions (locationRef)",
#     title = "Biallelic Positions with CN >= 2 (All DFT1 Tumors)"
#   ) +
#   theme_minimal()
# 
# 
# histoAllChromZoom <- ggplot(dfPlotAll %>% filter(nTumors > 3),
#                    aes(x = nTumors, 
#                        fill = CHROM)) +
#   geom_histogram(binwidth = 1, 
#                  alpha = 0.75) +
#   labs(
#     x = "Number of DFT1 tumors (n > 3)",
#     y = "Unique non-majority positions (locationRef)",
#     title = "Biallelic Positions with CN >= 2 (All DFT1 Tumors)"
#   ) +
#   theme_minimal()
# 
# 
# pdf(paste0(histoDir, fprefix, "_distributionFlippedSnps_biallelicCN2AllDft1.pdf"), 
#     height = 4, 
#     width = 8)
# 
# print(histoAll)
# print(histoAllChrom)
# print(histoAllChromZoom)
# 
# dev.off()
# 
# pdf(paste0(histoDir, fprefix, "_distributionFlippedSnps_biallelicCN2AllDft1_facet.pdf"), 
#     height = 10, 
#     width = 8)
# 
# print(histoSplitChromFacet)
# 
# dev.off()
# 
# write.xlsx(histoAll$data %>% as.data.frame() %>% mutate(varType = "biallelicCN2AllDft1"), 
#            paste0(resDirOut, fprefix, "_distributionFlippedSnps_biallelicCN2AllDft1.xlsx"))
```

####### 3.1.3.2.1.2.2 Sliding window - idx = snp position. 
```{r}
# Step 1: Prepare the data - flag flipping SNPs and get position
# allSNPs <- genotypeDft1Cn2AnnoAll %>%
#   filter(majorityStatus != "flag") %>%
#   mutate(POS = as.numeric(POS)) %>%
#   arrange(sampleTumor, 
#           CHROM, 
#           POS) %>%
#   left_join(metadata %>% 
#             dplyr::rename(sampleTumor=tumorId) %>%
#               dplyr::select(sampleTumor, 
#                             HOST.SEX, 
#                             X10_CLADE)) 
# 
# write_parquet(allSNPs,
#               paste0(resDir, fprefix, "_dft1Cn2_allSNPs.parquet"))
# 
# # Step 2: Sliding window per tumor per chromosome
# windowedFlips <- allSNPs %>%
#   group_by(sampleTumor, CHROM) %>%
#   arrange(POS) %>%
#   mutate(
#     # Slide index allows for irregular SNP positions (i.e., as they cluster in non-uniform manner), indexing by position rather than a set number of Mb. 
#     windowCount = slide_index_dbl(
#       majorityCheck,           # counts
#       .i = POS,                # index by position
#       .f = ~ sum(!.x),         # count majority status == F (number of flipping SNPs) per window
#       .before = 50000,         # 50kb each side = 100kb window
#       .after  = 50000
#     )) %>%
#   ungroup()
# 
# dim(windowedFlips)
# 
# write_parquet(windowedFlips,
#               paste0(resDir, fprefix, "_dft1Cn2_windowedFlips.parquet"))

allSNPS <- read.parquet(paste0(resDir, fprefix, "_dft1Cn2_allSNPs.parquet"))

windowedFlips <- read.parquet(paste0(resDir, fprefix, "_dft1Cn2_allSNPs.parquet"))

gc()

windowedFlipsFiltered <- windowedFlips %>%
  filter(windowCount > 0)

rm(windowedFlips)

dim(windowedFlipsFiltered)  
# 
# windowedFlips <- allSNPs %>%
#   group_by(sampleTumor, CHROM) %>%
#   arrange(POS) %>%
#   mutate(
#     # Slide index allows for irregular SNP positions (i.e., as they cluster in non-uniform manner), indexing by position rather than a set number of Mb. 
#     windowCount = slide_index_dbl(
#       majorityCheck,           # counts
#       .i = POS,                # index by position
#       .f = ~ sum(!.x),         # count majority status == F (number of flipping SNPs) per window
#       .before = 50000,         # 50kb each side = 100kb window
#       .after  = 50000
#     )) %>%
#   ungroup()
```

####### 3.1.3.2.1.2.3 Plot sliding window
```{r}
library(pracma)
library(zoo)

# ── 1. Extract peaks per tumor per chromosome ─────────────────────────────────
peaks <- windowedFlipsFiltered %>%
  group_by(sampleTumor, CHROM) %>%
  arrange(POS) %>%
  group_modify(~ {
    p <- findpeaks(
      .x$windowCount,
      minpeakheight   = mean(.x$windowCount, na.rm = TRUE) + 2 * sd(.x$windowCount, na.rm = TRUE),
      minpeakdistance = 10
    )
    if (is.null(p)) return(tibble())
    tibble(
      peak_idx    = p[, 2],
      windowCount = p[, 1]
    ) %>%
      mutate(POS = .x$POS[peak_idx])
  }) %>%
  ungroup()


# ── 2. Bin peaks into 500kb windows, count tumors sharing each bin ────────────
bin_size <- 500000  # adjust as needed

sharedPeaks <- peaks %>%
  mutate(bin = floor(POS / bin_size) * bin_size) %>%
  group_by(CHROM, bin) %>%
  summarise(
    n_tumors    = n_distinct(sampleTumor),
    tumor_list  = list(unique(sampleTumor)),
    mean_count  = mean(windowCount),
    .groups     = "drop"
  ) %>%
  filter(n_tumors > 1)   # only keep bins shared by 2+ tumors


# ── 3. Join clade info back onto peaks for plotting ───────────────────────────
# (assumes windowedFlips has X10_CLADE and sampleTumor columns)
cladeKey <- windowedFlipsFiltered %>%
  distinct(sampleTumor, X10_CLADE)

peaksWithClade <- peaks %>%
  left_join(cladeKey, by = "sampleTumor") %>%
  mutate(sampleTumor = factor(sampleTumor, levels = cladeOrder))

sharedBins <- sharedPeaks %>%
  mutate(
    bin_start = bin,
    bin_end   = bin + bin_size
  )

# ── 4. Plot per chromosome ────────────────────────────────────────────────────
for (chr in unique(windowedFlipsFiltered$CHROM)) {

  tempDf <- windowedFlipsFiltered %>%
    filter(CHROM == chr) %>%
    # left_join(cladeKey, by = "sampleTumor") %>%
    mutate(sampleTumor = factor(sampleTumor, 
                                levels = cladeOrder))

  tempPeaks  <- peaksWithClade  %>% filter(CHROM == chr)
  tempShared <- sharedBins      %>% filter(CHROM == chr)

  xmax <- max(tempDf$POS, na.rm = TRUE)

  tempPlotClade <- ggplot(tempDf,
      aes(x = POS, y = windowCount, colour = X10_CLADE)) +
    # ── shared-peak shading ──────────────────────────────────────────────────
    geom_rect(
      data        = tempShared,
      aes(xmin = bin_start, xmax = bin_end,
          ymin = -Inf,      ymax = Inf,
          alpha = n_tumors),
      inherit.aes = FALSE,
      fill        = "grey50"
    ) +
    scale_alpha_continuous(
      name   = "Tumors sharing peak",
      range  = c(0.10, 0.35)
    ) +
    # ── sliding-window lines ─────────────────────────────────────────────────
    geom_line(alpha = 0.6, linewidth = 0.4) +
    # ── individual peak points ───────────────────────────────────────────────
    geom_point(
      data        = tempPeaks,
      aes(x = POS, 
          y = windowCount, 
          colour = X10_CLADE),
      inherit.aes = FALSE,
      shape = 25, size = 1.2, stroke = 0.3
    ) +
    facet_grid(sampleTumor ~ CHROM, scales = "fixed") +
    labs(
      x        = "Genomic position",
      y        = "Flipping SNPs in 100kb window",
      colour   = "Clade",
      title    = "Flipping SNPs — DFT1 Tumors",
      subtitle = paste0("Chromosome: ", chr,
                        "  |  Shading = bins shared by ≥2 tumors")
    ) +
    scale_x_continuous(
  breaks = seq(0, xmax, by = 1e7),
  labels = scales::label_number(scale = 1e-3, suffix = " kb")
)+
    theme(
      legend.position  = "right",
      strip.text.x     = element_text(size = 7),
      axis.text.x      = element_text(angle = 45, hjust = 1, size = 6),
      axis.text.y      = element_text(size = 4)
    )

  pdf(
    paste0(barDir, "germlineDft1/biallelicCn2_countFlips/",
           fprefix, "_chr", chr,
           "_flippedSnps_biallelicCN2AllDft1_slidingWindow100Kb_sharedPeaks.pdf"),
    height = 40, width = 35
  )
  print(tempPlotClade)
  dev.off()
}
```

####### 3.1.3.2.1.2.4 Correct data
- Define a peak as any region where 


