---
title: "08_deconvTable_readVariants"
author: "IMEV2"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Environment
## 0.1 Packages
### 0.1.1 Library
```{r echo=F, message=F, warning=FALSE}
set.seed(123)
library(tidyverse)
library(openxlsx)
library(dplyr)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(data.table)
library(GenomicFeatures)
library(noisyr)
library(rtracklayer)
library(GenomicRanges)
```

## 0.2 Paths
```{r echo=F, message=FALSE, warning=FALSE}
wDir <- paste0("~/gitClones/08_deconvTable/")
fprefix <- "08_deconvTable_dft1Som"
dataDir <- paste0(wDir, "/081_Data/")
resDir <- paste0(wDir, "/083_Results/")
dir.create(resDir)
plotDir <- paste0(wDir, "/084_Plots/", fprefix, "/")
dir.create(plotDir)
```

## 0.3 Custom functions
### 0.3.1 Compute tAlt
```{r}
computeTAlt <- function(obsVaf, 
                         purityTumor, 
                         tTotal, 
                         hTotal, 
                         hAlt) {
  
    print("Solving for tAlt.")
    denominator = purityTumor*tTotal + (1-purityTumor)*hTotal
    tAlt = ((obsVaf*denominator)-(1-purityTumor)*hAlt)/purityTumor
  return(tAlt)
}
```

# 1 Data
## 1.1 VCF files
### 1.1.1 DFT1 somatic variants
- VCF file
```{r}
dft1Som <- read_table(paste0(dataDir, "Data_S1A_DFT1_somatic_variants.txt"), 
                      col_types = cols(.default = "c"))

dft1SomLong <- dft1Som %>%
  pivot_longer(
    cols = -c(CHROM, POS, REF, ALT),
    names_to = "tumor",
    values_to = "ratio"
  ) %>%
  mutate(chrom = as.character(CHROM), 
         location = paste0(CHROM, ":", POS), 
         position = as.numeric(POS)) %>%
  filter(nchar(REF) == 1, nchar(ALT) == 1) 

rm(dft1Som)
gc()
```

### 1.1.2 DFT2 somatic variants
```{r}
dft2Som <- read_table(paste0(dataDir, "Data_S1B_DFT2_somatic_variants.txt"),
                      col_types = cols(.default = "c"))
dft2SomLong <- dft2Som %>%
  pivot_longer(
    cols = -c(CHROM, POS, REF, ALT),
    names_to = "tumor",
    values_to = "ratio"
  ) %>%
  mutate(chrom = as.character(CHROM),
         location = paste0(CHROM, ":", POS),
         position = as.numeric(POS)) %>%
  filter(nchar(REF) == 1, nchar(ALT) == 1)

rm(dft2Som)
gc()
```

### 1.1.3 Germline variants
```{r}
# germline <- read_table(paste0(dataDir, "Data_S1D_Tasmanian_devil_germline_variants.txt"),
#                       col_types = cols(.default = "c"))
# 
# germlineLong <- head(germline) %>%
#   as.data.frame() %>%
#   pivot_longer(
#     cols = -c(CHROM, POS, REF, ALT),
#     names_to = "sample",
#     values_to = "ratio"
#   ) %>%
#   mutate(chrom = as.character(CHROM),
#          location = paste0(CHROM, ":", POS),
#          position = as.numeric(POS)) %>%
#   filter(nchar(REF) == 1, nchar(ALT) == 1)
# 
# rm(germline)
# gc()
```


## 1.2 CNVs
### 1.2.1 DFT1
```{r}
dft1Cnv <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s6.xlsx", 
                     sheet = "Table S6B", 
                     startRow = 3)

dft1CnvLong <- dft1Cnv %>%
  pivot_longer(-c(1:12),
              names_to = "tumor", 
              values_to = "presence") %>%
  mutate(presence = as.numeric(presence)) %>%
  mutate(
    START = as.numeric(START),
    END = as.numeric(END),
    CHROM = as.character(CHROM),
    tumor = as.character(tumor)
  )

rm(dft1Cnv)
gc()
```


## 1.3 Purities
```{r}
males <- germlinePuritiesPloidy %>%
  filter(sex == "male") 


dft1Purities <- read.xlsx("~/gitClones/07_suppDataPapers/suppScienceMax/science.abq6453_tables_s1_to_s8/science.abq6453_table_s2.xlsx", 
                          sheet = 1, 
                          startRow = 3)

firstRow <- as.character(dft1Purities[1, ])

# Create new column names
newNames <- ifelse(
  is.na(firstRow),
  colnames(dft1Purities),
  paste0(colnames(dft1Purities), "_", firstRow)
)

# Assign new names and remove the first row
colnames(dft1Purities) <- newNames
dft1Purities <- dft1Purities[-1, ]
```

## 1.4 Allele counter data
- Check for which tumors + hosts we have run alleleCounter (i.e., for which samples do we have RNAseq)?
- Capital R = TR
- a/b = BR
- Adjust names using metadata! 

### 1.4.1 Metadata
- Mapping AlCo results to VCF file names
- FOR now, filter where mapping = 1. There are sample names that I am certain match the name in Max's supplementary table. 
- Go back and check once workflow is done
```{r}
metadata <- read.xlsx(paste0(resDir,
                          "08_deconvTable_metadata_purityIrodsMergedManualEdit.xlsx")) %>%
  filter(mapping==1)  %>%
              dplyr::rename("tumorId"="TUMOUR.INFORMATION_TUMOUR.ID", 
                     "tumorIdRnaSeq"="rnaSeqManual")
```

### 1.4.2 DFT1
#### 1.4.2.1 Read in AlCo results
```{r}
dft1SomAlCo <- list.files(paste0(resDir, "/063_Results/06_deconvTable_alCo_dft1SomRefAltExport"), 
                          recursive = T)

# Read all files and combine
dft1SomAlCoDf <- rbindlist(lapply(dft1SomAlCo, function(f) {
  # Extract tumor ID between "dft1SomRefAltExport_" and "_realigned"
  tumorId <- sub(".*dft1SomRefAltExport_(.+)_realigned.*", "\\1", basename(f))
  
  # Read file
  dt <- fread(paste0(resDir, "/063_Results/06_deconvTable_alCo_dft1SomRefAltExport/", f),
              select = c("CHROM", "POS", "REF", "ALT", "TOTAL", "NREF", "NALT"))
  
  # Add tumor ID column
  dt[, tumorId := tumorId]
  
  return(dt)
})) 

length(unique(dft1SomAlCoDf$tumorId))
```
#### 1.4.2.2 Adjust tumorId of RNAseq data to match Max's metadata
- Make tumorId the metadata id.
- Save RNAseq data name.
- Adjust tumorLocationAlt
```{r}
dft1SomAlCoCombined <- dft1SomAlCoDf %>%
  filter(tumorId %in% metadata$tumorIdRnaSeq) %>%
  left_join(metadata %>%
              dplyr::select(tumorId,
                            tumorIdRnaSeq)) %>%
  mutate(tumorLocation = paste0(tumorId, "_", CHROM, ":", POS), 
         tumorLocationAlt = paste0(tumorId, "_", CHROM, ":", POS, "_", ALT))

length(unique(dft1SomAlCoCombined$tumorId))
```

# 2 Filter data
- Only variants for which the alternate allele is found exclusively in one of the samples (host or tumor), in at least one host-tumor pair.
- The same variant in each single host-tumor pair is considered independent (independent CNV + tumor purity value associated)
#===
<!-- ## 2.1 Toy example - 46T (diploid) -->
<!-- - Checking DFT1 mutations in 18T -->
<!-- - dft1Som --> By default the SNVs in dft1Som file are not present in any of the hosts -->
<!-- - germline --> filter where alternative allele is exclusive to host or tumor in a matched pair -->
<!-- - for tetraploid!!! --> trickier because we have pre and post tetraploidization CNVs?? -->
<!-- - to check, use vaf based purity or cnv based purity? -->
<!-- - to check, is the presence/absence the degree of cnv too? only had 0/1 so can't see if loss both copies?? -->

<!-- - if a variant falls in a region where cnv changes with presence = 1 multiple times, how do I account for this? --> sum up gains + losses! -->
<!-- - how to account for sex chromosomes? assume diploid for tumor, except in sex chromosomes! Y -->
<!-- ```{r} -->
<!-- toy <- "46T1" -->
<!-- ``` -->

<!-- ### 2.1.1 VCF + Purities + Ploidy -->
<!-- - Merge info on purity and ploidy from metadata -->
<!-- - Create seperate columns for Alt/Total -->
<!-- - Filter out Alt > 0, this variant was not present in the tumor (reduces the size of the table). -->
<!-- ```{r} -->
<!-- dft1PuritiesPloidy <- dft1Purities %>% -->
<!--   filter(`TUMOUR.INFORMATION_TUMOUR ID`==toy) %>% -->
<!--   mutate(ploidy = ifelse(X15_PLOIDY == "diploid", -->
<!--                  2,  -->
<!--                  ifelse(X15_PLOIDY == "tetraploid",  -->
<!--                         4,  -->
<!--                         "toCheck"))) %>% -->
<!--   filter(ploidy != "toCheck") -->


<!-- dft1SomLongVaf <- dft1SomLong %>% -->
<!--   filter(tumor==toy) %>% -->
<!--   mutate( -->
<!--     Alt = as.integer(sub("/.*", "", ratio)), -->
<!--     Total = as.integer(sub(".*/", "", ratio)) -->
<!--   ) %>% -->

<!--   # If nAlt = 0, this was a variant in a different tumor sample, so filter out any 0 vafs! -->
<!--   filter(Alt > 0) %>% -->
<!--   left_join(dft1PuritiesPloidy %>% -->
<!--               dplyr::rename("tumor"="TUMOUR.INFORMATION_TUMOUR ID",  -->
<!--                      "purity"="X14_PURITY (VAF-based)") %>% -->
<!--               mutate(purity = as.numeric(purity)) %>% -->
<!--               dplyr::select(tumor,  -->
<!--                             ploidy,  -->
<!--                             purity)) %>% -->
<!--   filter(tumor %in% dft1PuritiesPloidy$`TUMOUR.INFORMATION_TUMOUR ID`) %>% -->
<!--   mutate( -->
<!--     position = as.numeric(position), -->
<!--     CHROM = as.character(CHROM), -->
<!--     tumor = as.character(tumor) -->
<!--   ) -->
<!-- ``` -->

<!-- ### 2.1.2 Adjust ploidy at variant using CNV status.  -->
<!-- - Per variant per tumor, find all CNV events spanning that position -->
<!-- - This means that if you filter for presence > 0, there is a change in tumor CN.  -->
<!-- - For some variants, there may be more than one change in CN, to check which these are, filter for the rows where this is the case.  -->
<!-- ```{r} -->
<!-- cnvJoin <- dft1SomLongVaf %>% -->
<!--   left_join( -->
<!--     dft1CnvLong, -->
<!--     by = join_by( -->
<!--       CHROM, -->
<!--       tumor, -->
<!--       position >= START, -->
<!--       position <= END -->
<!--     ) -->
<!--   ) %>% -->
<!--   filter(presence > 0) %>% -->
<!--   distinct() %>% -->
<!--   mutate(tumorLocation = paste0(tumor, "_", location)) -->

<!-- check <- which(table(cnvJoin$tumorLocation)>1) -->

<!-- cnvJoinMultiplets <- cnvJoin %>% -->
<!--   mutate(multiplets = ifelse(tumorLocation %in% names(check),  -->
<!--                 "1",  -->
<!--                 "0"))  -->

<!-- write.xlsx(cnvJoinMultiplets,  -->
<!--            paste0(resDir, fprefix, "_cnvJoinMultiplets_46T.xlsx")) -->


<!-- ### Double checking ### -->
<!-- dft1CnvToy <- dft1Cnv %>% -->
<!--   dplyr::select(1:12, toy) %>% -->
<!--   filter(CNV.EVENT.ID %in% (cnvJoinMultiplets %>%  -->
<!--            filter(multiplets == 1) %>%  -->
<!--            pull(CNV.EVENT.ID))) -->
<!-- ``` -->












<!-- #=== -->
<!-- ## 2.2 Toy example.- 18T (tetraploid) -->
<!-- ```{r} -->
<!-- toy <- "18T" -->
<!-- ``` -->
<!-- ### 2.2.1 VCF + Purities + Ploidy -->
<!-- - Merge info on purity and ploidy from metadata -->
<!-- - Create seperate columns for Alt/Total -->
<!-- - Filter out Alt > 0, this variant was not present in the tumor (reduces the size of the table). -->
<!-- ```{r} -->
<!-- dft1PuritiesPloidy <- dft1Purities %>% -->
<!--   filter(`TUMOUR.INFORMATION_TUMOUR ID`==toy) %>% -->
<!--   mutate(ploidy = ifelse(X15_PLOIDY == "diploid", -->
<!--                  2,  -->
<!--                  ifelse(X15_PLOIDY == "tetraploid",  -->
<!--                         4,  -->
<!--                         "toCheck"))) %>% -->
<!--   filter(ploidy != "toCheck") -->


<!-- dft1SomLongVaf <- dft1SomLong %>% -->
<!--   filter(tumor==toy) %>% -->
<!--   mutate( -->
<!--     Alt = as.integer(sub("/.*", "", ratio)), -->
<!--     Total = as.integer(sub(".*/", "", ratio)) -->
<!--   ) %>% -->

<!--   # If nAlt = 0, this was a variant in a different tumor sample, so filter out any 0 vafs! -->
<!--   filter(Alt > 0) %>% -->
<!--   left_join(dft1PuritiesPloidy %>% -->
<!--               dplyr::rename("tumor"="TUMOUR.INFORMATION_TUMOUR ID",  -->
<!--                      "purity"="X14_PURITY (VAF-based)") %>% -->
<!--               mutate(purity = as.numeric(purity)) %>% -->
<!--               dplyr::select(tumor,  -->
<!--                             ploidy,  -->
<!--                             purity)) %>% -->
<!--   filter(tumor %in% dft1PuritiesPloidy$`TUMOUR.INFORMATION_TUMOUR ID`) %>% -->
<!--   mutate( -->
<!--     position = as.numeric(position), -->
<!--     CHROM = as.character(CHROM), -->
<!--     tumor = as.character(tumor) -->
<!--   ) -->
<!-- ``` -->

<!-- ### 2.2.2 Adjust ploidy at variant using CNV status.  -->
<!-- - Per variant per tumor, find all CNV events spanning that position -->
<!-- - This means that if you filter for presence > 0, there is a change in tumor CN.  -->
<!-- - For some variants, there may be more than one change in CN, to check which these are, filter for the rows where this is the case.  -->
<!-- ```{r} -->
<!-- cnvJoin <- dft1SomLongVaf %>% -->
<!--   left_join( -->
<!--     dft1CnvLong, -->
<!--     by = join_by( -->
<!--       CHROM, -->
<!--       tumor, -->
<!--       position >= START, -->
<!--       position <= END -->
<!--     ) -->
<!--   ) %>% -->
<!--   filter(presence > 0) %>% -->
<!--   distinct() %>% -->
<!--   mutate(tumorLocation = paste0(tumor, "_", location)) -->

<!-- check <- which(table(cnvJoin$tumorLocation)>1) -->

<!-- cnvJoinMultiplets <- cnvJoin %>% -->
<!--   mutate(multiplets = ifelse(tumorLocation %in% names(check),  -->
<!--                 "1",  -->
<!--                 "0"))  -->

<!-- write.xlsx(cnvJoinMultiplets,  -->
<!--            paste0(resDir,  -->
<!--                   fprefix,  -->
<!--                   "_cnvJoinMultiplets_18T.xlsx")) -->

<!-- ### Double checking ### -->
<!-- dft1CnvToy <- dft1Cnv %>% -->
<!--   dplyr::select(1:12, toy) %>% -->
<!--   filter(CNV.EVENT.ID %in% (cnvJoinMultiplets %>%  -->
<!--            filter(multiplets == 1) %>%  -->
<!--            pull(CNV.EVENT.ID))) -->
<!-- ``` -->
# Uncomment for toy example.
#===
## 2.1 VCF + Purities + Ploidy
- Merge info on purity and ploidy from metadata
- Create seperate columns for Alt/Total
- Filter out Alt > 0, this variant was not present in the tumor (reduces the size of the table).
```{r}
dft1PuritiesPloidy <- dft1Purities %>%
  mutate(ploidy = ifelse(X15_PLOIDY == "diploid",
                 2, 
                 ifelse(X15_PLOIDY == "tetraploid", 
                        4, 
                        "toCheck"))) %>%
  filter(ploidy != "toCheck") 


dft1SomLongVaf <- dft1SomLong %>%
  mutate(
    Alt = as.integer(sub("/.*", "", ratio)),
    Total = as.integer(sub(".*/", "", ratio))
  ) %>%
  # If nAlt = 0, this was a variant in a different tumor sample, so filter out any 0 vafs!
  filter(Alt > 0) %>%
  left_join(dft1PuritiesPloidy %>%
              dplyr::rename("tumor"="TUMOUR.INFORMATION_TUMOUR ID", 
                     "purity"="X14_PURITY (VAF-based)") %>%
              mutate(purity = as.numeric(purity)) %>%
              dplyr::select(tumor, 
                            ploidy, 
                            purity)) %>%
  filter(tumor %in% dft1PuritiesPloidy$`TUMOUR.INFORMATION_TUMOUR ID`) %>%
  mutate(
    position = as.numeric(position),
    CHROM = as.character(CHROM),
    tumor = as.character(tumor)
  ) 
```

## 2.2 Adjust ploidy at variant using CNV status. 
- Per variant per tumor, find all CNV events spanning that position
- This means that if you filter for presence > 0, there is a change in tumor CN. 
- For some variants, there may be more than one change in CN, to check which these are, filter for the rows where this is the case. 
### 2.2.1 Find multiplets
```{r}
dft1CnvJoin <- dft1SomLongVaf %>%
  left_join(
    dft1CnvLong,
    by = join_by(
      CHROM,
      tumor,
      position >= START,
      position <= END
    )
  ) %>%
  filter(presence > 0) %>%
  distinct() %>%
  mutate(tumorLocation = paste0(tumor, 
                                "_", 
                                location),
         tumorLocationAlt = paste0(tumor, 
                                "_", 
                                location, 
                                "_", 
                                ALT)) 

# Tumor_position combinations where there is more than one CNV event
check <- which(table(dft1CnvJoin$tumorLocation)>1)
checkAlt <- which(table(dft1CnvJoin$tumorLocationAlt)>1)

## Reconstruct table but keep all variants, even when presence == 0, because these are instances where no CNV exists at that variant. 
dft1CnvJoinMultiplets <- dft1SomLongVaf %>%
  left_join(
    dft1CnvLong,
    by = join_by(
      CHROM,
      tumor,
      position >= START,
      position <= END
    )
  ) %>%
  # filter(presence > 0) %>%
  distinct() %>%
  mutate(tumorLocation = paste0(tumor, 
                                "_", 
                                location), 
         tumorLocationAlt = paste0(tumor, 
                                "_", 
                                location, 
                                "_", 
                                ALT))  %>%
  mutate(multiplets = ifelse(tumorLocation %in% names(check), 
                1, 
                0), 
         multipletsAlt = ifelse(tumorLocationAlt %in% names(checkAlt), 
                1, 
                0)) %>%
  mutate(variantType = "dft1Som") 

## File is too large to save.
# write.xlsx(dft1CnvJoinMultiplets %>% filter(presence > 0), 
#            paste0(resDir, 
#                   fprefix, 
#                   "_dft1CnvJoinMultiplets.xlsx"))

rm(dft1CnvJoin)
gc()

# Special cases where looked like a multiplet, but this is actually because there are two different variants at the same position in the same tumor!
# checkMulti <- dft1CnvJoinMultiplets %>% 
#   filter(multiplets == "1" & multipletsAlt == "0" & presence > 0)
```

### 2.2.2 Test function for detecting multiple possible CN states
```{r}
possible_cn_absorbing <- function(start_cn, nGain, nLoss) {
  # Each state is a pair: CN, remaining events (gains and losses)
  states <- list(list(cn = start_cn, g = nGain, l = nLoss))
  
  final_cns <- c()
  
  while (length(states) > 0) {
    new_states <- list()
    
    for (s in states) {
      cn <- s$cn
      g  <- s$g
      l  <- s$l
      
      if (cn == 0 || (g == 0 && l == 0)) {
        # CN is 0 (absorbing) or no events left → record final CN
        final_cns <- c(final_cns, cn)
      } else {
        # Apply a gain if available
        if (g > 0 && cn > 0) {
          new_states <- append(new_states, list(list(cn = cn + 1, g = g - 1, l = l)))
        }
        # Apply a loss if available
        if (l > 0) {
          new_cn <- max(0, cn - 1)
          new_states <- append(new_states, list(list(cn = new_cn, g = g, l = l - 1)))
        }
      }
    }
    
    states <- new_states
  }
  sort(unique((final_cns)))
}

possible_cn_absorbing(3, 2, 3)

# # Filter on variants where I know ploidy = 2 and nGain = 1 and nLoss = 2. This is because this should lead to two possible final copy numbers (1 or 0).
# dft1Test <- dft1CnvJoinMultiplets %>%
#   
#   mutate(ploidy = as.numeric(ploidy)) %>%
#   
#   group_by(tumorLocationAlt) %>%
#   
#   filter(multipletsAlt == 1 & presence > 0) %>%
#   
#   mutate(
#     nGain = sum(EVENT.TYPE == "GAIN"),
#     nLoss = sum(EVENT.TYPE == "LOSS"),
#     nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
#     nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
#     nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
#     nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)")
#   ) %>%
#   
#   dplyr::select(-c(setdiff(colnames(dft1CnvLong), "tumor"))) %>%
#   distinct() %>%
#   rowwise()  %>% 
#   filter(ploidy == 3 & nGain == 1 & nLoss == 2) 
# 
# # Test functions
# dft1TestFunction <- dft1Test %>%
#   
#   mutate(
#     # ---- CASE 1: diploid ----
#     possibleCN = list(
#       if (ploidy == 2) {
#         possible_cn_absorbing(ploidy, nGain, nLoss)
#         
#       # ---- CASE 2: tetraploid with pre-events ----
#       } else if (ploidy == 4 & (nGainPre > 0 | nLossPre > 0)) {
#         
#         pre_cn <- possible_cn_absorbing(2, nGainPre, nLossPre)
#         pre_cn <- pre_cn * 2
#         
#         # apply post events to each possible pre state
#         unique(unlist(
#           lapply(pre_cn, function(cn) {
#             possible_cn_absorbing(
#               cn,
#               nGain + nGainPost,
#               nLoss + nLossPost
#             )
#           })
#         ))
#         
#       # ---- CASE 3: tetraploid without pre-events ----
#       } else if (ploidy == 4) {
#         
#         possible_cn_absorbing(
#           ploidy,
#           nGain + nGainPost,
#           nLoss + nLossPost
#         )
#         
#       } else {
#         NA_real_
#       }
#     ),
#     
#     minCN = min(possibleCN),
#     maxCN = max(possibleCN)
#     
#   ) %>%
#   ungroup() %>%
#   mutate(cnvAdjAnno = "multipleChangesTest")
```

### 2.2.3 Variants with no change
- copyNumberAdjusted = ploidy
- number of unique variants (taking into account alt) = number of rows, because no changes to cnv. YES? --> multiallelic variants possible (i.e., subclones, because these variants don't occur in any of the host samples.)
```{r}
dft1CnvJoinNoChanges <- dft1CnvJoinMultiplets %>%
  # Group by tumorLocation. If all rows presence = 0 (and, by default, not a multiplet), no change to CNV. 
  group_by(tumorLocationAlt) %>%
  filter(multipletsAlt == 0 & all(presence==0)) %>%
  mutate(
    nGain = sum(EVENT.TYPE == "GAIN"),
    nLoss = sum(EVENT.TYPE == "LOSS"),
    nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
    nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
    nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
    nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)"), 
    sumPresence = sum(presence)
  ) %>%
  ungroup() %>%
  mutate(copyNumberAdjusted = ploidy) %>%
  # remove CNV table columns
  dplyr::select(-c(setdiff(colnames(dft1CnvLong),
                           "tumor"))) %>%
  mutate(cnvAdjAnno = "noChanges")

# # length(unique(dft1CnvJoinNoChanges$tumorLocation))==nrow(dft1CnvJoinNoChanges)
# checkNoChange <- names(which(table(dft1CnvJoinNoChanges$tumorLocation)>1)) # 15 samples. This is because there are no changes, but you get duplicate rows if there is a diff Alt allele. 
# checkNoChangeAlt <- names(which(table(dft1CnvJoinNoChanges$tumorLocationAlt)>1)) # 0 samples. 
```

### 2.2.4 Variants with single change
- change ploidy according to single step
- number of unique variants = number of rows, because only 1 change to cnv per variant. 5 tumorLocation where more than one variant with a single CNV change.
- All presences should be 1 because single change
- If CN goes to 0, it should stay at 0. (Applies for multistep, not here.) Running same code as multiplets as a sanity check, none of the maxCn should differ from minCn. 
```{r}
dft1CnvJoinSingleChange <- dft1CnvJoinMultiplets %>%
  mutate(ploidy=as.numeric(ploidy)) %>%
  filter(presence != 0 & multipletsAlt == 0) %>%
  group_by(tumorLocationAlt) %>%
  mutate(
    nGain = sum(EVENT.TYPE == "GAIN"),
    nLoss = sum(EVENT.TYPE == "LOSS"),
    nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
    nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
    nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
    nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)"),
    sumPresence = sum(presence)
  ) %>%
   # remove CNV table columns, because this has been aggregated by summing rows.
  dplyr::select(-c(setdiff(colnames(dft1CnvLong),
                           "tumor"))) %>%
  mutate(
    copyNumberAdjusted = pmax(0,
    case_when(
      # Case 1: diploid
      ploidy == 2 ~ ploidy + nGain - nLoss,
      
      # Case 2: tetraploid with pre-tetra events present
      ploidy == 4 & (nGainPre > 0 | nLossPre > 0) ~ 
        (2 + nGainPre - nLossPre) * 2 + 
        (nGain + nGainPost) - (nLoss + nLossPost),
      
      # Case 3: tetraploid without pre-tetra events present
      ploidy == 4 & nGainPre == 0 & nLossPre == 0 ~
        ploidy + (nGain + nGainPost) - (nLoss + nLossPost),
      
      TRUE ~ NA_real_
    )
   )
  ) %>%
  mutate(cnvAdjAnno = "singleChange")
  
checkSingle <- names(which(table(dft1CnvJoinSingleChange$tumorLocation)>1)) # 0 
checkAltSingle <- names(which(table(dft1CnvJoinSingleChange$tumorLocationAlt)>1)) # 0 
```

### 2.2.5 Variants with multiple changes
- Aggregate steps using same logic as before, only this time ploidy will change using sum of all steps rather than single change of CNV status.
- Issue: we don't know the order of events. Depending on order of losses and gains, the final CN could differ. Adding a permutation function which returns how many times a different final CN state could occur. 
- None of the tetraploid tumors have nLossPre > 1, therefore cannot go to 0.
- The only variable cases are diploid tumors whose variants have two losses and one gain. 
- Filter out the five variants that could have been CN 0 or 1!
  - Check these variants against the counts + gtf file.
```{r}
dft1CnvJoinMultipleChanges <- dft1CnvJoinMultiplets %>%
  
  mutate(ploidy = as.numeric(ploidy)) %>%
  
  group_by(tumorLocationAlt) %>%
  
  filter(multipletsAlt == 1 & presence > 0) %>%
  
  mutate(
    nGain = sum(EVENT.TYPE == "GAIN"),
    nLoss = sum(EVENT.TYPE == "LOSS"),
    nGainPost = sum(EVENT.TYPE == "GAIN (post-tetraploidisation)"),
    nGainPre = sum(EVENT.TYPE == "GAIN (pre-tetraploidisation)"),
    nLossPost = sum(EVENT.TYPE == "LOSS (post-tetraploidisation)"),
    nLossPre = sum(EVENT.TYPE == "LOSS (pre-tetraploidisation)"),
    sumPresence = sum(presence)
  ) %>%
  
  dplyr::select(-c(setdiff(colnames(dft1CnvLong), "tumor"))) %>%
  distinct() %>%
  
  rowwise() %>%
  
  mutate(
    
    # ---- CASE 1: diploid ----
    possibleCn = list(
      if (ploidy == 2) {
        possible_cn_absorbing(ploidy, nGain, nLoss)
        
      # ---- CASE 2: tetraploid with pre-events ----
      } else if (ploidy == 4 & (nGainPre > 0 | nLossPre > 0)) {
        
        pre_cn <- possible_cn_absorbing(2, nGainPre, nLossPre)
        pre_cn <- pre_cn * 2
        
        # apply post events to each possible pre state
        unique(unlist(
          lapply(pre_cn, function(cn) {
            possible_cn_absorbing(
              cn,
              nGain + nGainPost,
              nLoss + nLossPost
            )
          })
        ))
        
      # ---- CASE 3: tetraploid without pre-events ----
      } else if (ploidy == 4) {
        
        possible_cn_absorbing(
          ploidy,
          nGain + nGainPost,
          nLoss + nLossPost
        )
        
      } else {
        NA_real_
      }
    ),
    
    minCn = min(possibleCn),
    maxCn = max(possibleCn)
    
  ) %>%
  ungroup() %>%
  mutate(cnvAdjAnno = "multipleChanges")

saveRDS(dft1CnvJoinMultipleChanges, 
        paste0(resDir, 
               fprefix, 
               "_dft1CnvJoinMultipleChanges_withMultipleCnvOptions.rds"))

multipleCns <- dft1CnvJoinMultipleChanges %>% 
  filter(minCn != maxCn)

dft1CnvJoinMultipleChangesFiltered <- dft1CnvJoinMultipleChanges %>%
  filter(minCn == maxCn) %>%
  mutate(copyNumberAdjusted = minCn) %>%
  dplyr::select(-c(possibleCn, 
                   maxCn, 
                   minCn))
```


### 2.2.6 Rejoin tables
- Adjust observed VAF using knowledge of purities. 
- TRIPLE check no variants are missing because of filtering!
- Add host columns for merging with germline variants later on. 
- TODO: Fix X chromosome in male samples
```{r}
dft1SomCnvCorrected <- rbind(dft1CnvJoinNoChanges,
                             dft1CnvJoinSingleChange,
                             dft1CnvJoinMultipleChangesFiltered) %>%
  dplyr::rename("purityTumor"="purity") %>%
  mutate(obsVaf = Alt/Total,
         nHostRef = 2, 
         nHostAlt = 0, 
         nTumorRef = NA, 
         nTumorAlt = NA, 
         nHostTotal = 2, 
         copyNumberAdjusted = as.numeric(copyNumberAdjusted)) %>%
  distinct() 

saveRDS(dft1SomCnvCorrected, 
        paste0(resDir, fprefix, "_dft1SomCnvCorrected.rds"))
```

# 3 Construct final table
- Join adjusted CNV status with AlleleCounter results
- Of note, NOT all samples have an associated RNA-seq file! See metadata script.
- Merge with gtf file on informative positions (exons)
- Compute tAlt from observed VAF.
## 3.1 Use VAF, purity, and copy number to compute nTumorRef and nTumorAlt
```{r}
dft1SomGenotyped <- dft1SomCnvCorrected %>%
  mutate(tAlt = computeTAlt(obsVaf, 
                            purityTumor, 
                            tTotal = copyNumberAdjusted, 
                            hTotal = nHostRef+nHostAlt, 
                            hAlt = nHostAlt), 
         nTumorAlt = round(tAlt),
         nTumorRef = copyNumberAdjusted-nTumorAlt
         )

saveRDS(dft1SomGenotyped, 
        paste0(resDir, 
               fprefix, 
               "_dft1SomGenotyped.rds"))
```




## 3.2 Merge AlleleCounter results
- Filter out rows for tumors where we don't have RNA seq data. 
- 24 DFT1 tumors/XYZ tumors with matched RNA-seq data by tumor id.
- Number of vars drops because not all vars in a tumor for which we have matched RNAseq.
```{r}
dft1SomGenotypedAlCo <- dft1SomGenotyped %>%
  # filter(tumorLocationAlt %in% dft1SomAlCoCombined$tumorLocationAlt) %>%
  left_join(dft1SomAlCoCombined %>%
              dplyr::select(TOTAL,
                            NREF, 
                            NALT, 
                            tumorLocationAlt,
                            tumorIdRnaSeq)) %>%
  distinct()

saveRDS(dft1SomGenotypedAlCo, 
        paste0(resDir, 
               fprefix, 
               "_dft1SomGenotypedAlCo.rds"))


nrow(dft1SomGenotypedAlCo %>% 
       filter(TOTAL > 0)) # 11290 variants

rm(dft1SomCnvCorrected)
gc()

### Only one variant genotyped??? ### --> therefore others we don't have RNAseq data for. (Two tumors for 5 variants, 4 came from one tumor for which we don't have RNA seq, therefore only one variant remains.) All have 0 aligned reads anyways (52T2_2:385998837_T)
checkMultipleCnsAlCo <- multipleCns %>%
  # filter(tumorLocationAlt %in% dft1SomAlCoCombined$tumorLocationAlt) %>%
  left_join(dft1SomAlCoCombined %>%
              dplyr::select(TOTAL,
                            NREF, 
                            NALT, 
                            tumorLocationAlt,
                            tumorIdRnaSeq)) %>%
  distinct() 
```


## 3.3 GTF annotations
- noisyr has a function to extract all exons and positions from a gtf file.
- alternatively annotate using 
```{r}
gtfFile <- "~/gitClones/01_Deconvolution/02_Deconvolution_DFT1/020_Data/RFS/Sarcophilus_harrisii.mSarHar1.11.115.gtf"

gtf <- import(gtfFile)

genes <- gtf[gtf$type == "gene"]

exons <- gtf[gtf$type == "exon"]

genes <- genes[genes$gene_biotype == "protein_coding"]


variants <- GRanges(
  seqnames = dft1SomGenotypedAlCo$chrom,
  ranges = IRanges(start = dft1SomGenotypedAlCo$position, 
                   end = dft1SomGenotypedAlCo$position)
)

# Which variants fall in an exon
exonHits <- findOverlaps(variants, exons)

# Map exon hits to gene IDs
exonGeneIds <- exons$gene_id[subjectHits(exonHits)]
exonGeneNames <- exons$gene_name[subjectHits(exonHits)]
exonBiotypes <- exons$gene_biotype[subjectHits(exonHits)]

hitsDf <- data.frame(
  query = queryHits(exonHits),
  gene_id = exonGeneIds,
  gene_name = exonGeneNames,
  biotype = exonBiotypes
)

collapseFun <- function(x) paste(unique(x), collapse = ";")

collapsed <- aggregate(. ~ query, 
                       data = hitsDf, 
                       FUN = collapseFun)

# Compare with assigned gene_id
dft1SomGenotypedAlCoAnno <- dft1SomGenotypedAlCo

dft1SomGenotypedAlCoAnno$exonGeneId <- NA
dft1SomGenotypedAlCoAnno$exonGeneName <- NA
dft1SomGenotypedAlCoAnno$exonBiotype <- NA

dft1SomGenotypedAlCoAnno$exonGeneId[collapsed$query] <- collapsed$gene_id
dft1SomGenotypedAlCoAnno$exonGeneName[collapsed$query] <- collapsed$gene_name
dft1SomGenotypedAlCoAnno$exonBiotype[collapsed$query] <- collapsed$biotype

length(unique(dft1SomGenotypedAlCoAnno %>%
         filter(is.na(exonGeneId)==F) %>%
         distinct() %>%
         pull(tumorLocationAlt)))

# [1] 2366 unique vars intersecting a gene

length(unique(unique(dft1SomGenotypedAlCoAnno %>%
         filter(is.na(exonGeneId)==F) %>%
         distinct() %>%
         pull(exonGeneId))))

# [1] 1255 genes

saveRDS(dft1SomGenotypedAlCoAnno, 
        paste0(resDir, 
               fprefix, 
               "_dft1SomGenotypedAlCoAnno.rds"))

write.xlsx(dft1SomGenotypedAlCoAnno %>%
             filter(is.na(TOTAL)==F & is.na(exonGeneId) ==F), 
        paste0(resDir, 
               fprefix, 
               "_dft1SomGenotypedAlCoAnnoCondensed.xlsx"))
```